---
title: "Between Cell Variability"
author: "Atreya Choudhury"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    code_folding: hide
    number_sections: no
    fig_crop: yes
---

## Strategy

The goal is to generate features with various heterogeneity levels for the purpose of between cell differentiation.
The goal here is to focus more on  the stochastic nature of methylation.

Our strategy here is to mix markov chains generated using a transition matrix of the form
$$
  \begin{bmatrix}
  \rho & 1 - \rho \\
  \rho & 1 - \rho
  \end{bmatrix}
$$
with different values of $\rho$ to create more or less heterogeneity across samples in a region.

We start by creating a template for a cell.
We have a fixed number of features, the length of each feature is taken to be the same and the number of simulated CpG sites is the same as well.
Every feature is given a **between cell Variability Index (bcVI)** between 1 and 10 and stored in the `bc_vi` column.
We have a list of 10 equally spaced values of $\rho$ from $0.1$ to $0.9$.
Now, for a feature with given variability index, say $x$, we take the first $x$ elements from our list of $\rho$ values and sample from it to generate $\rho$ values for the samples belonging to that feature.
These sampled values of $\rho$ are then stored in the `rho` column.

With the template, we now generate samples as described above.
At every intersection of sample and feature, we have a value of $\rho$ which can be used to sample a markov chain.
We observe that features with a lower bcVI see more homogenous samplings in `rho` which lead to lesser shannon entropy and vice-versa.
Also observe that chains generated with higher values of $\rho$ are also more methylated.
This leads to a relationship between average methylation and `bc_vi`, which is visible from scMET's performance.

```{r, setup}
suppressPackageStartupMessages({
  library(ggplot2)
  library(ggpubr)
  library(svglite)
})

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE, warning = FALSE,
  fig.path = snakemake@params[["plots_dir"]],
  dev = c("png", "svg"),
  dpi = 500
)

source("src/mi_table.R")
source("src/make_grid_plots.R")
source("src/ggtheme.R")

data_dir <- dirname(snakemake@input[["disp_scmet"]])
S_GRID <- snakemake@params[["S_GRID"]]
N_GRID <- snakemake@params[["N_GRID"]]
DISP <- snakemake@params[["DISP"]]
```

## Variation
For benchmarking against the number of samples, we used a grid with `r length(S_GRID$S)` sample sizes (S = `r paste(S_GRID$S, collapse = ", ")`), `r S_GRID$N` features where each feature has `r S_GRID$f` CpGs (all covered).
```{r}
s_plots <- make_grid_plots(S_GRID, "S", "bc_vi", data_dir)
```

For benchmarking against the number of features, we used a grid with `r N_GRID$S` samples, `r length(N_GRID$N)` feature sizes (N = `r paste(N_GRID$N, collapse = ", ")`) where each feature has `r N_GRID$f` CpGs (all covered).
```{r}
n_plots <- make_grid_plots(N_GRID, "N", "bc_vi", data_dir)
```

## One Example

We show one example with `r DISP$S` samples, `r DISP$N` features where each feature has `r DISP$f` CpGs (all covered).
```{r}
S <- DISP$S
N <- DISP$N
f <- DISP$f

intervals <- read.table(
  paste0(data_dir, paste("/intervals", S, N, f, "tsv", sep = ".")),
  header = T
)
intervals <- intervals["bc_vi"]
sampens <- read.table(
  paste0(data_dir, paste("/yamet", S, N, f, "det.out", sep = ".")),
  header = T
)
jnt <- cbind(sampens, intervals)
rownames(jnt) <- paste(jnt$start, jnt$end, sep = ":")
output_cols <- grep("^output", colnames(jnt), value = TRUE)
jnt$sampen_avg <- rowMeans(jnt[, output_cols])
jnt.long <- do.call(rbind, lapply(output_cols, function(col) {
  data.frame(
    start = jnt$start,
    end = jnt$end,
    sample = col,
    sampen = jnt[[col]],
    shannon = jnt$shannon,
    avg_meth = jnt$avg_meth,
    bc_vi = jnt$bc_vi
  )
}))

meth <- read.table(
  paste0(data_dir, paste("/yamet", S, N, f, "meth.out", sep = ".")),
  header = TRUE
)
meth.long <- do.call(rbind, lapply(output_cols, function(col) {
  data.frame(
    meth = meth[[col]]
  )
}))

all.long <- cbind(jnt.long, meth.long)

main_plots <- list()

main_plots$samp_meth <- ggplot(all.long, aes(x = meth, y = sampen, color = bc_vi)) +
  geom_point(alpha = 0.7) +
  theme_ng() +
  labs(
    ## title = "Sample Entropy in different bcVI values",
    x = "avg. methylation", y = "sample entropy", color = "bcVI"
  )

samp_vi <- ggplot(all.long, aes(x = as.factor(bc_vi), y = sampen)) +
  geom_point(alpha = 0.7, position = position_jitter(width = 0.2)) +
  geom_violin(alpha = 0.3, position = position_dodge(0.8)) +
  theme_ng() +
  labs(
    ## title = "Sample Entropy in different bcVI values",
    x = "bcVI", y = "sample entropy"
  )

shannon_meth <- ggplot(jnt, aes(x = avg_meth, y = shannon, color = bc_vi)) +
  geom_point(alpha = 0.7) +
  theme_ng() +
  labs(
    ## title = "Shannon Entropy in different bcVI values (per feature)",
    x = "avg. methylation", y = "Shannon entropy", color = "bcVI"
  )

shannon_vi <- ggplot(jnt, aes(x = as.factor(bc_vi), y = shannon, color = avg_meth)) +
  geom_point(alpha = 0.7) +
  theme_ng() +
  labs(
    ## title = "Shannon Entropy in different bcVI values (per feature)",
    x = "bcVI", y = "Shannon entropy", color = "avg. methylation"
  )

main_plots$samp_shan <- ggplot(all.long, aes(x = shannon, y = sampen, color = bc_vi)) +
  geom_point(alpha = 0.7) +
  theme_ng() +
  labs(
    ## title = "Sample Entropy in different bcVI values",
    x = "Shannon entropy", y = "sample entropy", color = "bcVI"
  )
```

```{r}
scmet_fit <- readRDS(paste0(data_dir, paste("/scmet", S, N, f, "rds", sep = ".")))
tmp <- colMeans(scmet_fit$posterior$mu)
jnt$scmet_mu <- tmp[match(rownames(jnt), names(tmp))]
tmp <- colMeans(scmet_fit$posterior$gamma)
jnt$scmet_gamma <- tmp[match(rownames(jnt), names(tmp))]
tmp <- colMeans(scmet_fit$posterior$epsilon)
jnt$scmet_epsilon <- tmp[match(rownames(jnt), names(tmp))]
```

```{r}
ggplot(jnt, aes(x = as.factor(bc_vi), y = scmet_mu, color = avg_meth)) +
  geom_point(alpha = 0.7) +
  theme_ng() +
  labs(
    ## title = "scMET (mu) in different bcVI values (per feature)",
    x = "bcVI", y = expression("scMET " ~ mu), color = "avg. methylation"
  )

main_plots$scmet_mu_meth <- ggplot(jnt, aes(x = avg_meth, y = scmet_mu, color = bc_vi)) +
  geom_point(alpha = 0.7) +
  theme_ng() +
  labs(
    ## title = "scMET (mu) in different meth values (per feature)",
    x = "avg. methylation", y = expression("scMET " ~ mu), color = "bcVI"
  )

main_plots$scmet_gamma_meth <- ggplot(jnt, aes(x = avg_meth, y = scmet_gamma, color = bc_vi)) +
  geom_point(alpha = 0.7) +
  theme_ng() +
  labs(
    ## title = "scMET (gamma) in different meth values (per feature)",
    x = "avg. methylation", y = expression("scMET " ~ gamma), color = "bcVI"
  )

main_plots$scmet_epsilon_meth <- ggplot(jnt, aes(x = avg_meth, y = scmet_epsilon, color = bc_vi)) +
  geom_point(alpha = 0.7) +
  theme_ng() +
  labs(
    ## title = "scMET (epsilon) in different meth values (per feature)",
    x = "avg. methylation", y = expression("scMET " ~ epislon), color = "bcVI"
  )

ggplot(jnt, aes(x = as.factor(bc_vi), y = sampen_avg, color = avg_meth)) +
  geom_point(alpha = 0.7) +
  theme_ng() +
  labs(
    ## title = "Sample Entropy in different bcVI values (per feature)",
    x = "bcVI", y = "avg. sample entropy", color = "avg. methylation"
  )

ggplot(jnt, aes(x = avg_meth, y = sampen_avg, color = bc_vi)) +
  geom_point(alpha = 0.7) +
  theme_ng() +
  labs(
    ## title = "Sample Entropy in different meth values (per feature)",
    x = "avg. methylation", y = "avg. sample entropy", color = "bcVI"
  )
```

```{r}
mi_table <- mi_table_gen(jnt, "bc_vi", table = T)
```

```{r fig.width=20, fig.height=15, out.width="100%"}
ggarrange(
  ggarrange(
    main_plots$samp_meth,
    ggarrange(
      ggarrange(
        s_plots$nmi, n_plots$nmi,
        legend = "right", common.legend = T,
        ncol = 2, widths = c(1, 1)
      ),
      ggarrange(
        s_plots$secmem, n_plots$secmem,
        legend = "top", common.legend = T,
        ncol = 2, widths = c(1, 1)
      ),
      ncol = 1, heights = c(1, 1)
    ),
    nrow = 2, heights = c(2, 3)
  ),
  ggarrange(
    main_plots$samp_shan,
    samp_vi,
    shannon_meth,
    shannon_vi,
    nrow = 4
  ),
  ggarrange(
    main_plots$scmet_mu_meth,
    main_plots$scmet_gamma_meth,
    main_plots$scmet_epsilon_meth,
    mi_table,
    common.legend = T,
    nrow = 4
  ),
  ncol = 3
)
```


# Simpler figure

```{r assembled, fig.width=15, fig.height=15, out.width="100%"}
# extract legends for bottom section
legends <- list()
legends$bcVI <- get_legend(main_plots$samp_meth)
legends$nmi <- get_legend(s_plots$nmi)
legends$perf <- get_legend(s_plots$secmem)

arranged <- ggarrange(
  ggarrange(
    # top row: sampen vs meth and shannon
    ggarrange(
      main_plots$samp_meth,
      main_plots$samp_shan,
      legend = "none",
      nrow = 1
    ),

    # Bottom row: scMET plots
    ggarrange(
      main_plots$scmet_mu_meth,
      main_plots$scmet_gamma_meth,
      main_plots$scmet_epsilon_meth,
      legend = "none",
      nrow = 1
    ),

    # bottom section: S-plots (row 1) and N-plots (row 2)
    ggarrange(
      s_plots$nmi + xlab("# cells") + ylab("performance (metric NMI)"),
      s_plots$secmem + xlab("# cells") + ylab("CPU time (s)"),
      n_plots$nmi + xlab("# features") + ylab("performance (metric NMI)"),
      n_plots$secmem + xlab("# features") + ylab("CPU time (s)"),
      legend = "none",
      ncol = 4
    ),
    ncol = 1,
    nrow = 3,
    heights = c(2, 1, 1)
  ),

  # shared legends stacked
  ggarrange(
    legends$bcVI,
    legends$nmi,
    legends$perf,
    ncol = 1,
    align = "v"
  ),
  ncol = 2,
  nrow = 1,
  widths = c(6, 1),
  align = "h"
)

saveRDS(list(
  main_plots = main_plots, s_plots = s_plots,
  n_plots = n_plots, legends = legends
), file = paste0(
  dirname(snakemake@output[["html"]]), "/06_plots.rds"
))

arranged

ggsave(
  filename = paste0(
    dirname(snakemake@output[["html"]]), "/06_between.svg"
  ),
  plot = arranged,
  width = 15,
  height = 15,
  units = "in"
)
```
