---
title: "Normalization, Emanuel's simuls"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Atreya, Emanuel, Izaskun"
params: 
  yamet_dir:
output:
  bookdown::html_document2:
    code_folding: hide
    number_sections: no
    fig_crop: yes
---

```{r, setup}
condaEnvPath <- file.path(Sys.getenv("CONDA_PREFIX"), "lib", "R", "library")
.libPaths(condaEnvPath)

source('ggtheme.R')

suppressPackageStartupMessages({
  library(ggplot2)
  library(data.table)
  library(knitr)
})

knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.path = "../output/03_plots/",
    dev = c("png", "svg"),
    dpi=500
)
```

Firstly, we import the different yamet output files.

```{r, import}
##data.dir <- '../output/yamet/out/'
data.dir <- file.path("..", params$yamet_dir)
data.paths <- list.files(data.dir, pattern = "\\.tsv$", full.names = TRUE)
data.list <- lapply(data.paths, fread)
data <- rbindlist(data.list)
str(data)
```

Then, we parse the different elements from the filenames and turn the appropriate columns into factors.

```{r, parse}
data[, c("no", "ignore", "ncpgs", "type", "coverage", "tr") :=
  tstrsplit(file, split = "_", keep = 4:9, type.convert = TRUE)]
data[, "tr" := sub("\\.tsv$", "", tr)]
#data[, c("file", "ignore") := NULL]
data[,cell:=sub("\\.tsv$", "", basename(file))]

factor_cols <- c("ncpgs", "type", "coverage", "tr")
data[, (factor_cols) := lapply(.SD, as.factor), .SDcols = factor_cols]
data <- data[sampen != -1 & avg_meth != -1]

print(unique(data$coverage))
print(unique(data$type))
data$coverage <- factor(data$coverage, levels = c("low", "lowReal", "real", "medium",
    "high", "complete"))
data$type <- factor(data$type, levels = c('nb', 'rand'))
data$tr <- factor(data$tr, levels = c('lmrCons', 'lmrRand', 'imrCons', 'imrRand', 'hmrRand', 'hmrCons'))
print(head(data))
```

Unadjusted sample entropy

```{r, plots, fig.width = 7, fig.height = 7}
unique.types <- unique(data$type)
plots.list <- list()

for (unique.type in unique.types) {
  p <- ggplot(data[type == unique.type], aes(x = avg_meth, y = sampen, color = tr)) +
    geom_point(alpha = 0.7) +
    ## facet_grid(ncpgs ~ coverage) +
      facet_grid(
         ncpgs ~ coverage,
          labeller = labeller(
              ncpgs   = function(x) paste("# CpGs:", x),
              coverage = function(x) paste("cov.:", x))) +
    theme_minimal() +
    labs(
      title = paste("Sample Entropy in different TR settings for type", unique.type),
      x = "avg. methylation", y = "sample entropy", color = "TR"
    ) +
      theme_ng_discrete() +
      theme( axis.text.x = element_text(angle = 45, hjust = 1))

  plots.list[[unique.type]] <- p
}
print(plots.list)
```


Adj sample entropy

```{r, norm_plots, fig.width = 7, fig.height = 7}
unique.types <- unique(data$type)
plots.list <- list()

for (unique.type in unique.types) {
  p <- ggplot(data[type == unique.type], aes(x = avg_meth, y = sampen_norm, color = tr)) +
    geom_point(alpha = 0.7) +
    ## facet_grid(ncpgs ~ coverage) +
      facet_grid(
         ncpgs ~ coverage,
          labeller = labeller(
              ncpgs   = function(x) paste("# CpGs:", x),
              coverage = function(x) paste("cov.:", x))) +
    theme_minimal() +
    labs(
      title = paste("Adjusted Sample Entropy in different TR settings for type", unique.type),
      x = "avg. methylation", y = "adj sample entropy (adjS)", color = "TR"
    ) +
      theme_ng_discrete() +
      theme( axis.text.x = element_text(angle = 45, hjust = 1))
  
  plots.list[[unique.type]] <- p
}
print(plots.list)
```
Scaled by dividing by adjS by the exp. SampEn to avoid the convex dependence of adjS. on the marginal methylation.   
```{r, norm_plots scaled, fig.width = 9, fig.height = 7}
# just quick test how simple scaling would perform
data[,exp_meth:=-log(avg_meth^2+(1-avg_meth)^2)]
data[,sampen_norm_scaled:=sampen_norm/exp_meth]

unique.types <- unique(data$type)
plots.list <- list()

for (unique.type in unique.types) {
  p <- ggplot(data[type == unique.type], aes(x = avg_meth, y = sampen_norm_scaled, color = tr)) +
    geom_point(alpha = 0.7) +
    ## facet_grid(ncpgs ~ coverage) +
      facet_grid(
         ncpgs ~ coverage,
          labeller = labeller(
              ncpgs   = function(x) paste("# CpGs:", x),
              coverage = function(x) paste("cov.:", x))) +
    theme_minimal() +
    labs(
      title = paste("scaled Adjusted Sample Entropy in different TR settings for type", unique.type),
      x = "avg. methylation", y = "adj sample entropy (adjS)", color = "TR"
    ) +
      theme_ng_discrete() +
      theme( axis.text.x = element_text(angle = 45, hjust = 1))
  
  plots.list[[unique.type]] <- p
}
print(plots.list)
```
Additive smoothing applied to the estimated marginal methylation probability $\hat p_{\text{m}}$ to avoid the "wings" occurring for sparse data.

$\hat p_{m,\,smoothed}=\frac{n_{m\,cpgs}+0.5}{n_{covered\,cpgs}+1}$

Currently, estimation of $\hat p_{m}$ corresponds to `--all-meth`.

```{r, norm_plots smoothed+scaled, fig.width = 9, fig.height = 7}
# quick & naive check reading the number of covered CpGs per file
simFiles <- list.files(file.path("..", "output/sim_data"), pattern="*.tsv", 
                       full.names=TRUE)
simFiles <- simFiles[!grepl("cpg", simFiles)]
simCovCpgs <- lapply(simFiles, 
                   function(f){
                     simDat <- fread(f)
                     covDt <- data.table(n_cov_cpgs=nrow(simDat))
                     if(covDt$n_cov_cpgs==0){
                       covDt$n_meth <- 0}
                     else{
                       covDt$n_meth <- sum(simDat$V3)
                     }
                     covDt
                   })
names(simCovCpgs) <- basename(simFiles)
simCovCpgs <- rbindlist(simCovCpgs, idcol="file")
simCovCpgs[,cell:=sub("\\.tsv$", "", basename(file))]

data2 <- merge(data, simCovCpgs, by=c("cell"), all.x=TRUE, all.y=FALSE)

# try some smoothing in dependence of the sparsity (additive smoothing)
data2[,p_meth_smooth:=(n_meth+0.5)/(n_cov_cpgs+1)]

# would correspond to the --all-meth atm. 
data2[,exp_meth_smooth:=-log(p_meth_smooth^2+(1-p_meth_smooth)^2)] 
data2[,sampen_norm_scaled2:=(sampen - exp_meth_smooth)/exp_meth_smooth]

unique.types <- unique(data2$type)
plots.list <- list()
for (unique.type in unique.types) {
  p <- ggplot(data2[type == unique.type], 
              aes(x = avg_meth, y = sampen_norm_scaled2, color = tr)) +
    geom_point(alpha = 0.7) +
    ## facet_grid(ncpgs ~ coverage) +
      facet_grid(
         ncpgs ~ coverage,
          labeller = labeller(
              ncpgs   = function(x) paste("# CpGs:", x),
              coverage = function(x) paste("cov.:", x))) +
    theme_minimal() +
    labs(
      title = paste("smoothed+scaled Adjusted Sample Entropy in different TR settings for type", unique.type),
      x = "avg. methylation", y = "smoothed+scaled adj sample entropy (adjS)", color = "TR"
    ) +
      theme_ng_discrete() +
      theme( axis.text.x = element_text(angle = 45, hjust = 1))
  
  plots.list[[unique.type]] <- p
}
print(plots.list)
```


```{r, norm_plots smoothed+scaled per group averages, fig.width = 9, fig.height = 7}
data3 <- data2[,.(avg_meth=mean(avg_meth),
                                       sampen_norm_scaled2=mean(sampen_norm_scaled2),
                                       sd_meth=sd(avg_meth),
                                       sd_sampen_norm_scaled2=sd(sampen_norm_scaled2)),
                                    by=.(ncpgs, coverage, tr, type)]

for (unique.type in unique.types) {
  p <- ggplot(data3[type == unique.type],
              aes(x = avg_meth, y = sampen_norm_scaled2, color = tr)) +
    geom_point(alpha = 0.7) +
    geom_errorbar(
      aes(ymin = sampen_norm_scaled2 - sd_sampen_norm_scaled2, 
          ymax = sampen_norm_scaled2 + sd_sampen_norm_scaled2))+
    geom_errorbarh(
      aes(xmin = avg_meth - sd_meth, 
          xmax = avg_meth + sd_meth))+
      facet_grid(
         ncpgs ~ coverage,
          labeller = labeller(
              ncpgs   = function(x) paste("# CpGs:", x),
              coverage = function(x) paste("cov.:", x))) +
    theme_minimal() +
    labs(
      title = paste("smoothed+scaled Adjusted Sample Entropy in different TR settings for type", unique.type),
      x = "Per group avg. methylation +/- SD", y = "Per group avg. smoothed+scaled adj sample entropy (adjS) +/- SD", 
      color = "TR")+
      theme_ng_discrete() +
      theme( axis.text.x = element_text(angle = 45, hjust = 1))
      #theme(axis.title.text=element_text(size=6))
  plots.list[[unique.type]] <- p
}
print(plots.list)
```
```


Plus some pairwise plots

```{r pairwise_norm, fig.width = 7, fig.height = 7}
unique.types <- unique(data$type)
plots.list <- list()

for (unique.type in unique.types) {
  p <- ggplot(data[type == unique.type],
              aes(x = sampen, y = sampen_norm, color = tr)) +
    geom_point(alpha = 0.7) +
    ## facet_grid(ncpgs ~ coverage) +
      facet_grid(
         ncpgs ~ coverage,
          labeller = labeller(
              ncpgs   = function(x) paste("# CpGs:", x),
              coverage = function(x) paste("cov.:", x))) +
    labs(
      title = paste("Sample Entropy vs Adjusted Sample Entropy for type", unique.type),
      x = "sample entropy (S)",
      y = "adj. sample entropy (adjS)",
      color = "TR"
    ) +
    theme_ng_discrete() +
    theme( axis.text.x = element_text(angle = 45, hjust = 1))


  
  plots.list[[unique.type]] <- p
}

print(plots.list)
```
