---
title: "Cometh simulation with tuple lengths"
params:
  seed: 1
author: "Izaskun Mallona"
output:
  html_document:
    keep_md: false
    toc: true
    toc_float: true
    toc_depth: 4

---

# Context


# SEt up

```{r tag}
analysis_tag <- 'cometh_simulation_tuple_aware'
```

```{r libraries, include=TRUE, cache = FALSE}
library(data.table)
library(entropy)
library(truncnorm) # truncated normal
library(ggplot2)
library(ggExtra)
```

```{r functions}
ac <-  function(col, alpha=1){

    apply(sapply(col, col2rgb)/255, 2, 
          function(x) 
              rgb(x[1], x[2], x[3], alpha=alpha))  
}
```

Test expand grid

```{r}
get_tuple_identifiers <- function(tuple_length) {
    bar <- list()
    bar[1:tuple_length] <- list(c('M', 'U'))

    apply(expand.grid(bar), 1, function(x) paste0(x, collapse = ''))    
}

```

```{r}

count_m_and_u <- function(strings) {
    ms <- lengths(regmatches(strings, gregexpr("M", strings)))
    us <- lengths(regmatches(strings, gregexpr("U", strings)))
    return(list(m = ms, u = us))
}

get_methylation_tuple_length_aware <- function(d, tuple_length) {
    ## get tuple length

    counts <- count_m_and_u(colnames(d))

    are_tuples <- (counts$m + counts$u) > 0

    counts <- lapply(counts, function(x) x[are_tuples])
    
    ## for ( i in 1:length(tmp$m)) {
    ##     if (tmp$m[i] + tmp$[i] == tuple_length) {
    ##         ugly[[colnames(d)[i]]]  <- (counts$m[i]*) / ()
    ##     }
    ## }

    apply(d[,colnames(d)[are_tuples]],
          1,
          function(x) return((sum(x * counts$m)) / (tuple_length*sum(x))))
}

## get_methylation <- function(mm, um, mu, uu) {
##     return (( (2 * mm) + um + mu) / (2* ( mm + um + mu + uu) ))
## }

get_methylation <- function(d, tuple_length) {
    get_methylation_tuple_length_aware(d, tuple_length)
}

# @deprecated does not work with tuple length >= 2 
get_tuple_name <- function(mm, um, mu, uu) {
    sprintf('MM%s;MU%s;UM%s;UU%s', mm, mu, um, uu)
}

fill_positions_no_gaps <- function(num_tuples, tuple_distance) {

    pos <- seq(from = 1, length.out = num_tuples, by = tuple_distance)
    
    end <- c(pos[2:length(pos)], pos[length(pos)] + tuple_distance)

    return(list(start = pos, end = end))
}

build_tuple_scaffold <- function(num_tuples, tuple_length) {

    ids <- get_tuple_identifiers(tuple_length)
    tmp <- as.data.frame(matrix(nrow = num_tuples, ncol = 6 + length(ids), data = NA))
    colnames(tmp) <- c('chrom', 'start', 'end', ids, 'entropy',
                       'strand', 'meth')
    tmp
}

dissect_tuple_name <- function(count_string) {

    cnames <- strsplit(gsub("[0-9]+", "", as.character(count_string)), ';')[[1]]
    tmp <- strsplit(gsub("[A-Z]+", "", as.character(count_string)), ';')

    tmp <- data.frame(matrix(unlist(tmp), nrow=length(count_string), byrow = TRUE))
    
    ## rownames(tmp) <- count_string
    colnames(tmp) <- cnames
    tmp <- sapply(tmp, function(x) as.numeric(as.character(x)))
    tmp
}

get_coverage <- function(count_string) {
    tmp <-  dissect_tuple_name(count_string)
    rowSums(tmp)
}


plot_simulation <- function(d) {
    d$coverage <- get_coverage(d$id)
    d$log10_distance <- log10(d$end - d$start)

    p1 <- ggplot(d, aes (x = coverage, y = meth, col = entropy)) +
        geom_point() + theme_bw(15)
    
    p1 <- ggExtra::ggMarginal(p1,
                              type = 'density',
                              margins = 'both',
                              size = 5)


    p1
    
    p2 <- ggplot(d, aes (x = entropy, y = meth, col = entropy)) +
        geom_point() + theme_bw(15)
    
    p2 <- ggExtra::ggMarginal(p2,
                              type = 'density',
                              margins = 'both',
                              size = 5)

    p2 
}

```

```{r}

write_simulation <- function(d, filename) {    
    write.table(x = d, file = gzfile(filename),
                row.names = FALSE, col.names = FALSE, quote = FALSE,
                sep = '\t')

}

```


# Simulation scenarios

Data structure:

```
chr, cpg1, cpg2, tuples, entropies, strand, methylation
```

```
18	10862	10870	MM50;MU0;UM0;UU0	0	+	1
18	10870	10876	MM50;MU0;UM0;UU0	0	+	1
18	10876	10888	MM49;MU0;UM0;UU0	0	+	1
18	48387	48396	MM57;MU0;UM38;UU0	0.673012	-	0.8
18	48396	48398	MM59;MU37;UM0;UU0	0.666654	-	0.807292
18	48398	48412	MM50;MU9;UM37;UU0	0.929138	-	0.760417
18	48412	48434	MM87;MU0;UM9;UU0	0.311129	-	0.953125
```

```{r}
sims <- list()
```

## Fully methylated

⏺⏺

⏺⏺

⏺⏺

⏺⏺


```{r}

## no gaps between tuples
simulate_full_meth <- function(chrom, num_tuples, average_depth, std_depth, tuple_distance, tuple_length) {
    d <- build_tuple_scaffold(num_tuples, tuple_length)
    d[,'chrom'] <- chrom
    
    ## one sided, rounded normal distribution for coverages
    set.seed(1)
    ## d[,'MM'] <- round(abs(rnorm(n = num_tuples,
    ##                           mean = average_depth,
    ##                           sd = std_depth)))
    ## d[d$MM == 0, 'MM'] <- average_depth

    d[,paste(rep('M', tuple_length), collapse = '')]  <- round(truncnorm::rtruncnorm(
        num_tuples, a = 1, b = Inf,
        mean = average_depth,
        sd = std_depth))

    
    foo <- fill_positions_no_gaps(num_tuples = num_tuples,
                                  tuple_distance = tuple_distance)

    d$start <- foo$start
    d$end <- foo$end

    ## at random
    set.seed(33)
    d[,'strand'] <- sample(c('+', '-'), size = num_tuples, replace = TRUE)

    unmethylated <- grep('U', names(d))
    d[,unmethylated] <- 0

    ## d[,'meth'] <- get_methylation(mm = d$MM, um = d$UM,
    d[,'meth'] <- get_methylation_tuple_length_aware(d, tuple_length)

    counts <- count_m_and_u(colnames(d))
    tuples <- colnames(d)[(counts$m + counts$u) > 0]
    
    d[, 'entropy'] <- apply(d[tuples], 1, entropy)
    
    d$id <- '.' # get_tuple_name(mm = d$MM, um = d$UM, mu = d$MU, uu = d$UU)
    
    d[c('chrom', 'start', 'end', 'id', 'entropy', 'strand', 'meth' )]
                    
}

```

```{r}
sims$meth <- simulate_full_meth(chrom = 'chr19', num_tuples = 10000,
                                average_depth = 10,
                                tuple_length = 5,
                                std_depth = 5,
                                tuple_distance = 20 * 5)

summary(sims$meth)

head(sims$meth)
## hist(get_coverage(sims$meth$id))
## lines(density(get_coverage(sims$meth$id)), lwd = 2)

```

## Fully unmethylated

○○

○○

○○

○○

```{r}

## chrom <- 'chrx'
## num_tuples <- 100
## average_depth <- 30
## std_depth <- 5
## tuple_distance <- 2
## tuple_length <- 5

simulate_full_unmeth <- function(chrom, num_tuples, average_depth, std_depth, tuple_distance,
                                 tuple_length) {
    d <- build_tuple_scaffold(num_tuples, tuple_length)
    d[,'chrom'] <- chrom
    
    ## one sided, rounded normal distribution for coverages
    set.seed(1)
    d[,paste(rep('U', tuple_length), collapse = '')] <- round(
        truncnorm::rtruncnorm(num_tuples, a = 1, b = Inf,
                              mean = average_depth,
                              sd = std_depth))

    foo <- fill_positions_no_gaps(num_tuples = num_tuples,
                                  tuple_distance = tuple_distance)

    d$start <- foo$start
    d$end <- foo$end
    
    ## at random
    set.seed(4)
    d[,'strand'] <- sample(c('+', '-'), size = num_tuples, replace = TRUE)

    methylated <- grep('M', names(d))
                       
    
    d[,methylated] <- 0

    ## here
    ## d[,'meth'] <- get_methylation(mm = d$MM, um = d$UM,
    ##                                 mu = d$MU, uu = d$UU)
    d[,'meth'] <- get_methylation_tuple_length_aware(d, tuple_length)

    counts <- count_m_and_u(colnames(d))
    tuples <- colnames(d)[(counts$m + counts$u) > 0]
    
    d[, 'entropy'] <- apply(d[tuples],
                              1,
                              entropy)
    
    d$id <- '.' # get_tuple_name(mm = d$MM, um = d$UM, mu = d$MU, uu = d$UU)
    
    d[c('chrom', 'start', 'end', 'id', 'entropy', 'strand', 'meth' )]            
}


```

```{r}
sims$unmeth <- simulate_full_unmeth(chrom = 'chr19',
                                    num_tuples = 10000,
                                    tuple_length = 8,
                                    average_depth = 10,
                                    std_depth = 5,
                                    tuple_distance = 20*8)

summary(sims$unmeth)

plot_simulation(sims$unmeth)

## hist(get_coverage(sims$unmeth$id))
## lines(density(get_coverage(sims$unmeth$id)), lwd = 2)
```



## Two haplotypes, half methylation, ordered

⏺⏺

⏺⏺

○○

○○


```{r}
simulate_two_haplotypes_lowest_entropy <- function(chrom, num_tuples, average_depth,
                                                   std_depth, tuple_distance, tuple_length) {

    dmeth <- simulate_full_meth(chrom = chrom,
                                num_tuples = round(num_tuples/2),
                                average_depth = average_depth,
                                std_depth = std_depth,
                                tuple_distance = tuple_distance,
                                tuple_length =  tuple_length)
    ## print(head(dmeth))

    dunmeth <-  simulate_full_unmeth(chrom = chrom,
                                     num_tuples = round(num_tuples/2),
                                     average_depth = average_depth,
                                     std_depth = std_depth,
                                     tuple_distance = tuple_distance,
                                     tuple_length =  tuple_length)


    ## print(head(dunmeth))
    
    d <- rbind(dmeth, dunmeth)

    d <- d[order(d$start),]
    d
    ## return(list(d = d, dmeth = dmeth, dunmeth = dunmeth))
}

```


```{r, eval = FALSE}
sims$imprint <- simulate_two_haplotypes_lowest_entropy(chrom = 'chr19', num_tuples = 10000,
                               average_depth = 10,
                               std_depth = 5,
                               tuple_distance = 20,
                               tuple_length=4)

summary(sims$imprint)

head(sims$imprint)
# hist(get_coverage(sims$imprint$id))
#lines(density(get_coverage(sims$imprint$id)), lwd = 2)

```


## Infinite haplotypes, max entropy

⏺○

○⏺

○○

⏺⏺


```{r entropic, eval = TRUE}

## chrom <- 'chrx'
## num_tuples <- 100
## average_depth <- 30
## std_depth <- 5
## tuple_distance <- 2
## tuple_length <- 5

simulate_infinite_haplotypes <- function(chrom, num_tuples, average_depth,
                                         std_depth, tuple_distance, tuple_length) {

    d <- build_tuple_scaffold(num_tuples = num_tuples, tuple_length = tuple_length)
    d[,'chrom'] <- chrom
    
    ## one sided, rounded normal distribution for coverages
    set.seed(1)

    tuples <- get_tuple_identifiers(tuple_length = tuple_length)
    for (tuple in tuples) {
        d[,tuple] <- round(truncnorm::rtruncnorm(num_tuples, a = 0, b = Inf,
                                                mean = average_depth,
                                                sd = std_depth))
    }

    head(d)
    foo <- fill_positions_no_gaps(num_tuples = num_tuples,
                                  tuple_distance = tuple_distance)    
    
    d$start <- foo$start
    d$end <- foo$end
    
    d[,'meth'] <- get_methylation_tuple_length_aware(d, tuple_length)
    
    d[, 'entropy'] <- apply(d[tuples], 1, entropy)
    ## at random
    set.seed(4)
    d[,'strand'] <- sample(c('+', '-'), size = num_tuples, replace = TRUE)

    
    d$id <- '.' # get_tuple_name(mm = d$MM, um = d$UM, mu = d$MU, uu = d$UU)
    
    d[c('chrom', 'start', 'end', 'id', 'entropy', 'strand', 'meth' )]    
    
}
```

```{r untested2, eval = TRUE}
sims$entropic <- simulate_infinite_haplotypes(chrom = 'chr19',
                                              num_tuples = 10000,
                                              average_depth = 10,
                                              std_depth = 10,
                                              tuple_length = 10,
                                              tuple_distance = 20)

summary(sims$entropic)

head(sims$entropic)

## hist(get_coverage(sims$entropic$id))
## lines(density(get_coverage(sims$entropic$id)), lwd = 2)

hist(sims$entropic$meth)
hist(sims$entropic$entropy)

plot_simulation(sims$entropic)

```



## Boundary simple


⏺○

⏺○

⏺○

⏺○


```{r, eval = TRUE}
simulate_boundary <- function(chrom, num_tuples, average_depth,
                                         std_depth, tuple_distance, tuple_length) {

    d <- build_tuple_scaffold(num_tuples = num_tuples, tuple_length = tuple_length)
    d[,'chrom'] <- chrom

    ## get midpoint
    tuples <- get_tuple_identifiers(tuple_length = tuple_length)
    first_half <- round(tuple_length/2)
    second_half <- tuple_length - first_half

    boundary <- paste(c(rep('U', first_half), rep('M', second_half)), collapse = '')
   
    ## one sided, rounded normal distribution for coverages
    set.seed(88)
    
    d[,boundary] <- round(truncnorm::rtruncnorm(num_tuples, a = 1, b = Inf,
                                            mean = average_depth,
                                            sd = std_depth))
    
    d[,setdiff(tuples, boundary)] <- 0

    foo <- fill_positions_no_gaps(num_tuples = num_tuples,
                                  tuple_distance = tuple_distance)    

    d$start <- foo$start
    d$end <- foo$end

    d[,'meth'] <- get_methylation(d, tuple_length = tuple_length)

    d[, 'entropy'] <- apply(d[tuples], 1, entropy)
    
    ## at random
    set.seed(4)
    d[,'strand'] <- sample(c('+', '-'), size = num_tuples, replace = TRUE)

    
    
    d$id <- '.' # get_tuple_name(mm = d$MM, um = d$UM, mu = d$MU, uu = d$UU)
    
    d[c('chrom', 'start', 'end', 'id', 'entropy', 'strand', 'meth' )]    
    
}

```


```{r}
sims$boundary <- simulate_boundary(chrom = 'chr19', num_tuples = 10000,
                                   average_depth = 10,
                                   std_depth = 5,
                                   tuple_distance = 20,
                                   tuple_length = 4)

summary(sims$boundary)

head(sims$boundary)

hist(get_coverage(sims$boundary$id))
lines(density(get_coverage(sims$boundary$id)), lwd = 2)

hist(sims$boundary$meth)
hist(sims$boundary$entropy)

plot_simulation(sims$boundary)

```


## Boundary opposed


⏺○

⏺○

○⏺

○⏺


```{r}
simulate_opposed_boundary <- function(chrom, num_tuples, average_depth,
                                         std_depth, tuple_distance, tuple_length) {
    d <- build_tuple_scaffold(num_tuples = num_tuples, tuple_length = tuple_length)
    d[,'chrom'] <- chrom

    ## get midpoint
    tuples <- get_tuple_identifiers(tuple_length = tuple_length)
    first_half <- round(tuple_length/2)
    second_half <- tuple_length - first_half

    boundary <- paste(c(rep('U', first_half), rep('M', second_half)), collapse = '')
   
    ## one sided, rounded normal distribution for coverages
    set.seed(88)
    
    d[,boundary] <- round(truncnorm::rtruncnorm(num_tuples/2, a = 1, b = Inf,
                                            mean = average_depth,
                                            sd = std_depth))

    rev_boundary <- paste(c(rep('M', first_half), rep('U', second_half)), collapse = '')

    d[,rev_boundary] <- d[,boundary]

    d[,setdiff(tuples, c(boundary, rev_boundary))] <- 0
    
    foo <- fill_positions_no_gaps(num_tuples = num_tuples,
                                  tuple_distance = tuple_distance)    

    d$start <- foo$start
    d$end <- foo$end

    ## at random
    set.seed(4)
    d[,'strand'] <- sample(c('+', '-'), size = num_tuples, replace = TRUE)

    
    d[,'meth'] <- get_methylation(d, tuple_length = tuple_length)
    d[, 'entropy'] <- apply(d[tuples], 1, entropy)

    d$id <- '.'
    
    d[c('chrom', 'start', 'end', 'id', 'entropy', 'strand', 'meth' )]    
    
}

```


```{r}
sims$opposed_boundary <- simulate_opposed_boundary(chrom = 'chr19', num_tuples = 10000,
                                         average_depth = 10,
                                         std_depth = 5,
                                         tuple_length = 5,
                                         tuple_distance = 20)

summary(sims$opposed_boundary)

head(sims$opposed_boundary)

plot_simulation(sims$opposed_boundary)

```

# Diffuse boundary, from ordered to disordered

```{r}
simulate_disordered_gradient <- function(chrom, num_tuples, average_depth,
                                         std_depth, tuple_distance, tuple_length) {
    ## d <- build_tuple_scaffold(num_tuples = num_tuples, tuple_length = tuple_length)
    ## d[,'chrom'] <- chrom

    ## ## get midpoint
    ## tuples <- get_tuple_identifiers(tuple_length = tuple_length)
    ## first_half <- round(tuple_length/2)
    ## second_half <- tuple_length - first_half

    ## boundary <- paste(c(rep('U', first_half), rep('M', second_half)), collapse = '')
   
    ## ## one sided, rounded normal distribution for coverages
    ## set.seed(88)
    
    ## d[,boundary] <- round(truncnorm::rtruncnorm(num_tuples/2, a = 1, b = Inf,
    ##                                         mean = average_depth,
    ##                                         sd = std_depth))

    ## rev_boundary <- paste(c(rep('M', first_half), rep('U', second_half)), collapse = '')

    ## d[,rev_boundary] <- d[,boundary]

    ## d[,setdiff(tuples, c(boundary, rev_boundary))] <- 0
    
    ## foo <- fill_positions_no_gaps(num_tuples = num_tuples,
    ##                               tuple_distance = tuple_distance)    

    ## d$start <- foo$start
    ## d$end <- foo$end

    ## ## at random
    ## set.seed(4)
    ## d[,'strand'] <- sample(c('+', '-'), size = num_tuples, replace = TRUE)

    
    ## d[,'meth'] <- get_methylation(d, tuple_length = tuple_length)
    ## d[, 'entropy'] <- apply(d[tuples], 1, entropy)

    ## d$id <- '.'
    
    ## d[c('chrom', 'start', 'end', 'id', 'entropy', 'strand', 'meth' )]    
    
}


```


## Write simulations

```{r}
dir.create('simulations', showWarnings = FALSE)

for (sim in names(sims)) {
    write_simulation(sims[sim],
                     filename = file.path('simulations', sprintf('%s.gz', sim)))
}
```

# Coverage simulations


# Todo

Simulating regions longer than 2 items?

Strand-specific simulations?

# Session

```{r sessionInfo, cache = 0}

date()
devtools::session_info()

```
