---
title: "Argelaguet 2019 mouse gastrulation: methylation entropy by feature (CpG context)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Izaskun Mallona"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
---

```{r logging_early, include = FALSE}
log <- file(snakemake@log$log, open = "wt")
sink(log)
sink(log, type = "message")
```

```{r setup}
suppressPackageStartupMessages({
  library(readr)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(BiocParallel)
  library(ggrepel)
  library(ggridges)
  library(ComplexHeatmap)
  library(circlize)
  library(uwot)
  library(irlba)
})

source("src/ggtheme.R")
source("src/embedding_utils.R")
param <- MulticoreParam(workers = snakemake@threads)
```

```{r opts}
knitr::opts_chunk$set(
  echo       = TRUE,
  fig.width  = 5,
  fig.height = 5,
  cache      = FALSE,
  include    = TRUE,
  fig.path   = "argelaguet_plots/",
  dev        = c("png", "svg"),
  dpi        = 500,
  cache.lazy = FALSE,
  warning    = TRUE,
  message    = TRUE
)
```

```{r git_commit, echo = FALSE, results = "asis"}
git_sha <- tryCatch(
  system("git rev-parse --short HEAD", intern = TRUE, ignore.stderr = TRUE),
  error = function(e) "unknown"
)
cat(sprintf("Git commit: `%s`\n\n", git_sha))
```

```{r save_snmk_debug, include = FALSE}
saveRDS(file = "snmk_argelaguet_features.rds", object = snakemake, compress = "xz")
```

Mouse (mm10 / GRCm38), CpG context. chr10 only: `r snakemake@params[["chr10_only"]]`.
TET TKO cells (Plate11-16) excluded. Only cells passing methylation QC included.

```{r helpers}
get_annotation <- function(fn) sub("^(.*)_[^_]+_[^_]+\\..*$", "\\1", fn)
get_stage      <- function(fn) sub("^.*_([^_]+)_[^_]+\\..*$", "\\1", fn)
get_lineage    <- function(fn) sub("^.*_([^_.]+)\\.[^_]*$",   "\\1", fn)

ann_labels <- c(
  "genes"            = "Genes",
  "promoters"        = "Promoters",
  "lines"            = "LINEs",
  "sines"            = "SINEs",
  "h3k4me3"          = "H3K4me3 (ENCODE)",
  "h3k9me3"          = "H3K9me3 (ENCODE)",
  "h3k27me3"         = "H3K27me3 (ENCODE)",
  "h3k4me1"          = "H3K4me1 (ENCODE)",
  "h3k27ac"          = "H3K27ac (ENCODE)",
  ## gastrulation-specific from GSE125318 (E7.5 ChIP-seq)
  "enh-E75-Ect"      = "Enh E7.5 Ect",
  "enh-E75-End"      = "Enh E7.5 End",
  "enh-E75-Mes"      = "Enh E7.5 Mes",
  "enh-E75-union"    = "Enh E7.5 union",
  "h3k4me3-E75-Ect"      = "H3K4me3 E7.5 Ect",
  "h3k4me3-E75-End"      = "H3K4me3 E7.5 End",
  "h3k4me3-E75-Mes"      = "H3K4me3 E7.5 Mes",
  "h3k4me3-E75-common"   = "H3K4me3 E7.5 common",
  ## ESC marks from ENCODE
  "esc-p300"         = "ESC p300",
  "esc-dhs"          = "ESC DHS"
)

harmonize_levels <- function(df) {
  missing_ann <- setdiff(unique(df$annotation), names(ann_labels))
  if (length(missing_ann) > 0)
    warning("unknown annotations: ", paste(missing_ann, collapse = ", "))
  df$annotation <- factor(df$annotation,
                           levels = names(ann_labels),
                           labels = unname(ann_labels))
  if ("stage" %in% names(df))
    df$stage <- factor(df$stage, levels = sort(unique(df$stage)))
  df
}

## Sanitize metadata strings the same way the smk does for wildcard values.
sanitize <- function(x) gsub("[ ._]", "-", x)
```

```{r load_metadata}
meta_raw <- read_tsv(snakemake@params[["meta_path"]],
                     na = "NA", show_col_types = FALSE)

meta_cells <- meta_raw %>%
  mutate(
    stage_san   = sanitize(stage),
    lineage_san = sanitize(lineage10x)
  )

meta_grp <- meta_cells %>%
  group_by(stage_san, lineage_san) %>%
  summarise(
    stage         = first(stage),
    lineage       = first(lineage10x),
    lineage_class = {
      vals <- lineage10x_2[!is.na(lineage10x_2) & lineage10x_2 != ""]
      if (length(vals) == 0) NA_character_ else first(vals)
    },
    n_meta        = n(),
    .groups       = "drop"
  )

cat("Groups (stage x lineage):", nrow(meta_grp), "\n")
cat("Stages:", paste(sort(unique(meta_grp$stage)), collapse = ", "), "\n")
cat("Lineage classes:", paste(sort(unique(meta_grp$lineage_class)), collapse = ", "), "\n")
```

# S - sample entropy {.tabset .tabset-pills}

Per-cell sample entropy (S) from `.out.gz` files.

```{r import_sampen}
short_paths <- list.files(snakemake@params[["output_path"]],
                           pattern = "\\.out\\.gz$", full.names = TRUE)
short_paths <- short_paths[!grepl("\\.det\\.out\\.gz$",  short_paths)]
short_paths <- short_paths[!grepl("\\.meth\\.out\\.gz$", short_paths)]

read_short <- function(fp) {
  fn <- basename(fp)
  dt <- read_tsv(fp, na = "-1", show_col_types = FALSE)
  dt <- dt[, c("sampen", "avg_meth")]
  dt$annotation  <- get_annotation(fn)
  dt$stage_san   <- get_stage(fn)
  dt$lineage_san <- get_lineage(fn)
  dt
}

sampens <- do.call(rbind, bplapply(short_paths, read_short, BPPARAM = param))
sampens <- harmonize_levels(sampens)
sampens <- sampens %>%
  left_join(meta_grp %>% select(stage_san, lineage_san, stage, lineage,
                                 lineage_class),
            by = c("stage_san", "lineage_san"))

cat("cell_class NA:", sum(is.na(sampens$lineage_class)), "/", nrow(sampens), "\n")
```

## S vs methylation

```{r scatter_sampen, fig.width = ng_fig_size(5, 9)$w, fig.height = ng_fig_size(5, 9)$h}
ggplot(sampens %>% filter(!is.na(lineage_class)),
       aes(x = avg_meth, y = sampen, color = lineage_class)) +
  geom_point(size = 0.3, alpha = 0.15) +
  scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95,
                         na.value = "grey70") +
  facet_grid(annotation ~ stage) +
  labs(x = "average methylation", y = "sample entropy (S)",
       color = "lineage class") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng()
```

## S distributions

```{r violin_sampen, fig.width = ng_fig_size(4, 9)$w, fig.height = ng_fig_size(4, 9)$h}
ggplot(sampens %>% filter(!is.na(lineage_class)),
       aes(x = lineage_class, y = sampen, fill = lineage_class)) +
  geom_violin(alpha = 0.5, draw_quantiles = 0.5) +
  scale_fill_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
  facet_grid(annotation ~ stage, scales = "free_y") +
  labs(y = "sample entropy (S)") +
  theme_ng() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

# adjS - methylation-adjusted sample entropy {.tabset .tabset-pills}

Per-cell normalised sample entropy from `.out.gz` files.

```{r import_adjS}
read_adjS <- function(fp) {
  fn <- basename(fp)
  dt <- read_tsv(fp, na = "-1", show_col_types = FALSE)
  dt <- dt[, c("sampen_norm", "avg_meth")]
  colnames(dt) <- c("sampen", "avg_meth")
  dt$annotation  <- get_annotation(fn)
  dt$stage_san   <- get_stage(fn)
  dt$lineage_san <- get_lineage(fn)
  dt
}

adjsampens <- do.call(rbind, bplapply(short_paths, read_adjS, BPPARAM = param))
adjsampens <- harmonize_levels(adjsampens)
adjsampens <- adjsampens %>%
  left_join(meta_grp %>% select(stage_san, lineage_san, stage, lineage,
                                 lineage_class),
            by = c("stage_san", "lineage_san"))
```

## adjS vs methylation

```{r scatter_adjS, fig.width = ng_fig_size(5, 9)$w, fig.height = ng_fig_size(5, 9)$h}
ggplot(adjsampens %>% filter(!is.na(lineage_class)),
       aes(x = avg_meth, y = sampen, color = lineage_class)) +
  geom_point(size = 0.3, alpha = 0.15) +
  scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95,
                         na.value = "grey70") +
  facet_grid(annotation ~ stage) +
  labs(x = "average methylation", y = "adj. sample entropy (adjS)",
       color = "lineage class") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng()
```

## adjS distributions

```{r ridges_adjS, fig.width = ng_fig_size(4, 9)$w, fig.height = ng_fig_size(4, 9)$h}
ggplot(adjsampens %>% filter(!is.na(lineage_class)),
       aes(x = sampen, y = lineage, fill = lineage_class)) +
  geom_density_ridges(alpha = 0.6, scale = 0.85) +
  scale_fill_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
  facet_grid(annotation ~ stage, scales = "free") +
  labs(x = "adj. sample entropy (adjS)") +
  theme_ng() +
  theme(legend.position = "none")
```

# H - Shannon entropy medians {.tabset .tabset-pills}

Per-locus Shannon entropy (H) from `.det.out.gz`. One median per group per annotation.

```{r import_shannon}
det_paths <- list.files(snakemake@params[["output_path"]],
                         pattern = "det\\.out\\.gz$", full.names = TRUE)
det_paths <- det_paths[!grepl("norm", det_paths)]

read_shannon <- function(fp) {
  fn  <- basename(fp)
  tmp <- read_tsv(fp, na = "-1", show_col_types = FALSE, n_max = 10e4)
  tibble(
    file            = fn,
    median_shannon  = median(tmp$shannon,  na.rm = TRUE),
    median_avg_meth = median(tmp$avg_meth, na.rm = TRUE),
    annotation      = get_annotation(fn),
    stage_san       = get_stage(fn),
    lineage_san     = get_lineage(fn)
  )
}

shannons <- bplapply(det_paths, read_shannon, BPPARAM = param) |> bind_rows()
shannons <- harmonize_levels(shannons)
shannons <- shannons %>%
  left_join(meta_grp %>% select(stage_san, lineage_san, stage, lineage,
                                 lineage_class),
            by = c("stage_san", "lineage_san"))
```

## Median H vs methylation

```{r scatter_shannon, fig.width = ng_fig_size(4, 9)$w, fig.height = ng_fig_size(4, 9)$h}
ggplot(shannons %>% filter(!is.na(lineage_class)),
       aes(x = median_avg_meth, y = median_shannon, color = lineage_class)) +
  geom_point(size = 1.5, alpha = 0.8) +
  scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95,
                         na.value = "grey70") +
  facet_grid(annotation ~ stage) +
  labs(x = "median avg. methylation", y = "median Shannon entropy (H)",
       color = "lineage class") +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  theme_ng()
```

# adjH - methylation-adjusted Shannon entropy {.tabset .tabset-pills}

Per-locus normalised Shannon entropy from `.norm.det.out.gz`.

```{r import_adjH}
norm_det_paths <- list.files(snakemake@params[["output_path"]],
                              pattern = "norm\\.det\\.out\\.gz$",
                              full.names = TRUE)

read_adjH <- function(fp) {
  fn  <- basename(fp)
  tmp <- read_tsv(fp, na = "-1", show_col_types = FALSE, n_max = 10e4)
  tibble(
    file            = fn,
    median_shannon  = median(tmp$shannon_norm, na.rm = TRUE),
    median_avg_meth = median(tmp$avg_meth,     na.rm = TRUE),
    annotation      = get_annotation(fn),
    stage_san       = get_stage(fn),
    lineage_san     = get_lineage(fn)
  )
}

adjshannons <- bplapply(norm_det_paths, read_adjH, BPPARAM = param) |> bind_rows()
adjshannons <- harmonize_levels(adjshannons)
adjshannons <- adjshannons %>%
  left_join(meta_grp %>% select(stage_san, lineage_san, stage, lineage,
                                 lineage_class),
            by = c("stage_san", "lineage_san"))
```

## Median adjH vs methylation

```{r scatter_adjH, fig.width = ng_fig_size(4, 9)$w, fig.height = ng_fig_size(4, 9)$h}
ggplot(adjshannons %>% filter(!is.na(lineage_class)),
       aes(x = median_avg_meth, y = median_shannon, color = lineage_class)) +
  geom_point(size = 1.5, alpha = 0.8) +
  scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95,
                         na.value = "grey70") +
  facet_grid(annotation ~ stage) +
  labs(x = "median avg. methylation", y = "median adj. Shannon entropy (adjH)",
       color = "lineage class") +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  theme_ng()
```

# Assembly {.tabset .tabset-pills}

S and adjS are per-cell. H and adjH are medians across loci per group.

```{r group_medians}
saveRDS(
  list(sampens     = sampens,
       adjsampens  = adjsampens,
       shannons    = shannons,
       adjshannons = adjshannons),
  file = "argelaguet_entropy.rds"
)

sampen_grp <- sampens %>%
  group_by(annotation, stage_san, lineage_san) %>%
  summarise(median_S    = median(sampen,   na.rm = TRUE),
            median_meth = median(avg_meth, na.rm = TRUE),
            .groups     = "drop")

adjS_grp <- adjsampens %>%
  group_by(annotation, stage_san, lineage_san) %>%
  summarise(median_adjS = median(sampen, na.rm = TRUE),
            .groups     = "drop")

all_grp <- sampen_grp %>%
  left_join(adjS_grp,
            by = c("annotation", "stage_san", "lineage_san")) %>%
  left_join(shannons %>% select(annotation, stage_san, lineage_san,
                                 median_H = median_shannon),
            by = c("annotation", "stage_san", "lineage_san")) %>%
  left_join(adjshannons %>% select(annotation, stage_san, lineage_san,
                                    median_adjH = median_shannon),
            by = c("annotation", "stage_san", "lineage_san"))

all_grp_meta <- all_grp %>%
  left_join(meta_grp, by = c("stage_san", "lineage_san"))

saveRDS(
  list(all_grp      = all_grp,
       meta_grp     = meta_grp,
       all_grp_meta = all_grp_meta),
  file = "argelaguet_groups_meta.rds"
)

cat("Groups with entropy data:", nrow(all_grp_meta), "\n")
```

## adjS vs adjH

```{r adjS_vs_adjH, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(4, 1)$h}
df_cmp <- all_grp_meta %>% filter(!is.na(lineage_class))
if (nrow(df_cmp) > 0) {
  ggplot(df_cmp, aes(x = median_adjH, y = median_adjS, color = annotation)) +
    geom_point(size = 1.5, alpha = 0.8) +
    scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
    facet_grid(. ~ lineage_class) +
    labs(x = "median adjH", y = "median adjS", color = "annotation") +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme_ng()
}
```

## S vs H

```{r S_vs_H, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(4, 1)$h}
if (nrow(df_cmp) > 0) {
  ggplot(df_cmp, aes(x = median_H, y = median_S, color = annotation)) +
    geom_point(size = 1.5, alpha = 0.8) +
    scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
    facet_grid(. ~ lineage_class) +
    labs(x = "median H", y = "median S", color = "annotation") +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme_ng()
}
```

# Time course {.tabset .tabset-pills}

Entropy trajectories across developmental stages. Each point is the median for one
(stage, lineage) group. The gastrulation time series spans E4.5 through E7.5, covering
the emergence of epiblast, primitive streak, and the three germ layers. Entropy at
enhancers and active chromatin marks is expected to rise in progenitor states and
fall upon lineage commitment.

```{r timecourse_data}
stage_order <- sort(unique(all_grp_meta$stage))
tc <- all_grp_meta %>%
  filter(!is.na(lineage_class), !is.na(stage)) %>%
  mutate(stage = factor(stage, levels = stage_order))
```

## adjS by stage and lineage class

```{r timecourse_adjS, fig.width = ng_fig_size(length(stage_order), 9)$w, fig.height = ng_fig_size(length(stage_order), 9)$h}
ggplot(tc, aes(x = stage, y = median_adjS,
               color = lineage_class, group = interaction(lineage_san, annotation))) +
  geom_line(alpha = 0.5, linewidth = 0.4) +
  geom_point(size = 1.2, alpha = 0.8) +
  scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
  facet_wrap(~ annotation, ncol = 3, scales = "free_y") +
  labs(x = "developmental stage", y = "median adjS",
       color = "lineage class") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1,
                                                   linewidth = 1))) +
  theme_ng()
```

## adjH by stage and lineage class

```{r timecourse_adjH, fig.width = ng_fig_size(length(stage_order), 9)$w, fig.height = ng_fig_size(length(stage_order), 9)$h}
ggplot(tc, aes(x = stage, y = median_adjH,
               color = lineage_class, group = interaction(lineage_san, annotation))) +
  geom_line(alpha = 0.5, linewidth = 0.4) +
  geom_point(size = 1.2, alpha = 0.8) +
  scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
  facet_wrap(~ annotation, ncol = 3, scales = "free_y") +
  labs(x = "developmental stage", y = "median adjH",
       color = "lineage class") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1,
                                                   linewidth = 1))) +
  theme_ng()
```

## Methylation by stage and lineage class

```{r timecourse_meth, fig.width = ng_fig_size(length(stage_order), 9)$w, fig.height = ng_fig_size(length(stage_order), 9)$h}
ggplot(tc, aes(x = stage, y = median_meth,
               color = lineage_class, group = interaction(lineage_san, annotation))) +
  geom_line(alpha = 0.5, linewidth = 0.4) +
  geom_point(size = 1.2, alpha = 0.8) +
  scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
  facet_wrap(~ annotation, ncol = 3, scales = "free_y") +
  labs(x = "developmental stage", y = "median avg. methylation",
       color = "lineage class") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1,
                                                   linewidth = 1))) +
  theme_ng()
```

# Enhancers vs promoters {.tabset .tabset-pills}

Comparing entropy at active enhancer marks (H3K27ac, H3K4me1) against promoter marks
(H3K4me3, Promoters). The gastrulation paper reports that meso- and endoderm-defining
enhancers undergo concerted demethylation upon germ-layer specification; adjS at
these marks captures the heterogeneity of that process. H3K27ac and H3K4me1
here come from ENCODE bulk mm10 data, not gastrulation-specific ChIP.

```{r enh_vs_prom_data}
enh_marks  <- c("Enh E7.5 Ect", "Enh E7.5 End", "Enh E7.5 Mes",
                "Enh E7.5 union", "ESC p300")
prom_marks <- c("H3K4me3 E7.5 common", "Promoters",
                "H3K4me3 (ENCODE)")

ep_df <- all_grp_meta %>%
  filter(annotation %in% c(enh_marks, prom_marks),
         !is.na(lineage_class)) %>%
  mutate(
    mark_type = case_when(
      annotation %in% enh_marks  ~ "enhancer",
      annotation %in% prom_marks ~ "promoter"
    ),
    stage = factor(stage, levels = stage_order)
  )
```

## adjS enhancers vs promoters across stage

```{r enh_vs_prom_adjS, fig.width = ng_fig_size(length(stage_order), 2)$w, fig.height = ng_fig_size(length(stage_order), 2)$h}
ggplot(ep_df, aes(x = stage, y = median_adjS,
                   color = lineage_class, shape = mark_type,
                   group = interaction(lineage_san, annotation))) +
  geom_line(alpha = 0.4, linewidth = 0.4) +
  geom_point(size = 1.5, alpha = 0.85) +
  scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
  scale_shape_manual(values = c("enhancer" = 16, "promoter" = 17)) +
  facet_grid(mark_type ~ lineage_class, scales = "free_y") +
  labs(x = "developmental stage", y = "median adjS",
       color = "lineage class", shape = "mark type") +
  guides(
    color = guide_legend(override.aes = list(size = 2.5, alpha = 1, linewidth = 1)),
    shape = guide_legend(override.aes = list(size = 2.5))
  ) +
  theme_ng()
```

## adjH enhancers vs promoters across stage

```{r enh_vs_prom_adjH, fig.width = ng_fig_size(length(stage_order), 2)$w, fig.height = ng_fig_size(length(stage_order), 2)$h}
ggplot(ep_df, aes(x = stage, y = median_adjH,
                   color = lineage_class, shape = mark_type,
                   group = interaction(lineage_san, annotation))) +
  geom_line(alpha = 0.4, linewidth = 0.4) +
  geom_point(size = 1.5, alpha = 0.85) +
  scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
  scale_shape_manual(values = c("enhancer" = 16, "promoter" = 17)) +
  facet_grid(mark_type ~ lineage_class, scales = "free_y") +
  labs(x = "developmental stage", y = "median adjH",
       color = "lineage class", shape = "mark type") +
  guides(
    color = guide_legend(override.aes = list(size = 2.5, alpha = 1, linewidth = 1)),
    shape = guide_legend(override.aes = list(size = 2.5))
  ) +
  theme_ng()
```

## Paired scatter: adjS enhancer vs adjS promoter

Each point is one (stage, lineage) group. Groups above the diagonal have higher
within-cell entropy at enhancers than at promoters.

```{r enh_prom_scatter, fig.width = ng_fig_size(2, 2)$w, fig.height = ng_fig_size(2, 2)$h}
ep_wide <- all_grp_meta %>%
  filter(annotation %in% c("Enh E7.5 union", "Promoters"),
         !is.na(lineage_class)) %>%
  select(stage_san, lineage_san, lineage_class, stage, annotation, median_adjS) %>%
  pivot_wider(names_from = annotation, values_from = median_adjS) %>%
  rename(adjS_enh = `Enh E7.5 union`, adjS_prom = Promoters) %>%
  mutate(stage = factor(stage, levels = stage_order))

if (all(c("adjS_enh", "adjS_prom") %in% names(ep_wide))) {
  ggplot(ep_wide, aes(x = adjS_prom, y = adjS_enh,
                       color = lineage_class, shape = stage)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                color = "grey60", linewidth = 0.4) +
    geom_point(size = 2, alpha = 0.8) +
    scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
    scale_shape_manual(values = seq_along(levels(ep_wide$stage))) +
    labs(x = "median adjS at promoters", y = "median adjS at E7.5 enhancers (union)",
         color = "lineage class", shape = "stage") +
    guides(
      color = guide_legend(override.aes = list(size = 3)),
      shape = guide_legend(override.aes = list(size = 3))
    ) +
    theme_ng()
}
```

# Variation across lineages {.tabset .tabset-pills}

SD and range of group medians per annotation, broken down by lineage class.

```{r variation_helpers}
safe_range <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  diff(range(x))
}
```

```{r variation_lineage, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(4, 1)$h}
var_lineage <- all_grp_meta %>%
  filter(!is.na(lineage_class)) %>%
  group_by(annotation, lineage_class) %>%
  summarise(
    adjS_sd    = sd(median_adjS, na.rm = TRUE),
    adjH_sd    = sd(median_adjH, na.rm = TRUE),
    adjS_range = safe_range(median_adjS),
    adjH_range = safe_range(median_adjH),
    .groups = "drop"
  )

if (any(!is.na(var_lineage$adjH_sd))) {
  ggplot(var_lineage, aes(x = adjH_sd, y = adjS_sd,
                           color = annotation, label = annotation)) +
    geom_point(size = 3) +
    geom_text_repel(size = 3) +
    scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
    facet_grid(. ~ lineage_class) +
    labs(x = "SD of adjH across groups",
         y = "SD of adjS across groups") +
    theme_ng()
}
```

# Driver categorization {.tabset .tabset-pills}

Whether entropy variation is driven within cells (adjS) or across cells (adjH),
based on variance across all (stage, lineage) groups. Upper-left points are within-cell
driven; lower-right are across-cell driven. The epiCHAOS paper found epiblast to have
the highest epigenetic heterogeneity; annotations that separate lineages cleanly
should appear lower-right here.

```{r drivers_sd, fig.width = ng_fig_size(3, 3)$w, fig.height = ng_fig_size(3, 3)$h}
var_all <- all_grp_meta %>%
  group_by(annotation) %>%
  summarise(
    adjS_sd    = sd(median_adjS, na.rm = TRUE),
    adjH_sd    = sd(median_adjH, na.rm = TRUE),
    adjS_range = safe_range(median_adjS),
    adjH_range = safe_range(median_adjH),
    .groups = "drop"
  )

if (any(!is.na(var_all$adjH_sd))) {
  shannon_sd_thr <- quantile(var_all$adjH_sd, 0.3, na.rm = TRUE)
  sampen_sd_thr  <- quantile(var_all$adjS_sd, 0.3, na.rm = TRUE)

  var_all <- var_all %>%
    mutate(
      driver = case_when(
        is.na(adjH_sd) | is.na(adjS_sd)                    ~ NA_character_,
        adjH_sd < shannon_sd_thr & adjS_sd < sampen_sd_thr ~ "not changing enough",
        adjH_sd >= adjS_sd * 1.5                            ~ "across-cell driven",
        adjS_sd >= adjH_sd * 1.5                            ~ "within-cell driven",
        TRUE                                                 ~ "both"
      )
    )

  cat("Driver table (SD-based):\n")
  print(table(var_all$driver, useNA = "always"))

  ggplot(var_all, aes(x = adjH_sd, y = adjS_sd,
                       color = driver, shape = driver, label = annotation)) +
    geom_point(size = 3) +
    geom_text_repel() +
    scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95,
                           na.value = "grey60") +
    labs(x = "SD of adjH across all groups",
         y = "SD of adjS across all groups") +
    theme_ng()
}
```

## R-squared by lineage class

Variance in group-level adjH and adjS explained by lineage class (R2). Upper-left =
within-cell driven; lower-right = across-cell (lineage-driven).

```{r drivers_r2, fig.width = ng_fig_size(3, 3)$w, fig.height = ng_fig_size(3, 3)$h}
safe_r2 <- function(y, grp) {
  tryCatch({
    r <- summary(lm(y ~ grp))$r.squared
    if (is.nan(r) || is.na(r)) 0 else r
  }, error = function(e) NA_real_)
}

r2_by_ann <- all_grp_meta %>%
  filter(!is.na(lineage_class)) %>%
  group_by(annotation) %>%
  summarise(
    r2_adjH = safe_r2(median_adjH, lineage_class),
    r2_adjS = safe_r2(median_adjS, lineage_class),
    .groups = "drop"
  )

if (any(!is.na(r2_by_ann$r2_adjH))) {
  r2_thr <- 0.3
  r2_by_ann <- r2_by_ann %>%
    mutate(
      driver_r2 = case_when(
        is.na(r2_adjH) | is.na(r2_adjS)              ~ NA_character_,
        r2_adjH < r2_thr & r2_adjS < r2_thr          ~ "not explained by lineage",
        r2_adjH >= r2_adjS * 1.5 & r2_adjH >= r2_thr ~ "across-cell driven",
        r2_adjS >= r2_adjH * 1.5 & r2_adjS >= r2_thr ~ "within-cell driven",
        TRUE                                           ~ "both"
      )
    )

  cat("Driver table (R2-based):\n")
  print(table(r2_by_ann$driver_r2, useNA = "always"))

  ggplot(r2_by_ann, aes(x = r2_adjH, y = r2_adjS,
                         color = driver_r2, shape = driver_r2,
                         label = annotation)) +
    geom_vline(xintercept = r2_thr, linetype = "dashed",
               color = "grey60") +
    geom_hline(yintercept = r2_thr, linetype = "dashed",
               color = "grey60") +
    geom_point(size = 3) +
    geom_text_repel(size = 3) +
    scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95,
                           na.value = "grey60") +
    scale_x_continuous(limits = c(0, 1)) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(x = "R2 of adjH explained by lineage class",
         y = "R2 of adjS explained by lineage class",
         color = "driver", shape = "driver") +
    theme_ng()
}
```

# Heatmaps {.tabset .tabset-pills}

Rows = lineage (clustered). Columns = annotation. Row bars show lineage class.

```{r heatmap_helpers}
make_heatmap <- function(df, value_col, title,
                          palette = c("navy", "white", "firebrick")) {
  wide <- df %>%
    select(lineage, annotation, lineage_class,
           value = all_of(value_col)) %>%
    filter(!is.na(lineage_class)) %>%
    group_by(lineage, annotation, lineage_class) %>%
    summarise(value = median(value, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = annotation, values_from = value)

  if (nrow(wide) < 2) {
    message("Too few rows for heatmap")
    return(invisible(NULL))
  }

  mat <- wide %>% select(-lineage, -lineage_class) %>% as.matrix()
  rownames(mat) <- wide$lineage

  breaks <- c(min(mat, na.rm = TRUE),
              mean(range(mat, na.rm = TRUE)),
              max(mat, na.rm = TRUE))
  if (any(!is.finite(breaks)) || breaks[1] == breaks[3]) {
    message("Skipping heatmap: degenerate data (all-NA or constant matrix)")
    return(invisible(NULL))
  }

  lc_levels <- sort(unique(wide$lineage_class))
  lc_pal    <- setNames(viridisLite::turbo(length(lc_levels),
                                            begin = 0.05, end = 0.95),
                         lc_levels)

  row_ann <- rowAnnotation(
    lineage_class = wide$lineage_class,
    col = list(lineage_class = lc_pal),
    annotation_legend_param = list(
      lineage_class = list(title = "lineage class")
    )
  )

  col_fun <- colorRamp2(breaks, palette)

  Heatmap(mat,
          name              = title,
          col               = col_fun,
          row_split         = wide$lineage_class,
          cluster_rows      = TRUE,
          cluster_columns   = FALSE,
          show_row_names    = TRUE,
          show_column_names = TRUE,
          row_names_gp      = gpar(fontsize = 7),
          right_annotation  = row_ann,
          column_title      = title,
          na_col            = "grey90")
}
```

## adjS

```{r heatmap_adjS, fig.width = 10, fig.height = 8}
heat_df_adjs <- adjsampens %>%
  group_by(annotation, stage_san, lineage_san, lineage, lineage_class) %>%
  summarise(median_adjS = median(sampen, na.rm = TRUE), .groups = "drop")

make_heatmap(heat_df_adjs, "median_adjS", "median adjS",
             palette = c("navy", "white", "firebrick"))
```

## adjH

```{r heatmap_adjH, fig.width = 10, fig.height = 8}
heat_df_adjh <- adjshannons %>%
  group_by(annotation, stage_san, lineage_san, lineage, lineage_class) %>%
  summarise(median_adjH = median(median_shannon, na.rm = TRUE), .groups = "drop")

make_heatmap(heat_df_adjh, "median_adjH", "median adjH",
             palette = c("navy", "white", "firebrick"))
```

# Paper UMAP + our entropy overlay {.tabset .tabset-pills}

All cells with MOFA-derived UMAP coordinates from the original Argelaguet et al.
metaccrna analysis (`metaccrna/mofa/all_stages/umap_coordinates.txt`) placed at
their published positions and colored by the entropy or methylation values
computed here. H and adjH are group medians per (stage, lineage) broadcast onto
each cell of the group; S and adjS are per-cell values from the `.out.gz`
files. One panel per genomic annotation.

```{r paper_umap_load}
paper_umap <- read_csv(snakemake@params[["paper_umap_path"]],
                       show_col_types = FALSE)
colnames(paper_umap) <- c("sample", "umap1", "umap2")
cat("Cells in MOFA UMAP file:", nrow(paper_umap), "\n")
cat("Cells in argelaguet meta:", nrow(meta_cells), "\n")
cat("Overlap:",
    sum(meta_cells$sample %in% paper_umap$sample), "\n")
```

```{r paper_umap_data_grp}
cells_overlay_grp <- meta_cells %>%
  inner_join(paper_umap, by = "sample") %>%
  left_join(
    all_grp_meta %>% select(stage_san, lineage_san, annotation,
                            median_adjH, median_meth, median_adjS, median_S),
    by = c("stage_san", "lineage_san"),
    relationship = "many-to-many"
  ) %>%
  filter(!is.na(annotation))

saveRDS(cells_overlay_grp, "argelaguet_paper_umap_overlay.rds")
cat("Rows in group-level overlay (cells x annotations):",
    nrow(cells_overlay_grp), "\n")
```

## Group adjH per annotation

```{r paper_umap_adjH, fig.width = 14, fig.height = 8}
df_adjH <- cells_overlay_grp %>% filter(!is.na(median_adjH)) %>% droplevels()
if (nrow(df_adjH) > 0) {
  ggplot(df_adjH, aes(x = umap1, y = umap2, color = median_adjH)) +
    geom_point(size = 0.4, alpha = 0.6) +
    scale_color_viridis_c(option = "plasma", na.value = "grey80") +
    facet_wrap(~ annotation, ncol = 4) +
    labs(title = "Paper MOFA UMAP - color = group adjH",
         x = "UMAP1 (Argelaguet MOFA)", y = "UMAP2 (Argelaguet MOFA)",
         color = "median adjH") +
    theme_ng() +
    theme(axis.text = element_blank(), axis.ticks = element_blank())
} else {
  message("No adjH overlay data available")
}
```

## Group avg methylation per annotation

```{r paper_umap_meth, fig.width = 14, fig.height = 8}
df_meth <- cells_overlay_grp %>% filter(!is.na(median_meth)) %>% droplevels()
if (nrow(df_meth) > 0) {
  ggplot(df_meth, aes(x = umap1, y = umap2, color = median_meth)) +
    geom_point(size = 0.4, alpha = 0.6) +
    scale_color_viridis_c(option = "viridis", na.value = "grey80") +
    facet_wrap(~ annotation, ncol = 4) +
    labs(title = "Paper MOFA UMAP - color = group avg. methylation",
         x = "UMAP1 (Argelaguet MOFA)", y = "UMAP2 (Argelaguet MOFA)",
         color = "median meth") +
    theme_ng() +
    theme(axis.text = element_blank(), axis.ticks = element_blank())
} else {
  message("No methylation overlay data available")
}
```

## Per-cell adjS and S per annotation

Unlike H and adjH (group medians broadcast to all cells in a group), S and adjS
are per-cell values. The `.out.gz` files carry a `file` column with the
harmonized cell path; the basename minus `.gz` is the `id_met` cell identifier
which is also the MOFA UMAP sample key.

```{r paper_umap_per_cell_load}
read_short_with_file_arg <- function(fp) {
  fn <- basename(fp)
  dt <- read_tsv(fp, na = "-1", show_col_types = FALSE)
  dt <- dt[, c("file", "sampen", "sampen_norm", "avg_meth")]
  dt$annotation <- get_annotation(fn)
  dt
}

cells_sampen_paper <- do.call(
  rbind,
  bplapply(short_paths, read_short_with_file_arg, BPPARAM = param)
) %>%
  mutate(id_met = sub("\\.gz$", "", basename(file))) %>%
  inner_join(paper_umap, by = c("id_met" = "sample"))

cells_sampen_paper <- harmonize_levels(cells_sampen_paper)

cat("Per-cell rows joined to paper UMAP:",
    nrow(cells_sampen_paper), "\n")
cat("Unique cells with paper UMAP coords:",
    length(unique(cells_sampen_paper$id_met)), "\n")
```

```{r paper_umap_adjS, fig.width = 14, fig.height = 8}
df_adjS_cells <- cells_sampen_paper %>%
  filter(!is.na(sampen_norm)) %>%
  droplevels()
if (nrow(df_adjS_cells) > 0) {
  ggplot(df_adjS_cells, aes(x = umap1, y = umap2, color = sampen_norm)) +
    geom_point(size = 0.4, alpha = 0.6) +
    scale_color_viridis_c(option = "plasma", na.value = "grey80") +
    facet_wrap(~ annotation, ncol = 4) +
    labs(title = "Paper MOFA UMAP - per-cell adjS",
         x = "UMAP1 (Argelaguet MOFA)", y = "UMAP2 (Argelaguet MOFA)",
         color = "adjS") +
    theme_ng() +
    theme(axis.text = element_blank(), axis.ticks = element_blank())
} else {
  message("No per-cell adjS overlay data available")
}
```

```{r paper_umap_S, fig.width = 14, fig.height = 8}
df_S_cells <- cells_sampen_paper %>%
  filter(!is.na(sampen)) %>%
  droplevels()
if (nrow(df_S_cells) > 0) {
  ggplot(df_S_cells, aes(x = umap1, y = umap2, color = sampen)) +
    geom_point(size = 0.4, alpha = 0.6) +
    scale_color_viridis_c(option = "magma", na.value = "grey80") +
    facet_wrap(~ annotation, ncol = 4) +
    labs(title = "Paper MOFA UMAP - per-cell S",
         x = "UMAP1 (Argelaguet MOFA)", y = "UMAP2 (Argelaguet MOFA)",
         color = "S") +
    theme_ng() +
    theme(axis.text = element_blank(), axis.ticks = element_blank())
} else {
  message("No per-cell S overlay data available")
}
```

## Lineage class reference

```{r paper_umap_lineage_class, fig.width = 6, fig.height = 5}
ref_df <- meta_cells %>% inner_join(paper_umap, by = "sample")
ggplot(ref_df, aes(x = umap1, y = umap2, color = lineage10x_2)) +
  geom_point(size = 0.5, alpha = 0.6) +
  scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95,
                        na.value = "grey80") +
  labs(title = "Paper MOFA UMAP - lineage class reference",
       x = "UMAP1 (Argelaguet MOFA)", y = "UMAP2 (Argelaguet MOFA)",
       color = "lineage class") +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  theme_ng() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

## Stage reference

```{r paper_umap_stage, fig.width = 6, fig.height = 5}
ggplot(ref_df, aes(x = umap1, y = umap2, color = stage)) +
  geom_point(size = 0.5, alpha = 0.6) +
  scale_color_viridis_d(option = "viridis") +
  labs(title = "Paper MOFA UMAP - stage reference",
       x = "UMAP1 (Argelaguet MOFA)", y = "UMAP2 (Argelaguet MOFA)",
       color = "stage") +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  theme_ng() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

# Cell-level UMAP {.tabset .tabset-pills}

```{r umap_helpers}

plot_umap <- function(df, color_col, title, subtitle = NULL) {
  if (is.null(df) || nrow(df) == 0) {
    message("No UMAP result for: ", title)
    return(invisible(NULL))
  }
  ggplot(df %>% filter(!is.na(.data[[color_col]])),
         aes(x = UMAP1, y = UMAP2, color = .data[[color_col]])) +
    geom_point(size = 1.5, alpha = 0.7) +
    scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
    labs(title = title, subtitle = subtitle,
         color = color_col) +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme_ng()
}
```

```{r cell_umap_load}
short_paths_umap <- list.files(snakemake@params[["output_path"]],
                                pattern = "\\.out\\.gz$", full.names = TRUE)
short_paths_umap <- short_paths_umap[!grepl("\\.det\\.out\\.gz$",  short_paths_umap)]
short_paths_umap <- short_paths_umap[!grepl("\\.meth\\.out\\.gz$", short_paths_umap)]

read_cells_long <- function(fps, col) {
  bind_rows(bplapply(fps, function(fp) {
    fn   <- basename(fp)
    vals <- read_tsv(fp, na = "-1", show_col_types = FALSE)[[col]]
    data.frame(
      stage_san   = get_stage(fn),
      lineage_san = get_lineage(fn),
      annotation  = get_annotation(fn),
      cell_idx    = seq_along(vals),
      value       = vals,
      stringsAsFactors = FALSE
    )
  }, BPPARAM = param))
}

cells_adjS_long <- read_cells_long(short_paths_umap, "sampen_norm")
cells_meth_long <- read_cells_long(short_paths_umap, "avg_meth")

meta_cols_cell <- c("stage_san", "lineage_san", "cell_idx",
                    "stage", "lineage", "lineage_class")

cells_adjS_wide <- pivot_wider(cells_adjS_long,
                                id_cols    = c(stage_san, lineage_san, cell_idx),
                                names_from = annotation,
                                values_from = value) %>%
  left_join(meta_grp %>% select(stage_san, lineage_san, stage, lineage,
                                 lineage_class),
            by = c("stage_san", "lineage_san"))

cells_meth_wide <- pivot_wider(cells_meth_long,
                                id_cols    = c(stage_san, lineage_san, cell_idx),
                                names_from = annotation,
                                values_from = value) %>%
  left_join(meta_grp %>% select(stage_san, lineage_san, stage, lineage,
                                 lineage_class),
            by = c("stage_san", "lineage_san"))

saveRDS(list(cells_adjS_wide = cells_adjS_wide,
             cells_meth_wide = cells_meth_wide),
        "argelaguet_cell_matrices.rds")
```

## Cell-level: adjS across annotations

```{r cell_umap_adjS, fig.width = ng_fig_size(2, 2)$w, fig.height = ng_fig_size(2, 2)$h}
umap_cell_adjS <- run_umap_wide(cells_adjS_wide, meta_cols_cell,
                                 n_neighbors = 15L)
saveRDS(umap_cell_adjS, "argelaguet_umap_cell_adjS.rds")

plot_umap(umap_cell_adjS, "lineage_class",
          title    = "Cell-level UMAP - adjS features",
          subtitle = "per-cell adjS per annotation as feature vector")
```

## Cell-level: adjS colored by stage

```{r cell_umap_adjS_stage, fig.width = ng_fig_size(2, 2)$w, fig.height = ng_fig_size(2, 2)$h}
plot_umap(umap_cell_adjS, "stage",
          title    = "Cell-level UMAP - adjS features",
          subtitle = "colored by developmental stage")
```

## Cell-level: avg methylation across annotations

```{r cell_umap_meth, fig.width = ng_fig_size(2, 2)$w, fig.height = ng_fig_size(2, 2)$h}
umap_cell_meth <- run_umap_wide(cells_meth_wide, meta_cols_cell,
                                 n_neighbors = 15L)
saveRDS(umap_cell_meth, "argelaguet_umap_cell_meth.rds")

plot_umap(umap_cell_meth, "lineage_class",
          title    = "Cell-level UMAP - avg. methylation features",
          subtitle = "per-cell avg. methylation per annotation as feature vector")
```

## Cell-level: S with marginal methylation regressed out

Unadjusted S per cell per annotation, with per-cell mean methylation regressed out
column-wise before embedding. This checks that cell separation reflects locus-level
heterogeneity rather than global methylation differences between cells.

```{r cell_umap_S_regress}
cells_S_long <- read_cells_long(short_paths_umap, "sampen")

cells_S_wide <- pivot_wider(cells_S_long,
                             id_cols    = c(stage_san, lineage_san, cell_idx),
                             names_from = annotation,
                             values_from = value) %>%
  left_join(meta_grp %>% select(stage_san, lineage_san, stage, lineage,
                                 lineage_class),
            by = c("stage_san", "lineage_san"))

id_cols <- c("stage_san", "lineage_san", "cell_idx")
shared_cells <- inner_join(
  cells_S_wide[, id_cols], cells_meth_wide[, id_cols], by = id_cols
)
cells_S_matched <- semi_join(cells_S_wide, shared_cells, by = id_cols)
cells_meth_matched <- semi_join(cells_meth_wide, shared_cells, by = id_cols)

annot_cols_S <- setdiff(colnames(cells_S_matched), meta_cols_cell)
annot_cols_meth <- setdiff(colnames(cells_meth_matched), meta_cols_cell)

S_resid <- regress_out_meth(
  as.matrix(cells_S_matched[, annot_cols_S, drop = FALSE]),
  as.matrix(cells_meth_matched[, annot_cols_meth, drop = FALSE])
)

cells_S_regressed <- bind_cols(
  cells_S_matched[, meta_cols_cell, drop = FALSE],
  as.data.frame(S_resid)
)
```

```{r cell_umap_S_regressed, fig.width = ng_fig_size(2, 2)$w, fig.height = ng_fig_size(2, 2)$h}
umap_cell_S_regressed <- run_umap_wide(cells_S_regressed, meta_cols_cell,
                                        n_neighbors = 15L)
saveRDS(umap_cell_S_regressed, "argelaguet_umap_cell_S_regressed.rds")

plot_umap(umap_cell_S_regressed, "lineage_class",
          title    = "Cell-level UMAP - S with marginal meth regressed out",
          subtitle = "residuals of S ~ cell mean methylation per annotation")
```

## Group-level: adjH across annotations

```{r group_umap_adjH, fig.width = ng_fig_size(2, 2)$w, fig.height = ng_fig_size(2, 2)$h}
adjH_wide_grp <- all_grp_meta %>%
  filter(!is.na(lineage_class)) %>%
  select(stage_san, lineage_san, stage, lineage, lineage_class,
         annotation, median_adjH) %>%
  pivot_wider(names_from = annotation, values_from = median_adjH)

umap_grp_adjH <- run_umap_wide(
  adjH_wide_grp,
  meta_cols   = c("stage_san", "lineage_san", "stage", "lineage", "lineage_class"),
  n_neighbors = 5L
)
saveRDS(umap_grp_adjH, "argelaguet_umap_grp_adjH.rds")

plot_umap(umap_grp_adjH, "lineage_class",
          title    = "Group-level UMAP - adjH features",
          subtitle = "median adjH per annotation as feature vector")
```

```{r group_umap_adjH_stage, fig.width = ng_fig_size(2, 2)$w, fig.height = ng_fig_size(2, 2)$h}
plot_umap(umap_grp_adjH, "stage",
          title    = "Group-level UMAP - adjH features",
          subtitle = "colored by developmental stage")
```

# UMAP: per-annotation locus-level {.tabset .tabset-pills}

```{r locus_umap_constants}
LOCUS_N_UMAP <- 2000L
```

For each annotation, groups (stage x lineage) are embedded using per-locus
normalised Shannon entropy (adjH) from the `.norm.det.out.gz` files as features.
This gives a richer feature space than the cross-annotation UMAP above.

Locus selection (two-pass, memory-efficient):
1. Pass 1 reads each file once, accumulating only a per-locus coverage count
   (whether the value is finite). Loci covered in >= 50% of groups are eligible.
2. The most-covered eligible loci (up to `r LOCUS_N_UMAP`) are selected.
3. Pass 2 reads each file again to extract values only for the selected loci.

Cells (stage x lineage groups) with too many uncovered loci (> 50% NA after
`read_tsv` converts the -1 sentinel to NA) are removed by cell-level QC in
`run_embedding` before HVF selection.

For well-covered annotations (e.g. broad marks): stage 1 of the embedding uses
complete features (0 NAs across all retained groups), keeping all groups.
For very sparse annotations: a coverage-weighted fallback selects HVFs from
features covered in up to 30% of groups, then drops groups with missing values
at those HVFs. Annotations that remain too sparse to embed return no plot.

```{r locus_umap_setup}
LOCUS_SEED <- 42L

meta_cols_grp_arg <- c("stage_san", "lineage_san", "stage", "lineage", "lineage_class")

## Locus selection for UMAP: -1 in shannon_norm means no coverage at that
## locus in that group; read_tsv(na="-1") converts it to NA. Random
## locus sampling yields almost no usable signal for sparse feature sets
## (SINEs, histone marks) because coverage is low and scattered.
##
## Two-pass strategy:
##   Pass 1 - coverage scan: read each file once, accumulate only a boolean
##     coverage count per locus. Memory cost: O(n_loci) integers, not the
##     full groups x loci matrix.
##   Pass 2 - value extraction: read each file again, extract only the loci
##     that are covered in >= min_locus_cov_frac of groups (most-covered first,
##     up to n_loci). This keeps peak RAM at one file at a time.
##
## Column names in norm.det.out files contain file paths with special
## characters (dots, slashes) which produce benign read_tsv type-guessing
## warnings; suppressWarnings is used here as only the named col is needed.
build_ann_locus_mat <- function(annotation_name, fps, col, n_loci = LOCUS_N_UMAP,
                                min_locus_cov_frac = 0.5) {
  fps_ann <- fps[get_annotation(basename(fps)) == annotation_name]
  if (length(fps_ann) < 4L) return(NULL)

  ## Pass 1: accumulate per-locus coverage counts, one file at a time.
  cov_counts <- NULL
  n_loci_total <- NULL
  for (fp in fps_ann) {
    vals <- suppressWarnings(
      read_tsv(fp, na = "-1", show_col_types = FALSE))[[col]]
    if (is.null(cov_counts)) {
      n_loci_total <- length(vals)
      cov_counts <- integer(n_loci_total)
    }
    cov_counts <- cov_counts + is.finite(vals)
  }
  locus_cov <- cov_counts / length(fps_ann)

  eligible <- which(locus_cov >= min_locus_cov_frac)
  if (length(eligible) < 2L) {
    cat(annotation_name, ": no loci covered in >=", min_locus_cov_frac * 100,
        "% of groups (total loci =", n_loci_total, "); skipping\n")
    return(NULL)
  }
  top_loci <- eligible[order(locus_cov[eligible], decreasing = TRUE)]
  idx <- sort(top_loci[seq_len(min(as.integer(n_loci), length(top_loci)))])

  ## Pass 2: extract values only for selected loci, one file at a time.
  rows <- lapply(fps_ann, function(fp) {
    suppressWarnings(
      read_tsv(fp, na = "-1", show_col_types = FALSE))[[col]][idx]
  })
  mat <- do.call(rbind, rows)
  colnames(mat) <- paste0("locus_", idx)

  cat(annotation_name, ": groups =", nrow(mat),
      "| loci used =", ncol(mat),
      "| eligible (cov >=", min_locus_cov_frac, ") =", length(eligible),
      "| total loci =", n_loci_total, "\n")

  df <- data.frame(
    stage_san   = get_stage(basename(fps_ann)),
    lineage_san = get_lineage(basename(fps_ann)),
    as.data.frame(mat),
    check.names = FALSE
  )
  left_join(df,
            meta_grp %>% select(stage_san, lineage_san, stage, lineage, lineage_class),
            by = c("stage_san", "lineage_san"))
}

ann_names_locus <- unique(get_annotation(basename(norm_det_paths)))
```

```{r locus_umap_adjH, fig.width = ng_fig_size(2, 2)$w, fig.height = ng_fig_size(2, 2)$h, results = "asis"}
for (ann in ann_names_locus) {
  cat("\n## ", ann, "\n\n")
  mat <- build_ann_locus_mat(ann, norm_det_paths, "shannon_norm")
  if (is.null(mat)) {
    cat("Not enough groups for", ann, "\n\n")
    next
  }
  umap_res <- run_umap_wide(mat, meta_cols_grp_arg, n_neighbors = 5L, seed = LOCUS_SEED)
  p <- plot_umap(umap_res, "lineage_class",
                 title    = paste0("Locus UMAP - adjH - ", ann),
                 subtitle = paste0(LOCUS_N_UMAP, " loci sampled"))
  if (!is.null(p)) print(p)
  cat("\n\n")
}
```
