---
title: "CRC by feature (no windows)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Izaskun Mallona"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default # kable
    theme: lumen
---

```{r, setup}
suppressPackageStartupMessages({
  library(readr)
  library(lmerTest)
  library(ggplot2)
  library(knitr)
  library(effects)
  library(ComplexHeatmap)
  library(viridis)
  library(BiocParallel)
  library(ggplot2)
  library(ggstance)  # for horizontal error bars
  library(dplyr)
})

source("src/ggtheme.R")
param <- MulticoreParam(workers = snakemake@threads)
```

```{r}
opts_chunk$set(
  fig.width = 7.5,
  fig.height = 7.5,
  cache = FALSE,
  include = TRUE,
  fig.path = "crc_plots/",
  dev = c("png", "svg"),
  dpi = 500,
  cache.lazy = FALSE,
  warning = TRUE,
  message = TRUE
)
```

```{r save_snmk_object_for_debugging, include = FALSE}
saveRDS(file = 'snmk_object_features.rds', object = snakemake)
```


```{r logging, include = FALSE}
log <- file(snakemake@log$log, open = "wt")
sink(log)
sink(log, type = "message")
```

# Short (global) reports by annotation  {.tabset .tabset-pills}

## Modelling

Firstly, we read in the yamet output files and create a table storing the biopsy location, patient id and genomic segmentation.

```{r, import_short_reports}
filepaths <- list.files(
  snakemake@params[["output_path"]],
  pattern = "\\.out\\.gz$", full.names = TRUE
)

filepaths <- filepaths[!grepl("\\.det\\.out\\.gz$", filepaths)]
filepaths <- filepaths[!grepl("\\.norm\\.out\\.gz$", filepaths)]
filepaths <- filepaths[!grepl("\\.meth\\.out\\.gz$", filepaths)]

regex <- "(.*)_(.*)_(.*)_(.*).out.gz"
process_file <- function(filepath) {
  filename <- basename(filepath)

  dt <- read_tsv(filepath, na = "-1", show_col_types = FALSE)
  dt <- dt[, c("sampen", "avg_meth")]

  dt$annotation <- gsub(regex, "\\1", filename)
  dt$patient <- gsub(regex, "\\3", filename)
  dt$location <- gsub(regex, "\\4", filename)

  return(dt)
}

files <- do.call(rbind, bplapply(filepaths, process_file, BPPARAM = param))

files$annotation <- as.factor(files$annotation)
files$patient <- as.factor(files$patient)
files$location <- factor(
  files$location,
  levels = c("NC", "PT", "LN", "ML", "MP", "MO")
)
```

Patient as random effect

```{r, lmer}
lmer.files <- lmer(
  sampen ~ location * avg_meth + (1 | patient),
  data = subset(files, annotation == "pmds")
)
summary(lmer.files)
anova(lmer.files, type = 3)
```

## Scatters by location and annotation
 
```{r}
ggplot(
  ## subset(files, patient == "CRC01"),
  files,
  aes(y = sampen, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation) +
  labs(
    x = "average methylation",
    y = "sample entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Scatters by location and annotation and patient
 
```{r, fig.height = 15}
ggplot(
  ## subset(files, patient == "CRC01"),
  files,
  aes(y = sampen, x = avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_grid(annotation ~ patient) +
  labs(
    x = "average methylation",
    y = "sample entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Scatters by location and annotation and patient (shorter)
 
```{r, fig.height = 3.2}
ggplot(
  subset(files, annotation %in% c("laminb1", "H3K27me3", "H3K9me3")),
  aes(y = sampen, x = avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_grid(annotation ~ patient) +
  labs(
    x = "average methylation",
    y = "sample entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## sampEn boxes by location and annotation - crc01

```{r, fig.height = 6, fig.width = 10}
ggplot(
  subset(files, patient == "CRC01"),
  aes(y = sampen, x = annotation, color = location)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = "sample entropy") +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## meth boxes by location and annotation - crc01

```{r, fig.height = 6, fig.width = 10}
ggplot(
  subset(files, patient == "CRC01"),
  aes(y = avg_meth, x = annotation, color = location)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = "average methylation") +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Range plots - by compartment

```{r, fig.height = 10, fig.width = 10}


ranges <- as.data.frame(files %>%
   filter(patient == "CRC01") %>%
   group_by(compartment = annotation, location) %>%
   summarise(
     sampen_mean = mean(sampen, na.rm = TRUE),
     sampen_min  = min(sampen, na.rm = TRUE),
     sampen_max  = max(sampen, na.rm = TRUE),
     meth_mean   = mean(avg_meth, na.rm = TRUE),
     meth_min    = min(avg_meth, na.rm = TRUE),
     meth_max    = max(avg_meth, na.rm = TRUE)))

## ggplot(ranges, aes(x = meth_mean, y = sampen_mean, color = location)) +
##   geom_point(shape = 3, size = 1.2) +  # "+" symbol
##   geom_errorbar(aes(ymin = sampen_min, ymax = sampen_max), width = 0.3, alpha = 1, size = 1) +
##   geom_errorbarh(aes(xmin = meth_min, xmax = meth_max), height = 0.3, alpha = 1, size = 0) +
##   scale_color_brewer(palette = "Dark2") +
##   labs(
##     x = "DNA methylation",
##     y = "Sample entropy"
##   ) +
##   facet_wrap(~ compartment) +
##   theme_ng()

# I think this is the correct one
ggplot(ranges, aes(color = location)) +
  ## horizontal segment: range of methylations for that location/compartment
  geom_segment(
    aes(x = meth_min, xend = meth_max, y = sampen_mean, yend = sampen_mean),
    size = 0.6, alpha = 1
  ) +
  # vertical segment: range of sampens for that location/compartment
  geom_segment(
    aes(x = meth_mean, xend = meth_mean, y = sampen_min, yend = sampen_max),
    size = 0.6, alpha = 1
  ) +
  # center point
  geom_point(
    aes(x = meth_mean, y = sampen_mean),
    shape = 3, size = 1.2
  ) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "DNA methylation", y = "Sample entropy") +
  facet_wrap(~ compartment) +
  theme_ng()

```

## Others

```{r, fig.height = 10, fig.width = 10}

ggplot(ranges, aes(x = compartment, y = location, fill = sampen_mean)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))


ggplot(ranges, aes(x = compartment, y = location, fill = meth_mean)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))

```

# Full (local) reports for **downsampled** PMDs and CRC01

Caution downsampled to some rows only.

```{r, import_log_reports}
DOWNSAMPLING <- 5e4

annotation <- "pmds"
patient <- "CRC01"

local <- list()
for (location in c("NC", "PT", "LN", "ML", "MP")) {
  filepaths <- list.files(
    snakemake@params[["output_path"]],
    pattern = "\\.det\\.out\\.gz$", full.names = TRUE
  )
  filepaths <- grep(location, filepaths, value = TRUE)
  filepaths <- grep(annotation, filepaths, value = TRUE)
  filepaths <- grep(patient, filepaths, value = TRUE)

  d <- bplapply(filepaths, \(filepath) {
    dt <- read_tsv(
      filepath,
      na = "-1", n_max = DOWNSAMPLING, show_col_types = FALSE
    )
    colnames(dt) <- basename(colnames(dt))
    dt
  }, BPPARAM = param)

  d2 <- bplapply(d, \(x) {
    avg_sampen <- apply(
      x[, grep(location, colnames(x))], 1, \(y) mean(y, na.rm = TRUE)
    )
    median_sampen <- apply(
      x[, grep(location, colnames(x))], 1, \(y) median(y, na.rm = TRUE)
    )
    q25_sampen <- apply(
      x[, grep(location, colnames(x))], 1, \(y) quantile(y, 0.25, na.rm = TRUE)
    )
    q75_sampen <- apply(
      x[, grep(location, colnames(x))], 1, \(y) quantile(y, 0.75, na.rm = TRUE)
    )
    cbind(x, data.frame(
      avg_sampen = avg_sampen, median_sampen = median_sampen,
      q25_sampen = q25_sampen, q75_sampen = q75_sampen,
      location = location,
      patient = patient,
      annotation = annotation
    ))
  }, BPPARAM = param)

  local[[location]] <- d2[[1]]
  rm(d, d2)
}
```

## Scatters {.tabset .tabset-pills}

### By meth NC

```{r, fig.width = 10, fig.height = 10}
pal <- viridis(10, alpha = 0.8)

beta2m <- function(betas, epsilon = 1e-05) {
  if (!is.numeric(betas)) {
    stop("Non-numeric betas")
  }
  if (!(is.numeric(epsilon) && (!is.na(epsilon)))) {
    stop("Epison not valid")
  }
  if (epsilon < 0 || epsilon >= 0.5) {
    stop("Not 0 < epsilon < 0.5")
  }
  betas[betas < epsilon] <- epsilon
  betas[betas > (1 - epsilon)] <- 1 - epsilon
  return(log2(betas / (1 - betas)))
}

pmds <- data.frame(
  shannon_NC = local$NC$shannon,
  meth_NC = local$NC$avg_meth,
  med_sampen_NC = local$NC$median_sampen,
  shannon_PT = local$PT$shannon,
  meth_PT = local$PT$avg_meth,
  med_sampen_PT = local$PT$median_sampen
)
pmds$meth_change <- pmds$meth_PT - pmds$meth_NC
pmds <- na.omit(pmds)

pmds$by_meth_nc <- cut(pmds$meth_NC, breaks = 10)
pmds$by_meth_pt <- cut(pmds$meth_PT, breaks = 10)

plot(pmds[, grep("by", colnames(pmds), invert = TRUE)],
  pch = ".", col = pal[as.numeric(pmds$by_meth_nc)], main = "by_meth_nc"
)
```

### By median sampen PT


```{r, fig.width = 10, fig.height = 10}
pmds$by_med_sampen_pt <- cut(pmds$med_sampen_PT, breaks = 10)
plot(pmds[, grep("by", colnames(pmds), invert = TRUE)],
  pch = ".", col = pal[as.numeric(pmds$by_med_sampen_pt)],
  main = "by_med_sampen_PT"
)
```

### By median sampen NC

```{r, fig.width = 10, fig.height = 10, results = 'asis'}
pmds$by_med_sampen_nc <- cut(pmds$med_sampen_NC, breaks = 10)
plot(pmds[, grep("by", colnames(pmds), invert = TRUE)],
  pch = ".", col = pal[as.numeric(pmds$by_med_sampen_nc)],
  main = "by_med_sampen_nc"
)
```

### By methylation change

```{r, fig.width = 10, fig.height = 10, results = 'asis'}
pmds$by_meth_change <- cut(pmds$meth_change, breaks = 10)
plot(pmds[, grep("by", colnames(pmds), invert = TRUE)],
  pch = ".", col = pal[as.numeric(pmds$by_meth_change)],
  main = "by_meth_change"
)
```

Add extra discretization.

```{r}
pmds$by_sampen_pt <- cut(pmds$med_sampen_PT, breaks = 10)
pmds$by_sampen_nc <- cut(pmds$med_sampen_NC, breaks = 10)
pmds$by_shannon_pt <- cut(pmds$shannon_PT, breaks = 10)
pmds$by_shannon_nc <- cut(pmds$shannon_NC, breaks = 10)
```



# Session

```{r, cache = FALSE}
sessionInfo()
```
