---
title: "CRC"
author: "Atreya Choudhury, Izaskun Mallona"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    code_folding: show
    theme: flatly
    number_sections: no
    fig_crop: yes
---

We require `readr`, `lmerTest` and `ggplot2` for our analysis.
```{r, setup}
suppressPackageStartupMessages({
  library(readr)
  library(lmerTest)
  library(ggplot2)
  library(Matrix)
  library(ComplexHeatmap)
  library(BiocParallel)
  library(limma)
})
```

Firstly, we read in the yamet output files and create a table storing the stage of cancer, patient and annotation type against the yamet output.
```{r, import}
filepaths <- list.files(
  snakemake@params[["yamet"]],
  pattern = "\\.out$", full.names = T
)
filepaths <- filepaths[!grepl("\\.(det|meth)\\.out$|bookended\\.custom\\.", filepaths)]
process_file <- function(filepath) {
  filename <- basename(filepath)

  parts <- strsplit(filename, "\\.")[[1]]
  annotation <- parts[1]
  patient <- parts[3]
  stage <- parts[4]

  dt <- read_delim(filepath, delim = "\t", show_col_types = FALSE)
  dt <- dt[, c("sampen", "avg_meth")]

  dt$annotation <- annotation
  dt$patient <- patient
  dt$stage <- stage

  return(dt)
}

files <- do.call(rbind, lapply(filepaths, process_file))

files$annotation <- as.factor(files$annotation)
files$patient <- as.factor(files$patient)
files$stage <- as.factor(files$stage)
files[files == -1] <- NA
```

We account for fixed effects of stage and average methylation and a random effect from the patient.
```{r, lmer}
lmer.files <- lmer(
  sampen ~ stage * avg_meth + (1 | patient),
  data = subset(files, annotation == "pmds")
)
summary(lmer.files)
anova(lmer.files, type = 3)
```
We observe that the effect of the different stages on sample entropy is strongly confounded by the average methylation.

```{r, plot, out.width="100%"}
ggplot(
  subset(files, patient == "CRC01"),
  aes(y = sampen, x = avg_meth, color = stage)
) +
  geom_point(size = 0.9, alpha = 0.7) +
  scale_color_brewer(palette = "Dark2") +
  facet_wrap(~annotation) +
  labs(
    x = "average methylation",
    y = "sample entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2)))

ggplot(
  subset(files, patient == "CRC01"),
  aes(y = sampen, x = annotation, color = stage)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = "sample entropy") +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggplot(
  subset(files, patient == "CRC01"),
  aes(y = avg_meth, x = annotation, color = stage)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = "average methylation") +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Heatmap

```{r}
filepaths <- list.files(
  snakemake@params[["yamet"]],
  pattern = "bookended\\.custom\\..*\\.det\\.out$", full.names = T
)

process_reports <- function(filepath) {
  out_path <- gsub("\\.det", "", filepath)
  out <- read_tsv(out_path, show_col_types = FALSE, na = "-1")
  d_out <- read_tsv(filepath, show_col_types = FALSE, na = "-1")

  se <- as(as.matrix(d_out[, grep("singleC", colnames(d_out))]), "dgCMatrix")
  rownames(se) <- paste0(d_out$chr, ":", d_out$start, "-", d_out$end)

  return(
    list(
      shannon = d_out[, c("shannon", "avg_meth")],
      sampen = se, cell_meth = out$avg_meth
    )
  )
}

param <- MulticoreParam(workers = snakemake@threads)
all_data <- bplapply(filepaths, process_reports, BPPARAM = param)
sampens <- do.call(cbind, lapply(all_data, \(x) x$sampen))

meta <- data.frame(
  colname = colnames(sampens),
  patient = sapply(strsplit(colnames(sampens), "_"), \(x) x[3]),
  subloc = sapply(strsplit(colnames(sampens), "_"), \(x) x[4]),
  cell_meth = unlist(lapply(all_data, \(x) x$cell_meth), recursive = FALSE)
)
meta$loc <- substr(meta$subloc, 1, 2)
meta$loc <- relevel(factor(meta$loc), ref = "NC")
meta <- meta[order(as.integer(meta$loc), meta$subloc, meta$patient), ]
sampens <- sampens[, meta$colname]

groups <- unique(meta[c("subloc", "patient")])
groups <- groups[order(as.integer(substr(groups$subloc, 1, 2)), groups$subloc, groups$patient), ]
sub_sampens <- do.call(cbind, lapply(seq_len(nrow(groups)), \(i) {
  cols <- meta$colname[meta$patient == groups$patient[i] & meta$subloc == groups$subloc[i]]
  rowMeans(sampens[, cols, drop = FALSE], na.rm = TRUE)
}))
colnames(sub_sampens) <- paste(groups$subloc, groups$patient, sep = "_")

design <- model.matrix(~ patient + loc + cell_meth, meta)
fit <- lmFit(sampens, design)
fit <- eBayes(fit)
regions <- rownames(topTable(fit, coef = "locPT", number = 20000))
topTable(fit, coef = "locPT")

dim(sub_sampens)
sub_sampens <- na.omit(sub_sampens[regions, ])
regions <- rownames(sub_sampens)
dim(sub_sampens)

annot <- read.table(gzfile(snakemake@input[["annotation"]]), header = T)
rownames(annot) <- rownames(sampens)
row_ha <- rowAnnotation(tile_annotation = as.matrix(data.frame(
  pmd = annot[regions, "pmds_pmd"],
  hmd = annot[regions, "hmds_pmd"],
  H3K27me3 = annot[regions, "H3K27me3_chip"],
  H3K9me3 = annot[regions, "H3K9me3_chip"],
  H3K4me3 = annot[regions, "H3K4me3_chip"],
  laminb1 = annot[regions, "laminb1_lad"],
  X8_Quiescent = annot[regions, "X8_Quiescent_hmm"]
)))

column_ha <- HeatmapAnnotation(
  loc = substr(groups$subloc, 1, 2),
  subloc = groups$subloc,
  patient = groups$patient
)

Heatmap(sub_sampens,
  name = "sampens",
  top_annotation = column_ha,
  right_annotation = row_ha,
  show_row_names = FALSE,
  show_column_names = FALSE,
  cluster_columns = FALSE
)
```