---
title: "CRC exploratory"
author: "Atreya Choudhury, Izaskun Mallona"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    code_folding: show
    theme: flatly
    number_sections: no
    fig_crop: yes
---


```{r, setup}
suppressPackageStartupMessages({
  library(readr)
  library(lmerTest)
  library(ggplot2)
  library(knitr)
})
```

```{r}
opts_chunk$set(fig.width = 5,
               fig.height = 5,
               cache = TRUE,
               include = TRUE,
               dev = "png",
               cache.lazy = FALSE,
               warning = TRUE,
               message = TRUE)
```

Firstly, we read in the yamet output files and create a table storing the biopsy location, patient id and genomic segmentation.

```{r, import}
filepaths <- list.files(
  snakemake@params[["yamet"]],
  pattern = "\\.out$", full.names = T
)

filepaths <- filepaths[!grepl("\\.det\\.out$", filepaths)]

process_file <- function(filepath) {
  filename <- basename(filepath)

  parts <- strsplit(filename, "_")[[1]] ## here! filepaths have changed
  annotation <- parts[1]
  patient <- parts[3]
  location <- parts[4]

  dt <- read_delim(filepath, delim = "\t", show_col_types = FALSE, na.strings = "-1")
  dt <- dt[, c("sampen", "avg_meth")]

  dt$annotation <- annotation
  dt$patient <- patient
  dt$location <- location

  return(dt)
}

files <- do.call(rbind, lapply(filepaths, process_file))

files$annotation <- as.factor(files$annotation)
files$patient <- as.factor(files$patient)
files$location <- as.factor(files$location)
## files[files == -1] <- NA
```

We account for fixed effects of location and average methylation and a random effect from the patient.

```{r, lmer}
lmer.files <- lmer(
  sampen ~ location * avg_meth + (1 | patient),
  data = subset(files, annotation == "pmds")
)
summary(lmer.files)
anova(lmer.files, type = 3)
```

We observe that the effect of the different biopsy locations for patient `CRC01` on sample entropy is strongly confounded by the average methylation.

```{r, plot, out.width="100%"}
ggplot(
  subset(files, patient == "CRC01"),
  aes(y = sampen, x = avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.7) +
  scale_color_brewer(palette = "Dark2") +
  facet_wrap(~annotation) +
  labs(
    x = "average methylation",
    y = "sample entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2)))

ggplot(
  subset(files, patient == "CRC01"),
  aes(y = sampen, x = annotation, color = location)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = "sample entropy") +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggplot(
  subset(files, patient == "CRC01"),
  aes(y = avg_meth, x = annotation, color = location)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = "average methylation") +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

```{r, cache = FALSE}
sessionInfo()
```
