---
title: "CRC by feature (no windows)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Izaskun Mallona"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default # kable
    theme: lumen
---

```{r, setup}
suppressPackageStartupMessages({
  library(readr)
  library(lmerTest)
  library(ggplot2)
  library(knitr)
  library(effects)
  library(ComplexHeatmap)
  library(viridis)
  library(BiocParallel)
  library(ggplot2)
  library(ggstance)  # for horizontal error bars
  library(dplyr)
})

source("src/ggtheme.R")
param <- MulticoreParam(workers = snakemake@threads)
```

```{r}
opts_chunk$set(
  fig.width = 7.5,
  fig.height = 7.5,
  cache = FALSE,
  include = TRUE,
  fig.path = "crc_plots/",
  dev = c("png", "svg"),
  dpi = 500,
  cache.lazy = FALSE,
  warning = TRUE,
  message = TRUE
)
```

```{r save_snmk_object_for_debugging, include = FALSE}
saveRDS(file = 'snmk_object_features.rds', object = snakemake)
## snakemake <- readRDS('../../snmk_object_features.rds')
```


```{r logging, include = FALSE}
log <- file(snakemake@log$log, open = "wt")
sink(log)
sink(log, type = "message")
```



These are not det.out.gz, but only out.gz files.

```{r, import_short_reports}
filepaths <- list.files(
  snakemake@params[["output_path"]],
  pattern = "\\.out\\.gz$", full.names = TRUE
)

filepaths <- filepaths[!grepl("\\.det\\.out\\.gz$", filepaths)]
filepaths <- filepaths[!grepl("\\.meth\\.out\\.gz$", filepaths)]


regex <- "(.*)_(.*)_(.*)_(.*).out.gz"
process_file <- function(filepath) {
  filename <- basename(filepath)

  dt <- read_tsv(filepath, na = "-1", show_col_types = FALSE)
  dt <- dt[, c("sampen", "avg_meth")]

  dt$annotation <- gsub(regex, "\\1", filename)
  dt$patient <- gsub(regex, "\\3", filename)
  dt$location <- gsub(regex, "\\4", filename)

  return(dt)
}

files <- do.call(rbind, bplapply(filepaths, process_file, BPPARAM = param))

files$annotation <- as.factor(files$annotation)
files$patient <- as.factor(files$patient)
files$location <- factor(
  files$location,
  levels = c("NC", "PT", "LN", "ML", "MP", "MO")
)
```

## Scatters by location and annotation
 
```{r}
ggplot(
  ## subset(files, patient == "CRC01"),
  files,
  aes(y = sampen, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation) +
  labs(
    x = "average methylation",
    y = "sample entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Scatters by location and annotation and patient
 
```{r, fig.height = 15}
ggplot(
  ## subset(files, patient == "CRC01"),
  files,
  aes(y = sampen, x = avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_grid(annotation ~ patient) +
  labs(
    x = "average methylation",
    y = "sample entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Scatters by location and annotation and patient (shorter)
 
```{r, fig.height = 3.2}
ggplot(
  subset(files, annotation %in% c("laminb1", "H3K27me3", "H3K9me3")),
  aes(y = sampen, x = avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_grid(annotation ~ patient) +
  labs(
    x = "average methylation",
    y = "sample entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## sampEn boxes by location and annotation - crc01

```{r, fig.height = 6, fig.width = 10}
ggplot(
  subset(files, patient == "CRC01"),
  aes(y = sampen, x = annotation, color = location)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = "sample entropy") +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## meth boxes by location and annotation - crc01

```{r, fig.height = 6, fig.width = 10}
ggplot(
  subset(files, patient == "CRC01"),
  aes(y = avg_meth, x = annotation, color = location)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = "average methylation") +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Range plots - by compartment

```{r, fig.height = 10, fig.width = 10}


ranges <- as.data.frame(files %>%
   filter(patient == "CRC01") %>%
   group_by(compartment = annotation, location) %>%
   summarise(
     sampen_mean = mean(sampen, na.rm = TRUE),
     sampen_min  = min(sampen, na.rm = TRUE),
     sampen_max  = max(sampen, na.rm = TRUE),
     meth_mean   = mean(avg_meth, na.rm = TRUE),
     meth_min    = min(avg_meth, na.rm = TRUE),
     meth_max    = max(avg_meth, na.rm = TRUE)))

## ggplot(ranges, aes(x = meth_mean, y = sampen_mean, color = location)) +
##   geom_point(shape = 3, size = 1.2) +  # "+" symbol
##   geom_errorbar(aes(ymin = sampen_min, ymax = sampen_max), width = 0.3, alpha = 1, size = 1) +
##   geom_errorbarh(aes(xmin = meth_min, xmax = meth_max), height = 0.3, alpha = 1, size = 0) +
##   scale_color_brewer(palette = "Dark2") +
##   labs(
##     x = "DNA methylation",
##     y = "Sample entropy"
##   ) +
##   facet_wrap(~ compartment) +
##   theme_ng()

# I think this is the correct one
ggplot(ranges, aes(color = location)) +
  ## horizontal segment: range of methylations for that location/compartment
  geom_segment(
    aes(x = meth_min, xend = meth_max, y = sampen_mean, yend = sampen_mean),
    size = 0.6, alpha = 1
  ) +
  # vertical segment: range of sampens for that location/compartment
  geom_segment(
    aes(x = meth_mean, xend = meth_mean, y = sampen_min, yend = sampen_max),
    size = 0.6, alpha = 1
  ) +
  # center point
  geom_point(
    aes(x = meth_mean, y = sampen_mean),
    shape = 3, size = 1.2
  ) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "DNA methylation", y = "Sample entropy") +
  facet_wrap(~ compartment) +
  theme_ng()

```

# After normalization - same plots (the yamet-way)

That is, not by locus but by cell norm sampEns (that are calculated according to some regionset anyway)

```{r, import_short_reports_norm}
filepaths <- list.files(
  snakemake@params[["output_path"]],
  pattern = "out\\.gz$", full.names = TRUE)


filepaths <- filepaths[!grepl("\\.det\\.out\\.gz$", filepaths)]
filepaths <- filepaths[!grepl("\\.norm\\.out\\.gz$", filepaths)]
filepaths <- filepaths[!grepl("\\.meth\\.out\\.gz$", filepaths)]

head(filepaths)
regex <- "(.*)_(.*)_(.*)_(.*).out.gz"
process_file <- function(filepath) {
  filename <- basename(filepath)

  dt <- read_tsv(filepath, na = "-1", show_col_types = FALSE)
  dt <- dt[, c("sampen_norm", "avg_meth")]
  colnames(dt) <- c('sampen', 'avg_meth') ## caution they're normalized though

  dt$annotation <- gsub(regex, "\\1", filename)
  dt$patient <- gsub(regex, "\\3", filename)
  dt$location <- gsub(regex, "\\4", filename)

  return(dt)
}

head(filepaths)

files <- do.call(rbind, bplapply(filepaths, process_file, BPPARAM = param))

files$annotation <- as.factor(files$annotation)
files$patient <- as.factor(files$patient)
files$location <- factor(
  files$location,
  levels = c("NC", "PT", "LN", "ML", "MP", "MO")
)
```

## Scatters by location and annotation
 
```{r}
ggplot(
  ## subset(files, patient == "CRC01"),
  files,
  aes(y = sampen, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation) +
  labs(
    x = "average methylation",
    y = "norm sample entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Scatters by location and annotation and patient
 
```{r, fig.height = 15}
ggplot(
  ## subset(files, patient == "CRC01"),
  files,
  aes(y = sampen, x = avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_grid(annotation ~ patient) +
  labs(
    x = "average methylation",
    y = "norm sample entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Scatters by location and annotation and patient (shorter)
 
```{r, fig.height = 3.2}
ggplot(
  subset(files, annotation %in% c("laminb1", "H3K27me3", "H3K9me3")),
  aes(y = sampen, x = avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_grid(annotation ~ patient) +
  labs(
    x = "average methylation",
    y = "norm sample entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## sampEn boxes by location and annotation - crc01

```{r, fig.height = 6, fig.width = 10}
ggplot(
  subset(files, patient == "CRC01"),
  aes(y = sampen, x = annotation, color = location)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = "norm sample entropy") +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## meth boxes by location and annotation - crc01

```{r, fig.height = 6, fig.width = 10}
ggplot(
  subset(files, patient == "CRC01"),
  aes(y = avg_meth, x = annotation, color = location)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = "average methylation") +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Range plots - by compartment

```{r, fig.height = 10, fig.width = 10}


ranges <- as.data.frame(files %>%
   filter(patient == "CRC01") %>%
   group_by(compartment = annotation, location) %>%
   summarise(
     sampen_mean = mean(sampen, na.rm = TRUE),
     sampen_min  = min(sampen, na.rm = TRUE),
     sampen_max  = max(sampen, na.rm = TRUE),
     meth_mean   = mean(avg_meth, na.rm = TRUE),
     meth_min    = min(avg_meth, na.rm = TRUE),
     meth_max    = max(avg_meth, na.rm = TRUE)))

## ggplot(ranges, aes(x = meth_mean, y = sampen_mean, color = location)) +
##   geom_point(shape = 3, size = 1.2) +  # "+" symbol
##   geom_errorbar(aes(ymin = sampen_min, ymax = sampen_max), width = 0.3, alpha = 1, size = 1) +
##   geom_errorbarh(aes(xmin = meth_min, xmax = meth_max), height = 0.3, alpha = 1, size = 0) +
##   scale_color_brewer(palette = "Dark2") +
##   labs(
##     x = "DNA methylation",
##     y = "Sample entropy"
##   ) +
##   facet_wrap(~ compartment) +
##   theme_ng()

# I think this is the correct one
ggplot(ranges, aes(color = location)) +
  ## horizontal segment: range of methylations for that location/compartment
  geom_segment(
    aes(x = meth_min, xend = meth_max, y = sampen_mean, yend = sampen_mean),
    size = 0.6, alpha = 1
  ) +
  # vertical segment: range of sampens for that location/compartment
  geom_segment(
    aes(x = meth_mean, xend = meth_mean, y = sampen_min, yend = sampen_max),
    size = 0.6, alpha = 1
  ) +
  # center point
  geom_point(
    aes(x = meth_mean, y = sampen_mean),
    shape = 3, size = 1.2
  ) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "DNA methylation", y = "Normalized sample entropy") +
  facet_wrap(~ compartment) +
  theme_ng()

```

# Unnormalized Shannon's MEDIANS DOWNSAMPLED {.tabset .tabset-pills}

This cannot be done on short `out.gz` files, but on detailed ones only. Median values, not all loci included!

```{r, eval = TRUE}
filepaths <- list.files(
  snakemake@params[["output_path"]],
  pattern = "det\\.out\\.gz$", full.names = TRUE
)

filepaths <- filepaths[!grepl("norm", filepaths)]

head(filepaths)

## todo extract Shannon's here instead

regex <- "(.*)_(.*)_(.*)_(.*).det.out.gz"

process_file <- function(filepath) {
  filename <- basename(filepath)

  tmp <- read_tsv(filepath, na = "-1", show_col_types = FALSE, n_max = 10e4) ## caution downsampling

  tibble(
    file = filename,
    median_shannon   = median(tmp$shannon,   na.rm = TRUE),
    median_avg_meth  = median(tmp$avg_meth,  na.rm = TRUE),
    annotation = gsub(regex, "\\1", filename),
    patient    = gsub(regex, "\\3", filename),
    location   = gsub(regex, "\\4", filename)
  )
}

files <- bplapply(filepaths, process_file, BPPARAM = param) |> bind_rows()

files$annotation <- factor(files$annotation)
files$patient    <- factor(files$patient)
files$location   <- factor(files$location,
                           levels = c("NC", "PT", "LN", "ML", "MP", "MO"))

```

## Scatters by location and annotation
 
```{r}
ggplot(
  files,
  aes(y = median_shannon, x = median_avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation) +
  labs(
    x = "median average methylation",
    y = "median Shannon entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Scatters by location and annotation and patient
 
```{r, fig.height = 15}
ggplot(
  files,
  aes(y = median_shannon, x = median_avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  ## scale_color_brewer(palette = "Set1") +
  ## facet_grid(annotation ~ patient) +
  labs(
    x = "median average methylation",
    y = "median Shannon entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```



# Meth-normalized Shannons (the yamet way) MEDIANS {.tabset .tabset-pills}

This cannot be done on short `out.gz` files, but on detailed ones only. Median values, not all loci included!

```{r, eval = TRUE}
filepaths <- list.files(
  snakemake@params[["output_path"]],
  pattern = "norm\\.det\\.out\\.gz$", full.names = TRUE
)

head(filepaths)

## todo extract Shannon's here instead

regex <- "(.*)_(.*)_(.*)_(.*).norm.det.out.gz"

process_file <- function(filepath) {
  filename <- basename(filepath)

  tmp <- read_tsv(filepath, na = "-1", show_col_types = FALSE, n_max = 10e4) # caution downsampling

  tibble(
    file = filename,
    median_shannon   = median(tmp$shannon_norm,   na.rm = TRUE),
    median_avg_meth  = median(tmp$avg_meth,  na.rm = TRUE),
    annotation = gsub(regex, "\\1", filename),
    patient    = gsub(regex, "\\3", filename),
    location   = gsub(regex, "\\4", filename)
  )
}

files <- bplapply(filepaths, process_file, BPPARAM = param) |> bind_rows()

files$annotation <- factor(files$annotation)
files$patient    <- factor(files$patient)
files$location   <- factor(files$location,
                           levels = c("NC", "PT", "LN", "ML", "MP", "MO"))

```

## Scatters by location and annotation
 
```{r}
ggplot(
  files,
  aes(y = median_shannon, x = median_avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation) +
  labs(
    x = "median average methylation",
    y = "median norm Shannon entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Scatters by location and annotation and patient
 
```{r, fig.height = 15}
ggplot(
  files,
  aes(y = median_shannon, x = median_avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  ## scale_color_brewer(palette = "Set1") +
  ## facet_grid(annotation ~ patient) +
  labs(
    x = "median average methylation",
    y = "median norm Shannon entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```



# Session

```{r, cache = FALSE}
sessionInfo()
```
