---
title: "CRC by feature (no windows)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Izaskun Mallona"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default # kable
    theme: lumen
---

```{r logging_early, include = FALSE}
log <- file(snakemake@log$log, open = "wt")
sink(log)
sink(log, type = "message")
```

```{r, setup}
suppressPackageStartupMessages({
  library(readr)
  library(ggplot2)
  library(knitr)
  library(ComplexHeatmap)
  library(viridis)
  library(BiocParallel)
  library(dplyr)
  library(tidyr)
  library(ggrepel)
})

source("src/ggtheme.R")
source("src/palettes.R")
source("src/driver_utils.R")
param <- MulticoreParam(workers = snakemake@threads)
```

```{r}
opts_chunk$set(
  fig.width = 5,
  fig.height = 5,
  cache = FALSE,
  include = TRUE,
  fig.path = "crc_plots/",
  dev = c("png", "svg"),
  dpi = 500,
  cache.lazy = FALSE,
  warning = TRUE,
  message = TRUE
)
```

```{r save_snmk_object_for_debugging, include = FALSE}
saveRDS(file = 'snmk_object_features.rds', object = snakemake)
## snakemake <- readRDS('../../snmk_object_features.rds')
## setwd('../..')
```




These are not det.out.gz, but only out.gz files.

```{r, import_short_reports}
filepaths <- list.files(
  snakemake@params[["output_path"]],
  pattern = "\\.out\\.gz$", full.names = TRUE
)

filepaths <- filepaths[!grepl("\\.det\\.out\\.gz$", filepaths)]
filepaths <- filepaths[!grepl("\\.meth\\.out\\.gz$", filepaths)]


regex <- "(.*)_(.*)_(.*)_(.*).out.gz"
process_file <- function(filepath) {
  filename <- basename(filepath)

  dt <- read_tsv(filepath, na = "-1", show_col_types = FALSE)
  dt <- dt[, c("sampen", "avg_meth")]

  dt$annotation <- gsub(regex, "\\1", filename)
  dt$patient <- gsub(regex, "\\3", filename)
  dt$location <- gsub(regex, "\\4", filename)

  return(dt)
}

files <- do.call(rbind, bplapply(filepaths, process_file, BPPARAM = param))

# mind this is per cell, not per locus. This is not something you can do for Shannon.
# imallona@mlshainan:~/src/yamet/workflow$ zcat data/crc/output/sines_sines_CRC01_LN.out.gz | head
#file	sampen	sampen_norm	avg_meth
#data/crc/harmonized/GSM2697511_scTrioSeq2Met_CRC01_LN3_268.singleC.txt.gz	0.0984465	0.311104	0.838184
#data/crc/harmonized/GSM2697889_scTrioSeq2Met_CRC01_LN1_164.singleC.txt.gz	0.0985076	0.290518	0.825905
#data/crc/harmonized/GSM2697915_scTrioSeq2Met_CRC01_LN3_447.singleC.txt.gz	0.098421	0.307972	0.836493
## data/crc/harmonized/GSM2697508_scTrioSeq2Met_CRC01_LN2_456.singleC.txt.gz	0.108082	0.316819	0.824773
## data/crc/harmonized/GSM2697493_scTrioSeq2Met_CRC01_LN1_175.singleC.txt.gz	0.114474	0.312642	0.810966
## data/crc/harmonized/GSM2697905_scTrioSeq2Met_CRC01_LN2_453.singleC.txt.gz	0.100684	0.305491	0.831078
## data/crc/harmonized/GSM2697898_scTrioSeq2Met_CRC01_LN2_192.singleC.txt.gz	0.117015	0.359772	0.833428
## data/crc/harmonized/GSM2697908_scTrioSeq2Met_CRC01_LN3_244.singleC.txt.gz	0.101921	0.28672	0.816892
## data/crc/harmonized/GSM2697909_scTrioSeq2Met_CRC01_LN3_245.singleC.txt.gz	0.0962967	0.299841	0.835639


## files$annotation <- factor(files$annotation,
##                            levels = c("0_Enhancer",
##                                       "1_Transcribed",
##                                       "2_Enhancer",
##                                       "3_Quiescent",
##                                       "4_Transcribed",     
##                                       "5_RegPermissive",
##                                       "6_LowConfidence",
##                                       "7_RegPermissive",
##                                       "8_Quiescent",
##                                       "9_ConstitutiveHet",
##                                       "10_Quiescent",
##                                       "11_Promoter",
##                                       "12_Promoter",
##                                       "13_ConstitutiveHet",                                                                            
##                                       "cpgIslandExt", "H3K4me3", "genes",
##                                       "H3K27me3", "H3K9me3", "laminb1",
##                                       "pmds", "lines", "sines", "hmds",
##                                       "crc01_lost_scna", "crc01_nc_scna", "crc01_gain_scna"))


## files$patient <- as.factor(files$patient)
## files$location <- factor(
##   files$location,
##   levels = c("NC", "PT", "LN", "ML", "MP", "MO")
## )

## head(files)

## harmonize_levels <- function(df) {

##   df$annotation <- factor(
##       df$annotation,
##       levels = c(
##           "0_Enhancer", "1_Transcribed", "2_Enhancer", "3_Quiescent",
##           "4_Transcribed", "5_RegPermissive", "6_LowConfidence",
##           "7_RegPermissive", "8_Quiescent", "9_ConstitutiveHet",
##           "10_Quiescent", "11_Promoter", "12_Promoter", "13_ConstitutiveHet",
##           "cpgIslandExt", "H3K4me3", "genes", "H3K27me3", "H3K9me3",
##           "laminb1", "pmds", "lines", "sines", "hmds",
##           "crc01_lost_scna", "crc01_nc_scna", "crc01_gain_scna"
##     )
##   )

##   df$patient <- as.factor(df$patient)
##   df$location <- factor(df$location, levels = c("NC", "PT", "LN", "ML", "MP", "MO"))

##   return(df)
## }


## prettifies and sorts categorical vars
harmonize_levels <- function(df) { 
    pretty_hmm <- c(
        "0_Enhancer"        = "0 Enh.",
        "1_Transcribed"     = "1 Transc. ",
        "2_Enhancer"        = "2 Enh.",
        "3_Quiescent"       = "3 Quies.",
        "4_Transcribed"     = "4 Transc.",
        "5_RegPermissive"   = "5 Reg. perm.",
        "6_LowConfidence"   = "6 Low conf.",
        "7_RegPermissive"   = "7 Reg. perm.",
        "8_Quiescent"       = "8 Quies.",
        "9_ConstitutiveHet" = "9 Cons. het.",
        "10_Quiescent"      = "10 Quies.",
        "11_Promoter"       = "11 Prom.",
        "12_Promoter"       = "12 Prom.",
        "13_ConstitutiveHet"= "13 Cons. het"
    )

    pretty_other <- c(
        "cpgIslandExt"      = "CpGi",
        "H3K4me3"           = "H3K4me3",
        "genes"             = "Genes",
        "H3K27me3"          = "H3K27me3",
        "H3K9me3"           = "H3K9me3",
        "laminb1"           = "Lamin B1",
        "pmds"              = "PMDs",
        "lines"             = "LINEs",
        "sines"             = "SINEs",
        "hmds"              = "HMDs",
        "crc01_lost_scna"   = "SCNA loss",
        "crc01_nc_scna"     = "SCNA neutral",
        "crc01_gain_scna"   = "SCNA gain")

    all_levels <- c(pretty_hmm, pretty_other)

    missing <- setdiff(unique(df$annotation), names(all_levels))
    if (length(missing) > 0) {
        stop(missing)
    }
    

  df$annotation <- factor(df$annotation,
                          levels = names(all_levels),
                          labels = unname(all_levels))

  df$patient  <- as.factor(df$patient)
  df$location <- factor(df$location, levels = c("NC","PT","LN","ML","MP","MO"))

  df
}

files <- harmonize_levels(files)
table(files$annotation, useNA = 'always')

```

# Bulk unnormalized sampen  {.tabset .tabset-pills}

## Scatters by location and annotation
 
```{r}
ggplot(
  ## subset(files, patient == "CRC01"),
  files,
  aes(y = sampen, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation) +
  labs(
    x = "average methylation",
    y = "sample entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Scatters by location and annotation and patient
 
```{r, fig.width = 8, fig.height = 20}
ggplot(
  ## subset(files, patient == "CRC01"),
  files,
  aes(y = sampen, x = avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_grid(annotation ~ patient) +
  labs(
    x = "average methylation",
    y = "sample entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Scatters by location and annotation and patient (shorter)
 
```{r, fig.width = 8, fig.height = 6}
ggplot(
  subset(files, annotation %in% c("laminb1", "H3K27me3", "H3K9me3")),
  aes(y = sampen, x = avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_grid(annotation ~ patient) +
  labs(
    x = "average methylation",
    y = "sample entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Segmentations  {.tabset .tabset-pills} 

### Scatters by location and annotation
 
```{r}
ggplot(
    subset(files, grepl("^[0-9]{1,2}", annotation)),
  aes(y = sampen, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "avg. methylation",
    y = "sample entropy (S)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```


## Nonsegmentations  {.tabset .tabset-pills}

### Scatters by location and annotation
 
```{r}
ggplot(
    subset(files, !grepl("^[0-9]{1,2}", annotation)),
  aes(y = sampen, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "avg. methylation",
    y = "sample entropy (S)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Others  {.tabset .tabset-pills}

### sampEn boxes by location and annotation - crc01

```{r, fig.width = 9, fig.height = 5}
ggplot(
  subset(files, patient == "CRC01"),
  aes(y = sampen, x = annotation, color = location)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = "sample entropy") +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### meth boxes by location and annotation - crc01

```{r, fig.width = 9, fig.height = 5}
ggplot(
  subset(files, patient == "CRC01"),
  aes(y = avg_meth, x = annotation, color = location)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = "average methylation") +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### Range plots - by compartment

```{r, fig.width = 8, fig.height = 8}


ranges <- as.data.frame(files %>%
   filter(patient == "CRC01") %>%
   group_by(compartment = annotation, location) %>%
   summarise(
     sampen_mean = mean(sampen, na.rm = TRUE),
     sampen_min  = min(sampen, na.rm = TRUE),
     sampen_max  = max(sampen, na.rm = TRUE),
     meth_mean   = mean(avg_meth, na.rm = TRUE),
     meth_min    = min(avg_meth, na.rm = TRUE),
     meth_max    = max(avg_meth, na.rm = TRUE)))

## ggplot(ranges, aes(x = meth_mean, y = sampen_mean, color = location)) +
##   geom_point(shape = 3, size = 1.2) +  # "+" symbol
##   geom_errorbar(aes(ymin = sampen_min, ymax = sampen_max), width = 0.3, alpha = 1, size = 1) +
##   geom_errorbarh(aes(xmin = meth_min, xmax = meth_max), height = 0.3, alpha = 1, size = 0) +
##   scale_color_brewer(palette = "Dark2") +
##   labs(
##     x = "DNA methylation",
##     y = "Sample entropy"
##   ) +
##   facet_wrap(~ compartment) +
##   theme_ng()

# I think this is the correct one
ggplot(ranges, aes(color = location)) +
  ## horizontal segment: range of methylations for that location/compartment
  geom_segment(
    aes(x = meth_min, xend = meth_max, y = sampen_mean, yend = sampen_mean),
    linewidth = 0.6, alpha = 1
  ) +
  # vertical segment: range of sampens for that location/compartment
  geom_segment(
    aes(x = meth_mean, xend = meth_mean, y = sampen_min, yend = sampen_max),
    linewidth = 0.6, alpha = 1
  ) +
  # center point
  geom_point(
    aes(x = meth_mean, y = sampen_mean),
    shape = 3, size = 1.2
  ) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "DNA methylation", y = "Sample entropy") +
  facet_wrap(~ compartment) +
  theme_ng()

```

```{r}
sampens <- files
```

# With adjS meth-norm sampen {.tabset .tabset-pills}

That is, not by locus but by cell norm sampEns (that are calculated according to some regionset anyway)

```{r, import_short_reports_norm}
filepaths <- list.files(
  snakemake@params[["output_path"]],
  pattern = "out\\.gz$", full.names = TRUE)


filepaths <- filepaths[!grepl("\\.det\\.out\\.gz$", filepaths)]
filepaths <- filepaths[!grepl("\\.norm\\.out\\.gz$", filepaths)]
filepaths <- filepaths[!grepl("\\.meth\\.out\\.gz$", filepaths)]

head(filepaths)
regex <- "(.*)_(.*)_(.*)_(.*).out.gz"
process_file <- function(filepath) {
  filename <- basename(filepath)

  dt <- read_tsv(filepath, na = "-1", show_col_types = FALSE)
  dt <- dt[, c("sampen_norm", "avg_meth")]
  colnames(dt) <- c('sampen', 'avg_meth') ## caution they're normalized though

  dt$annotation <- gsub(regex, "\\1", filename)
  dt$patient <- gsub(regex, "\\3", filename)
  dt$location <- gsub(regex, "\\4", filename)

  return(dt)
}


files <- do.call(rbind, bplapply(filepaths, process_file, BPPARAM = param))

files <- harmonize_levels(files)

```

## Scatters by location and annotation
 
```{r}
ggplot(
  ## subset(files, patient == "CRC01"),
  files,
  aes(y = sampen, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation) +
  labs(
    x = "avg. methylation",
    y = "adj. sample entropy (adjS)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Scatters by location and annotation and patient
 
```{r, fig.width = 8, fig.height = 20}
ggplot(
  ## subset(files, patient == "CRC01"),
  files,
  aes(y = sampen, x = avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_grid(annotation ~ patient) +
  labs(
    x = "avg. methylation",
    y = "adj. sample entropy (adjS)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Scatters by location and annotation and patient (shorter)
 
```{r, fig.width = 8, fig.height = 6}
ggplot(
  subset(files, annotation %in% c("laminb1", "H3K27me3", "H3K9me3")),
  aes(y = sampen, x = avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_grid(annotation ~ patient) +
  labs(
    x = "avg. methylation",
    y = "adj. sample entropy (adjS)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```


## Segmentations {.tabset .tabset-pills}

### Scatters by location and annotation
 
```{r}
ggplot(
    subset(files, grepl("^[0-9]{1,2}", annotation)),
  aes(y = sampen, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "avg. methylation",
    y = "adj. sample entropy (adjS)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```


## Nonsegmentations {.tabset .tabset-pills}

### Scatters by location and annotation
 
```{r}
ggplot(
    subset(files, !grepl("^[0-9]{1,2}", annotation)),
  aes(y = sampen, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "avg. methylation",
    y = "adj. sample entropy (adjS)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Others  {.tabset .tabset-pills}

### sampEn boxes by location and annotation - crc01

```{r, fig.width = 9, fig.height = 5}
ggplot(
  subset(files, patient == "CRC01"),
  aes(y = sampen, x = annotation, color = location)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = "adj sample entropy (adjS)") +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### meth boxes by location and annotation - crc01

```{r, fig.width = 9, fig.height = 5}
ggplot(
  subset(files, patient == "CRC01"),
  aes(y = avg_meth, x = annotation, color = location)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = "avg. methylation") +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### Range plots - by compartment

```{r, fig.width = 8, fig.height = 8}
ranges <- as.data.frame(files %>%
   filter(patient == "CRC01") %>%
   group_by(compartment = annotation, location) %>%
   summarise(
     sampen_mean = mean(sampen, na.rm = TRUE),
     sampen_min  = min(sampen, na.rm = TRUE),
     sampen_max  = max(sampen, na.rm = TRUE),
     meth_mean   = mean(avg_meth, na.rm = TRUE),
     meth_min    = min(avg_meth, na.rm = TRUE),
     meth_max    = max(avg_meth, na.rm = TRUE)))

## ggplot(ranges, aes(x = meth_mean, y = sampen_mean, color = location)) +
##   geom_point(shape = 3, size = 1.2) +  # "+" symbol
##   geom_errorbar(aes(ymin = sampen_min, ymax = sampen_max), width = 0.3, alpha = 1, size = 1) +
##   geom_errorbarh(aes(xmin = meth_min, xmax = meth_max), height = 0.3, alpha = 1, size = 0) +
##   scale_color_brewer(palette = "Dark2") +
##   labs(
##     x = "DNA methylation",
##     y = "Sample entropy"
##   ) +
##   facet_wrap(~ compartment) +
##   theme_ng()

# I think this is the correct one
ggplot(ranges, aes(color = location)) +
  ## horizontal segment: range of methylations for that location/compartment
  geom_segment(
    aes(x = meth_min, xend = meth_max, y = sampen_mean, yend = sampen_mean),
    linewidth = 0.6, alpha = 1
  ) +
  # vertical segment: range of sampens for that location/compartment
  geom_segment(
    aes(x = meth_mean, xend = meth_mean, y = sampen_min, yend = sampen_max),
    linewidth = 0.6, alpha = 1
  ) +
  # center point
  geom_point(
    aes(x = meth_mean, y = sampen_mean),
    shape = 3, size = 1.2
  ) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "DNA methylation", y = "adj. sample entropy (adjS)") +
  facet_wrap(~ compartment) +
  theme_ng()

```

```{r}
adjsampens <- files
```

## adjS within methylation bins {.tabset .tabset-pills}

Boxplots of adjS stratified by 0.1-wide average methylation bins, colored by
location. Persistent differences within bins show that adjS captures
heterogeneity beyond average methylation level.

### All annotations

```{r adjS_meth_bins_all, fig.width = ng_fig_size(5, 6)$w, fig.height = ng_fig_size(5, 6)$h}
adjsampens %>%
  mutate(meth_bin = cut(avg_meth, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>%
  filter(!is.na(meth_bin), !is.na(sampen)) %>%
  ggplot(aes(x = meth_bin, y = sampen, fill = location)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(~annotation) +
  labs(x = "avg. methylation bin", y = "adj. sample entropy (adjS)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### CRC01 only

```{r adjS_meth_bins_crc01, fig.width = ng_fig_size(5, 6)$w, fig.height = ng_fig_size(5, 6)$h}
adjsampens %>%
  filter(patient == "CRC01") %>%
  mutate(meth_bin = cut(avg_meth, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>%
  filter(!is.na(meth_bin), !is.na(sampen)) %>%
  ggplot(aes(x = meth_bin, y = sampen, fill = location)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(~annotation) +
  labs(x = "avg. methylation bin", y = "adj. sample entropy (adjS)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

# Unnormalized Shannon's MEDIANS DOWNSAMPLED {.tabset .tabset-pills}

This cannot be done on short `out.gz` files, but on detailed ones only. Median values, not all loci included!

meaning, here we'll have one value per biopsy+patient (the "group") only, because the individual loci are averaged. And there are no individual cells anymore (it's Shannon).

```{r, eval = TRUE}
filepaths <- list.files(
  snakemake@params[["output_path"]],
  pattern = "det\\.out\\.gz$", full.names = TRUE
)

filepaths <- filepaths[!grepl("norm", filepaths)]

head(filepaths)

## todo extract Shannon's here instead

regex <- "(.*)_(.*)_(.*)_(.*).det.out.gz"

process_file <- function(filepath) {
  filename <- basename(filepath)

  tmp <- read_tsv(filepath, na = "-1", show_col_types = FALSE, n_max = 10e4) ## caution downsampling

  tibble(
    file = filename,
    median_shannon   = median(tmp$shannon,   na.rm = TRUE),
    median_avg_meth  = median(tmp$avg_meth,  na.rm = TRUE),
    annotation = gsub(regex, "\\1", filename),
    patient    = gsub(regex, "\\3", filename),
    location   = gsub(regex, "\\4", filename)
  )
}

files <- bplapply(filepaths, process_file, BPPARAM = param) |> bind_rows()

files <- harmonize_levels(files)

head(files)
```

## Scatters by location and annotation
 
```{r}
ggplot(
  files,
  aes(y = median_shannon, x = median_avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation) +
  labs(
    x = "median average methylation",
    y = "median Shannon entropy (H)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```



## Segmentations {.tabset .tabset-pills}

### Scatters by location and annotation
 
```{r}
ggplot(
    subset(files, grepl("^[0-9]{1,2}", annotation)),
  aes(y = median_shannon, x = median_avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "median avg. methylation",
    y = "median Shannon entropy (H)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```


## Nonsegmentations {.tabset .tabset-pills}

### Scatters by location and annotation
 
```{r}
ggplot(
    subset(files, !grepl("^[0-9]{1,2}", annotation)),
  aes(y = median_shannon, x = median_avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "median avg. methylation",
    y = "median Shannon entropy (H)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```



```{r}
shannons <- files
```

# Meth-normalized Shannons (the yamet way) MEDIANS {.tabset .tabset-pills}

This cannot be done on short `out.gz` files, but on detailed ones only. Median values, not all loci included!

```{r, eval = TRUE}
filepaths <- list.files(
  snakemake@params[["output_path"]],
  pattern = "norm\\.det\\.out\\.gz$", full.names = TRUE
)

head(filepaths)

## todo extract Shannon's here instead

regex <- "(.*)_(.*)_(.*)_(.*).norm.det.out.gz"

process_file <- function(filepath) {
  filename <- basename(filepath)

  tmp <- read_tsv(filepath, na = "-1", show_col_types = FALSE, n_max = 10e4) # caution downsampling

  tibble(
    file = filename,
    median_shannon   = median(tmp$shannon_norm,   na.rm = TRUE),
    median_avg_meth  = median(tmp$avg_meth,  na.rm = TRUE),
    annotation = gsub(regex, "\\1", filename),
    patient    = gsub(regex, "\\3", filename),
    location   = gsub(regex, "\\4", filename)
  )
}

files <- bplapply(filepaths, process_file, BPPARAM = param) |> bind_rows()

files <- harmonize_levels(files)

```

## Scatters by location and annotation
 
```{r}
ggplot(
  files,
  aes(y = median_shannon, x = median_avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation) +
  labs(
    x = "median average methylation",
    y = "median norm Shannon entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Scatters by location and annotation and patient
 
```{r}
ggplot(
  files,
  aes(y = median_shannon, x = median_avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  ## scale_color_brewer(palette = "Set1") +
  ## facet_grid(annotation ~ patient) +
  labs(
    x = "median average methylation",
    y = "median norm Shannon entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```



## Segmentations {.tabset .tabset-pills}

### Scatters by location and annotation
 
```{r}
ggplot(
    subset(files, grepl("^[0-9]{1,2}", annotation)),
  aes(y = median_shannon, x = median_avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "median avg. methylation",
    y = "median adj Shannon entropy (adjH)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```


## Nonsegmentations {.tabset .tabset-pills}

### Scatters by location and annotation
 
```{r}
ggplot(
    subset(files, !grepl("^[0-9]{1,2}", annotation)),
  aes(y = median_shannon, x = median_avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "median avg. methylation",
    y = "median adj Shannon entropy (adjH)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```



```{r}
adjshannons <- files
```

# Attempt to assemble these all

Being some medians across loci (for a group of cells), and others actually values per cell across loci (!)

```{r}

str(sampens)
str(shannons)
str(adjshannons)
str(adjsampens)
```

```{r save_for_emanuel}
saveRDS(list(sampens    = sampens,
             shannons   = shannons,
             adjshannons = adjshannons,
             adjsampens = adjsampens),
        file = "crc_entropy_summaries.rds")
```


```{r summary_heatmap, fig.width = 10, fig.height = 8}
library(ComplexHeatmap)
library(circlize)

make_hm_mat <- function(df, value_col) {
  wide <- df %>%
    group_by(annotation, location) %>%
    summarise(v = median(.data[[value_col]], na.rm = TRUE), .groups = "drop") %>%
    tidyr::pivot_wider(names_from = location, values_from = v)
  mat <- as.matrix(wide[, -1])
  rownames(mat) <- wide$annotation
  mat
}

hm_adjS <- make_hm_mat(adjsampens, "sampen")
hm_adjH <- make_hm_mat(adjshannons, "median_shannon")

col_adjS <- colorRamp2(
  quantile(hm_adjS, c(0.05, 0.5, 0.95), na.rm = TRUE),
  c("#2166AC", "white", "#B2182B"))
col_adjH <- colorRamp2(
  quantile(hm_adjH, c(0.05, 0.5, 0.95), na.rm = TRUE),
  c("#4DAC26", "white", "#D01C8B"))

draw(
  Heatmap(hm_adjS, name = "adjS", col = col_adjS,
          cluster_columns = FALSE,
          row_names_gp = gpar(fontsize = 7),
          column_names_gp = gpar(fontsize = 7)) +
  Heatmap(hm_adjH, name = "adjH", col = col_adjH,
          cluster_rows = FALSE, cluster_columns = FALSE,
          show_row_names = FALSE,
          column_names_gp = gpar(fontsize = 7)),
  column_title = "Median adjS and adjH per annotation x location"
)
```



```{r, fig.width = 9, fig.height = 12}

library(dplyr)

adj_df <- adjsampens %>%
  left_join(adjshannons, 
            by = c("annotation", "patient", "location"),
            suffix = c("_sampen", "_shannon"))

head(adj_df)
ggplot(
    adj_df,
    aes(x = median_shannon, y = sampen,
        color = location, shape = patient)
) +
    scale_shape_manual(values = 1:nlevels(adj_df$patient)) +
    geom_point(size = 0.8) +
    facet_wrap(~annotation, ncol = 4) +
    labs(
        x = "median adj Shannon entropy (adjH)",
        y = "adj sample entropy (adjS)"
    ) +
  theme_ng()

```


```{r, fig.width = 9, fig.height = 12}
df <- sampens %>%
  left_join(shannons, 
            by = c("annotation", "patient", "location"),
            suffix = c("_sampen", "_shannon"))

head(df)
ggplot(
    df,
    aes(x = median_shannon, y = sampen,
        color = location, shape = patient)
) +
    geom_point(size = 0.8) +
     scale_shape_manual(values = 1:nlevels(df$patient)) +
    facet_wrap(~annotation, ncol = 4) +
    labs(
        x = "median Shannon entropy (H)",
        y = "sample entropy (S)"
    ) +
    theme_ng()
```

```{r}
variation <- adj_df %>%
    group_by(annotation) %>%
    summarise(
        shannon_range = max(median_shannon, na.rm = TRUE) -
            min(median_shannon, na.rm = TRUE),
        sampen_range  = max(sampen, na.rm = TRUE) -
            min(sampen, na.rm = TRUE),
        shannon_sd = sd(median_shannon, na.rm = TRUE),
        sampen_sd  = sd(sampen, na.rm = TRUE)
    )

head(variation)

ggplot(variation, aes(x = shannon_range, y = sampen_range, label = annotation)) +
    geom_point() +
    ggrepel::geom_text_repel() +
    labs(
        x = "adjH range aggregating all biopsy sites",
        y = "adjS range aggregating all biopsy sites",
        ) +
  theme_ng()


## same per location, 

## variation_loc <- adj_df %>%
##     group_by(annotation, location) %>%
##     summarise(
##         shannon_median = median(median_shannon, na.rm = TRUE),
##         sampen_median  = median(sampen, na.rm = TRUE)
##     ) %>%
##     ungroup()

variation_loc <- adj_df %>%
    group_by(annotation, location) %>%
    summarise(
        shannon_median = median(median_shannon, na.rm = TRUE),
        sampen_median  = median(sampen, na.rm = TRUE),
        shannon_sd = sd(median_shannon, na.rm = TRUE),
        sampen_sd  = sd(sampen, na.rm = TRUE),
        shannon_range = max(median_shannon, na.rm = TRUE) -
            min(median_shannon, na.rm = TRUE),
        sampen_range  = max(sampen, na.rm = TRUE) -
            min(sampen, na.rm = TRUE)
    ) %>%
    ungroup()

ggplot(variation_loc,
       aes(x = shannon_median, y = sampen_median,
           color = location)) +
    geom_point() +
    facet_wrap(~annotation) +
    theme_ng() +
    labs(
         x = "median adjH across biopsy sites",
         y = "median adjS across biopsy sites")

ggplot(variation_loc,
       aes(x = shannon_sd, y = sampen_sd,
           color = location)) +
    geom_point() +
    facet_wrap(~annotation) +
    theme_ng() +
    labs(
         x = "sd adjH across biopsy sites",
         y = "sd adjS across biopsy sites")


ggplot(variation_loc,
       aes(x = shannon_range, y = sampen_range,
           color = location)) +
    geom_point() +
    facet_wrap(~annotation) +
    theme_ng() +
    labs(
         x = "adjH range across biopsy sites",
         y = "adjS range across biopsy sites")


```


# Driver categorization

For each annotation, we compute the SD of group-level median adjH (across-cell
heterogeneity) and median adjS (within-cell heterogeneity) across biopsy
locations. If one SD is at least 1.5x the other, that entropy component
dominates. Annotations where both SDs fall below the 30th percentile are
labelled "neither".

```{r drivers, fig.width = 5, fig.height = 5}
crc_grp_for_driver <- adj_df %>%
  rename(median_adjH = median_shannon, median_adjS = sampen)

driver_df <- categorize_drivers(crc_grp_for_driver, "location")

cat("Driver table:\n")
print(table(driver_df$driver))

saveRDS(driver_df, file = "crc_driver_sd_range.rds")

plot_driver_scatter(driver_df,
                    x_label = "SD of adjH across biopsy locations",
                    y_label = "SD of adjS across biopsy locations")
```


# Session

```{r, cache = FALSE}
sessionInfo()
```
