---
title: "CRC by feature (no windows)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Izaskun Mallona"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default # kable
    theme: lumen
---

```{r, setup}
suppressPackageStartupMessages({
  library(readr)
  library(lmerTest)
  library(ggplot2)
  library(knitr)
  #library(effects)
  library(ComplexHeatmap)
  library(viridis)
  library(BiocParallel)
  library(ggplot2)
  library(ggstance)  # for horizontal error bars
  library(dplyr)
})

source("src/ggtheme.R")
param <- MulticoreParam(workers = snakemake@threads)
```

```{r}
opts_chunk$set(
  fig.width = 7.5,
  fig.height = 7.5,
  cache = FALSE,
  include = TRUE,
  fig.path = "crc_plots/",
  dev = c("png", "svg"),
  dpi = 500,
  cache.lazy = FALSE,
  warning = TRUE,
  message = TRUE
)
```

```{r save_snmk_object_for_debugging, include = FALSE}
saveRDS(file = 'snmk_object_features.rds', object = snakemake)
## snakemake <- readRDS('../../snmk_object_features.rds')
## setwd('../..')
```

```{r logging, include = FALSE}
log <- file(snakemake@log$log, open = "wt")
sink(log)
sink(log, type = "message")
```



These are not det.out.gz, but only out.gz files.

```{r, import_short_reports}
filepaths <- list.files(
  snakemake@params[["output_path"]],
  pattern = "\\.out\\.gz$", full.names = TRUE
)

filepaths <- filepaths[!grepl("\\.det\\.out\\.gz$", filepaths)]
filepaths <- filepaths[!grepl("\\.meth\\.out\\.gz$", filepaths)]


regex <- "(.*)_(.*)_(.*)_(.*).out.gz"
process_file <- function(filepath) {
  filename <- basename(filepath)

  dt <- read_tsv(filepath, na = "-1", show_col_types = FALSE)
  dt <- dt[, c("sampen", "avg_meth")]

  dt$annotation <- gsub(regex, "\\1", filename)
  dt$patient <- gsub(regex, "\\3", filename)
  dt$location <- gsub(regex, "\\4", filename)

  return(dt)
}

files <- do.call(rbind, bplapply(filepaths, process_file, BPPARAM = param))

# mind this is per cell, not per locus. This is not something you can do for Shannon.
# imallona@mlshainan:~/src/yamet/workflow$ zcat data/crc/output/sines_sines_CRC01_LN.out.gz | head
#file	sampen	sampen_norm	avg_meth
#data/crc/harmonized/GSM2697511_scTrioSeq2Met_CRC01_LN3_268.singleC.txt.gz	0.0984465	0.311104	0.838184
#data/crc/harmonized/GSM2697889_scTrioSeq2Met_CRC01_LN1_164.singleC.txt.gz	0.0985076	0.290518	0.825905
#data/crc/harmonized/GSM2697915_scTrioSeq2Met_CRC01_LN3_447.singleC.txt.gz	0.098421	0.307972	0.836493
## data/crc/harmonized/GSM2697508_scTrioSeq2Met_CRC01_LN2_456.singleC.txt.gz	0.108082	0.316819	0.824773
## data/crc/harmonized/GSM2697493_scTrioSeq2Met_CRC01_LN1_175.singleC.txt.gz	0.114474	0.312642	0.810966
## data/crc/harmonized/GSM2697905_scTrioSeq2Met_CRC01_LN2_453.singleC.txt.gz	0.100684	0.305491	0.831078
## data/crc/harmonized/GSM2697898_scTrioSeq2Met_CRC01_LN2_192.singleC.txt.gz	0.117015	0.359772	0.833428
## data/crc/harmonized/GSM2697908_scTrioSeq2Met_CRC01_LN3_244.singleC.txt.gz	0.101921	0.28672	0.816892
## data/crc/harmonized/GSM2697909_scTrioSeq2Met_CRC01_LN3_245.singleC.txt.gz	0.0962967	0.299841	0.835639


## files$annotation <- factor(files$annotation,
##                            levels = c("0_Enhancer",
##                                       "1_Transcribed",
##                                       "2_Enhancer",
##                                       "3_Quiescent",
##                                       "4_Transcribed",     
##                                       "5_RegPermissive",
##                                       "6_LowConfidence",
##                                       "7_RegPermissive",
##                                       "8_Quiescent",
##                                       "9_ConstitutiveHet",
##                                       "10_Quiescent",
##                                       "11_Promoter",
##                                       "12_Promoter",
##                                       "13_ConstitutiveHet",                                                                            
##                                       "cpgIslandExt", "H3K4me3", "genes",
##                                       "H3K27me3", "H3K9me3", "laminb1",
##                                       "pmds", "lines", "sines", "hmds",
##                                       "crc01_lost_scna", "crc01_nc_scna", "crc01_gain_scna"))


## files$patient <- as.factor(files$patient)
## files$location <- factor(
##   files$location,
##   levels = c("NC", "PT", "LN", "ML", "MP", "MO")
## )

## head(files)

## harmonize_levels <- function(df) {

##   df$annotation <- factor(
##       df$annotation,
##       levels = c(
##           "0_Enhancer", "1_Transcribed", "2_Enhancer", "3_Quiescent",
##           "4_Transcribed", "5_RegPermissive", "6_LowConfidence",
##           "7_RegPermissive", "8_Quiescent", "9_ConstitutiveHet",
##           "10_Quiescent", "11_Promoter", "12_Promoter", "13_ConstitutiveHet",
##           "cpgIslandExt", "H3K4me3", "genes", "H3K27me3", "H3K9me3",
##           "laminb1", "pmds", "lines", "sines", "hmds",
##           "crc01_lost_scna", "crc01_nc_scna", "crc01_gain_scna"
##     )
##   )

##   df$patient <- as.factor(df$patient)
##   df$location <- factor(df$location, levels = c("NC", "PT", "LN", "ML", "MP", "MO"))

##   return(df)
## }


## prettifies and sorts categorical vars
harmonize_levels <- function(df) { 
    pretty_hmm <- c(
        "0_Enhancer"        = "0 Enh.",
        "1_Transcribed"     = "1 Transc. ",
        "2_Enhancer"        = "2 Enh.",
        "3_Quiescent"       = "3 Quies.",
        "4_Transcribed"     = "4 Transc.",
        "5_RegPermissive"   = "5 Reg. perm.",
        "6_LowConfidence"   = "6 Low conf.",
        "7_RegPermissive"   = "7 Reg. perm.",
        "8_Quiescent"       = "8 Quies.",
        "9_ConstitutiveHet" = "9 Cons. het.",
        "10_Quiescent"      = "10 Quies.",
        "11_Promoter"       = "11 Prom.",
        "12_Promoter"       = "12 Prom.",
        "13_ConstitutiveHet"= "13 Cons. het"
    )

    pretty_other <- c(
        "cpgIslandExt"      = "CpGi",
        "H3K4me3"           = "H3K4me3",
        "genes"             = "Genes",
        "H3K27me3"          = "H3K27me3",
        "H3K9me3"           = "H3K9me3",
        "laminb1"           = "Lamin B1",
        "pmds"              = "PMDs",
        "lines"             = "LINEs",
        "sines"             = "SINEs",
        "hmds"              = "HMDs",
        "crc01_lost_scna"   = "SCNA loss",
        "crc01_nc_scna"     = "SCNA neutral",
        "crc01_gain_scna"   = "SCNA gain")

    all_levels <- c(pretty_hmm, pretty_other)

    missing <- setdiff(unique(df$annotation), names(all_levels))
    if (length(missing) > 0) {
        stop(missing)
    }
    

  df$annotation <- factor(df$annotation,
                          levels = names(all_levels),
                          labels = unname(all_levels))

  df$patient  <- as.factor(df$patient)
  df$location <- factor(df$location, levels = c("NC","PT","LN","ML","MP","MO"))

  df
}

files <- harmonize_levels(files)
table(files$annotation, useNA = 'always')

```

# Bulk unnormalized sampen  {.tabset .tabset-pills}

## Scatters by location and annotation
 
```{r}
ggplot(
  ## subset(files, patient == "CRC01"),
  files,
  aes(y = sampen, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation) +
  labs(
    x = "average methylation",
    y = "sample entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Scatters by location and annotation and patient
 
```{r, fig.height = 15}
ggplot(
  ## subset(files, patient == "CRC01"),
  files,
  aes(y = sampen, x = avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_grid(annotation ~ patient) +
  labs(
    x = "average methylation",
    y = "sample entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Scatters by location and annotation and patient (shorter)
 
```{r, fig.height = 3.2}
ggplot(
  subset(files, annotation %in% c("laminb1", "H3K27me3", "H3K9me3")),
  aes(y = sampen, x = avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_grid(annotation ~ patient) +
  labs(
    x = "average methylation",
    y = "sample entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Segmentations  {.tabset .tabset-pills} 

### Scatters by location and annotation
 
```{r}
ggplot(
    subset(files, grepl("^[0-9]{1,2}", annotation)),
  aes(y = sampen, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "avg. methylation",
    y = "sample entropy (S)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```


## Nonsegmentations  {.tabset .tabset-pills}

### Scatters by location and annotation
 
```{r}
ggplot(
    subset(files, !grepl("^[0-9]{1,2}", annotation)),
  aes(y = sampen, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "avg. methylation",
    y = "sample entropy (S)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Others  {.tabset .tabset-pills}

### sampEn boxes by location and annotation - crc01

```{r, fig.height = 6, fig.width = 10}
ggplot(
  subset(files, patient == "CRC01"),
  aes(y = sampen, x = annotation, color = location)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = "sample entropy") +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### meth boxes by location and annotation - crc01

```{r, fig.height = 6, fig.width = 10}
ggplot(
  subset(files, patient == "CRC01"),
  aes(y = avg_meth, x = annotation, color = location)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = "average methylation") +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### Range plots - by compartment

```{r, fig.height = 10, fig.width = 10}


ranges <- as.data.frame(files %>%
   filter(patient == "CRC01") %>%
   group_by(compartment = annotation, location) %>%
   summarise(
     sampen_mean = mean(sampen, na.rm = TRUE),
     sampen_min  = min(sampen, na.rm = TRUE),
     sampen_max  = max(sampen, na.rm = TRUE),
     meth_mean   = mean(avg_meth, na.rm = TRUE),
     meth_min    = min(avg_meth, na.rm = TRUE),
     meth_max    = max(avg_meth, na.rm = TRUE)))

## ggplot(ranges, aes(x = meth_mean, y = sampen_mean, color = location)) +
##   geom_point(shape = 3, size = 1.2) +  # "+" symbol
##   geom_errorbar(aes(ymin = sampen_min, ymax = sampen_max), width = 0.3, alpha = 1, size = 1) +
##   geom_errorbarh(aes(xmin = meth_min, xmax = meth_max), height = 0.3, alpha = 1, size = 0) +
##   scale_color_brewer(palette = "Dark2") +
##   labs(
##     x = "DNA methylation",
##     y = "Sample entropy"
##   ) +
##   facet_wrap(~ compartment) +
##   theme_ng()

# I think this is the correct one
ggplot(ranges, aes(color = location)) +
  ## horizontal segment: range of methylations for that location/compartment
  geom_segment(
    aes(x = meth_min, xend = meth_max, y = sampen_mean, yend = sampen_mean),
    size = 0.6, alpha = 1
  ) +
  # vertical segment: range of sampens for that location/compartment
  geom_segment(
    aes(x = meth_mean, xend = meth_mean, y = sampen_min, yend = sampen_max),
    size = 0.6, alpha = 1
  ) +
  # center point
  geom_point(
    aes(x = meth_mean, y = sampen_mean),
    shape = 3, size = 1.2
  ) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "DNA methylation", y = "Sample entropy") +
  facet_wrap(~ compartment) +
  theme_ng()

```

```{r}
sampens <- files
```

# With adjS meth-norm sampen {.tabset .tabset-pills}

That is, not by locus but by cell norm sampEns (that are calculated according to some regionset anyway)

```{r, import_short_reports_norm}
filepaths <- list.files(
  snakemake@params[["output_path"]],
  pattern = "out\\.gz$", full.names = TRUE)


filepaths <- filepaths[!grepl("\\.det\\.out\\.gz$", filepaths)]
filepaths <- filepaths[!grepl("\\.norm\\.out\\.gz$", filepaths)]
filepaths <- filepaths[!grepl("\\.meth\\.out\\.gz$", filepaths)]

head(filepaths)
regex <- "(.*)_(.*)_(.*)_(.*).out.gz"
process_file <- function(filepath) {
  filename <- basename(filepath)

  dt <- read_tsv(filepath, na = "-1", show_col_types = FALSE)
  dt <- dt[, c("sampen_norm", "avg_meth")]
  colnames(dt) <- c('sampen', 'avg_meth') ## caution they're normalized though

  dt$annotation <- gsub(regex, "\\1", filename)
  dt$patient <- gsub(regex, "\\3", filename)
  dt$location <- gsub(regex, "\\4", filename)

  return(dt)
}


files <- do.call(rbind, bplapply(filepaths, process_file, BPPARAM = param))

files <- harmonize_levels(files)

```

## Scatters by location and annotation
 
```{r}
ggplot(
  ## subset(files, patient == "CRC01"),
  files,
  aes(y = sampen, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation) +
  labs(
    x = "avg. methylation",
    y = "adj. sample entropy (adjS)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Scatters by location and annotation and patient
 
```{r, fig.height = 15}
ggplot(
  ## subset(files, patient == "CRC01"),
  files,
  aes(y = sampen, x = avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_grid(annotation ~ patient) +
  labs(
    x = "avg. methylation",
    y = "adj. sample entropy (adjS)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Scatters by location and annotation and patient (shorter)
 
```{r, fig.height = 3.2}
ggplot(
  subset(files, annotation %in% c("laminb1", "H3K27me3", "H3K9me3")),
  aes(y = sampen, x = avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_grid(annotation ~ patient) +
  labs(
    x = "avg. methylation",
    y = "adj. sample entropy (adjS)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```


## Segmentations {.tabset .tabset-pills}

### Scatters by location and annotation
 
```{r}
ggplot(
    subset(files, grepl("^[0-9]{1,2}", annotation)),
  aes(y = sampen, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "avg. methylation",
    y = "adj. sample entropy (adjS)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```


## Nonsegmentations {.tabset .tabset-pills}

### Scatters by location and annotation
 
```{r}
ggplot(
    subset(files, !grepl("^[0-9]{1,2}", annotation)),
  aes(y = sampen, x = avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "avg. methylation",
    y = "adj. sample entropy (adjS)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Others  {.tabset .tabset-pills}

### sampEn boxes by location and annotation - crc01

```{r, fig.height = 6, fig.width = 10}
ggplot(
  subset(files, patient == "CRC01"),
  aes(y = sampen, x = annotation, color = location)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = "adj sample entropy (adjS)") +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### meth boxes by location and annotation - crc01

```{r, fig.height = 6, fig.width = 10}
ggplot(
  subset(files, patient == "CRC01"),
  aes(y = avg_meth, x = annotation, color = location)
) +
  geom_boxplot(
    outlier.size = 0.6,
    outlier.alpha = 0.3
  ) +
  labs(y = "avg. methylation") +
  geom_jitter(
    position = position_dodge(width = 0.75),
    size = 0.6, alpha = 0.3
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### Range plots - by compartment

```{r, fig.height = 10, fig.width = 10}
ranges <- as.data.frame(files %>%
   filter(patient == "CRC01") %>%
   group_by(compartment = annotation, location) %>%
   summarise(
     sampen_mean = mean(sampen, na.rm = TRUE),
     sampen_min  = min(sampen, na.rm = TRUE),
     sampen_max  = max(sampen, na.rm = TRUE),
     meth_mean   = mean(avg_meth, na.rm = TRUE),
     meth_min    = min(avg_meth, na.rm = TRUE),
     meth_max    = max(avg_meth, na.rm = TRUE)))

## ggplot(ranges, aes(x = meth_mean, y = sampen_mean, color = location)) +
##   geom_point(shape = 3, size = 1.2) +  # "+" symbol
##   geom_errorbar(aes(ymin = sampen_min, ymax = sampen_max), width = 0.3, alpha = 1, size = 1) +
##   geom_errorbarh(aes(xmin = meth_min, xmax = meth_max), height = 0.3, alpha = 1, size = 0) +
##   scale_color_brewer(palette = "Dark2") +
##   labs(
##     x = "DNA methylation",
##     y = "Sample entropy"
##   ) +
##   facet_wrap(~ compartment) +
##   theme_ng()

# I think this is the correct one
ggplot(ranges, aes(color = location)) +
  ## horizontal segment: range of methylations for that location/compartment
  geom_segment(
    aes(x = meth_min, xend = meth_max, y = sampen_mean, yend = sampen_mean),
    size = 0.6, alpha = 1
  ) +
  # vertical segment: range of sampens for that location/compartment
  geom_segment(
    aes(x = meth_mean, xend = meth_mean, y = sampen_min, yend = sampen_max),
    size = 0.6, alpha = 1
  ) +
  # center point
  geom_point(
    aes(x = meth_mean, y = sampen_mean),
    shape = 3, size = 1.2
  ) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "DNA methylation", y = "adj. sample entropy (adjS)") +
  facet_wrap(~ compartment) +
  theme_ng()

```

```{r}
adjsampens <- files
```

# Unnormalized Shannon's MEDIANS DOWNSAMPLED {.tabset .tabset-pills}

This cannot be done on short `out.gz` files, but on detailed ones only. Median values, not all loci included!

meaning, here we'll have one value per biopsy+patient (the "group") only, because the individual loci are averaged. And there are no individual cells anymore (it's Shannon).

```{r, eval = TRUE}
filepaths <- list.files(
  snakemake@params[["output_path"]],
  pattern = "det\\.out\\.gz$", full.names = TRUE
)

filepaths <- filepaths[!grepl("norm", filepaths)]

head(filepaths)

## todo extract Shannon's here instead

regex <- "(.*)_(.*)_(.*)_(.*).det.out.gz"

process_file <- function(filepath) {
  filename <- basename(filepath)

  tmp <- read_tsv(filepath, na = "-1", show_col_types = FALSE, n_max = 10e4) ## caution downsampling

  tibble(
    file = filename,
    median_shannon   = median(tmp$shannon,   na.rm = TRUE),
    median_avg_meth  = median(tmp$avg_meth,  na.rm = TRUE),
    annotation = gsub(regex, "\\1", filename),
    patient    = gsub(regex, "\\3", filename),
    location   = gsub(regex, "\\4", filename)
  )
}

files <- bplapply(filepaths, process_file, BPPARAM = param) |> bind_rows()

files <- harmonize_levels(files)

head(files)
```

## Scatters by location and annotation
 
```{r}
ggplot(
  files,
  aes(y = median_shannon, x = median_avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation) +
  labs(
    x = "median average methylation",
    y = "median Shannon entropy (H)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```



## Segmentations {.tabset .tabset-pills}

### Scatters by location and annotation
 
```{r}
ggplot(
    subset(files, grepl("^[0-9]{1,2}", annotation)),
  aes(y = median_shannon, x = median_avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "median avg. methylation",
    y = "median Shannon entropy (H)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```


## Nonsegmentations {.tabset .tabset-pills}

### Scatters by location and annotation
 
```{r}
ggplot(
    subset(files, !grepl("^[0-9]{1,2}", annotation)),
  aes(y = median_shannon, x = median_avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "median avg. methylation",
    y = "median Shannon entropy (H)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```



```{r}
shannons <- files
```

# Meth-normalized Shannons (the yamet way) MEDIANS {.tabset .tabset-pills}

This cannot be done on short `out.gz` files, but on detailed ones only. Median values, not all loci included!

```{r, eval = TRUE}
filepaths <- list.files(
  snakemake@params[["output_path"]],
  pattern = "norm\\.det\\.out\\.gz$", full.names = TRUE
)

head(filepaths)

## todo extract Shannon's here instead

regex <- "(.*)_(.*)_(.*)_(.*).norm.det.out.gz"

process_file <- function(filepath) {
  filename <- basename(filepath)

  tmp <- read_tsv(filepath, na = "-1", show_col_types = FALSE, n_max = 10e4) # caution downsampling

  tibble(
    file = filename,
    median_shannon   = median(tmp$shannon_norm,   na.rm = TRUE),
    median_avg_meth  = median(tmp$avg_meth,  na.rm = TRUE),
    annotation = gsub(regex, "\\1", filename),
    patient    = gsub(regex, "\\3", filename),
    location   = gsub(regex, "\\4", filename)
  )
}

files <- bplapply(filepaths, process_file, BPPARAM = param) |> bind_rows()

files <- harmonize_levels(files)

```

## Scatters by location and annotation
 
```{r}
ggplot(
  files,
  aes(y = median_shannon, x = median_avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation) +
  labs(
    x = "median average methylation",
    y = "median norm Shannon entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Scatters by location and annotation and patient
 
```{r, fig.height = 15}
ggplot(
  files,
  aes(y = median_shannon, x = median_avg_meth, color = location)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  ## scale_color_brewer(palette = "Set1") +
  ## facet_grid(annotation ~ patient) +
  labs(
    x = "median average methylation",
    y = "median norm Shannon entropy"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```



## Segmentations {.tabset .tabset-pills}

### Scatters by location and annotation
 
```{r}
ggplot(
    subset(files, grepl("^[0-9]{1,2}", annotation)),
  aes(y = median_shannon, x = median_avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "median avg. methylation",
    y = "median adj Shannon entropy (adjH)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```


## Nonsegmentations {.tabset .tabset-pills}

### Scatters by location and annotation
 
```{r}
ggplot(
    subset(files, !grepl("^[0-9]{1,2}", annotation)),
  aes(y = median_shannon, x = median_avg_meth, color = location, shape = patient)
) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  scale_shape_manual(values = 1:nlevels(files$patient)) +
  facet_wrap(~annotation, ncol = 3, nrow = 5) +
  labs(
    x = "median avg. methylation",
    y = "median adj Shannon entropy (adjH)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```



```{r}
adjshannons <- files
```

# Attempt to assemble these all

Being some medians across loci (for a group of cells), and others actually values per cell across loci (!)

```{r}

str(sampens)
str(shannons)
str(adjshannons)
str(adjsampens)
```


```{r, fig.height = 12, fig.width = 6}

library(dplyr)

adj_df <- adjsampens %>%
  left_join(adjshannons, 
            by = c("annotation", "patient", "location"),
            suffix = c("_sampen", "_shannon"))

head(adj_df)
ggplot(
    adj_df,
    aes(x = median_shannon, y = sampen,
        color = location, shape = patient)
) +
    scale_shape_manual(values = 1:nlevels(adj_df$patient)) +
    geom_point(size = 0.8) +
    facet_wrap(~annotation, ncol = 4) +
    labs(
        x = "median adj Shannon entropy (adjH)",
        y = "adj sample entropy (adjS)"
    ) +
  theme_ng()

```


```{r, fig.height = 12, fig.width = 6}
df <- sampens %>%
  left_join(shannons, 
            by = c("annotation", "patient", "location"),
            suffix = c("_sampen", "_shannon"))

head(df)
ggplot(
    df,
    aes(x = median_shannon, y = sampen,
        color = location, shape = patient)
) +
    geom_point(size = 0.8) +
     scale_shape_manual(values = 1:nlevels(df$patient)) +
    facet_wrap(~annotation, ncol = 4) +
    labs(
        x = "median Shannon entropy (H)",
        y = "sample entropy (S)"
    ) +
    theme_ng()
```

```{r}
variation <- adj_df %>%
    group_by(annotation) %>%
    summarise(
        shannon_range = max(median_shannon, na.rm = TRUE) -
            min(median_shannon, na.rm = TRUE),
        sampen_range  = max(sampen, na.rm = TRUE) -
            min(sampen, na.rm = TRUE),
        shannon_sd = sd(median_shannon, na.rm = TRUE),
        sampen_sd  = sd(sampen, na.rm = TRUE)
    )

head(variation)

ggplot(variation, aes(x = shannon_range, y = sampen_range, label = annotation)) +
    geom_point() +
    ggrepel::geom_text_repel() +
    labs(
        x = "adjH range aggregating all biopsy sites",
        y = "adjS range aggregating all biopsy sites",
        ) +
  theme_ng()


## same per location, 

## variation_loc <- adj_df %>%
##     group_by(annotation, location) %>%
##     summarise(
##         shannon_median = median(median_shannon, na.rm = TRUE),
##         sampen_median  = median(sampen, na.rm = TRUE)
##     ) %>%
##     ungroup()

variation_loc <- adj_df %>%
    group_by(annotation, location) %>%
    summarise(
        shannon_median = median(median_shannon, na.rm = TRUE),
        sampen_median  = median(sampen, na.rm = TRUE),
        shannon_sd = sd(median_shannon, na.rm = TRUE),
        sampen_sd  = sd(sampen, na.rm = TRUE),
        shannon_range = max(median_shannon, na.rm = TRUE) -
            min(median_shannon, na.rm = TRUE),
        sampen_range  = max(sampen, na.rm = TRUE) -
            min(sampen, na.rm = TRUE)
    ) %>%
    ungroup()

ggplot(variation_loc,
       aes(x = shannon_median, y = sampen_median,
           color = location)) +
    geom_point() +
    facet_wrap(~annotation) +
    theme_ng() +
    labs(
         x = "median adjH across biopsy sites",
         y = "median adjS across biopsy sites")

ggplot(variation_loc,
       aes(x = shannon_sd, y = sampen_sd,
           color = location)) +
    geom_point() +
    facet_wrap(~annotation) +
    theme_ng() +
    labs(
         x = "sd adjH across biopsy sites",
         y = "sd adjS across biopsy sites")


ggplot(variation_loc,
       aes(x = shannon_range, y = sampen_range,
           color = location)) +
    geom_point() +
    facet_wrap(~annotation) +
    theme_ng() +
    labs(
         x = "adjH range across biopsy sites",
         y = "adjS range across biopsy sites")


```


Clustering the behaviours/drivers, according to the sample entropy vs shannon entropy ratio (can we standardize this somehow)? or shall we quantilize in low, high etc? Or, rather, quantilize and categorize by the within-het or across-het prevalence


Categorizing the drivers - this is ongoing and not final

```{r}
## categorizing in shannon- or sample- driven

variation <- variation %>%
    mutate(
        driver = case_when(
            shannon_range >= sampen_range * 1.5 ~ "across cells heterogeneity-driven",
            sampen_range  >= shannon_range * 1.5 ~ "within cells heterogeneity‑driven",
            TRUE ~ "both"))

ggplot(variation, aes(x = shannon_range, y = sampen_range, color = driver, shape = driver, label = annotation)) +
    geom_point(size = 3) +
    ggrepel::geom_text_repel() +
    labs(
        x = "adjH range across biopsy sites",
        y = "adjS range across biopsy sites",
        ) +
    theme_ng()


```

What about small changes? some of these don't change much in either adjS nor adjH, so we could label them accordingly. Why is this flaring for the SCNA things, if they were no changes?


Caution this actually makes no sense

```{r}
## this actually makes no sense
sds <- df %>%
    group_by(annotation) %>%
    summarise(
        shannon_sd = sd(median_shannon, na.rm = TRUE),
        sampen_sd  = sd(sampen, na.rm = TRUE)
    )

sds

shannon_thr <- quantile(sds$shannon_sd, 0.3, na.rm = TRUE)
sampen_thr  <- quantile(sds$sampen_sd, 0.3, na.rm = TRUE)

sds <- sds %>%
    mutate(
        driver = case_when(
            shannon_sd < shannon_thr & sampen_sd < sampen_thr ~ "not changing enough",
            shannon_sd >= sampen_sd * 1.5 ~ "across cells heterogeneity-driven",
            sampen_sd  >= shannon_sd * 1.5 ~ "within cells heterogeneity‑driven",
            TRUE ~ "both"
        )
    )

table(sds$driver)

ggplot(sds,
       aes(x = shannon_sd, y = sampen_sd,
           color = driver, shape = driver, label = annotation)) +
    geom_point(size = 3) +
    ggrepel::geom_text_repel() +
    labs(
        x = "SD of adjH across biopsy sites",
        y = "SD of adjS across biopsy sites"
    ) +
    theme_ng()


```


Hmm, this makes no sense, why the SCNA is there? would it make sense to plot the ranges instead?


```{r}
variation <- adj_df %>%
    group_by(annotation) %>%
    summarise(
        shannon_range = max(median_shannon, na.rm = TRUE) - min(median_shannon, na.rm = TRUE),
        sampen_range  = max(sampen, na.rm = TRUE) - min(sampen, na.rm = TRUE)
    )

shannon_thr <- quantile(variation$shannon_range, 0.5, na.rm = TRUE)
sampen_thr  <- quantile(variation$sampen_range, 0.5, na.rm = TRUE)

variation <- variation %>%
    mutate(
        driver = case_when(
            shannon_range < shannon_thr & sampen_range < sampen_thr ~ "not changing enough",
            shannon_range >= sampen_range * 1.5 ~ "across cells heterogeneity-driven",
            sampen_range  >= shannon_range * 1.5 ~ "within cells heterogeneity‑driven",
            TRUE ~ "both"
        )
    )

table(variation$driver)
ggplot(variation,
       aes(x = shannon_range, y = sampen_range,
           color = driver, shape = driver, label = annotation)) +
    geom_point(size = 3) +
    ggrepel::geom_text_repel() +
    labs(
        x = "adjH range across biopsy sites",
        y = "adjS range across biopsy sites"
    ) +
    theme_ng()



```


This might make sense, or not, but again it does more sense to do only in PT vs NC, and each metastasis vs PT separately, todo



# Session

```{r, cache = FALSE}
sessionInfo()
```
