---
title: "CRC: cell embeddings from epigenetic heterogeneity"
author: "Izaskun Mallona"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
---

```{r setup}
suppressPackageStartupMessages({
  library(readr)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(BiocParallel)
  library(uwot)
  library(matrixStats)
  library(SingleCellExperiment)
  library(SummarizedExperiment)
  library(scater)
  library(cluster)
  library(ComplexHeatmap)
  library(circlize)
  library(ggrepel)
  library(viridis)
  library(cowplot)
})

source("src/ggtheme.R")
param <- MulticoreParam(workers = snakemake@threads)
```

```{r opts}
knitr::opts_chunk$set(
  echo       = TRUE,
  fig.width  = ng_fig_size(1, 1)$w,
  fig.height = ng_fig_size(1, 1)$h,
  cache      = FALSE,
  include    = TRUE,
  fig.path   = "crc_embeddings_plots/",
  dev        = c("png", "svg"),
  dpi        = 500,
  cache.lazy = FALSE,
  warning    = TRUE,
  message    = TRUE
)
```

```{r logging, include = FALSE}
log <- file(snakemake@log$log, open = "wt")
sink(log)
sink(log, type = "message")
```

Two complementary feature spaces are compared for cell embedding:

1. **Per-window**: cells embedded using entropy across all 10 kb genomic tiles (high-
   dimensional, spatially resolved). HVW selection applied before PCA and UMAP.
2. **Per-annotation**: cells embedded using mean entropy within each named genomic
   annotation (compact, biologically interpretable). Also run with HVW -> PCA -> UMAP
   independently for each annotation.

Four metrics are tested in both spaces: S (raw sample entropy), adjS (feature-level
methylation correction), S_cellmeth (cell-level methylation regressed out from S as
an independent check), and DNA methylation as baseline.

```{r constants}
loc_levels <- c("NC", "PT", "LN", "ML", "MP", "MO")
loc_colors <- setNames(cbbPalette[seq_along(loc_levels)], loc_levels)

N_HVW <- 500   ## top variable windows to retain before PCA
N_PCS <- 10    ## PCs to compute
```

# Load data

```{r load_sce}
sce <- readRDS(snakemake@input[["corrected_sce"]])
colData(sce)$location <- factor(colData(sce)$location, levels = loc_levels)

cat("Cells:", ncol(sce), "\n")
cat("Windows:", nrow(sce), "\n")
cat("Assays:", paste(assayNames(sce), collapse = ", "), "\n")
cat("Locations:\n")
print(table(colData(sce)$location, colData(sce)$patient, useNA = "ifany"))
```

```{r filter_impute}
na_frac <- rowMeans(is.na(assay(sce, "sampen")))
sce <- sce[na_frac < 0.3, ]

impute_rowmeans <- function(mat) {
  row_means <- rowMeans(mat, na.rm = TRUE)
  for (i in seq_len(nrow(mat))) {
    nas <- is.na(mat[i, ])
    if (any(nas)) mat[i, nas] <- row_means[i]
  }
  mat
}

assay(sce, "sampen")           <- impute_rowmeans(as.matrix(assay(sce, "sampen")))
assay(sce, "meth")             <- impute_rowmeans(as.matrix(assay(sce, "meth")))
assay(sce, "sampen_corrected") <- impute_rowmeans(as.matrix(assay(sce, "sampen_corrected")))

cat("Windows after NA filter:", nrow(sce), "\n")
```

```{r cell_meth_correction}
## S_cellmeth: for each window regress out each cell's global mean methylation from
## S. This corrects at the cell level rather than at the window level (adjS), and
## tests whether separation survives independent of a cell-level meth gradient.
cell_mean_meth <- colMeans(assay(sce, "meth"), na.rm = TRUE)

fit_residuals <- function(i, y_mat, x) {
  y <- y_mat[i, ]
  fit <- try(lm(y ~ x), silent = TRUE)
  if (inherits(fit, "try-error")) return(y)
  residuals(fit)
}

s_mat <- as.matrix(assay(sce, "sampen"))
s_cellmeth_mat <- do.call(rbind,
  bplapply(seq_len(nrow(s_mat)), fit_residuals,
           y_mat = s_mat, x = cell_mean_meth, BPPARAM = param))
rownames(s_cellmeth_mat) <- rownames(sce)
colnames(s_cellmeth_mat) <- colnames(sce)
assay(sce, "sampen_cellmeth") <- s_cellmeth_mat
rm(s_mat, s_cellmeth_mat)
```

# Shared helpers

```{r helpers}
## Select indices of the top n_hvw most variable rows (windows) by across-cell variance.
select_hvw <- function(mat, n_hvw = N_HVW) {
  vars <- rowVars(mat, useNames = FALSE)
  order(vars, decreasing = TRUE)[seq_len(min(n_hvw, length(vars)))]
}

## PCA on a features x cells matrix; returns cells x n_pcs score matrix.
run_pca_mat <- function(mat, n_pcs = N_PCS) {
  n_pcs <- min(n_pcs, nrow(mat) - 1, ncol(mat) - 1)
  if (n_pcs < 2) return(NULL)
  prcomp(t(mat), center = TRUE, scale. = FALSE, rank. = n_pcs)$x
}

## UMAP on a features x cells matrix (transposes internally).
run_umap_feat <- function(mat, seed = 123, n_neighbors = 15, min_dist = 0.3) {
  set.seed(seed)
  umap(t(mat), n_neighbors = min(n_neighbors, ncol(mat) - 1),
       min_dist = min_dist, metric = "euclidean", verbose = FALSE)
}

## UMAP on an already-reduced cells x dims matrix (no transpose).
run_umap_scores <- function(scores, seed = 123, n_neighbors = 15, min_dist = 0.3) {
  set.seed(seed)
  umap(scores, n_neighbors = min(n_neighbors, nrow(scores) - 1),
       min_dist = min_dist, metric = "euclidean", verbose = FALSE)
}

## Full HVW -> PCA -> UMAP pipeline on a features x cells matrix.
hvw_pca_umap <- function(mat, n_hvw = N_HVW, n_pcs = N_PCS) {
  idx  <- select_hvw(mat, n_hvw)
  pca  <- run_pca_mat(mat[idx, , drop = FALSE], n_pcs)
  if (is.null(pca)) return(NULL)
  list(umap = run_umap_scores(pca), pca = pca, hvw_idx = idx)
}

sil_score <- function(coords, labels) {
  lv <- as.integer(as.factor(labels))
  if (length(unique(lv)) < 2) return(NA_real_)
  mean(silhouette(lv, dist(coords))[, "sil_width"])
}

plot_pair <- function(coords, sce_obj, title_prefix) {
  df <- data.frame(
    umap1    = coords[, 1], umap2 = coords[, 2],
    location = colData(sce_obj)$location,
    patient  = colData(sce_obj)$patient
  )
  p_loc <- ggplot(df, aes(x = umap1, y = umap2, color = location)) +
    geom_point(size = 0.6, alpha = 0.7) +
    scale_color_manual(values = loc_colors) +
    labs(title = paste(title_prefix, "- location"), x = "UMAP 1", y = "UMAP 2") +
    guides(x = guide_x_nolap(),
           color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
    theme_ng()
  p_pat <- ggplot(df, aes(x = umap1, y = umap2, color = patient)) +
    geom_point(size = 0.6, alpha = 0.7) +
    scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
    labs(title = paste(title_prefix, "- patient"), x = "UMAP 1", y = "UMAP 2") +
    guides(x = guide_x_nolap(),
           color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
    theme_ng()
  plot_grid(p_loc, p_pat, nrow = 1)
}

## Named list mapping assay names to display labels, including S_cellmeth.
assay_map <- c(
  "sampen"           = "S",
  "sampen_corrected" = "adjS",
  "sampen_cellmeth"  = "S_cellmeth",
  "meth"             = "methylation"
)
```

# Per-window cell embeddings (HVW -> PCA -> UMAP)

Top `r N_HVW` most variable windows are selected per assay before PCA and UMAP.

```{r win_hvw_pca_umap}
win_embeds <- lapply(names(assay_map), function(akey) {
  hvw_pca_umap(as.matrix(assay(sce, akey)))
})
names(win_embeds) <- assay_map
```

## By location {.tabset .tabset-pills}

```{r win_loc_S, fig.width = ng_fig_size(2, 1)$w, fig.height = ng_fig_size(2, 1)$h}
### S
print(plot_pair(win_embeds[["S"]]$umap, sce, "S (windows)"))
```

```{r win_loc_adjS, fig.width = ng_fig_size(2, 1)$w, fig.height = ng_fig_size(2, 1)$h}
### adjS
print(plot_pair(win_embeds[["adjS"]]$umap, sce, "adjS (windows)"))
```

```{r win_loc_scm, fig.width = ng_fig_size(2, 1)$w, fig.height = ng_fig_size(2, 1)$h}
### S_cellmeth
print(plot_pair(win_embeds[["S_cellmeth"]]$umap, sce, "S_cellmeth (windows)"))
```

```{r win_loc_meth, fig.width = ng_fig_size(2, 1)$w, fig.height = ng_fig_size(2, 1)$h}
### methylation
print(plot_pair(win_embeds[["methylation"]]$umap, sce, "methylation (windows)"))
```

## Four-panel comparison (location)

```{r win_panel, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(4, 1)$h}
df_win_panel <- bind_rows(lapply(names(win_embeds), function(lbl) {
  data.frame(umap1 = win_embeds[[lbl]]$umap[, 1],
             umap2 = win_embeds[[lbl]]$umap[, 2],
             location = colData(sce)$location,
             assay = lbl)
}))
df_win_panel$assay <- factor(df_win_panel$assay, levels = assay_map)

ggplot(df_win_panel, aes(x = umap1, y = umap2, color = location)) +
  geom_point(size = 0.4, alpha = 0.6) +
  scale_color_manual(values = loc_colors) +
  facet_wrap(~ assay, nrow = 1) +
  labs(x = "UMAP 1", y = "UMAP 2", title = "Per-window embeddings (HVW -> PCA -> UMAP)") +
  guides(x = guide_x_nolap(),
         color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng()
```

## PCA (first two PCs, per assay)

```{r win_pca_plot, fig.width = ng_fig_size(4, 2)$w, fig.height = ng_fig_size(4, 2)$h}
df_win_pca <- bind_rows(lapply(names(win_embeds), function(lbl) {
  pca <- win_embeds[[lbl]]$pca
  data.frame(PC1 = pca[, 1], PC2 = pca[, 2],
             location = colData(sce)$location,
             patient  = colData(sce)$patient,
             assay    = lbl)
}))
df_win_pca$assay <- factor(df_win_pca$assay, levels = assay_map)

ggplot(df_win_pca, aes(x = PC1, y = PC2, color = location)) +
  geom_point(size = 0.5, alpha = 0.6) +
  scale_color_manual(values = loc_colors) +
  facet_grid(assay ~ .) +
  labs(x = "PC 1", y = "PC 2", title = "PCA: per-window HVW") +
  guides(x = guide_x_nolap(),
         color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng()
```

## Variance explained

```{r win_varexp, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(2, 1)$h}
## Add HVW assays as separate SCE entries for getVarianceExplained
hvw_sce <- SingleCellExperiment(
  assays  = lapply(names(assay_map), function(lbl) {
    akey <- names(assay_map)[assay_map == lbl]
    mat  <- as.matrix(assay(sce, akey))
    mat[win_embeds[[lbl]]$hvw_idx, , drop = FALSE]
  }) |> setNames(assay_map),
  colData = colData(sce)
)

ve_win <- bind_rows(lapply(assay_map, function(lbl) {
  ve <- getVarianceExplained(hvw_sce, exprs_values = lbl,
                             variables = c("location", "patient"))
  data.frame(assay    = lbl,
             location = median(ve["location", ], na.rm = TRUE),
             patient  = median(ve["patient",  ], na.rm = TRUE))
}))
ve_win_long <- pivot_longer(ve_win, c("location", "patient"),
                            names_to = "variable", values_to = "median_ve")
ve_win_long$assay <- factor(ve_win_long$assay, levels = assay_map)

ggplot(ve_win_long, aes(x = assay, y = median_ve, fill = variable)) +
  geom_col(position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = c("location" = "#0072B2", "patient" = "#E69F00")) +
  labs(x = NULL, y = "median % variance explained",
       title = "Variance explained (per-window HVW)") +
  theme_ng()
```

## Silhouette scores

```{r win_sil}
sil_win <- data.frame(
  assay         = assay_map,
  silhouette    = sapply(names(win_embeds), function(lbl)
    sil_score(win_embeds[[lbl]]$umap, colData(sce)$location)),
  feature_space = "windows"
)
print(sil_win)
```

# Per-annotation cell embeddings (mean entropy as features)

Each cell is embedded using its mean entropy within each named genomic annotation
derived from SCE rowData. This compresses the window space into ~10 annotation-level
values per cell.

```{r annot_cols}
all_bin     <- grep("_bin$", colnames(rowData(sce)), value = TRUE)
de_flags    <- grep("_vs_|_up_bin|_down_bin", all_bin, value = TRUE)
annot_cols  <- setdiff(all_bin, de_flags)
cat("Annotation columns:\n", paste(annot_cols, collapse = "\n"), "\n")
```

```{r build_annot_matrix}
build_cell_x_annot <- function(assay_name, sce_obj, annot_columns) {
  mat <- as.matrix(assay(sce_obj, assay_name))
  rd  <- as.data.frame(rowData(sce_obj))
  res <- vapply(annot_columns, function(col) {
    idx <- which(rd[[col]] == 1)
    if (length(idx) == 0) return(rep(NA_real_, ncol(sce_obj)))
    colMeans(mat[idx, , drop = FALSE], na.rm = TRUE)
  }, numeric(ncol(sce_obj)))
  colnames(res) <- annot_columns
  res
}

annot_mats <- lapply(names(assay_map), function(akey) {
  m <- build_cell_x_annot(akey, sce, annot_cols)
  m <- m[, colSums(!is.na(m)) > 0, drop = FALSE]
  m
})
names(annot_mats) <- assay_map

cat("Cell x annotation matrix:", nrow(annot_mats[[1]]), "x", ncol(annot_mats[[1]]), "\n")
```

```{r annot_umap_compute}
## UMAP on cell x annotation matrices (small so no HVW needed; all annotations kept)
annot_umaps <- lapply(annot_mats, function(m) run_umap_scores(m))
```

## Four-panel comparison (location)

```{r annot_panel, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(4, 1)$h}
df_annot_panel <- bind_rows(lapply(names(annot_umaps), function(lbl) {
  data.frame(umap1 = annot_umaps[[lbl]][, 1],
             umap2 = annot_umaps[[lbl]][, 2],
             location = colData(sce)$location,
             assay = lbl)
}))
df_annot_panel$assay <- factor(df_annot_panel$assay, levels = assay_map)

ggplot(df_annot_panel, aes(x = umap1, y = umap2, color = location)) +
  geom_point(size = 0.4, alpha = 0.6) +
  scale_color_manual(values = loc_colors) +
  facet_wrap(~ assay, nrow = 1) +
  labs(x = "UMAP 1", y = "UMAP 2",
       title = "Per-annotation embeddings (mean entropy per annotation)") +
  guides(x = guide_x_nolap(),
         color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng()
```

## PCA

```{r annot_pca_plot, fig.width = ng_fig_size(4, 2)$w, fig.height = ng_fig_size(4, 2)$h}
annot_pcas <- lapply(annot_mats, function(m) run_pca_mat(t(m)))

df_annot_pca <- bind_rows(lapply(names(annot_pcas), function(lbl) {
  pca <- annot_pcas[[lbl]]
  if (is.null(pca)) return(NULL)
  data.frame(PC1 = pca[, 1], PC2 = pca[, 2],
             location = colData(sce)$location,
             patient  = colData(sce)$patient,
             assay    = lbl)
}))
df_annot_pca$assay <- factor(df_annot_pca$assay, levels = assay_map)

ggplot(df_annot_pca, aes(x = PC1, y = PC2, color = location)) +
  geom_point(size = 0.5, alpha = 0.6) +
  scale_color_manual(values = loc_colors) +
  facet_grid(assay ~ .) +
  labs(x = "PC 1", y = "PC 2", title = "PCA: per-annotation") +
  guides(x = guide_x_nolap(),
         color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng()
```

## Variance explained

```{r annot_varexp, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(2, 1)$h}
annot_sce <- SingleCellExperiment(
  assays  = lapply(annot_mats, t) |> setNames(assay_map),
  colData = colData(sce)
)

ve_annot <- bind_rows(lapply(assay_map, function(lbl) {
  ve <- getVarianceExplained(annot_sce, exprs_values = lbl,
                             variables = c("location", "patient"))
  data.frame(assay    = lbl,
             location = median(ve["location", ], na.rm = TRUE),
             patient  = median(ve["patient",  ], na.rm = TRUE))
}))
ve_annot_long <- pivot_longer(ve_annot, c("location", "patient"),
                              names_to = "variable", values_to = "median_ve")
ve_annot_long$assay <- factor(ve_annot_long$assay, levels = assay_map)

ggplot(ve_annot_long, aes(x = assay, y = median_ve, fill = variable)) +
  geom_col(position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = c("location" = "#0072B2", "patient" = "#E69F00")) +
  labs(x = NULL, y = "median % variance explained",
       title = "Variance explained (per-annotation)") +
  theme_ng()
```

## Annotation heatmap (mean adjS per location)

```{r annot_heatmap, fig.width = ng_fig_size(ncol(annot_mats[["adjS"]]), 1, panel_mm = 30)$w, fig.height = 4}
loc_means <- do.call(rbind, lapply(loc_levels, function(l) {
  idx <- which(colData(sce)$location == l)
  if (length(idx) == 0) return(NULL)
  colMeans(annot_mats[["adjS"]][idx, , drop = FALSE], na.rm = TRUE)
}))
present <- sapply(loc_levels, function(l) any(colData(sce)$location == l))
loc_means <- loc_means[present, , drop = FALSE]
rownames(loc_means) <- loc_levels[present]
colnames(loc_means) <- gsub("_bin$", "", colnames(loc_means))

col_fun <- colorRamp2(
  quantile(loc_means, c(0.05, 0.5, 0.95), na.rm = TRUE),
  c("navy", "white", "firebrick")
)
ha <- rowAnnotation(
  location = rownames(loc_means),
  col = list(location = loc_colors[rownames(loc_means)])
)
Heatmap(loc_means, name = "mean adjS", col = col_fun,
        right_annotation = ha, show_row_names = TRUE, show_column_names = TRUE,
        cluster_rows = TRUE, cluster_columns = TRUE,
        row_names_gp = gpar(fontsize = 8), column_names_gp = gpar(fontsize = 7),
        column_title = "Mean adjS per annotation per location")
```

## Silhouette scores

```{r annot_sil}
sil_annot <- data.frame(
  assay         = assay_map,
  silhouette    = sapply(names(annot_umaps), function(lbl)
    sil_score(annot_umaps[[lbl]], colData(sce)$location)),
  feature_space = "annotations"
)
print(sil_annot)
```

# Per-annotation HVW -> PCA -> UMAP

For each annotation separately: take the subset of windows in that annotation,
select the top `r N_HVW` highly variable windows, run PCA, then UMAP.
This is the full single-cell pipeline per genomic context.

```{r per_annot_loop, results = "asis", fig.width = ng_fig_size(2, 1)$w, fig.height = ng_fig_size(2, 1)$h}
per_annot_sil <- list()

for (col in annot_cols) {
  idx   <- which(rowData(sce)[[col]] == 1)
  label <- gsub("_bin$", "", col)

  if (length(idx) < 50) {
    per_annot_sil[[label]] <- data.frame(annotation = label)
    next
  }

  cat(sprintf("\n## %s {.tabset .tabset-pills}\n\n", label))
  cat(sprintf("Windows in annotation: %d. HVW kept: %d.\n\n",
              length(idx), min(N_HVW, length(idx))))

  sil_row <- data.frame(annotation = label)

  for (akey in names(assay_map)) {
    alabel  <- assay_map[[akey]]
    sub_mat <- as.matrix(assay(sce, akey))[idx, , drop = FALSE]
    res     <- hvw_pca_umap(sub_mat)
    if (is.null(res)) next

    sil_row[[paste0("sil_", alabel)]] <-
      sil_score(res$umap, colData(sce)$location)

    cat(sprintf("\n### %s\n\n", alabel))
    print(plot_pair(res$umap, sce,
                    sprintf("%s / %s (HVW -> PCA -> UMAP)", label, alabel)))
    cat("\n")
  }

  per_annot_sil[[label]] <- sil_row
}
```

## Silhouette summary {-}

```{r per_annot_sil_summary, fig.width = ng_fig_size(length(annot_cols), 1, panel_mm = 35)$w, fig.height = ng_fig_size(length(annot_cols), 1, panel_mm = 35)$h}
sil_pa_df <- bind_rows(per_annot_sil)

sil_pa_long <- pivot_longer(sil_pa_df, starts_with("sil_"),
                            names_to = "metric", values_to = "silhouette")
sil_pa_long$metric <- sub("^sil_", "", sil_pa_long$metric)
sil_pa_long$metric <- factor(sil_pa_long$metric, levels = assay_map)

ggplot(sil_pa_long, aes(x = annotation, y = silhouette, fill = metric)) +
  geom_col(position = position_dodge(width = 0.8)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_fill_viridis_d(option = "turbo", begin = 0.1, end = 0.9) +
  labs(x = "annotation", y = "mean silhouette (location clusters)",
       title = "Location separation per annotation (HVW -> PCA -> UMAP)") +
  theme_ng()
```

# Summary: all feature spaces and metrics

```{r combined_sil, fig.width = ng_fig_size(4, 3)$w, fig.height = ng_fig_size(4, 1)$h}
sil_all <- bind_rows(sil_win, sil_annot)
sil_all$assay         <- factor(sil_all$assay, levels = assay_map)
sil_all$feature_space <- factor(sil_all$feature_space,
                                levels = c("windows", "annotations"))

ggplot(sil_all, aes(x = assay, y = silhouette, fill = feature_space)) +
  geom_col(position = position_dodge(width = 0.8)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c("windows" = "#0072B2", "annotations" = "#D55E00")) +
  labs(x = NULL, y = "mean silhouette (location clusters)",
       title = "Location separation: windows vs annotation-level features") +
  theme_ng()
```

# Per-cell entropy summaries

```{r cell_summary_df}
cell_df <- data.frame(
  location  = colData(sce)$location,
  patient   = colData(sce)$patient,
  mean_S    = colMeans(assay(sce, "sampen"),           na.rm = TRUE),
  mean_adjS = colMeans(assay(sce, "sampen_corrected"), na.rm = TRUE),
  mean_meth = colMeans(assay(sce, "meth"),             na.rm = TRUE)
)
```

```{r violins, fig.width = ng_fig_size(6, 1)$w, fig.height = ng_fig_size(6, 3)$h}
cell_long <- pivot_longer(cell_df, c("mean_S", "mean_adjS", "mean_meth"),
                          names_to = "metric", values_to = "value")
cell_long$metric <- factor(cell_long$metric,
                           levels = c("mean_S", "mean_adjS", "mean_meth"))

ggplot(cell_long, aes(x = location, y = value, fill = location)) +
  geom_violin(trim = FALSE, scale = "width", alpha = 0.7) +
  geom_boxplot(width = 0.1, outlier.size = 0.3, alpha = 0.6) +
  scale_fill_manual(values = loc_colors) +
  facet_grid(metric ~ patient, scales = "free_y") +
  labs(x = "biopsy location", y = "per-cell mean") +
  theme_ng() +
  theme(legend.position = "none", axis.text.x = element_text(size = 6))
```

```{r scatter_S_meth, fig.width = ng_fig_size(6, 1)$w, fig.height = ng_fig_size(6, 1)$h}
ggplot(cell_df, aes(x = mean_meth, y = mean_S, color = location)) +
  geom_point(size = 0.5, alpha = 0.6) +
  geom_smooth(aes(group = location), method = "lm", se = FALSE, linewidth = 0.5) +
  scale_color_manual(values = loc_colors) +
  facet_wrap(~ patient, nrow = 1) +
  labs(x = "cell mean methylation", y = "cell mean S") +
  guides(x = guide_x_nolap(),
         color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng()
```

```{r scatter_adjS_meth, fig.width = ng_fig_size(6, 1)$w, fig.height = ng_fig_size(6, 1)$h}
ggplot(cell_df, aes(x = mean_meth, y = mean_adjS, color = location)) +
  geom_point(size = 0.5, alpha = 0.6) +
  geom_smooth(aes(group = location), method = "lm", se = FALSE, linewidth = 0.5) +
  scale_color_manual(values = loc_colors) +
  facet_wrap(~ patient, nrow = 1) +
  labs(x = "cell mean methylation", y = "cell mean adjS") +
  guides(x = guide_x_nolap(),
         color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng()
```

# Session info

```{r session}
sessionInfo()
```
