---
title: "Ecker 2021 mouse brain: methylation entropy by feature (CpG context)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Izaskun Mallona"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
---

```{r setup}
suppressPackageStartupMessages({
  library(readr)
  library(ggplot2)
  library(dplyr)
  library(BiocParallel)
  library(ggstance)
  library(ggrepel)
})

source("src/ggtheme.R")
param <- MulticoreParam(workers = snakemake@threads)
```

```{r opts}
opts_chunk$set(
  fig.width  = 7.5,
  fig.height = 7.5,
  cache      = FALSE,
  include    = TRUE,
  fig.path   = "ecker_plots/",
  dev        = c("png", "svg"),
  dpi        = 500,
  cache.lazy = FALSE,
  warning    = TRUE,
  message    = TRUE
)
```

```{r save_snmk_debug, include = FALSE}
saveRDS(file = "snmk_ecker_features.rds", object = snakemake)
```

```{r logging, include = FALSE}
log <- file(snakemake@log$log, open = "wt")
sink(log)
sink(log, type = "message")
```

Mouse (mm10), CpG context only. chr10 only: `r snakemake@params[["chr10_only"]]`.

```{r helpers}
## filenames: {annotation}_{cell_type}.*.gz
## cell_type is always the last underscore-delimited token
get_annotation <- function(fn) sub("(.*)_[^_]+\\..*", "\\1", fn)
get_cell_type  <- function(fn) sub(".*_([^_]+)\\..*", "\\1", fn)

ann_labels <- c(
  "genes"    = "Genes",
  "promoters"= "Promoters",
  "lines"    = "LINEs",
  "sines"    = "SINEs",
  "h3k4me3"  = "H3K4me3",
  "h3k9me3"  = "H3K9me3",
  "h3k27me3" = "H3K27me3",
  "h3k4me1"  = "H3K4me1",
  "h3k27ac"  = "H3K27ac"
)

cell_type_order <- c("Exc", "Inh", "ASC", "OPC", "ODC", "MGC")

harmonize_levels <- function(df) {
  missing_ann <- setdiff(unique(df$annotation), names(ann_labels))
  if (length(missing_ann) > 0)
    warning("unknown annotations: ", paste(missing_ann, collapse = ", "))

  df$annotation <- factor(df$annotation,
                          levels = names(ann_labels),
                          labels = unname(ann_labels))
  df$cell_type  <- factor(df$cell_type,
                          levels = intersect(cell_type_order, unique(df$cell_type)))
  df
}
```

# Unnormalized sampEn {.tabset .tabset-pills}

Per-cell sample entropy (S) from `.out.gz` files.

```{r import_sampen}
short_paths <- list.files(snakemake@params[["output_path"]],
                          pattern = "\\.out\\.gz$", full.names = TRUE)
short_paths <- short_paths[!grepl("\\.det\\.out\\.gz$",  short_paths)]
short_paths <- short_paths[!grepl("\\.meth\\.out\\.gz$", short_paths)]

read_short <- function(fp) {
  fn <- basename(fp)
  dt <- read_tsv(fp, na = "-1", show_col_types = FALSE)
  dt <- dt[, c("sampen", "avg_meth")]
  dt$annotation <- get_annotation(fn)
  dt$cell_type  <- get_cell_type(fn)
  dt
}

sampens <- do.call(rbind, bplapply(short_paths, read_short, BPPARAM = param))
sampens <- harmonize_levels(sampens)
head(sampens)
```

## Scatter by annotation

```{r scatter_sampen}
ggplot(sampens, aes(x = avg_meth, y = sampen, color = cell_type)) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~annotation) +
  labs(x = "average methylation", y = "sample entropy (S)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Boxes by annotation

```{r boxes_sampen, fig.width = 10, fig.height = 6}
ggplot(sampens, aes(x = annotation, y = sampen, color = cell_type)) +
  geom_boxplot(outlier.size = 0.6, outlier.alpha = 0.3) +
  geom_jitter(position = position_dodge(width = 0.75), size = 0.5, alpha = 0.3) +
  scale_color_brewer(palette = "Dark2") +
  labs(y = "sample entropy (S)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## Methylation boxes by annotation

```{r boxes_meth_sampen, fig.width = 10, fig.height = 6}
ggplot(sampens, aes(x = annotation, y = avg_meth, color = cell_type)) +
  geom_boxplot(outlier.size = 0.6, outlier.alpha = 0.3) +
  geom_jitter(position = position_dodge(width = 0.75), size = 0.5, alpha = 0.3) +
  scale_color_brewer(palette = "Dark2") +
  labs(y = "average methylation") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## Range plots by compartment

```{r ranges_sampen, fig.width = 10, fig.height = 10}
ranges_s <- sampens %>%
  group_by(compartment = annotation, cell_type) %>%
  summarise(
    sampen_mean = mean(sampen,    na.rm = TRUE),
    sampen_min  = min(sampen,     na.rm = TRUE),
    sampen_max  = max(sampen,     na.rm = TRUE),
    meth_mean   = mean(avg_meth,  na.rm = TRUE),
    meth_min    = min(avg_meth,   na.rm = TRUE),
    meth_max    = max(avg_meth,   na.rm = TRUE),
    .groups = "drop"
  )

ggplot(ranges_s, aes(color = cell_type)) +
  geom_segment(aes(x = meth_min,  xend = meth_max,
                   y = sampen_mean, yend = sampen_mean),
               linewidth = 0.6) +
  geom_segment(aes(x = meth_mean, xend = meth_mean,
                   y = sampen_min, yend = sampen_max),
               linewidth = 0.6) +
  geom_point(aes(x = meth_mean, y = sampen_mean), shape = 3, size = 1.2) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "DNA methylation", y = "sample entropy (S)") +
  facet_wrap(~compartment) +
  theme_ng()
```

# adjS (meth-normalised sampEn) {.tabset .tabset-pills}

Per-cell normalised sample entropy from `.out.gz` files.

```{r import_adjS}
norm_paths <- list.files(snakemake@params[["output_path"]],
                         pattern = "out\\.gz$", full.names = TRUE)
norm_paths <- norm_paths[!grepl("\\.det\\.out\\.gz$",  norm_paths)]
norm_paths <- norm_paths[!grepl("\\.norm\\.out\\.gz$", norm_paths)]
norm_paths <- norm_paths[!grepl("\\.meth\\.out\\.gz$", norm_paths)]

read_adjS <- function(fp) {
  fn <- basename(fp)
  dt <- read_tsv(fp, na = "-1", show_col_types = FALSE)
  dt <- dt[, c("sampen_norm", "avg_meth")]
  colnames(dt) <- c("sampen", "avg_meth")
  dt$annotation <- get_annotation(fn)
  dt$cell_type  <- get_cell_type(fn)
  dt
}

adjsampens <- do.call(rbind, bplapply(norm_paths, read_adjS, BPPARAM = param))
adjsampens <- harmonize_levels(adjsampens)
head(adjsampens)
```

## Scatter by annotation

```{r scatter_adjS}
ggplot(adjsampens, aes(x = avg_meth, y = sampen, color = cell_type)) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~annotation) +
  labs(x = "average methylation", y = "adj. sample entropy (adjS)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Boxes by annotation

```{r boxes_adjS, fig.width = 10, fig.height = 6}
ggplot(adjsampens, aes(x = annotation, y = sampen, color = cell_type)) +
  geom_boxplot(outlier.size = 0.6, outlier.alpha = 0.3) +
  geom_jitter(position = position_dodge(width = 0.75), size = 0.5, alpha = 0.3) +
  scale_color_brewer(palette = "Dark2") +
  labs(y = "adj. sample entropy (adjS)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## Range plots by compartment

```{r ranges_adjS, fig.width = 10, fig.height = 10}
ranges_adjs <- adjsampens %>%
  group_by(compartment = annotation, cell_type) %>%
  summarise(
    sampen_mean = mean(sampen,   na.rm = TRUE),
    sampen_min  = min(sampen,    na.rm = TRUE),
    sampen_max  = max(sampen,    na.rm = TRUE),
    meth_mean   = mean(avg_meth, na.rm = TRUE),
    meth_min    = min(avg_meth,  na.rm = TRUE),
    meth_max    = max(avg_meth,  na.rm = TRUE),
    .groups = "drop"
  )

ggplot(ranges_adjs, aes(color = cell_type)) +
  geom_segment(aes(x = meth_min,  xend = meth_max,
                   y = sampen_mean, yend = sampen_mean),
               linewidth = 0.6) +
  geom_segment(aes(x = meth_mean, xend = meth_mean,
                   y = sampen_min, yend = sampen_max),
               linewidth = 0.6) +
  geom_point(aes(x = meth_mean, y = sampen_mean), shape = 3, size = 1.2) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "DNA methylation", y = "adj. sample entropy (adjS)") +
  facet_wrap(~compartment) +
  theme_ng()
```

# Shannon H - medians, downsampled {.tabset .tabset-pills}

Per-locus Shannon entropy (H) from `.det.out.gz`. One median value per cell type per annotation; loci downsampled to 10e4 rows.

```{r import_shannon}
det_paths <- list.files(snakemake@params[["output_path"]],
                        pattern = "det\\.out\\.gz$", full.names = TRUE)
det_paths <- det_paths[!grepl("norm", det_paths)]

read_shannon <- function(fp) {
  fn  <- basename(fp)
  tmp <- read_tsv(fp, na = "-1", show_col_types = FALSE, n_max = 10e4)
  tibble(
    file            = fn,
    median_shannon  = median(tmp$shannon,  na.rm = TRUE),
    median_avg_meth = median(tmp$avg_meth, na.rm = TRUE),
    annotation      = get_annotation(fn),
    cell_type       = get_cell_type(fn)
  )
}

shannons <- bplapply(det_paths, read_shannon, BPPARAM = param) |> bind_rows()
shannons <- harmonize_levels(shannons)
head(shannons)
```

## Scatter by annotation

```{r scatter_shannon}
ggplot(shannons, aes(x = median_avg_meth, y = median_shannon, color = cell_type)) +
  geom_point(size = 1.5, alpha = 0.8) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~annotation) +
  labs(x = "median avg. methylation", y = "median Shannon entropy (H)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Boxes by annotation

```{r boxes_shannon, fig.width = 10, fig.height = 6}
ggplot(shannons, aes(x = annotation, y = median_shannon, color = cell_type)) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  scale_color_brewer(palette = "Dark2") +
  labs(y = "median Shannon entropy (H)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

# adjH (meth-normalised Shannon) - medians {.tabset .tabset-pills}

Per-locus normalised Shannon entropy from `.norm.det.out.gz`. Same downsampling as above.

```{r import_adjH}
norm_det_paths <- list.files(snakemake@params[["output_path"]],
                             pattern = "norm\\.det\\.out\\.gz$", full.names = TRUE)

read_adjH <- function(fp) {
  fn  <- basename(fp)
  tmp <- read_tsv(fp, na = "-1", show_col_types = FALSE, n_max = 10e4)
  tibble(
    file            = fn,
    median_shannon  = median(tmp$shannon_norm, na.rm = TRUE),
    median_avg_meth = median(tmp$avg_meth,     na.rm = TRUE),
    annotation      = get_annotation(fn),
    cell_type       = get_cell_type(fn)
  )
}

adjshannons <- bplapply(norm_det_paths, read_adjH, BPPARAM = param) |> bind_rows()
adjshannons <- harmonize_levels(adjshannons)
head(adjshannons)
```

## Scatter by annotation

```{r scatter_adjH}
ggplot(adjshannons, aes(x = median_avg_meth, y = median_shannon, color = cell_type)) +
  geom_point(size = 1.5, alpha = 0.8) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~annotation) +
  labs(x = "median avg. methylation", y = "median adj. Shannon entropy (adjH)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

## Boxes by annotation

```{r boxes_adjH, fig.width = 10, fig.height = 6}
ggplot(adjshannons, aes(x = annotation, y = median_shannon, color = cell_type)) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  scale_color_brewer(palette = "Dark2") +
  labs(y = "median adj. Shannon entropy (adjH)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

# Assembly

sampEn/adjS are per-cell values; Shannon/adjH are medians across loci for each cell type group.

```{r str_check}
str(sampens)
str(adjsampens)
str(shannons)
str(adjshannons)
```

```{r save_rds}
saveRDS(
  list(sampens    = sampens,
       adjsampens = adjsampens,
       shannons   = shannons,
       adjshannons= adjshannons),
  file = "ecker_entropy.rds"
)
```

## adjS vs adjH by cell type

```{r adjS_vs_adjH, fig.height = 10, fig.width = 8}
adj_df <- adjsampens %>%
  left_join(adjshannons,
            by      = c("annotation", "cell_type"),
            suffix  = c("_sampen", "_shannon"))

ggplot(adj_df, aes(x = median_shannon, y = sampen, color = cell_type)) +
  geom_point(size = 0.9) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~annotation, ncol = 3) +
  labs(x = "median adjH", y = "adjS") +
  theme_ng()
```

## S vs H by cell type

```{r S_vs_H, fig.height = 10, fig.width = 8}
raw_df <- sampens %>%
  left_join(shannons,
            by     = c("annotation", "cell_type"),
            suffix = c("_sampen", "_shannon"))

ggplot(raw_df, aes(x = median_shannon, y = sampen, color = cell_type)) +
  geom_point(size = 0.9) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~annotation, ncol = 3) +
  labs(x = "median H", y = "S") +
  theme_ng()
```

# Variation across cell types {.tabset .tabset-pills}

```{r variation_range}
variation <- adj_df %>%
  group_by(annotation) %>%
  summarise(
    shannon_range = max(median_shannon, na.rm = TRUE) - min(median_shannon, na.rm = TRUE),
    sampen_range  = max(sampen,         na.rm = TRUE) - min(sampen,         na.rm = TRUE),
    shannon_sd    = sd(median_shannon,  na.rm = TRUE),
    sampen_sd     = sd(sampen,          na.rm = TRUE),
    .groups = "drop"
  )

head(variation)

ggplot(variation, aes(x = shannon_range, y = sampen_range, label = annotation)) +
  geom_point() +
  geom_text_repel() +
  labs(x = "adjH range across cell types",
       y = "adjS range across cell types") +
  theme_ng()
```

## Per cell type

```{r variation_per_cell_type}
variation_ct <- adj_df %>%
  group_by(annotation, cell_type) %>%
  summarise(
    shannon_median = median(median_shannon, na.rm = TRUE),
    sampen_median  = median(sampen,         na.rm = TRUE),
    shannon_sd     = sd(median_shannon,     na.rm = TRUE),
    sampen_sd      = sd(sampen,             na.rm = TRUE),
    shannon_range  = max(median_shannon, na.rm = TRUE) - min(median_shannon, na.rm = TRUE),
    sampen_range   = max(sampen,         na.rm = TRUE) - min(sampen,         na.rm = TRUE),
    .groups = "drop"
  )

ggplot(variation_ct, aes(x = shannon_median, y = sampen_median, color = cell_type)) +
  geom_point() +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~annotation) +
  labs(x = "median adjH", y = "median adjS") +
  theme_ng()

ggplot(variation_ct, aes(x = shannon_sd, y = sampen_sd, color = cell_type)) +
  geom_point() +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~annotation) +
  labs(x = "sd adjH", y = "sd adjS") +
  theme_ng()
```

# Driver categorization

Categorizing annotations as within-cell (adjS-driven), across-cell (adjH-driven), or both. Ongoing, not final.

```{r drivers_range}
shannon_thr <- quantile(variation$shannon_range, 0.5, na.rm = TRUE)
sampen_thr  <- quantile(variation$sampen_range,  0.5, na.rm = TRUE)

variation <- variation %>%
  mutate(
    driver = case_when(
      shannon_range < shannon_thr & sampen_range < sampen_thr ~ "not changing enough",
      shannon_range >= sampen_range * 1.5 ~ "across-cell driven",
      sampen_range  >= shannon_range * 1.5 ~ "within-cell driven",
      TRUE ~ "both"
    )
  )

table(variation$driver)

ggplot(variation,
       aes(x = shannon_range, y = sampen_range,
           color = driver, shape = driver, label = annotation)) +
  geom_point(size = 3) +
  geom_text_repel() +
  labs(x = "adjH range across cell types",
       y = "adjS range across cell types") +
  theme_ng()
```

```{r drivers_sd}
sds <- raw_df %>%
  group_by(annotation) %>%
  summarise(
    shannon_sd = sd(median_shannon, na.rm = TRUE),
    sampen_sd  = sd(sampen,         na.rm = TRUE),
    .groups = "drop"
  )

shannon_sd_thr <- quantile(sds$shannon_sd, 0.3, na.rm = TRUE)
sampen_sd_thr  <- quantile(sds$sampen_sd,  0.3, na.rm = TRUE)

sds <- sds %>%
  mutate(
    driver = case_when(
      shannon_sd < shannon_sd_thr & sampen_sd < sampen_sd_thr ~ "not changing enough",
      shannon_sd >= sampen_sd * 1.5 ~ "across-cell driven",
      sampen_sd  >= shannon_sd * 1.5 ~ "within-cell driven",
      TRUE ~ "both"
    )
  )

table(sds$driver)

ggplot(sds,
       aes(x = shannon_sd, y = sampen_sd,
           color = driver, shape = driver, label = annotation)) +
  geom_point(size = 3) +
  geom_text_repel() +
  labs(x = "SD of adjH across cell types",
       y = "SD of adjS across cell types") +
  theme_ng()
```

# Session

```{r session, cache = FALSE}
sessionInfo()
```
