---
title: "Ecker 2021 mouse brain: methylation entropy by feature (CpG context)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Izaskun Mallona"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
---

```{r setup}
suppressPackageStartupMessages({
  library(readr)
  library(ggplot2)
  library(dplyr)
  library(BiocParallel)
  library(ggstance)
  library(ggrepel)
})

source("src/ggtheme.R")
param <- MulticoreParam(workers = snakemake@threads)
```

```{r opts}
opts_chunk$set(
  echo       = TRUE,
  fig.width  = 7.5,
  fig.height = 7.5,
  cache      = FALSE,
  include    = TRUE,
  fig.path   = "ecker_plots/",
  dev        = c("png", "svg"),
  dpi        = 500,
  cache.lazy = FALSE,
  warning    = TRUE,
  message    = TRUE
)
```

```{r save_snmk_debug, include = FALSE}
saveRDS(file = "snmk_ecker_features.rds", object = snakemake)
```

```{r logging, include = FALSE}
log <- file(snakemake@log$log, open = "wt")
sink(log)
sink(log, type = "message")
```

Mouse (mm10), CpG context only. chr10 only: `r snakemake@params[["chr10_only"]]`.

```{r helpers}
## filenames: {annotation}_{major_region}_{cell_class}.*.gz
## annotation has no underscores; neither do major_region or cell_class values
get_annotation   <- function(fn) sub("^(.*)_[^_]+_[^_]+\\..*$", "\\1", fn)
get_major_region <- function(fn) sub("^.*_([^_]+)_[^_]+\\..*$", "\\1", fn)
get_cell_class   <- function(fn) sub("^.*_([^_]+)\\..*$",        "\\1", fn)

ann_labels <- c(
  "genes"    = "Genes",
  "promoters"= "Promoters",
  "lines"    = "LINEs",
  "sines"    = "SINEs",
  "h3k4me3"  = "H3K4me3",
  "h3k9me3"  = "H3K9me3",
  "h3k27me3" = "H3K27me3",
  "h3k4me1"  = "H3K4me1",
  "h3k27ac"  = "H3K27ac"
)

harmonize_levels <- function(df) {
  missing_ann <- setdiff(unique(df$annotation), names(ann_labels))
  if (length(missing_ann) > 0)
    warning("unknown annotations: ", paste(missing_ann, collapse = ", "))
  df$annotation   <- factor(df$annotation,
                             levels = names(ann_labels),
                             labels = unname(ann_labels))
  df$major_region <- factor(df$major_region)
  df$cell_class   <- factor(df$cell_class)
  df
}
```

# Unnormalized sampEn (S) {.tabset .tabset-pills}

Per-cell sample entropy (S) from `.out.gz` files.

```{r import_sampen}
short_paths <- list.files(snakemake@params[["output_path"]],
                           pattern = "\\.out\\.gz$", full.names = TRUE)
short_paths <- short_paths[!grepl("\\.det\\.out\\.gz$",  short_paths)]
short_paths <- short_paths[!grepl("\\.meth\\.out\\.gz$", short_paths)]

read_short <- function(fp) {
  fn <- basename(fp)
  dt <- read_tsv(fp, na = "-1", show_col_types = FALSE)
  dt <- dt[, c("sampen", "avg_meth")]
  dt$annotation   <- get_annotation(fn)
  dt$major_region <- get_major_region(fn)
  dt$cell_class   <- get_cell_class(fn)
  dt
}

sampens <- do.call(rbind, bplapply(short_paths, read_short, BPPARAM = param))
sampens <- harmonize_levels(sampens)
head(sampens)
```

## By cell class

```{r scatter_sampen_class}
ggplot(sampens, aes(x = avg_meth, y = sampen, color = cell_class)) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~annotation) +
  labs(x = "average methylation", y = "sample entropy (S)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

```{r boxes_sampen_class, fig.width = 10, fig.height = 6}
ggplot(sampens, aes(x = annotation, y = sampen, color = cell_class)) +
  geom_boxplot(outlier.size = 0.6, outlier.alpha = 0.3) +
  scale_color_brewer(palette = "Dark2") +
  labs(y = "sample entropy (S)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## By major region

```{r scatter_sampen_region}
ggplot(sampens, aes(x = avg_meth, y = sampen, color = major_region)) +
  geom_point(size = 0.9, alpha = 0.2) +
  facet_wrap(~annotation) +
  labs(x = "average methylation", y = "sample entropy (S)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

```{r boxes_sampen_region, fig.width = 10, fig.height = 6}
ggplot(sampens, aes(x = annotation, y = sampen, color = major_region)) +
  geom_boxplot(outlier.size = 0.6, outlier.alpha = 0.3) +
  labs(y = "sample entropy (S)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## Region by class

```{r boxes_sampen_combo, fig.width = 12, fig.height = 8}
ggplot(sampens, aes(x = cell_class, y = sampen, color = major_region)) +
  geom_boxplot(outlier.size = 0.6, outlier.alpha = 0.3) +
  facet_wrap(~annotation) +
  labs(y = "sample entropy (S)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## Range plots

```{r ranges_sampen, fig.width = 10, fig.height = 10}
ranges_s <- sampens %>%
  group_by(compartment = annotation, major_region, cell_class) %>%
  summarise(
    sampen_mean = mean(sampen,   na.rm = TRUE),
    sampen_min  = min(sampen,    na.rm = TRUE),
    sampen_max  = max(sampen,    na.rm = TRUE),
    meth_mean   = mean(avg_meth, na.rm = TRUE),
    meth_min    = min(avg_meth,  na.rm = TRUE),
    meth_max    = max(avg_meth,  na.rm = TRUE),
    .groups = "drop"
  )

ggplot(ranges_s, aes(color = cell_class, shape = major_region)) +
  geom_segment(aes(x = meth_min,  xend = meth_max,
                   y = sampen_mean, yend = sampen_mean),
               linewidth = 0.6) +
  geom_segment(aes(x = meth_mean, xend = meth_mean,
                   y = sampen_min, yend = sampen_max),
               linewidth = 0.6) +
  geom_point(aes(x = meth_mean, y = sampen_mean), size = 1.2) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "DNA methylation", y = "sample entropy (S)") +
  facet_wrap(~compartment) +
  theme_ng()
```

# adjS (methylation-normalised sample entropy) {.tabset .tabset-pills}

Per-cell normalised sample entropy from `.out.gz` files.

```{r import_adjS}
norm_paths <- list.files(snakemake@params[["output_path"]],
                          pattern = "out\\.gz$", full.names = TRUE)
norm_paths <- norm_paths[!grepl("\\.det\\.out\\.gz$",  norm_paths)]
norm_paths <- norm_paths[!grepl("\\.norm\\.out\\.gz$", norm_paths)]
norm_paths <- norm_paths[!grepl("\\.meth\\.out\\.gz$", norm_paths)]

read_adjS <- function(fp) {
  fn <- basename(fp)
  dt <- read_tsv(fp, na = "-1", show_col_types = FALSE)
  dt <- dt[, c("sampen_norm", "avg_meth")]
  colnames(dt) <- c("sampen", "avg_meth")
  dt$annotation   <- get_annotation(fn)
  dt$major_region <- get_major_region(fn)
  dt$cell_class   <- get_cell_class(fn)
  dt
}

adjsampens <- do.call(rbind, bplapply(norm_paths, read_adjS, BPPARAM = param))
adjsampens <- harmonize_levels(adjsampens)
head(adjsampens)
```

## By cell class

```{r scatter_adjS_class}
ggplot(adjsampens, aes(x = avg_meth, y = sampen, color = cell_class)) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~annotation) +
  labs(x = "average methylation", y = "adj. sample entropy (adjS)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

```{r boxes_adjS_class, fig.width = 10, fig.height = 6}
ggplot(adjsampens, aes(x = annotation, y = sampen, color = cell_class)) +
  geom_boxplot(outlier.size = 0.6, outlier.alpha = 0.3) +
  scale_color_brewer(palette = "Dark2") +
  labs(y = "adj. sample entropy (adjS)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## By major region

```{r scatter_adjS_region}
ggplot(adjsampens, aes(x = avg_meth, y = sampen, color = major_region)) +
  geom_point(size = 0.9, alpha = 0.2) +
  facet_wrap(~annotation) +
  labs(x = "average methylation", y = "adj. sample entropy (adjS)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

```{r boxes_adjS_region, fig.width = 10, fig.height = 6}
ggplot(adjsampens, aes(x = annotation, y = sampen, color = major_region)) +
  geom_boxplot(outlier.size = 0.6, outlier.alpha = 0.3) +
  labs(y = "adj. sample entropy (adjS)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## Region by class

```{r boxes_adjS_combo, fig.width = 12, fig.height = 8}
ggplot(adjsampens, aes(x = cell_class, y = sampen, color = major_region)) +
  geom_boxplot(outlier.size = 0.6, outlier.alpha = 0.3) +
  facet_wrap(~annotation) +
  labs(y = "adj. sample entropy (adjS)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## Range plots

```{r ranges_adjS, fig.width = 10, fig.height = 10}
ranges_adjs <- adjsampens %>%
  group_by(compartment = annotation, major_region, cell_class) %>%
  summarise(
    sampen_mean = mean(sampen,   na.rm = TRUE),
    sampen_min  = min(sampen,    na.rm = TRUE),
    sampen_max  = max(sampen,    na.rm = TRUE),
    meth_mean   = mean(avg_meth, na.rm = TRUE),
    meth_min    = min(avg_meth,  na.rm = TRUE),
    meth_max    = max(avg_meth,  na.rm = TRUE),
    .groups = "drop"
  )

ggplot(ranges_adjs, aes(color = cell_class, shape = major_region)) +
  geom_segment(aes(x = meth_min,  xend = meth_max,
                   y = sampen_mean, yend = sampen_mean),
               linewidth = 0.6) +
  geom_segment(aes(x = meth_mean, xend = meth_mean,
                   y = sampen_min, yend = sampen_max),
               linewidth = 0.6) +
  geom_point(aes(x = meth_mean, y = sampen_mean), size = 1.2) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "DNA methylation", y = "adj. sample entropy (adjS)") +
  facet_wrap(~compartment) +
  theme_ng()
```

# Shannon H - medians {.tabset .tabset-pills}

Per-locus Shannon entropy (H) from `.det.out.gz`. One median per group per annotation; loci downsampled to 10e4 rows.

```{r import_shannon}
det_paths <- list.files(snakemake@params[["output_path"]],
                         pattern = "det\\.out\\.gz$", full.names = TRUE)
det_paths <- det_paths[!grepl("norm", det_paths)]

read_shannon <- function(fp) {
  fn  <- basename(fp)
  tmp <- read_tsv(fp, na = "-1", show_col_types = FALSE, n_max = 10e4)
  tibble(
    file            = fn,
    median_shannon  = median(tmp$shannon,  na.rm = TRUE),
    median_avg_meth = median(tmp$avg_meth, na.rm = TRUE),
    annotation      = get_annotation(fn),
    major_region    = get_major_region(fn),
    cell_class      = get_cell_class(fn)
  )
}

shannons <- bplapply(det_paths, read_shannon, BPPARAM = param) |> bind_rows()
shannons <- harmonize_levels(shannons)
head(shannons)
```

## By cell class

```{r scatter_shannon_class}
ggplot(shannons, aes(x = median_avg_meth, y = median_shannon, color = cell_class)) +
  geom_point(size = 1.5, alpha = 0.8) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~annotation) +
  labs(x = "median avg. methylation", y = "median Shannon entropy (H)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

```{r boxes_shannon_class, fig.width = 10, fig.height = 6}
ggplot(shannons, aes(x = annotation, y = median_shannon, color = cell_class)) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  scale_color_brewer(palette = "Dark2") +
  labs(y = "median Shannon entropy (H)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## By major region

```{r scatter_shannon_region}
ggplot(shannons, aes(x = median_avg_meth, y = median_shannon, color = major_region)) +
  geom_point(size = 1.5, alpha = 0.8) +
  facet_wrap(~annotation) +
  labs(x = "median avg. methylation", y = "median Shannon entropy (H)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

```{r boxes_shannon_region, fig.width = 10, fig.height = 6}
ggplot(shannons, aes(x = annotation, y = median_shannon, color = major_region)) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  labs(y = "median Shannon entropy (H)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## Region by class

```{r boxes_shannon_combo, fig.width = 12, fig.height = 8}
ggplot(shannons, aes(x = cell_class, y = median_shannon, color = major_region)) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  facet_wrap(~annotation) +
  labs(y = "median Shannon entropy (H)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

# adjH (methylation-normalised Shannon entropy) - medians {.tabset .tabset-pills}

Per-locus normalised Shannon entropy from `.norm.det.out.gz`. Same downsampling as above.

```{r import_adjH}
norm_det_paths <- list.files(snakemake@params[["output_path"]],
                              pattern = "norm\\.det\\.out\\.gz$", full.names = TRUE)

read_adjH <- function(fp) {
  fn  <- basename(fp)
  tmp <- read_tsv(fp, na = "-1", show_col_types = FALSE, n_max = 10e4)
  tibble(
    file            = fn,
    median_shannon  = median(tmp$shannon_norm, na.rm = TRUE),
    median_avg_meth = median(tmp$avg_meth,     na.rm = TRUE),
    annotation      = get_annotation(fn),
    major_region    = get_major_region(fn),
    cell_class      = get_cell_class(fn)
  )
}

adjshannons <- bplapply(norm_det_paths, read_adjH, BPPARAM = param) |> bind_rows()
adjshannons <- harmonize_levels(adjshannons)
head(adjshannons)
```

## By cell class

```{r scatter_adjH_class}
ggplot(adjshannons, aes(x = median_avg_meth, y = median_shannon, color = cell_class)) +
  geom_point(size = 1.5, alpha = 0.8) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~annotation) +
  labs(x = "median avg. methylation", y = "median adj. Shannon entropy (adjH)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

```{r boxes_adjH_class, fig.width = 10, fig.height = 6}
ggplot(adjshannons, aes(x = annotation, y = median_shannon, color = cell_class)) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  scale_color_brewer(palette = "Dark2") +
  labs(y = "median adj. Shannon entropy (adjH)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## By major region

```{r scatter_adjH_region}
ggplot(adjshannons, aes(x = median_avg_meth, y = median_shannon, color = major_region)) +
  geom_point(size = 1.5, alpha = 0.8) +
  facet_wrap(~annotation) +
  labs(x = "median avg. methylation", y = "median adj. Shannon entropy (adjH)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

```{r boxes_adjH_region, fig.width = 10, fig.height = 6}
ggplot(adjshannons, aes(x = annotation, y = median_shannon, color = major_region)) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  labs(y = "median adj. Shannon entropy (adjH)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## Region by class

```{r boxes_adjH_combo, fig.width = 12, fig.height = 8}
ggplot(adjshannons, aes(x = cell_class, y = median_shannon, color = major_region)) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  facet_wrap(~annotation) +
  labs(y = "median adj. Shannon entropy (adjH)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

# Assembly

S and adjS are per-cell values. H and adjH are medians across loci for each group.

```{r str_check}
str(sampens)
str(adjsampens)
str(shannons)
str(adjshannons)
```

```{r save_rds}
saveRDS(
  list(sampens     = sampens,
       adjsampens  = adjsampens,
       shannons    = shannons,
       adjshannons = adjshannons),
  file = "ecker_entropy.rds"
)
```

Aggregate per-cell S and adjS to per-group medians so they are comparable to H and adjH.

```{r group_medians}
sampen_grp <- sampens %>%
  group_by(annotation, major_region, cell_class) %>%
  summarise(median_S    = median(sampen,   na.rm = TRUE),
            median_meth = median(avg_meth, na.rm = TRUE),
            .groups = "drop")

adjS_grp <- adjsampens %>%
  group_by(annotation, major_region, cell_class) %>%
  summarise(median_adjS = median(sampen, na.rm = TRUE),
            .groups = "drop")

all_grp <- sampen_grp %>%
  left_join(adjS_grp,    by = c("annotation", "major_region", "cell_class")) %>%
  left_join(shannons  %>% select(annotation, major_region, cell_class,
                                  median_H    = median_shannon),
            by = c("annotation", "major_region", "cell_class")) %>%
  left_join(adjshannons %>% select(annotation, major_region, cell_class,
                                    median_adjH = median_shannon),
            by = c("annotation", "major_region", "cell_class"))

head(all_grp)
```

## adjS vs adjH

```{r adjS_vs_adjH, fig.height = 10, fig.width = 8}
ggplot(all_grp, aes(x = median_adjH, y = median_adjS,
                     color = cell_class, shape = major_region)) +
  geom_point(size = 1.5) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~annotation, ncol = 3) +
  labs(x = "median adjH", y = "median adjS") +
  theme_ng()
```

## S vs H

```{r S_vs_H, fig.height = 10, fig.width = 8}
ggplot(all_grp, aes(x = median_H, y = median_S,
                     color = cell_class, shape = major_region)) +
  geom_point(size = 1.5) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~annotation, ncol = 3) +
  labs(x = "median H", y = "median S") +
  theme_ng()
```

# Variation {.tabset .tabset-pills}

Range and SD of group medians for each metric, broken down by major region, cell class, and their combination.

```{r variation_helpers}
## Summarise variation of a numeric vector: range and SD
var_stats <- function(x) {
  x <- x[!is.na(x)]
  list(range = diff(range(x)), sd = sd(x))
}
```

## Across cell classes (per region and annotation)

How much do entropies differ between cell classes within the same region?

```{r variation_class, fig.width = 10, fig.height = 8}
var_class <- all_grp %>%
  group_by(annotation, major_region) %>%
  summarise(
    S_range    = diff(range(median_S,    na.rm = TRUE)),
    adjS_range = diff(range(median_adjS, na.rm = TRUE)),
    H_range    = diff(range(median_H,    na.rm = TRUE)),
    adjH_range = diff(range(median_adjH, na.rm = TRUE)),
    S_sd       = sd(median_S,    na.rm = TRUE),
    adjS_sd    = sd(median_adjS, na.rm = TRUE),
    H_sd       = sd(median_H,    na.rm = TRUE),
    adjH_sd    = sd(median_adjH, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(var_class, aes(x = adjH_range, y = adjS_range,
                       color = major_region, label = annotation)) +
  geom_point(size = 2) +
  geom_text_repel(size = 3) +
  labs(x = "adjH range across cell classes",
       y = "adjS range across cell classes",
       title = "Variation across cell classes (per region)") +
  theme_ng()

ggplot(var_class, aes(x = H_range, y = S_range,
                       color = major_region, label = annotation)) +
  geom_point(size = 2) +
  geom_text_repel(size = 3) +
  labs(x = "H range across cell classes",
       y = "S range across cell classes",
       title = "Variation across cell classes (per region), unnormalised") +
  theme_ng()
```

## Across major regions (per cell class and annotation)

How much do entropies differ between major regions within the same cell class?

```{r variation_region, fig.width = 10, fig.height = 8}
var_region <- all_grp %>%
  group_by(annotation, cell_class) %>%
  summarise(
    S_range    = diff(range(median_S,    na.rm = TRUE)),
    adjS_range = diff(range(median_adjS, na.rm = TRUE)),
    H_range    = diff(range(median_H,    na.rm = TRUE)),
    adjH_range = diff(range(median_adjH, na.rm = TRUE)),
    S_sd       = sd(median_S,    na.rm = TRUE),
    adjS_sd    = sd(median_adjS, na.rm = TRUE),
    H_sd       = sd(median_H,    na.rm = TRUE),
    adjH_sd    = sd(median_adjH, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(var_region, aes(x = adjH_range, y = adjS_range,
                        color = cell_class, label = annotation)) +
  geom_point(size = 2) +
  geom_text_repel(size = 3) +
  labs(x = "adjH range across major regions",
       y = "adjS range across major regions",
       title = "Variation across major regions (per cell class)") +
  theme_ng()

ggplot(var_region, aes(x = H_range, y = S_range,
                        color = cell_class, label = annotation)) +
  geom_point(size = 2) +
  geom_text_repel(size = 3) +
  labs(x = "H range across major regions",
       y = "S range across major regions",
       title = "Variation across major regions (per cell class), unnormalised") +
  theme_ng()
```

## Across all groups (annotation-level)

Overall variation collapsing region and class together.

```{r variation_all, fig.width = 8, fig.height = 6}
var_all <- all_grp %>%
  group_by(annotation) %>%
  summarise(
    S_range    = diff(range(median_S,    na.rm = TRUE)),
    adjS_range = diff(range(median_adjS, na.rm = TRUE)),
    H_range    = diff(range(median_H,    na.rm = TRUE)),
    adjH_range = diff(range(median_adjH, na.rm = TRUE)),
    S_sd       = sd(median_S,    na.rm = TRUE),
    adjS_sd    = sd(median_adjS, na.rm = TRUE),
    H_sd       = sd(median_H,    na.rm = TRUE),
    adjH_sd    = sd(median_adjH, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(var_all, aes(x = adjH_range, y = adjS_range, label = annotation)) +
  geom_point(size = 3) +
  geom_text_repel() +
  labs(x = "adjH range across all groups",
       y = "adjS range across all groups",
       title = "Overall variation (region by class)") +
  theme_ng()
```

## Region vs class: which drives more variation?

Mean range across groups for cell class and major region, per metric.

```{r class_vs_region_comparison, fig.width = 10, fig.height = 6}
mean_var <- bind_rows(
  var_class  %>% summarise(across(ends_with("_range"), mean, na.rm = TRUE)) %>%
    mutate(driver = "cell class"),
  var_region %>% summarise(across(ends_with("_range"), mean, na.rm = TRUE)) %>%
    mutate(driver = "major region")
)

mean_var_long <- mean_var %>%
  tidyr::pivot_longer(-driver, names_to = "metric", values_to = "mean_range") %>%
  mutate(metric = sub("_range$", "", metric))

ggplot(mean_var_long, aes(x = metric, y = mean_range, fill = driver)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = "mean range across groups",
       title = "Average entropy range: cell class vs major region") +
  theme_ng()
```

# Driver categorization {.tabset .tabset-pills}

Annotations are labelled as within-cell (adjS-driven), across-cell (adjH-driven),
or both, based on overall variation across all region by class groups.

```{r drivers_range}
shannon_thr <- quantile(var_all$adjH_range, 0.5, na.rm = TRUE)
sampen_thr  <- quantile(var_all$adjS_range, 0.5, na.rm = TRUE)

var_all <- var_all %>%
  mutate(
    driver = case_when(
      adjH_range < shannon_thr & adjS_range < sampen_thr ~ "not changing enough",
      adjH_range >= adjS_range * 1.5                      ~ "across-cell driven",
      adjS_range >= adjH_range * 1.5                      ~ "within-cell driven",
      TRUE                                                 ~ "both"
    )
  )

table(var_all$driver)

ggplot(var_all,
       aes(x = adjH_range, y = adjS_range,
           color = driver, shape = driver, label = annotation)) +
  geom_point(size = 3) +
  geom_text_repel() +
  labs(x = "adjH range across all groups",
       y = "adjS range across all groups") +
  theme_ng()
```

```{r drivers_sd}
shannon_sd_thr <- quantile(var_all$adjH_sd, 0.3, na.rm = TRUE)
sampen_sd_thr  <- quantile(var_all$adjS_sd, 0.3, na.rm = TRUE)

var_all <- var_all %>%
  mutate(
    driver_sd = case_when(
      adjH_sd < shannon_sd_thr & adjS_sd < sampen_sd_thr ~ "not changing enough",
      adjH_sd >= adjS_sd * 1.5                            ~ "across-cell driven",
      adjS_sd >= adjH_sd * 1.5                            ~ "within-cell driven",
      TRUE                                                 ~ "both"
    )
  )

table(var_all$driver_sd)

ggplot(var_all,
       aes(x = adjH_sd, y = adjS_sd,
           color = driver_sd, shape = driver_sd, label = annotation)) +
  geom_point(size = 3) +
  geom_text_repel() +
  labs(x = "SD of adjH across all groups",
       y = "SD of adjS across all groups") +
  theme_ng()
```

# Session

```{r session, cache = FALSE}
sessionInfo()
```
