---
title: "Ecker 2021 mouse brain: methylation entropy by feature (CpG context)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Izaskun Mallona"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
---

```{r logging_early, include = FALSE}
log <- file(snakemake@log$log, open = "wt")
sink(log)
sink(log, type = "message")
```

```{r setup}
suppressPackageStartupMessages({
  library(readr)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(BiocParallel)
  library(ggrepel)
  library(ggridges)
  library(ComplexHeatmap)
  library(circlize)
  library(irlba)
})

source("src/ggtheme.R")
param <- MulticoreParam(workers = snakemake@threads)
```

```{r opts}
knitr::opts_chunk$set(
  echo       = TRUE,
  fig.width  = 5,
  fig.height = 5,
  cache      = FALSE,
  include    = TRUE,
  fig.path   = "ecker_plots/",
  dev        = c("png", "svg"),
  dpi        = 500,
  cache.lazy = FALSE,
  warning    = TRUE,
  message    = TRUE
)
```

```{r git_commit, echo = FALSE, results = "asis"}
git_sha <- tryCatch(
  system("git rev-parse --short HEAD", intern = TRUE, ignore.stderr = TRUE),
  error = function(e) "unknown"
)
cat(sprintf("Git commit: `%s`\n\n", git_sha))
```

```{r save_snmk_debug, include = FALSE}
saveRDS(file = "snmk_ecker_features.rds", object = snakemake, compress = "xz")
```

Mouse (mm10), CpG context only. chr10 only: `r snakemake@params[["chr10_only"]]`.

```{r helpers}
## filenames: {annotation}_{sub_region}_{sub_type}.*.gz
## annotation has no underscores; neither do sub_region or sub_type values
get_annotation <- function(fn) sub("^(.*)_[^_]+_[^_]+\\..*$", "\\1", fn)
get_sub_region <- function(fn) sub("^.*_([^_]+)_[^_]+\\..*$", "\\1", fn)
get_sub_type   <- function(fn) sub("^.*_([^_.]+)\\.[^_]*$",   "\\1", fn)

ann_labels <- c(
  "genes"    = "Genes",
  "promoters"= "Promoters",
  "lines"    = "LINEs",
  "sines"    = "SINEs",
  "h3k4me3"  = "H3K4me3",
  "h3k9me3"  = "H3K9me3",
  "h3k27me3" = "H3K27me3",
  "h3k4me1"  = "H3K4me1",
  "h3k27ac"  = "H3K27ac"
)

harmonize_levels <- function(df) {
  missing_ann <- setdiff(unique(df$annotation), names(ann_labels))
  if (length(missing_ann) > 0)
    warning("unknown annotations: ", paste(missing_ann, collapse = ", "))
  df$annotation <- factor(df$annotation,
                           levels = names(ann_labels),
                           labels = unname(ann_labels))
  if ("sub_region" %in% names(df)) df$sub_region <- factor(df$sub_region)
  if ("sub_type"   %in% names(df)) df$sub_type   <- factor(df$sub_type)
  df
}
```

```{r load_metadata_early}
## Load cell-level metadata early so cell_class / major_type are available
## in all data-loading chunks below.
meta_full <- read_tsv(snakemake@params[["meta_path"]],
                       na = "NA", show_col_types = FALSE) %>%
  mutate(
    sub_region   = gsub(" ", "-", SubRegion),
    sub_type     = gsub(" ", "-", SubType),
    major_region = gsub(" ", "-", MajorRegion)
  )

meta_grp <- meta_full %>%
  group_by(sub_region, sub_type) %>%
  summarise(
    cell_class   = first(na.omit(CellClass)),
    major_type   = first(na.omit(MajorType)),
    major_region = first(na.omit(major_region)),
    umap1        = median(L1UMAP_0, na.rm = TRUE),
    umap2        = median(L1UMAP_1, na.rm = TRUE),
    n_meta       = n(),
    .groups      = "drop"
  )

meta_cells <- meta_full %>%
  select(sub_region, sub_type, major_region,
         cell_class = CellClass, major_type = MajorType,
         umap1 = L1UMAP_0, umap2 = L1UMAP_1, basename)

cat("meta_grp rows:", nrow(meta_grp), "\n")
cat("major_region levels:", paste(sort(unique(meta_grp$major_region)), collapse = ", "), "\n")
cat("cell_class levels:", paste(sort(unique(meta_grp$cell_class)), collapse = ", "), "\n")
```

# Unnormalized sampEn (S) {.tabset .tabset-pills}

Per-cell sample entropy (S) from `.out.gz` files.

```{r import_sampen}
short_paths <- list.files(snakemake@params[["output_path"]],
                           pattern = "\\.out\\.gz$", full.names = TRUE)
short_paths <- short_paths[!grepl("\\.det\\.out\\.gz$",  short_paths)]
short_paths <- short_paths[!grepl("\\.meth\\.out\\.gz$", short_paths)]

read_short <- function(fp) {
  fn <- basename(fp)
  dt <- read_tsv(fp, na = "-1", show_col_types = FALSE)
  dt <- dt[, c("sampen", "avg_meth")]
  dt$annotation <- get_annotation(fn)
  dt$sub_region <- get_sub_region(fn)
  dt$sub_type   <- get_sub_type(fn)
  dt
}

sampens <- do.call(rbind, bplapply(short_paths, read_short, BPPARAM = param))
sampens <- harmonize_levels(sampens)
sampens <- sampens %>%
  left_join(meta_grp %>% select(sub_region, sub_type, cell_class, major_type, major_region),
            by = c("sub_region" = "sub_region", "sub_type" = "sub_type"))
cat("cell_class NA:", sum(is.na(sampens$cell_class)), "/", nrow(sampens), "\n")
head(sampens)
```

## Entropy vs methylation

```{r scatter_sampen, fig.width = 10, fig.height = 15}
ggplot(sampens %>% filter(!is.na(cell_class)),
       aes(x = avg_meth, y = sampen, color = major_type)) +
  geom_point(size = 0.4, alpha = 0.15) +
  scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95, na.value = "grey70") +
  facet_grid(annotation ~ cell_class) +
  labs(x = "average methylation", y = "sample entropy (S)", color = "cell type") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng()
```

## Distributions

```{r violin_sampen, fig.width = 10, fig.height = 15}
ggplot(sampens %>% filter(!is.na(cell_class)),
       aes(x = major_type, y = sampen, fill = major_type)) +
  geom_violin(alpha = 0.5, draw_quantiles = 0.5) +
  scale_fill_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
  facet_grid(annotation ~ cell_class, scales = "free_y") +
  labs(y = "sample entropy (S)", fill = "cell type") +
  theme_ng() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

```{r ridges_sampen, fig.width = 10, fig.height = 15}
ggplot(sampens %>% filter(!is.na(cell_class)),
       aes(x = sampen, y = major_type, fill = major_type)) +
  geom_density_ridges(alpha = 0.6, scale = 0.85) +
  scale_fill_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
  facet_grid(annotation ~ cell_class, scales = "free") +
  labs(x = "sample entropy (S)") +
  theme_ng() +
  theme(legend.position = "none")
```

# adjS (methylation-normalised sample entropy) {.tabset .tabset-pills}

Per-cell normalised sample entropy from `.out.gz` files.

```{r import_adjS}
norm_paths <- list.files(snakemake@params[["output_path"]],
                          pattern = "out\\.gz$", full.names = TRUE)
norm_paths <- norm_paths[!grepl("\\.det\\.out\\.gz$",  norm_paths)]
norm_paths <- norm_paths[!grepl("\\.norm\\.out\\.gz$", norm_paths)]
norm_paths <- norm_paths[!grepl("\\.meth\\.out\\.gz$", norm_paths)]

read_adjS <- function(fp) {
  fn <- basename(fp)
  dt <- read_tsv(fp, na = "-1", show_col_types = FALSE)
  dt <- dt[, c("sampen_norm", "avg_meth")]
  colnames(dt) <- c("sampen", "avg_meth")
  dt$annotation <- get_annotation(fn)
  dt$sub_region <- get_sub_region(fn)
  dt$sub_type   <- get_sub_type(fn)
  dt
}

adjsampens <- do.call(rbind, bplapply(norm_paths, read_adjS, BPPARAM = param))
adjsampens <- harmonize_levels(adjsampens)
adjsampens <- adjsampens %>%
  left_join(meta_grp %>% select(sub_region, sub_type, cell_class, major_type, major_region),
            by = c("sub_region" = "sub_region", "sub_type" = "sub_type"))
head(adjsampens)
```

## Entropy vs methylation

```{r scatter_adjS, fig.width = 10, fig.height = 15}
ggplot(adjsampens %>% filter(!is.na(cell_class)),
       aes(x = avg_meth, y = sampen, color = major_type)) +
  geom_point(size = 0.4, alpha = 0.15) +
  scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95, na.value = "grey70") +
  facet_grid(annotation ~ cell_class) +
  labs(x = "average methylation", y = "adj. sample entropy (adjS)", color = "cell type") +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng()
```

## Distributions

```{r violin_adjS, fig.width = 10, fig.height = 15}
ggplot(adjsampens %>% filter(!is.na(cell_class)),
       aes(x = major_type, y = sampen, fill = major_type)) +
  geom_violin(alpha = 0.5, draw_quantiles = 0.5) +
  scale_fill_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
  facet_grid(annotation ~ cell_class, scales = "free_y") +
  labs(y = "adj. sample entropy (adjS)", fill = "cell type") +
  theme_ng() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

```{r ridges_adjS, fig.width = 10, fig.height = 15}
ggplot(adjsampens %>% filter(!is.na(cell_class)),
       aes(x = sampen, y = major_type, fill = major_type)) +
  geom_density_ridges(alpha = 0.6, scale = 0.85) +
  scale_fill_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
  facet_grid(annotation ~ cell_class, scales = "free") +
  labs(x = "adj. sample entropy (adjS)") +
  theme_ng() +
  theme(legend.position = "none")
```

# Shannon H - medians {.tabset .tabset-pills}

Per-locus Shannon entropy (H) from `.det.out.gz`. One median per group per annotation.

```{r import_shannon}
det_paths <- list.files(snakemake@params[["output_path"]],
                         pattern = "det\\.out\\.gz$", full.names = TRUE)
det_paths <- det_paths[!grepl("norm", det_paths)]

read_shannon <- function(fp) {
  fn  <- basename(fp)
  tmp <- read_tsv(fp, na = "-1", show_col_types = FALSE, n_max = 10e4)
  tibble(
    file            = fn,
    median_shannon  = median(tmp$shannon,  na.rm = TRUE),
    median_avg_meth = median(tmp$avg_meth, na.rm = TRUE),
    annotation      = get_annotation(fn),
    sub_region      = get_sub_region(fn),
    sub_type        = get_sub_type(fn)
  )
}

shannons <- bplapply(det_paths, read_shannon, BPPARAM = param) |> bind_rows()
shannons <- harmonize_levels(shannons)
shannons <- shannons %>%
  left_join(meta_grp %>% select(sub_region, sub_type, cell_class, major_type, major_region),
            by = c("sub_region" = "sub_region", "sub_type" = "sub_type"))
head(shannons)
```

## Median H vs methylation

```{r scatter_shannon, fig.width = 10, fig.height = 15}
ggplot(shannons %>% filter(!is.na(cell_class)),
       aes(x = median_avg_meth, y = median_shannon, color = major_type)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95, na.value = "grey70") +
  facet_grid(annotation ~ cell_class) +
  labs(x = "median avg. methylation", y = "median Shannon entropy (H)", color = "cell type") +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  theme_ng()
```

# adjH (methylation-normalised Shannon entropy) - medians {.tabset .tabset-pills}

Per-locus normalised Shannon entropy from `.norm.det.out.gz`.

```{r import_adjH}
norm_det_paths <- list.files(snakemake@params[["output_path"]],
                              pattern = "norm\\.det\\.out\\.gz$", full.names = TRUE)

read_adjH <- function(fp) {
  fn  <- basename(fp)
  tmp <- read_tsv(fp, na = "-1", show_col_types = FALSE, n_max = 10e4)
  tibble(
    file            = fn,
    median_shannon  = median(tmp$shannon_norm, na.rm = TRUE),
    median_avg_meth = median(tmp$avg_meth,     na.rm = TRUE),
    annotation      = get_annotation(fn),
    sub_region      = get_sub_region(fn),
    sub_type        = get_sub_type(fn)
  )
}

adjshannons <- bplapply(norm_det_paths, read_adjH, BPPARAM = param) |> bind_rows()
adjshannons <- harmonize_levels(adjshannons)
adjshannons <- adjshannons %>%
  left_join(meta_grp %>% select(sub_region, sub_type, cell_class, major_type, major_region),
            by = c("sub_region" = "sub_region", "sub_type" = "sub_type"))
head(adjshannons)
```

## Median adjH vs methylation

```{r scatter_adjH, fig.width = 10, fig.height = 15}
ggplot(adjshannons %>% filter(!is.na(cell_class)),
       aes(x = median_avg_meth, y = median_shannon, color = major_type)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95, na.value = "grey70") +
  facet_grid(annotation ~ cell_class) +
  labs(x = "median avg. methylation", y = "median adj. Shannon entropy (adjH)", color = "cell type") +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  theme_ng()
```

# Assembly

S and adjS are per-cell values. H and adjH are medians across loci for each group.

```{r str_check}
str(sampens)
str(adjsampens)
str(shannons)
str(adjshannons)
```

```{r save_rds}
saveRDS(
  list(sampens     = sampens,
       adjsampens  = adjsampens,
       shannons    = shannons,
       adjshannons = adjshannons),
  file = "ecker_entropy.rds"
)
```

Aggregate per-cell S and adjS to per-group medians so they are comparable to H and adjH.

```{r group_medians}
sampen_grp <- sampens %>%
  group_by(annotation, sub_region, sub_type) %>%
  summarise(median_S    = median(sampen,   na.rm = TRUE),
            median_meth = median(avg_meth, na.rm = TRUE),
            .groups = "drop")

adjS_grp <- adjsampens %>%
  group_by(annotation, sub_region, sub_type) %>%
  summarise(median_adjS = median(sampen, na.rm = TRUE),
            .groups = "drop")

all_grp <- sampen_grp %>%
  left_join(adjS_grp, by = c("annotation", "sub_region", "sub_type")) %>%
  left_join(shannons   %>% select(annotation, sub_region, sub_type,
                                   median_H   = median_shannon),
            by = c("annotation", "sub_region", "sub_type")) %>%
  left_join(adjshannons %>% select(annotation, sub_region, sub_type,
                                    median_adjH = median_shannon),
            by = c("annotation", "sub_region", "sub_type"))

head(all_grp)
```

```{r load_metadata}
## Attach CellClass, MajorType, and paper UMAP coords to the group-level table.
all_grp_meta <- all_grp %>%
  mutate(sub_region = as.character(sub_region),
         sub_type   = as.character(sub_type)) %>%
  left_join(meta_grp, by = c("sub_region", "sub_type"))

saveRDS(
  list(all_grp      = all_grp,
       meta_grp     = meta_grp,
       all_grp_meta = all_grp_meta,
       meta_cells   = meta_cells),
  file = "ecker_groups_meta.rds"
)

cat("all_grp_meta rows:", nrow(all_grp_meta), "\n")
cat("cell_class non-NA:", sum(!is.na(all_grp_meta$cell_class)), "\n")
cat("median_H non-NA:", sum(!is.na(all_grp_meta$median_H)), "\n")
```

## adjS vs adjH

```{r adjS_vs_adjH, fig.width = 13, fig.height = 4}
df_cmp <- all_grp_meta %>% filter(!is.na(cell_class))
if (nrow(df_cmp) > 0) {
  ggplot(df_cmp, aes(x = median_adjH, y = median_adjS, color = annotation)) +
    geom_point(size = 1.5, alpha = 0.8) +
    scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
    facet_grid(. ~ cell_class) +
    labs(x = "median adjH", y = "median adjS", color = "annotation") +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme_ng()
} else {
  message("No data with cell_class assigned - check metadata join")
}
```

## S vs H

```{r S_vs_H, fig.width = 13, fig.height = 4}
if (nrow(df_cmp) > 0) {
  ggplot(df_cmp, aes(x = median_H, y = median_S, color = annotation)) +
    geom_point(size = 1.5, alpha = 0.8) +
    scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
    facet_grid(. ~ cell_class) +
    labs(x = "median H", y = "median S", color = "annotation") +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme_ng()
}
```

# Variation across cell types {.tabset .tabset-pills}

SD and range of group medians per annotation, broken down by cell class.

```{r variation_helpers}
safe_range <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  diff(range(x))
}
```

```{r variation_class, fig.width = 13, fig.height = 4}
var_class <- all_grp_meta %>%
  filter(!is.na(cell_class)) %>%
  group_by(annotation, cell_class) %>%
  summarise(
    adjS_sd    = sd(median_adjS, na.rm = TRUE),
    adjH_sd    = sd(median_adjH, na.rm = TRUE),
    adjS_range = safe_range(median_adjS),
    adjH_range = safe_range(median_adjH),
    .groups = "drop"
  )

if (nrow(var_class) > 0 && any(!is.na(var_class$adjH_sd))) {
  ggplot(var_class, aes(x = adjH_sd, y = adjS_sd,
                         color = annotation, label = annotation)) +
    geom_point(size = 3) +
    geom_text_repel(size = 3) +
    scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
    facet_grid(. ~ cell_class) +
    labs(x = "SD of adjH across cell sub-types",
         y = "SD of adjS across cell sub-types") +
    theme_ng()
} else {
  message("No variation data available")
}
```

# Driver categorization {.tabset .tabset-pills}

Annotations labelled as within-cell (adjS-driven), across-cell (adjH-driven), or both, based on overall variation across all groups.

```{r drivers_sd, fig.width = 5, fig.height = 5}
var_all <- all_grp_meta %>%
  group_by(annotation) %>%
  summarise(
    adjS_sd    = sd(median_adjS, na.rm = TRUE),
    adjH_sd    = sd(median_adjH, na.rm = TRUE),
    adjS_range = safe_range(median_adjS),
    adjH_range = safe_range(median_adjH),
    .groups = "drop"
  )

if (any(!is.na(var_all$adjH_sd))) {
  shannon_sd_thr <- quantile(var_all$adjH_sd, 0.3, na.rm = TRUE)
  sampen_sd_thr  <- quantile(var_all$adjS_sd, 0.3, na.rm = TRUE)

  var_all <- var_all %>%
    mutate(
      driver = case_when(
        is.na(adjH_sd) | is.na(adjS_sd)                    ~ NA_character_,
        adjH_sd < shannon_sd_thr & adjS_sd < sampen_sd_thr ~ "not changing enough",
        adjH_sd >= adjS_sd * 1.5                            ~ "across-cell driven",
        adjS_sd >= adjH_sd * 1.5                            ~ "within-cell driven",
        TRUE                                                 ~ "both"
      )
    )

  cat("Driver table (SD-based):\n")
  print(table(var_all$driver, useNA = "always"))

  ggplot(var_all, aes(x = adjH_sd, y = adjS_sd,
                       color = driver, shape = driver, label = annotation)) +
    geom_point(size = 3) +
    geom_text_repel() +
    scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95, na.value = "grey60") +
    labs(x = "SD of adjH across all groups",
         y = "SD of adjS across all groups") +
    theme_ng()
} else {
  message("No variation data available - check group medians and metadata join")
}
```

## R-squared by cell class

Variance in group-level adjH and adjS explained by cell class (R²), one point per annotation. More principled than SD comparison: points upper-left are within-cell driven, lower-right are across-cell driven.

```{r drivers_r2, fig.width = 5, fig.height = 5}
r2_by_ann <- all_grp_meta %>%
  filter(!is.na(cell_class)) %>%
  group_by(annotation) %>%
  summarise(
    r2_adjH = tryCatch({
      m <- lm(median_adjH ~ cell_class, data = cur_data())
      summary(m)$r.squared
    }, error = function(e) NA_real_),
    r2_adjS = tryCatch({
      m <- lm(median_adjS ~ cell_class, data = cur_data())
      summary(m)$r.squared
    }, error = function(e) NA_real_),
    .groups = "drop"
  )

if (any(!is.na(r2_by_ann$r2_adjH))) {
  r2_thr <- 0.3

  r2_by_ann <- r2_by_ann %>%
    mutate(
      driver_r2 = case_when(
        is.na(r2_adjH) | is.na(r2_adjS)              ~ NA_character_,
        r2_adjH < r2_thr & r2_adjS < r2_thr          ~ "not explained by cell class",
        r2_adjH >= r2_adjS * 1.5 & r2_adjH >= r2_thr ~ "across-cell driven",
        r2_adjS >= r2_adjH * 1.5 & r2_adjS >= r2_thr ~ "within-cell driven",
        TRUE                                           ~ "both"
      )
    )

  cat("Driver table (R2-based):\n")
  print(table(r2_by_ann$driver_r2, useNA = "always"))

  ggplot(r2_by_ann, aes(x = r2_adjH, y = r2_adjS,
                         color = driver_r2, shape = driver_r2, label = annotation)) +
    geom_vline(xintercept = r2_thr, linetype = "dashed", color = "grey60") +
    geom_hline(yintercept = r2_thr, linetype = "dashed", color = "grey60") +
    geom_point(size = 3) +
    geom_text_repel(size = 3) +
    scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95, na.value = "grey60") +
    scale_x_continuous(limits = c(0, 1)) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(x = "R² of adjH explained by cell class",
         y = "R² of adjS explained by cell class",
         color = "driver", shape = "driver") +
    theme_ng()
} else {
  message("No R2 data available - check group medians and metadata join")
}
```

# Heatmaps of median entropy per group {.tabset .tabset-pills}

Rows = sub-type (clustered). Columns = annotation. Row annotation bars show cell class and major type.

```{r heatmap_helpers}
make_heatmap <- function(df, value_col, title,
                          palette = c("navy", "white", "firebrick")) {
  ## wide matrix: rows = sub_type, cols = annotation
  wide <- df %>%
    select(sub_type, annotation, cell_class, major_type,
           value = all_of(value_col)) %>%
    filter(!is.na(cell_class)) %>%
    group_by(sub_type, annotation, cell_class, major_type) %>%
    summarise(value = median(value, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = annotation, values_from = value)

  if (nrow(wide) < 2) { message("Too few rows for heatmap"); return(invisible(NULL)) }

  mat <- wide %>% select(-sub_type, -cell_class, -major_type) %>% as.matrix()
  rownames(mat) <- wide$sub_type

  ## row annotations
  cc_levels   <- sort(unique(wide$cell_class))
  mt_levels   <- sort(unique(wide$major_type))
  cc_pal      <- setNames(viridisLite::viridis(length(cc_levels), begin = 0.1, end = 0.9), cc_levels)
  mt_pal      <- setNames(viridisLite::turbo(length(mt_levels), begin = 0.05, end = 0.95), mt_levels)

  row_ann <- rowAnnotation(
    cell_class = wide$cell_class,
    major_type = wide$major_type,
    col = list(cell_class = cc_pal, major_type = mt_pal),
    annotation_legend_param = list(
      cell_class = list(title = "cell class"),
      major_type = list(title = "major type")
    )
  )

  col_fun <- colorRamp2(
    c(min(mat, na.rm = TRUE),
      mean(range(mat, na.rm = TRUE)),
      max(mat, na.rm = TRUE)),
    palette
  )

  Heatmap(mat,
          name            = title,
          col             = col_fun,
          row_split       = wide$cell_class,
          cluster_rows    = TRUE,
          cluster_columns = FALSE,
          show_row_names  = TRUE,
          show_column_names = TRUE,
          row_names_gp    = gpar(fontsize = 7),
          right_annotation = row_ann,
          column_title    = title,
          na_col          = "grey90")
}
```

## adjS

```{r heatmap_adjS, fig.width = 10, fig.height = 8}
heat_df_adjs <- adjsampens %>%
  group_by(annotation, sub_region, sub_type, cell_class, major_type) %>%
  summarise(median_adjS = median(sampen, na.rm = TRUE), .groups = "drop")

make_heatmap(heat_df_adjs, "median_adjS", "median adjS",
             palette = c("navy", "white", "firebrick"))
```

## adjH

```{r heatmap_adjH, fig.width = 10, fig.height = 8}
heat_df_adjh <- adjshannons %>%
  group_by(annotation, sub_region, sub_type, cell_class, major_type) %>%
  summarise(median_adjH = median(median_shannon, na.rm = TRUE), .groups = "drop")

make_heatmap(heat_df_adjh, "median_adjH", "median adjH",
             palette = c("navy", "white", "firebrick"))
```

## S

```{r heatmap_S, fig.width = 10, fig.height = 8}
heat_df_s <- sampens %>%
  group_by(annotation, sub_region, sub_type, cell_class, major_type) %>%
  summarise(median_S = median(sampen, na.rm = TRUE), .groups = "drop")

make_heatmap(heat_df_s, "median_S", "median S",
             palette = c("navy", "white", "firebrick"))
```

# Locus-level H distributions {.tabset .tabset-pills}

Per-locus Shannon entropy (H) distributions from `.det.out.gz`. Same row indices sampled across all groups.

```{r locus_level_load}
LOCUS_ANNOTATIONS <- c("genes", "h3k4me3")
LOCUS_N           <- 5000L
LOCUS_SEED        <- 42L

det_paths_locus <- list.files(snakemake@params[["output_path"]],
                               pattern = "det\\.out\\.gz$", full.names = TRUE)
det_paths_locus <- det_paths_locus[!grepl("norm", det_paths_locus)]
det_paths_locus <- det_paths_locus[
  get_annotation(basename(det_paths_locus)) %in% LOCUS_ANNOTATIONS
]

locus_indices <- lapply(LOCUS_ANNOTATIONS, function(ann) {
  fps <- det_paths_locus[get_annotation(basename(det_paths_locus)) == ann]
  if (length(fps) == 0) return(NULL)
  nrows <- nrow(read_tsv(fps[[1]], na = "-1", show_col_types = FALSE))
  n     <- min(LOCUS_N, nrows)
  set.seed(LOCUS_SEED)
  sort(sample.int(nrows, n))
})
names(locus_indices) <- LOCUS_ANNOTATIONS

read_locus <- function(fp) {
  fn  <- basename(fp)
  ann <- get_annotation(fn)
  idx <- locus_indices[[ann]]
  if (is.null(idx)) return(NULL)
  tmp <- read_tsv(fp, na = "-1", show_col_types = FALSE)
  tmp <- tmp[idx, , drop = FALSE]
  tibble(
    shannon    = tmp$shannon,
    avg_meth   = tmp$avg_meth,
    annotation = ann,
    sub_region = get_sub_region(fn),
    sub_type   = get_sub_type(fn)
  )
}

locus_data <- bplapply(det_paths_locus, read_locus, BPPARAM = param) |> bind_rows()
locus_data <- harmonize_levels(locus_data)
locus_data <- locus_data %>%
  left_join(meta_grp %>% select(sub_region, sub_type, cell_class, major_type, major_region),
            by = c("sub_region" = "sub_region", "sub_type" = "sub_type"))
head(locus_data)
```

## Violin

```{r locus_violin, fig.width = 11, fig.height = 5}
ggplot(locus_data %>% filter(!is.na(cell_class)),
       aes(x = major_type, y = shannon, fill = major_type)) +
  geom_violin(alpha = 0.5, draw_quantiles = 0.5) +
  scale_fill_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
  facet_grid(annotation ~ cell_class, scales = "free_y") +
  labs(y = "per-locus Shannon entropy (H)", fill = "cell type") +
  theme_ng() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

## Ridgelines

```{r locus_ridges, fig.width = 11, fig.height = 5}
ggplot(locus_data %>% filter(!is.na(cell_class)),
       aes(x = shannon, y = major_type, fill = major_type)) +
  geom_density_ridges(alpha = 0.6, scale = 0.85) +
  scale_fill_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
  facet_grid(annotation ~ cell_class, scales = "free") +
  labs(x = "per-locus Shannon entropy (H)") +
  theme_ng() +
  theme(legend.position = "none")
```

## Methylation vs H

```{r locus_meth_vs_H, fig.width = 11, fig.height = 5}
locus_sub <- locus_data %>%
  filter(!is.na(cell_class)) %>%
  slice_sample(n = min(3000, nrow(.)), replace = FALSE)

ggplot(locus_sub, aes(x = avg_meth, y = shannon, color = major_type)) +
  geom_point(size = 0.4, alpha = 0.2) +
  scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
  facet_grid(annotation ~ cell_class) +
  labs(x = "avg. methylation", y = "per-locus Shannon entropy (H)", color = "cell type") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  theme_ng()
```

# UMAP: cell-type separability {.tabset .tabset-pills}

```{r umap_helpers}
library(uwot)
source("src/embedding_utils.R")

plot_umap_class <- function(df, title, subtitle = NULL) {
  if (is.null(df) || nrow(df) == 0) {
    message("No UMAP result for: ", title)
    return(invisible(NULL))
  }
  if (all(is.na(df$cell_class))) {
    message("All cell_class values NA in UMAP for: ", title,
            " - check metadata join")
    return(invisible(NULL))
  }
  ggplot(df %>% filter(!is.na(cell_class)),
         aes(x = UMAP1, y = UMAP2, color = cell_class)) +
    geom_point(size = 2, alpha = 0.8) +
    scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
    labs(title = title, subtitle = subtitle, color = "cell class") +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme_ng()
}
```

## Cell-level: adjS across annotations

```{r cell_umap_load}
short_paths_umap <- list.files(snakemake@params[["output_path"]],
                                pattern = "\\.out\\.gz$", full.names = TRUE)
short_paths_umap <- short_paths_umap[!grepl("\\.det\\.out\\.gz$",  short_paths_umap)]
short_paths_umap <- short_paths_umap[!grepl("\\.meth\\.out\\.gz$", short_paths_umap)]

read_cells_long <- function(fps, col) {
  bind_rows(bplapply(fps, function(fp) {
    fn   <- basename(fp)
    vals <- read_tsv(fp, na = "-1", show_col_types = FALSE)[[col]]
    data.frame(
      sub_region = get_sub_region(fn),
      sub_type   = get_sub_type(fn),
      annotation = get_annotation(fn),
      cell_idx   = seq_along(vals),
      value      = vals,
      stringsAsFactors = FALSE
    )
  }, BPPARAM = param))
}

cells_adjS_long <- read_cells_long(short_paths_umap, "sampen_norm")
cells_meth_long <- read_cells_long(short_paths_umap, "avg_meth")

cells_adjS_wide <- pivot_wider(cells_adjS_long,
                                id_cols     = c(sub_region, sub_type, cell_idx),
                                names_from  = annotation,
                                values_from = value) %>%
  left_join(meta_grp %>% select(sub_region, sub_type, cell_class, major_type, major_region),
            by = c("sub_region", "sub_type"))

cells_meth_wide <- pivot_wider(cells_meth_long,
                                id_cols     = c(sub_region, sub_type, cell_idx),
                                names_from  = annotation,
                                values_from = value) %>%
  left_join(meta_grp %>% select(sub_region, sub_type, cell_class, major_type, major_region),
            by = c("sub_region", "sub_type"))

saveRDS(list(cells_adjS_wide = cells_adjS_wide,
             cells_meth_wide = cells_meth_wide),
        "ecker_cell_matrices.rds")

cat("Cells (adjS):", nrow(cells_adjS_wide),
    "| Annotations:", ncol(cells_adjS_wide) - 5L, "\n")
cat("Cell class breakdown:\n")
print(table(cells_adjS_wide$cell_class, useNA = "always"))
```

```{r cell_umap_adjS, fig.width = 5, fig.height = 5}
meta_cols_cell <- c("sub_region", "sub_type", "cell_idx", "cell_class", "major_type", "major_region")

umap_cell_adjS <- run_umap_wide(cells_adjS_wide, meta_cols_cell, n_neighbors = 15L)
saveRDS(umap_cell_adjS, "ecker_umap_cell_adjS.rds")

plot_umap_class(umap_cell_adjS,
                title    = "Cell-level UMAP - adjS features",
                subtitle = "per-cell adjS per annotation as feature vector")
```

## Cell-level: avg_meth across annotations

```{r cell_umap_meth, fig.width = 5, fig.height = 5}
umap_cell_meth <- run_umap_wide(cells_meth_wide, meta_cols_cell, n_neighbors = 15L)
saveRDS(umap_cell_meth, "ecker_umap_cell_meth.rds")

plot_umap_class(umap_cell_meth,
                title    = "Cell-level UMAP - avg. methylation features",
                subtitle = "per-cell avg. methylation per annotation as feature vector")
```

## Cell-level: S with marginal methylation regressed out

Unadjusted S per cell per annotation, with the per-cell mean methylation (averaged
across all annotations for that cell) regressed out column-wise before embedding.
This decouples entropy from the global methylation level of each cell while keeping
locus-level heterogeneity signal, as a check that cell-type separation is not merely
reflecting methylation differences.

```{r cell_umap_S_load}
cells_S_long <- read_cells_long(short_paths_umap, "sampen")

cells_S_wide <- pivot_wider(cells_S_long,
                            id_cols = c(sub_region, sub_type, cell_idx),
                            names_from = annotation,
                            values_from = value) %>%
  left_join(meta_grp %>% select(sub_region, sub_type, cell_class, major_type, major_region),
            by = c("sub_region", "sub_type"))

cat("Cells (S):", nrow(cells_S_wide), "\n")
```

```{r cell_umap_S_regress}
id_cols <- c("sub_region", "sub_type", "cell_idx")
shared_cells <- inner_join(
  cells_S_wide[, id_cols], cells_meth_wide[, id_cols], by = id_cols
)
cells_S_matched <- semi_join(cells_S_wide, shared_cells, by = id_cols)
cells_meth_matched <- semi_join(cells_meth_wide, shared_cells, by = id_cols)

annot_cols_S <- setdiff(colnames(cells_S_matched), meta_cols_cell)
annot_cols_meth <- setdiff(colnames(cells_meth_matched), meta_cols_cell)

S_resid <- regress_out_meth(
  as.matrix(cells_S_matched[, annot_cols_S, drop = FALSE]),
  as.matrix(cells_meth_matched[, annot_cols_meth, drop = FALSE])
)

cells_S_regressed <- bind_cols(
  cells_S_matched[, meta_cols_cell, drop = FALSE],
  as.data.frame(S_resid)
)
```

```{r cell_umap_S_regressed, fig.width = 5, fig.height = 5}
umap_cell_S_regressed <- run_umap_wide(cells_S_regressed, meta_cols_cell,
                                        n_neighbors = 15L)
saveRDS(umap_cell_S_regressed, "ecker_umap_cell_S_regressed.rds")

plot_umap_class(umap_cell_S_regressed,
                title    = "Cell-level UMAP - S with marginal meth regressed out",
                subtitle = "residuals of S ~ cell mean methylation per annotation")
```

## Group-level: adjH across annotations

```{r group_umap_adjH, fig.width = 5, fig.height = 5}
adjH_wide_grp <- all_grp_meta %>%
  filter(!is.na(cell_class)) %>%
  select(sub_region, sub_type, cell_class, major_type, annotation, median_adjH) %>%
  pivot_wider(names_from = annotation, values_from = median_adjH)

umap_grp_adjH <- run_umap_wide(adjH_wide_grp,
                                meta_cols = c("sub_region", "sub_type",
                                              "cell_class", "major_type"),
                                n_neighbors = 5L)
saveRDS(umap_grp_adjH, "ecker_umap_grp_adjH.rds")

plot_umap_class(umap_grp_adjH,
                title    = "Group-level UMAP - adjH features",
                subtitle = "median adjH per annotation as feature vector")
```

## Group-level: avg_meth across annotations

```{r group_umap_meth, fig.width = 5, fig.height = 5}
meth_wide_grp <- all_grp_meta %>%
  filter(!is.na(cell_class)) %>%
  select(sub_region, sub_type, cell_class, major_type, annotation, median_meth) %>%
  pivot_wider(names_from = annotation, values_from = median_meth)

umap_grp_meth <- run_umap_wide(meth_wide_grp,
                                meta_cols = c("sub_region", "sub_type",
                                              "cell_class", "major_type"),
                                n_neighbors = 5L)
plot_umap_class(umap_grp_meth,
                title    = "Group-level UMAP - avg. methylation features",
                subtitle = "median avg. methylation per annotation as feature vector")
```

## Group-level: gene loci, adjH

```{r gene_locus_umap_setup}
LOCUS_N_UMAP <- 2000L

norm_det_paths_genes <- list.files(snakemake@params[["output_path"]],
                                    pattern = "norm\\.det\\.out\\.gz$",
                                    full.names = TRUE)
det_paths_genes <- list.files(snakemake@params[["output_path"]],
                               pattern = "det\\.out\\.gz$", full.names = TRUE)
det_paths_genes <- det_paths_genes[!grepl("norm", det_paths_genes)]

## Two-pass coverage-based locus selection (same strategy as argelaguet.Rmd):
## Pass 1 scans all files to find loci covered in >= min_locus_cov_frac of groups.
## Pass 2 extracts values for the most-covered eligible loci (up to n_loci).
## is.finite() handles both NA (-1 converted on read) and Inf (norm.det.out missing).
build_gene_mat <- function(col, fps, n_loci = LOCUS_N_UMAP, min_locus_cov_frac = 0.5) {
  fps_g <- fps[get_annotation(basename(fps)) == "genes"]
  if (length(fps_g) < 4L) return(NULL)

  cov_counts <- NULL
  n_loci_total <- NULL
  for (fp in fps_g) {
    vals <- suppressWarnings(
      read_tsv(fp, na = "-1", show_col_types = FALSE))[[col]]
    if (is.null(cov_counts)) {
      n_loci_total <- length(vals)
      cov_counts <- integer(n_loci_total)
    }
    cov_counts <- cov_counts + is.finite(vals)
  }
  locus_cov <- cov_counts / length(fps_g)

  eligible <- which(locus_cov >= min_locus_cov_frac)
  if (length(eligible) < 2L) {
    cat("genes: no loci covered in >=", min_locus_cov_frac * 100,
        "% of groups (total loci =", n_loci_total, "); skipping\n")
    return(NULL)
  }
  top_loci <- eligible[order(locus_cov[eligible], decreasing = TRUE)]
  idx <- sort(top_loci[seq_len(min(as.integer(n_loci), length(top_loci)))])

  rows <- bplapply(fps_g, function(fp)
    suppressWarnings(
      read_tsv(fp, na = "-1", show_col_types = FALSE))[[col]][idx],
    BPPARAM = param)
  mat <- do.call(rbind, rows)
  colnames(mat) <- paste0("gene_", idx)

  cat("genes: groups =", nrow(mat),
      "| loci used =", ncol(mat),
      "| eligible (cov >=", min_locus_cov_frac, ") =", length(eligible),
      "| total loci =", n_loci_total, "\n")

  df <- data.frame(
    sub_region = get_sub_region(basename(fps_g)),
    sub_type   = get_sub_type(basename(fps_g)),
    as.data.frame(mat),
    check.names = FALSE
  )
  left_join(df,
            meta_grp %>% select(sub_region, sub_type, cell_class, major_type, major_region),
            by = c("sub_region", "sub_type"))
}

gene_mat_adjH <- build_gene_mat("shannon_norm", fps = norm_det_paths_genes)
gene_mat_meth <- build_gene_mat("avg_meth",     fps = det_paths_genes)

saveRDS(list(gene_mat_adjH = gene_mat_adjH, gene_mat_meth = gene_mat_meth),
        "ecker_gene_matrices.rds")
```

```{r gene_locus_umap_adjH, fig.width = 5, fig.height = 5}
meta_cols_grp <- c("sub_region", "sub_type", "cell_class", "major_type", "major_region")

umap_gene_adjH <- run_umap_wide(gene_mat_adjH, meta_cols_grp, n_neighbors = 5L)
plot_umap_class(umap_gene_adjH,
                title    = "Gene-locus UMAP - adjH features",
                subtitle = paste0(LOCUS_N_UMAP, " genes (coverage-based selection)"))
```

## Group-level: gene loci, avg_meth

```{r gene_locus_umap_meth, fig.width = 5, fig.height = 5}
umap_gene_meth <- run_umap_wide(gene_mat_meth, meta_cols_grp, n_neighbors = 5L)
plot_umap_class(umap_gene_meth,
                title    = "Gene-locus UMAP - avg. methylation features",
                subtitle = paste0(LOCUS_N_UMAP, " genes (coverage-based selection)"))
```

# Paper UMAP + our entropy overlay {.tabset .tabset-pills}

All cells from the original Ecker et al. metadata placed at their published UMAP coordinates (L1UMAP), colored by the group-level entropy or methylation value computed here. One panel per genomic annotation.

```{r paper_umap_data}
cells_overlay <- meta_cells %>%
  left_join(
    all_grp_meta %>%
      select(sub_region, sub_type, annotation, median_adjH, median_meth,
             median_adjS, cell_class, major_type),
    by      = c("sub_region", "sub_type"),
    relationship = "many-to-many"
  ) %>%
  filter(!is.na(umap1), !is.na(annotation))

saveRDS(cells_overlay, "ecker_paper_umap_overlay.rds")
cat("Rows in overlay table:", nrow(cells_overlay), "\n")
cat("Non-NA median_adjH:", sum(!is.na(cells_overlay$median_adjH)), "\n")
```

## adjH per annotation

```{r paper_umap_adjH, fig.width = 14, fig.height = 6}
df_adjH <- cells_overlay %>% filter(!is.na(median_adjH)) %>% droplevels()
if (nrow(df_adjH) > 0) {
  ggplot(df_adjH, aes(x = umap1, y = umap2, color = median_adjH)) +
    geom_point(size = 0.4, alpha = 0.6) +
    scale_color_viridis_c(option = "plasma", na.value = "grey80") +
    facet_wrap(~annotation, nrow = 2) +
    labs(title = "Paper UMAP coordinates - color = group adjH",
         x = "UMAP1 (Ecker et al.)", y = "UMAP2 (Ecker et al.)",
         color = "median adjH") +
    theme_ng() +
    theme(axis.text = element_blank(), axis.ticks = element_blank())
} else {
  message("No adjH overlay data available - check ecker_paper_umap_overlay.rds")
}
```

## avg_meth per annotation

```{r paper_umap_meth, fig.width = 14, fig.height = 6}
df_meth <- cells_overlay %>% filter(!is.na(median_meth)) %>% droplevels()
if (nrow(df_meth) > 0) {
  ggplot(df_meth, aes(x = umap1, y = umap2, color = median_meth)) +
    geom_point(size = 0.4, alpha = 0.6) +
    scale_color_viridis_c(option = "viridis", na.value = "grey80") +
    facet_wrap(~annotation, nrow = 2) +
    labs(title = "Paper UMAP coordinates - color = group avg. methylation",
         x = "UMAP1 (Ecker et al.)", y = "UMAP2 (Ecker et al.)",
         color = "median meth") +
    theme_ng() +
    theme(axis.text = element_blank(), axis.ticks = element_blank())
} else {
  message("No methylation overlay data available - check ecker_paper_umap_overlay.rds")
}
```

## Per-cell S and adjS per annotation

Unlike H/adjH (which are group medians broadcast to all cells in a group), S and
adjS are per-cell values. The `.out.gz` file column maps directly to cell basenames,
so each cell is colored by its own entropy value.

```{r paper_umap_sampen_load}
read_short_with_file <- function(fp) {
  fn <- basename(fp)
  dt <- read_tsv(fp, na = "-1", show_col_types = FALSE)
  dt <- dt[, c("file", "sampen", "sampen_norm", "avg_meth")]
  dt$annotation <- get_annotation(fn)
  dt
}

cells_sampen_long <- do.call(
  rbind,
  bplapply(short_paths, read_short_with_file, BPPARAM = param)
)

cells_sampen_umap <- cells_sampen_long %>%
  mutate(basename = basename(file)) %>%
  inner_join(meta_cells %>% select(basename, umap1, umap2),
             by = "basename") %>%
  filter(!is.na(umap1)) %>%
  harmonize_levels()

cat("Cells with UMAP coords and S values:", nrow(cells_sampen_umap), "\n")
cat("Non-NA adjS:", sum(!is.na(cells_sampen_umap$sampen_norm)), "\n")
```

```{r paper_umap_adjS, fig.width = 14, fig.height = 6}
df_adjS_umap <- cells_sampen_umap %>% filter(!is.na(sampen_norm)) %>% droplevels()
if (nrow(df_adjS_umap) > 0) {
  ggplot(df_adjS_umap, aes(x = umap1, y = umap2, color = sampen_norm)) +
    geom_point(size = 0.4, alpha = 0.6) +
    scale_color_viridis_c(option = "plasma", na.value = "grey80") +
    facet_wrap(~annotation, nrow = 2) +
    labs(title = "Paper UMAP coordinates - per-cell adjS",
         x = "UMAP1 (Ecker et al.)", y = "UMAP2 (Ecker et al.)",
         color = "adjS") +
    theme_ng() +
    theme(axis.text = element_blank(), axis.ticks = element_blank())
} else {
  message("No adjS overlay data available")
}
```

```{r paper_umap_S, fig.width = 14, fig.height = 6}
df_S_umap <- cells_sampen_umap %>% filter(!is.na(sampen)) %>% droplevels()
if (nrow(df_S_umap) > 0) {
  ggplot(df_S_umap, aes(x = umap1, y = umap2, color = sampen)) +
    geom_point(size = 0.4, alpha = 0.6) +
    scale_color_viridis_c(option = "magma", na.value = "grey80") +
    facet_wrap(~annotation, nrow = 2) +
    labs(title = "Paper UMAP coordinates - per-cell S",
         x = "UMAP1 (Ecker et al.)", y = "UMAP2 (Ecker et al.)",
         color = "S") +
    theme_ng() +
    theme(axis.text = element_blank(), axis.ticks = element_blank())
} else {
  message("No S overlay data available")
}
```

## Cell class reference

```{r paper_umap_cellclass, fig.width = 5, fig.height = 5}
meta_cells %>%
  filter(!is.na(umap1)) %>%
  distinct(basename, .keep_all = TRUE) %>%
  ggplot(aes(x = umap1, y = umap2, color = cell_class)) +
  geom_point(size = 0.5, alpha = 0.6) +
  scale_color_viridis_d(option = "turbo", begin = 0.05, end = 0.95) +
  labs(title = "Paper UMAP coordinates - cell class reference",
       x = "UMAP1 (Ecker et al.)", y = "UMAP2 (Ecker et al.)",
       color = "cell class") +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  theme_ng() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

# Session

```{r session, cache = FALSE}
sessionInfo()
```
