---
title: "Ecker: cell embeddings from genomic windows entropy"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
---

```{r logging_early, include = FALSE}
log <- file(snakemake@log$log, open = "wt")
sink(log)
sink(log, type = "message")
```

```{r setup}
suppressPackageStartupMessages({
  library(readr)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(BiocParallel)
  library(uwot)
  library(irlba)
  library(matrixStats)
  library(SingleCellExperiment)
  library(SummarizedExperiment)

  library(cluster)
  library(viridis)
  library(viridisLite)
  library(cowplot)
  library(ggrastr)
})

source("src/ggtheme.R")
source("src/palettes.R")
source("src/embedding_utils.R")

param <- SerialParam()

knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  cache = FALSE,
  fig.path = "ecker_embeddings_plots/"
)
```

```{r save_snmk_debug, include = FALSE}
saveRDS(snakemake, file = "snmk_ecker_embeddings.rds", compress = "xz")
```

```{r constants}
N_HVW <- 1000
N_PCS <- 10

assay_map <- c(
  "sampen"           = "S",
  "sampen_corrected" = "adjS",
  "adjS"             = "adjS_yamet",
  "meth"             = "methylation"
)
```

## Cell selection and NA-aware embedding

The cells embedded here are not a uniformly random subsample. For each
(SubRegion, SubType) annotation we keep up to a fixed number of cells, picking
the highest-coverage cells (file size on disk is monotonic in CpG count) and
distributing the slots round-robin across plates. Stratifying by plate first
keeps a single high-coverage plate from dominating an annotation. The visible
cost is that any residual plate effect (cell cycle, batch) is preserved rather
than averaged out across plates; the alternative of random sampling produced
embeddings collapsed to a handful of cells per group because of how few windows
were complete.

scBS-seq methylation and per-cell entropy values have a meaningful zero
(unmethylated CpG, zero local entropy) which is distinct from missing coverage.
Imputing NAs to zero would bias both the variance signal and the embedding
geometry, so the pipeline keeps NAs as NAs throughout: HVF selection scores
features by variance times coverage fraction, and when too few windows have
complete coverage the PCA step uses NIPALS (`pcaMethods`) which iterates over
the observed entries without filling in missing values. Cells are dropped only
when their per-cell NA fraction is above the QC threshold; once they pass that
filter they are never dropped to make a feature block dense.

# Load data

```{r load_sce}
sce <- readRDS(snakemake@params$corrected_sce)

## ensure clean factor columns
colData(sce)$cell_class  <- factor(colData(sce)$CellClass)
colData(sce)$major_type  <- factor(colData(sce)$MajorType)
colData(sce)$sub_region  <- factor(colData(sce)$SubRegion)

cc_lvls <- sort(unique(as.character(colData(sce)$cell_class)))

cat("Cells:", ncol(sce), "\n")
cat("Windows:", nrow(sce), "\n")
cat("Assays:", paste(assayNames(sce), collapse = ", "), "\n")
cat("Cell classes:\n")
print(table(colData(sce)$cell_class, useNA = "ifany"))
```

# Shared helpers

```{r helpers}
hvw_pca_umap <- function(mat) {
  run_embedding(mat, n_hvf = N_HVW, n_pcs = N_PCS,
                n_neighbors = 15L, min_dist = 0.3, seed = 42L)
}

plot_umap_cellclass <- function(coords, sce_obj, title_prefix) {
  df <- data.frame(
    umap1      = coords[, 1], umap2 = coords[, 2],
    cell_class = colData(sce_obj)$cell_class
  )
  ggplot(df, aes(x = umap1, y = umap2, color = cell_class)) +
    geom_point_rast(size = 0.3, alpha = 0.5, raster.dpi = 300) +
    scale_color_manual(values = ecker_cell_class_pal) +
    labs(title = paste(title_prefix, "- cell class"),
         x = "UMAP 1", y = "UMAP 2") +
    guides(color = guide_legend(override.aes = list(size = 2, alpha = 1), ncol = 2)) +
    theme_ng()
}

plot_umap_majortype <- function(coords, sce_obj, title_prefix) {
  df <- data.frame(
    umap1      = coords[, 1], umap2 = coords[, 2],
    major_type = colData(sce_obj)$major_type
  )
  ggplot(df, aes(x = umap1, y = umap2, color = major_type)) +
    geom_point_rast(size = 0.3, alpha = 0.5, raster.dpi = 300) +
    scale_color_manual(values = ecker_major_type_pal) +
    labs(title = paste(title_prefix, "- major type"),
         x = "UMAP 1", y = "UMAP 2") +
    guides(color = guide_legend(override.aes = list(size = 2, alpha = 1), ncol = 2)) +
    theme_ng()
}
```

# Per-window embeddings (HVW -> PCA -> UMAP)

HVW selection and NA handling: see `embedding_utils.R`.
Top `r N_HVW` most variable windows retained per assay before PCA and UMAP.

```{r win_hvw_pca_umap}
win_embeds <- lapply(names(assay_map), function(akey) {
  if (!akey %in% assayNames(sce)) return(NULL)
  hvw_pca_umap(as.matrix(assay(sce, akey)))
})
names(win_embeds) <- assay_map

failed <- names(which(sapply(win_embeds, is.null)))
if (length(failed) > 0)
  warning("run_embedding returned NULL for: ", paste(failed, collapse = ", "))

cat("\nHVF selection summary:\n")
for (nm in names(win_embeds)) {
  em <- win_embeds[[nm]]
  if (is.null(em)) {
    cat(nm, ": NULL\n")
  } else {
    cat(nm, ": HVF =", length(em$hvf_idx),
        "| cells =", sum(em$kept_cols), "/", ncol(sce), "\n")
  }
}

saveRDS(
  list(col_data   = as.data.frame(colData(sce)),
       sce_dim    = dim(sce),
       assay_names = assayNames(sce),
       win_embeds = win_embeds,
       assay_map  = assay_map),
  "ecker_umap_windows_adjS.rds", compress = "xz"
)
```

## By cell class {.tabset .tabset-pills}

```{r win_cc_S, fig.width = ng_fig_size(2, 1, panel_mm = 60)$w, fig.height = ng_fig_size(2, 1, panel_mm = 60)$h}
### S
if (!is.null(win_embeds[["S"]])) {
  p1 <- plot_umap_cellclass(win_embeds[["S"]]$umap,
                            sce[, win_embeds[["S"]]$kept_cols], "S")
  p2 <- plot_umap_majortype(win_embeds[["S"]]$umap,
                            sce[, win_embeds[["S"]]$kept_cols], "S")
  print(plot_grid(p1, p2, nrow = 1))
}
```

```{r win_cc_adjS, fig.width = ng_fig_size(2, 1, panel_mm = 60)$w, fig.height = ng_fig_size(2, 1, panel_mm = 60)$h}
### adjS
if (!is.null(win_embeds[["adjS"]])) {
  p1 <- plot_umap_cellclass(win_embeds[["adjS"]]$umap,
                            sce[, win_embeds[["adjS"]]$kept_cols], "adjS")
  p2 <- plot_umap_majortype(win_embeds[["adjS"]]$umap,
                            sce[, win_embeds[["adjS"]]$kept_cols], "adjS")
  print(plot_grid(p1, p2, nrow = 1))
}
```

```{r win_cc_meth, fig.width = ng_fig_size(2, 1, panel_mm = 60)$w, fig.height = ng_fig_size(2, 1, panel_mm = 60)$h}
### methylation
if (!is.null(win_embeds[["methylation"]])) {
  p1 <- plot_umap_cellclass(win_embeds[["methylation"]]$umap,
                            sce[, win_embeds[["methylation"]]$kept_cols], "methylation")
  p2 <- plot_umap_majortype(win_embeds[["methylation"]]$umap,
                            sce[, win_embeds[["methylation"]]$kept_cols], "methylation")
  print(plot_grid(p1, p2, nrow = 1))
}
```

## Four-panel comparison (cell class)

```{r win_panel_cc, fig.width = ng_fig_size(4, 1, panel_mm = 55)$w, fig.height = ng_fig_size(4, 1, panel_mm = 55)$h}
ok_embeds <- Filter(Negate(is.null), win_embeds)
df_panel <- bind_rows(lapply(names(ok_embeds), function(lbl) {
  em <- ok_embeds[[lbl]]
  data.frame(umap1      = em$umap[, 1], umap2 = em$umap[, 2],
             cell_class = colData(sce[, em$kept_cols])$cell_class,
             assay      = lbl)
}))
df_panel$assay <- factor(df_panel$assay, levels = assay_map)

ggplot(df_panel, aes(x = umap1, y = umap2, color = cell_class)) +
  geom_point_rast(size = 0.2, alpha = 0.5, raster.dpi = 300) +
  scale_color_manual(values = ecker_cell_class_pal) +
  facet_wrap(~ assay, nrow = 1) +
  labs(x = "UMAP 1", y = "UMAP 2",
       title = "Per-window embeddings (HVW -> PCA -> UMAP)") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1), ncol = 2)) +
  theme_ng()
```

## Variance explained

```{r win_varexp, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(2, 1)$h}
ve_win <- bind_rows(lapply(names(assay_map), function(akey) {
  lbl <- assay_map[[akey]]
  if (is.null(win_embeds[[lbl]]) || !akey %in% assayNames(sce)) return(NULL)
  kept <- win_embeds[[lbl]]$kept_cols
  hvw_mat <- as.matrix(assay(sce, akey))[win_embeds[[lbl]]$hvf_idx, kept, drop = FALSE]
  data.frame(
    assay      = lbl,
    cell_class = median(row_variance_explained(hvw_mat, colData(sce[, kept])$cell_class), na.rm = TRUE),
    sub_region = median(row_variance_explained(hvw_mat, colData(sce[, kept])$sub_region), na.rm = TRUE))
}))

ve_long <- pivot_longer(ve_win, c("cell_class", "sub_region"),
                        names_to = "variable", values_to = "median_ve")
ve_long$assay <- factor(ve_long$assay, levels = assay_map)

write.csv(ve_long, "ecker_win_varexp.csv", row.names = FALSE)

ggplot(ve_long, aes(x = assay, y = median_ve, fill = variable)) +
  geom_col(position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = c("cell_class" = "#0072B2", "sub_region" = "#E69F00")) +
  labs(x = NULL, y = "median % variance explained",
       title = "Variance explained (per-window HVW)") +
  theme_ng()
```

## Embedding diagnostics

Why does per-window variance explained stay low? Candidates are high NA
fractions per window, few HVFs surviving QC, cells dropped by the sparse
fallback, and per-HVF variance being low in absolute terms. The diagnostic
below logs these quantities per assay to `ecker_embedding_diagnostics.csv`
and a compact text summary.

```{r embedding_diagnostic}
diag_rows <- list()

na_stats <- function(mat) {
  na_frac_row <- rowMeans(is.na(mat))
  na_frac_col <- colMeans(is.na(mat))
  list(
    row_na_median = median(na_frac_row, na.rm = TRUE),
    row_na_p90    = quantile(na_frac_row, 0.9, na.rm = TRUE),
    col_na_median = median(na_frac_col, na.rm = TRUE),
    col_na_p90    = quantile(na_frac_col, 0.9, na.rm = TRUE),
    n_complete_rows = sum(na_frac_row == 0, na.rm = TRUE)
  )
}

for (akey in names(assay_map)) {
  lbl <- assay_map[[akey]]
  if (!akey %in% assayNames(sce)) next
  em  <- win_embeds[[lbl]]
  if (is.null(em)) next

  mat  <- as.matrix(assay(sce, akey))
  ns   <- na_stats(mat)

  hvf_mat <- mat[em$hvf_idx, em$kept_cols, drop = FALSE]
  hvf_row_var <- matrixStats::rowVars(hvf_mat, na.rm = TRUE)

  diag_rows[[lbl]] <- data.frame(
    assay                  = lbl,
    stage                  = em$stage,
    n_cells_total          = ncol(sce),
    n_cells_used           = sum(em$kept_cols),
    cells_dropped_fraction = 1 - sum(em$kept_cols) / ncol(sce),
    n_windows_total        = nrow(sce),
    n_complete_rows        = ns$n_complete_rows,
    row_na_median          = ns$row_na_median,
    row_na_p90             = as.numeric(ns$row_na_p90),
    col_na_median          = ns$col_na_median,
    col_na_p90             = as.numeric(ns$col_na_p90),
    n_hvf                  = length(em$hvf_idx),
    hvf_var_median         = median(hvf_row_var, na.rm = TRUE),
    hvf_var_q25            = quantile(hvf_row_var, 0.25, na.rm = TRUE),
    hvf_var_q75            = quantile(hvf_row_var, 0.75, na.rm = TRUE),
    stringsAsFactors       = FALSE
  )
}

diag_df <- bind_rows(diag_rows)
write.csv(diag_df, "ecker_embedding_diagnostics.csv", row.names = FALSE)

diag_lines <- c(
  "Ecker embedding diagnostic summary",
  paste0("Total cells in SCE: ", ncol(sce)),
  paste0("Total windows in SCE: ", nrow(sce)),
  ""
)
for (lbl in diag_df$assay) {
  r <- diag_df[diag_df$assay == lbl, ]
  diag_lines <- c(diag_lines, sprintf(
    "[%s] stage=%s | cells used=%d/%d (%.1f%% dropped) | HVF=%d | complete windows=%d | row NA median=%.3f | HVF row variance median=%.4f",
    lbl, r$stage,
    r$n_cells_used, r$n_cells_total,
    100 * r$cells_dropped_fraction,
    r$n_hvf, r$n_complete_rows, r$row_na_median, r$hvf_var_median
  ))
}
writeLines(diag_lines, "ecker_embedding_diagnostics.txt")

cat(paste(diag_lines, collapse = "\n"), "\n")
```

## Silhouette scores

```{r win_sil, fig.width = ng_fig_size(2, 1)$w, fig.height = ng_fig_size(1, 1)$h}
sil_win <- data.frame(
  assay = names(ok_embeds),
  silhouette_cell_class = sapply(names(ok_embeds), function(lbl)
    sil_score(ok_embeds[[lbl]]$umap,
              colData(sce[, ok_embeds[[lbl]]$kept_cols])$cell_class)),
  silhouette_sub_region = sapply(names(ok_embeds), function(lbl)
    sil_score(ok_embeds[[lbl]]$umap,
              colData(sce[, ok_embeds[[lbl]]$kept_cols])$sub_region))
)

sil_long <- pivot_longer(sil_win, c("silhouette_cell_class", "silhouette_sub_region"),
                         names_to = "grouping", values_to = "silhouette")
sil_long$grouping <- sub("silhouette_", "", sil_long$grouping)
sil_long$assay    <- factor(sil_long$assay, levels = assay_map)

ggplot(sil_long, aes(x = assay, y = silhouette, fill = grouping)) +
  geom_col(position = position_dodge(width = 0.8)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c("cell_class" = "#0072B2", "sub_region" = "#E69F00")) +
  labs(x = NULL, y = "mean silhouette",
       title = "Cluster separation by assay") +
  theme_ng()
```

# Per-cell entropy summaries

```{r cell_df}
cell_df <- data.frame(
  cell_class = colData(sce)$cell_class,
  sub_region = colData(sce)$sub_region,
  mean_S     = colMeans(assay(sce, "sampen"),           na.rm = TRUE),
  mean_adjS  = colMeans(assay(sce, "sampen_corrected"), na.rm = TRUE),
  mean_meth  = colMeans(assay(sce, "meth"),             na.rm = TRUE)
)
```

```{r violins, fig.width = ng_fig_size(4, 1)$w, fig.height = ng_fig_size(4, 2)$h}
cell_long <- pivot_longer(cell_df, c("mean_S", "mean_adjS", "mean_meth"),
                          names_to = "metric", values_to = "value")

ggplot(cell_long, aes(x = cell_class, y = value, fill = cell_class)) +
  geom_violin(trim = FALSE, scale = "width", alpha = 0.7) +
  geom_boxplot(width = 0.1, outlier.size = 0.3, alpha = 0.6) +
  scale_fill_manual(values = ecker_cell_class_pal) +
  facet_grid(metric ~ ., scales = "free_y") +
  labs(x = "cell class", y = "per-cell mean") +
  theme_ng() +
  theme(legend.position = "none")
```

```{r session}
sessionInfo()
```
