---
title: "Ecker 2021 mouse brain: methylation entropy by feature (CpG context)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Izaskun Mallona"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: default
    theme: lumen
---

```{r setup}
suppressPackageStartupMessages({
  library(readr)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(BiocParallel)
  library(ggstance)
  library(ggrepel)
  library(ggridges)
})

source("src/ggtheme.R")
param <- MulticoreParam(workers = snakemake@threads)
```

```{r opts}
knitr::opts_chunk$set(
  echo       = TRUE,
  fig.width  = 7.5,
  fig.height = 7.5,
  cache      = FALSE,
  include    = TRUE,
  fig.path   = "ecker_plots/",
  dev        = c("png", "svg"),
  dpi        = 500,
  cache.lazy = FALSE,
  warning    = TRUE,
  message    = TRUE
)
```

```{r save_snmk_debug, include = FALSE}
saveRDS(file = "snmk_ecker_features.rds", object = snakemake)
```

```{r logging, include = FALSE}
log <- file(snakemake@log$log, open = "wt")
sink(log)
sink(log, type = "message")
```

Mouse (mm10), CpG context only. chr10 only: `r snakemake@params[["chr10_only"]]`.

```{r helpers}
## filenames: {annotation}_{sub_region}_{sub_type}.*.gz
## annotation has no underscores; neither do sub_region or sub_type values
get_annotation   <- function(fn) sub("^(.*)_[^_]+_[^_]+\\..*$", "\\1", fn)
get_sub_region <- function(fn) sub("^.*_([^_]+)_[^_]+\\..*$", "\\1", fn)
get_sub_type   <- function(fn) sub("^.*_([^_]+)\\..*$",        "\\1", fn)

ann_labels <- c(
  "genes"    = "Genes",
  "promoters"= "Promoters",
  "lines"    = "LINEs",
  "sines"    = "SINEs",
  "h3k4me3"  = "H3K4me3",
  "h3k9me3"  = "H3K9me3",
  "h3k27me3" = "H3K27me3",
  "h3k4me1"  = "H3K4me1",
  "h3k27ac"  = "H3K27ac"
)

harmonize_levels <- function(df) {
  missing_ann <- setdiff(unique(df$annotation), names(ann_labels))
  if (length(missing_ann) > 0)
    warning("unknown annotations: ", paste(missing_ann, collapse = ", "))
  df$annotation   <- factor(df$annotation,
                             levels = names(ann_labels),
                             labels = unname(ann_labels))
  df$sub_region <- factor(df$sub_region)
  df$sub_type   <- factor(df$sub_type)
  df
}
```

# Unnormalized sampEn (S) {.tabset .tabset-pills}

Per-cell sample entropy (S) from `.out.gz` files.

```{r import_sampen}
short_paths <- list.files(snakemake@params[["output_path"]],
                           pattern = "\\.out\\.gz$", full.names = TRUE)
short_paths <- short_paths[!grepl("\\.det\\.out\\.gz$",  short_paths)]
short_paths <- short_paths[!grepl("\\.meth\\.out\\.gz$", short_paths)]

read_short <- function(fp) {
  fn <- basename(fp)
  dt <- read_tsv(fp, na = "-1", show_col_types = FALSE)
  dt <- dt[, c("sampen", "avg_meth")]
  dt$annotation   <- get_annotation(fn)
  dt$sub_region <- get_sub_region(fn)
  dt$sub_type   <- get_sub_type(fn)
  dt
}

sampens <- do.call(rbind, bplapply(short_paths, read_short, BPPARAM = param))
sampens <- harmonize_levels(sampens)
head(sampens)
```

## By cell class

```{r scatter_sampen_class}
ggplot(sampens, aes(x = avg_meth, y = sampen, color = sub_type)) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~annotation) +
  labs(x = "average methylation", y = "sample entropy (S)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

```{r boxes_sampen_class, fig.width = 10, fig.height = 6}
ggplot(sampens, aes(x = annotation, y = sampen, color = sub_type)) +
  geom_boxplot(outlier.size = 0.6, outlier.alpha = 0.3) +
  scale_color_brewer(palette = "Dark2") +
  labs(y = "sample entropy (S)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

```{r violin_sampen_class, fig.width = 12, fig.height = 10}
ggplot(sampens, aes(x = sub_type, y = sampen, fill = sub_type)) +
  geom_violin(alpha = 0.5, draw_quantiles = 0.5) +
  geom_jitter(width = 0.2, size = 0.25, alpha = 0.1) +
  facet_wrap(~annotation, scales = "free_y") +
  labs(y = "sample entropy (S)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        legend.position = "none")
```

```{r ridges_sampen_class, fig.width = 12, fig.height = 10}
ggplot(sampens, aes(x = sampen, y = sub_type, fill = sub_type)) +
  geom_density_ridges(alpha = 0.6, scale = 0.85) +
  facet_wrap(~annotation, scales = "free_x") +
  labs(x = "sample entropy (S)") +
  theme_ng() +
  theme(legend.position = "none")
```

## By major region

```{r scatter_sampen_region}
ggplot(sampens, aes(x = avg_meth, y = sampen, color = sub_region)) +
  geom_point(size = 0.9, alpha = 0.2) +
  facet_wrap(~annotation) +
  labs(x = "average methylation", y = "sample entropy (S)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

```{r boxes_sampen_region, fig.width = 10, fig.height = 6}
ggplot(sampens, aes(x = annotation, y = sampen, color = sub_region)) +
  geom_boxplot(outlier.size = 0.6, outlier.alpha = 0.3) +
  labs(y = "sample entropy (S)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## Region by class

```{r boxes_sampen_combo, fig.width = 12, fig.height = 8}
ggplot(sampens, aes(x = sub_type, y = sampen, color = sub_region)) +
  geom_boxplot(outlier.size = 0.6, outlier.alpha = 0.3) +
  facet_wrap(~annotation) +
  labs(y = "sample entropy (S)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## Range plots

```{r ranges_sampen, fig.width = 10, fig.height = 10}
ranges_s <- sampens %>%
  group_by(compartment = annotation, sub_region, sub_type) %>%
  summarise(
    sampen_mean = mean(sampen,   na.rm = TRUE),
    sampen_min  = min(sampen,    na.rm = TRUE),
    sampen_max  = max(sampen,    na.rm = TRUE),
    meth_mean   = mean(avg_meth, na.rm = TRUE),
    meth_min    = min(avg_meth,  na.rm = TRUE),
    meth_max    = max(avg_meth,  na.rm = TRUE),
    .groups = "drop"
  )

ggplot(ranges_s, aes(color = sub_type, shape = sub_region)) +
  geom_segment(aes(x = meth_min,  xend = meth_max,
                   y = sampen_mean, yend = sampen_mean),
               linewidth = 0.6) +
  geom_segment(aes(x = meth_mean, xend = meth_mean,
                   y = sampen_min, yend = sampen_max),
               linewidth = 0.6) +
  geom_point(aes(x = meth_mean, y = sampen_mean), size = 1.2) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "DNA methylation", y = "sample entropy (S)") +
  facet_wrap(~compartment) +
  theme_ng()
```

# adjS (methylation-normalised sample entropy) {.tabset .tabset-pills}

Per-cell normalised sample entropy from `.out.gz` files.

```{r import_adjS}
norm_paths <- list.files(snakemake@params[["output_path"]],
                          pattern = "out\\.gz$", full.names = TRUE)
norm_paths <- norm_paths[!grepl("\\.det\\.out\\.gz$",  norm_paths)]
norm_paths <- norm_paths[!grepl("\\.norm\\.out\\.gz$", norm_paths)]
norm_paths <- norm_paths[!grepl("\\.meth\\.out\\.gz$", norm_paths)]

read_adjS <- function(fp) {
  fn <- basename(fp)
  dt <- read_tsv(fp, na = "-1", show_col_types = FALSE)
  dt <- dt[, c("sampen_norm", "avg_meth")]
  colnames(dt) <- c("sampen", "avg_meth")
  dt$annotation   <- get_annotation(fn)
  dt$sub_region <- get_sub_region(fn)
  dt$sub_type   <- get_sub_type(fn)
  dt
}

adjsampens <- do.call(rbind, bplapply(norm_paths, read_adjS, BPPARAM = param))
adjsampens <- harmonize_levels(adjsampens)
head(adjsampens)
```

## By cell class

```{r scatter_adjS_class}
ggplot(adjsampens, aes(x = avg_meth, y = sampen, color = sub_type)) +
  geom_point(size = 0.9, alpha = 0.2) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~annotation) +
  labs(x = "average methylation", y = "adj. sample entropy (adjS)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

```{r boxes_adjS_class, fig.width = 10, fig.height = 6}
ggplot(adjsampens, aes(x = annotation, y = sampen, color = sub_type)) +
  geom_boxplot(outlier.size = 0.6, outlier.alpha = 0.3) +
  scale_color_brewer(palette = "Dark2") +
  labs(y = "adj. sample entropy (adjS)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

```{r violin_adjS_class, fig.width = 12, fig.height = 10}
ggplot(adjsampens, aes(x = sub_type, y = sampen, fill = sub_type)) +
  geom_violin(alpha = 0.5, draw_quantiles = 0.5) +
  geom_jitter(width = 0.2, size = 0.25, alpha = 0.1) +
  facet_wrap(~annotation, scales = "free_y") +
  labs(y = "adj. sample entropy (adjS)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        legend.position = "none")
```

```{r ridges_adjS_class, fig.width = 12, fig.height = 10}
ggplot(adjsampens, aes(x = sampen, y = sub_type, fill = sub_type)) +
  geom_density_ridges(alpha = 0.6, scale = 0.85) +
  facet_wrap(~annotation, scales = "free_x") +
  labs(x = "adj. sample entropy (adjS)") +
  theme_ng() +
  theme(legend.position = "none")
```

## By major region

```{r scatter_adjS_region}
ggplot(adjsampens, aes(x = avg_meth, y = sampen, color = sub_region)) +
  geom_point(size = 0.9, alpha = 0.2) +
  facet_wrap(~annotation) +
  labs(x = "average methylation", y = "adj. sample entropy (adjS)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

```{r boxes_adjS_region, fig.width = 10, fig.height = 6}
ggplot(adjsampens, aes(x = annotation, y = sampen, color = sub_region)) +
  geom_boxplot(outlier.size = 0.6, outlier.alpha = 0.3) +
  labs(y = "adj. sample entropy (adjS)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## Region by class

```{r boxes_adjS_combo, fig.width = 12, fig.height = 8}
ggplot(adjsampens, aes(x = sub_type, y = sampen, color = sub_region)) +
  geom_boxplot(outlier.size = 0.6, outlier.alpha = 0.3) +
  facet_wrap(~annotation) +
  labs(y = "adj. sample entropy (adjS)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## Range plots

```{r ranges_adjS, fig.width = 10, fig.height = 10}
ranges_adjs <- adjsampens %>%
  group_by(compartment = annotation, sub_region, sub_type) %>%
  summarise(
    sampen_mean = mean(sampen,   na.rm = TRUE),
    sampen_min  = min(sampen,    na.rm = TRUE),
    sampen_max  = max(sampen,    na.rm = TRUE),
    meth_mean   = mean(avg_meth, na.rm = TRUE),
    meth_min    = min(avg_meth,  na.rm = TRUE),
    meth_max    = max(avg_meth,  na.rm = TRUE),
    .groups = "drop"
  )

ggplot(ranges_adjs, aes(color = sub_type, shape = sub_region)) +
  geom_segment(aes(x = meth_min,  xend = meth_max,
                   y = sampen_mean, yend = sampen_mean),
               linewidth = 0.6) +
  geom_segment(aes(x = meth_mean, xend = meth_mean,
                   y = sampen_min, yend = sampen_max),
               linewidth = 0.6) +
  geom_point(aes(x = meth_mean, y = sampen_mean), size = 1.2) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "DNA methylation", y = "adj. sample entropy (adjS)") +
  facet_wrap(~compartment) +
  theme_ng()
```

# Shannon H - medians {.tabset .tabset-pills}

Per-locus Shannon entropy (H) from `.det.out.gz`. One median per group per annotation; loci downsampled to 10e4 rows.

```{r import_shannon}
det_paths <- list.files(snakemake@params[["output_path"]],
                         pattern = "det\\.out\\.gz$", full.names = TRUE)
det_paths <- det_paths[!grepl("norm", det_paths)]

read_shannon <- function(fp) {
  fn  <- basename(fp)
  tmp <- read_tsv(fp, na = "-1", show_col_types = FALSE, n_max = 10e4)
  tibble(
    file            = fn,
    median_shannon  = median(tmp$shannon,  na.rm = TRUE),
    median_avg_meth = median(tmp$avg_meth, na.rm = TRUE),
    annotation      = get_annotation(fn),
    sub_region    = get_sub_region(fn),
    sub_type      = get_sub_type(fn)
  )
}

shannons <- bplapply(det_paths, read_shannon, BPPARAM = param) |> bind_rows()
shannons <- harmonize_levels(shannons)
head(shannons)
```

## By cell class

```{r scatter_shannon_class}
ggplot(shannons, aes(x = median_avg_meth, y = median_shannon, color = sub_type)) +
  geom_point(size = 1.5, alpha = 0.8) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~annotation) +
  labs(x = "median avg. methylation", y = "median Shannon entropy (H)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

```{r boxes_shannon_class, fig.width = 10, fig.height = 6}
ggplot(shannons, aes(x = annotation, y = median_shannon, color = sub_type)) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  scale_color_brewer(palette = "Dark2") +
  labs(y = "median Shannon entropy (H)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## By major region

```{r scatter_shannon_region}
ggplot(shannons, aes(x = median_avg_meth, y = median_shannon, color = sub_region)) +
  geom_point(size = 1.5, alpha = 0.8) +
  facet_wrap(~annotation) +
  labs(x = "median avg. methylation", y = "median Shannon entropy (H)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

```{r boxes_shannon_region, fig.width = 10, fig.height = 6}
ggplot(shannons, aes(x = annotation, y = median_shannon, color = sub_region)) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  labs(y = "median Shannon entropy (H)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## Region by class

```{r boxes_shannon_combo, fig.width = 12, fig.height = 8}
ggplot(shannons, aes(x = sub_type, y = median_shannon, color = sub_region)) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  facet_wrap(~annotation) +
  labs(y = "median Shannon entropy (H)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

# adjH (methylation-normalised Shannon entropy) - medians {.tabset .tabset-pills}

Per-locus normalised Shannon entropy from `.norm.det.out.gz`. Same downsampling as above.

```{r import_adjH}
norm_det_paths <- list.files(snakemake@params[["output_path"]],
                              pattern = "norm\\.det\\.out\\.gz$", full.names = TRUE)

read_adjH <- function(fp) {
  fn  <- basename(fp)
  tmp <- read_tsv(fp, na = "-1", show_col_types = FALSE, n_max = 10e4)
  tibble(
    file            = fn,
    median_shannon  = median(tmp$shannon_norm, na.rm = TRUE),
    median_avg_meth = median(tmp$avg_meth,     na.rm = TRUE),
    annotation      = get_annotation(fn),
    sub_region    = get_sub_region(fn),
    sub_type      = get_sub_type(fn)
  )
}

adjshannons <- bplapply(norm_det_paths, read_adjH, BPPARAM = param) |> bind_rows()
adjshannons <- harmonize_levels(adjshannons)
head(adjshannons)
```

## By cell class

```{r scatter_adjH_class}
ggplot(adjshannons, aes(x = median_avg_meth, y = median_shannon, color = sub_type)) +
  geom_point(size = 1.5, alpha = 0.8) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~annotation) +
  labs(x = "median avg. methylation", y = "median adj. Shannon entropy (adjH)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

```{r boxes_adjH_class, fig.width = 10, fig.height = 6}
ggplot(adjshannons, aes(x = annotation, y = median_shannon, color = sub_type)) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  scale_color_brewer(palette = "Dark2") +
  labs(y = "median adj. Shannon entropy (adjH)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## By major region

```{r scatter_adjH_region}
ggplot(adjshannons, aes(x = median_avg_meth, y = median_shannon, color = sub_region)) +
  geom_point(size = 1.5, alpha = 0.8) +
  facet_wrap(~annotation) +
  labs(x = "median avg. methylation", y = "median adj. Shannon entropy (adjH)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

```{r boxes_adjH_region, fig.width = 10, fig.height = 6}
ggplot(adjshannons, aes(x = annotation, y = median_shannon, color = sub_region)) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  labs(y = "median adj. Shannon entropy (adjH)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## Region by class

```{r boxes_adjH_combo, fig.width = 12, fig.height = 8}
ggplot(adjshannons, aes(x = sub_type, y = median_shannon, color = sub_region)) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  facet_wrap(~annotation) +
  labs(y = "median adj. Shannon entropy (adjH)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

# Assembly

S and adjS are per-cell values. H and adjH are medians across loci for each group.

```{r str_check}
str(sampens)
str(adjsampens)
str(shannons)
str(adjshannons)
```

```{r save_rds}
saveRDS(
  list(sampens     = sampens,
       adjsampens  = adjsampens,
       shannons    = shannons,
       adjshannons = adjshannons),
  file = "ecker_entropy.rds"
)
```

Aggregate per-cell S and adjS to per-group medians so they are comparable to H and adjH.

```{r group_medians}
sampen_grp <- sampens %>%
  group_by(annotation, sub_region, sub_type) %>%
  summarise(median_S    = median(sampen,   na.rm = TRUE),
            median_meth = median(avg_meth, na.rm = TRUE),
            .groups = "drop")

adjS_grp <- adjsampens %>%
  group_by(annotation, sub_region, sub_type) %>%
  summarise(median_adjS = median(sampen, na.rm = TRUE),
            .groups = "drop")

all_grp <- sampen_grp %>%
  left_join(adjS_grp,    by = c("annotation", "sub_region", "sub_type")) %>%
  left_join(shannons  %>% select(annotation, sub_region, sub_type,
                                  median_H    = median_shannon),
            by = c("annotation", "sub_region", "sub_type")) %>%
  left_join(adjshannons %>% select(annotation, sub_region, sub_type,
                                    median_adjH = median_shannon),
            by = c("annotation", "sub_region", "sub_type"))

head(all_grp)
```

```{r load_metadata}
## Load the full cell-level metadata to get CellClass, MajorType, and the
## paper's own UMAP coordinates (L1UMAP_0/1).
meta_full <- read_tsv(snakemake@params[["meta_path"]],
                       na = "NA", show_col_types = FALSE) %>%
  mutate(
    sub_region = gsub(" ", "-", SubRegion),
    sub_type   = gsub(" ", "-", SubType)
  )

## Per-group summary: one row per (sub_region, sub_type)
meta_grp <- meta_full %>%
  group_by(sub_region, sub_type) %>%
  summarise(
    cell_class  = first(na.omit(CellClass)),
    major_type  = first(na.omit(MajorType)),
    umap1       = median(L1UMAP_0, na.rm = TRUE),
    umap2       = median(L1UMAP_1, na.rm = TRUE),
    n_meta      = n(),
    .groups     = "drop"
  )

## Attach CellClass, MajorType, and UMAP coords to the group-level entropy table.
## Convert sub_region/sub_type to character first to avoid factor-join issues.
all_grp_meta <- all_grp %>%
  mutate(sub_region = as.character(sub_region),
         sub_type   = as.character(sub_type)) %>%
  left_join(meta_grp, by = c("sub_region", "sub_type"))

## Individual-cell table with paper UMAP coords, for cell-level overlay plots
meta_cells <- meta_full %>%
  select(sub_region, sub_type, cell_class = CellClass, major_type = MajorType,
         umap1 = L1UMAP_0, umap2 = L1UMAP_1, basename)

saveRDS(
  list(all_grp      = all_grp,
       meta_grp     = meta_grp,
       all_grp_meta = all_grp_meta,
       meta_cells   = meta_cells),
  file = "ecker_groups_meta.rds"
)

str(all_grp_meta)
table(all_grp_meta$cell_class, useNA = "always")
```

## adjS vs adjH

```{r adjS_vs_adjH, fig.height = 10, fig.width = 8}
ggplot(all_grp, aes(x = median_adjH, y = median_adjS,
                     color = sub_type, shape = sub_region)) +
  geom_point(size = 1.5) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~annotation, ncol = 3) +
  labs(x = "median adjH", y = "median adjS") +
  theme_ng()
```

## S vs H

```{r S_vs_H, fig.height = 10, fig.width = 8}
ggplot(all_grp, aes(x = median_H, y = median_S,
                     color = sub_type, shape = sub_region)) +
  geom_point(size = 1.5) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~annotation, ncol = 3) +
  labs(x = "median H", y = "median S") +
  theme_ng()
```

# Variation {.tabset .tabset-pills}

Range and SD of group medians for each metric, broken down by major region, cell class, and their combination.

```{r variation_helpers}
## diff(range()) returns -Inf when all values are NA; return NA instead.
safe_range <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 2) return(NA_real_)
  diff(range(x))
}
```

## Across cell classes (per region and annotation)

How much do entropies differ between cell classes within the same region?

```{r variation_class, fig.width = 10, fig.height = 8}
var_class <- all_grp %>%
  group_by(annotation, sub_region) %>%
  summarise(
    S_range    = safe_range(median_S),
    adjS_range = safe_range(median_adjS),
    H_range    = safe_range(median_H),
    adjH_range = safe_range(median_adjH),
    S_sd       = sd(median_S,    na.rm = TRUE),
    adjS_sd    = sd(median_adjS, na.rm = TRUE),
    H_sd       = sd(median_H,    na.rm = TRUE),
    adjH_sd    = sd(median_adjH, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(var_class, aes(x = adjH_range, y = adjS_range,
                       color = sub_region, label = annotation)) +
  geom_point(size = 2) +
  geom_text_repel(size = 3) +
  labs(x = "adjH range across cell classes",
       y = "adjS range across cell classes",
       title = "Variation across cell classes (per region)") +
  theme_ng()

ggplot(var_class, aes(x = H_range, y = S_range,
                       color = sub_region, label = annotation)) +
  geom_point(size = 2) +
  geom_text_repel(size = 3) +
  labs(x = "H range across cell classes",
       y = "S range across cell classes",
       title = "Variation across cell classes (per region), unnormalised") +
  theme_ng()
```

## Across major regions (per cell class and annotation)

How much do entropies differ between major regions within the same cell class?

```{r variation_region, fig.width = 10, fig.height = 8}
var_region <- all_grp %>%
  group_by(annotation, sub_type) %>%
  summarise(
    S_range    = safe_range(median_S),
    adjS_range = safe_range(median_adjS),
    H_range    = safe_range(median_H),
    adjH_range = safe_range(median_adjH),
    S_sd       = sd(median_S,    na.rm = TRUE),
    adjS_sd    = sd(median_adjS, na.rm = TRUE),
    H_sd       = sd(median_H,    na.rm = TRUE),
    adjH_sd    = sd(median_adjH, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(var_region, aes(x = adjH_range, y = adjS_range,
                        color = sub_type, label = annotation)) +
  geom_point(size = 2) +
  geom_text_repel(size = 3) +
  labs(x = "adjH range across major regions",
       y = "adjS range across major regions",
       title = "Variation across major regions (per cell class)") +
  theme_ng()

ggplot(var_region, aes(x = H_range, y = S_range,
                        color = sub_type, label = annotation)) +
  geom_point(size = 2) +
  geom_text_repel(size = 3) +
  labs(x = "H range across major regions",
       y = "S range across major regions",
       title = "Variation across major regions (per cell class), unnormalised") +
  theme_ng()
```

## Across all groups (annotation-level)

Overall variation collapsing region and class together.

```{r variation_all, fig.width = 8, fig.height = 6}
var_all <- all_grp %>%
  group_by(annotation) %>%
  summarise(
    S_range    = safe_range(median_S),
    adjS_range = safe_range(median_adjS),
    H_range    = safe_range(median_H),
    adjH_range = safe_range(median_adjH),
    S_sd       = sd(median_S,    na.rm = TRUE),
    adjS_sd    = sd(median_adjS, na.rm = TRUE),
    H_sd       = sd(median_H,    na.rm = TRUE),
    adjH_sd    = sd(median_adjH, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(var_all, aes(x = adjH_range, y = adjS_range, label = annotation)) +
  geom_point(size = 3) +
  geom_text_repel() +
  labs(x = "adjH range across all groups",
       y = "adjS range across all groups",
       title = "Overall variation (region by class)") +
  theme_ng()
```

## Region vs class: which drives more variation?

Mean range across groups for cell class and major region, per metric.

```{r class_vs_region_comparison, fig.width = 10, fig.height = 6}
mean_var <- bind_rows(
  var_class  %>% summarise(across(ends_with("_range"), \(x) mean(x, na.rm = TRUE))) %>%
    mutate(driver = "cell class"),
  var_region %>% summarise(across(ends_with("_range"), \(x) mean(x, na.rm = TRUE))) %>%
    mutate(driver = "major region")
)

mean_var_long <- mean_var %>%
  tidyr::pivot_longer(-driver, names_to = "metric", values_to = "mean_range") %>%
  mutate(metric = sub("_range$", "", metric))

ggplot(mean_var_long, aes(x = metric, y = mean_range, fill = driver)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = "mean range across groups",
       title = "Average entropy range: cell class vs major region") +
  theme_ng()
```

# Driver categorization {.tabset .tabset-pills}

Annotations are labelled as within-cell (adjS-driven), across-cell (adjH-driven),
or both, based on overall variation across all region by class groups.

```{r drivers_range}
shannon_thr <- quantile(var_all$adjH_range, 0.5, na.rm = TRUE)
sampen_thr  <- quantile(var_all$adjS_range, 0.5, na.rm = TRUE)

var_all <- var_all %>%
  mutate(
    driver = case_when(
      is.na(adjH_range) | is.na(adjS_range)               ~ NA_character_,
      adjH_range < shannon_thr & adjS_range < sampen_thr   ~ "not changing enough",
      adjH_range >= adjS_range * 1.5                       ~ "across-cell driven",
      adjS_range >= adjH_range * 1.5                       ~ "within-cell driven",
      TRUE                                                  ~ "both"
    )
  )

table(var_all$driver)

ggplot(var_all,
       aes(x = adjH_range, y = adjS_range,
           color = driver, shape = driver, label = annotation)) +
  geom_point(size = 3) +
  geom_text_repel() +
  labs(x = "adjH range across all groups",
       y = "adjS range across all groups") +
  theme_ng()
```

```{r drivers_sd}
shannon_sd_thr <- quantile(var_all$adjH_sd, 0.3, na.rm = TRUE)
sampen_sd_thr  <- quantile(var_all$adjS_sd, 0.3, na.rm = TRUE)

var_all <- var_all %>%
  mutate(
    driver_sd = case_when(
      adjH_sd < shannon_sd_thr & adjS_sd < sampen_sd_thr ~ "not changing enough",
      adjH_sd >= adjS_sd * 1.5                            ~ "across-cell driven",
      adjS_sd >= adjH_sd * 1.5                            ~ "within-cell driven",
      TRUE                                                 ~ "both"
    )
  )

table(var_all$driver_sd)

ggplot(var_all,
       aes(x = adjH_sd, y = adjS_sd,
           color = driver_sd, shape = driver_sd, label = annotation)) +
  geom_point(size = 3) +
  geom_text_repel() +
  labs(x = "SD of adjH across all groups",
       y = "SD of adjS across all groups") +
  theme_ng()
```

# Heatmaps of median entropy per group {.tabset .tabset-pills}

Median entropy per annotation and sub-type combination. Useful for spotting which annotations show the most consistent differences across groups.

## adjS

```{r heatmap_adjS, fig.width = 10, fig.height = 7}
heat_adjs <- adjsampens %>%
  group_by(annotation, sub_type) %>%
  summarise(median_adjS = median(sampen, na.rm = TRUE), .groups = "drop")

ggplot(heat_adjs, aes(x = annotation, y = sub_type, fill = median_adjS)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = NULL, y = NULL, fill = "median adjS") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        aspect.ratio = NULL)
```

## adjH

```{r heatmap_adjH, fig.width = 10, fig.height = 7}
heat_adjh <- adjshannons %>%
  group_by(annotation, sub_type) %>%
  summarise(median_adjH = median(median_shannon, na.rm = TRUE), .groups = "drop")

ggplot(heat_adjh, aes(x = annotation, y = sub_type, fill = median_adjH)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma") +
  labs(x = NULL, y = NULL, fill = "median adjH") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        aspect.ratio = NULL)
```

## S

```{r heatmap_S, fig.width = 10, fig.height = 7}
heat_s <- sampens %>%
  group_by(annotation, sub_type) %>%
  summarise(median_S = median(sampen, na.rm = TRUE), .groups = "drop")

ggplot(heat_s, aes(x = annotation, y = sub_type, fill = median_S)) +
  geom_tile() +
  scale_fill_viridis_c(option = "viridis") +
  labs(x = NULL, y = NULL, fill = "median S") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        aspect.ratio = NULL)
```

# Locus-level H distributions {.tabset .tabset-pills}

Per-locus Shannon entropy (H) distributions from `.det.out.gz` for a small fixed set of annotations. The same row indices are sampled across all groups so comparisons are on the same loci. Uses annotations `genes` and `h3k4me3`; 5000 loci sampled with seed 42.

```{r locus_level_load}
LOCUS_ANNOTATIONS <- c("genes", "h3k4me3")
LOCUS_N           <- 5000L
LOCUS_SEED        <- 42L

det_paths_locus <- list.files(snakemake@params[["output_path"]],
                               pattern = "det\\.out\\.gz$", full.names = TRUE)
det_paths_locus <- det_paths_locus[!grepl("norm", det_paths_locus)]

## keep only target annotations
det_paths_locus <- det_paths_locus[
  get_annotation(basename(det_paths_locus)) %in% LOCUS_ANNOTATIONS
]

## for each annotation, determine sampling indices from the first file found
locus_indices <- lapply(LOCUS_ANNOTATIONS, function(ann) {
  fps <- det_paths_locus[get_annotation(basename(det_paths_locus)) == ann]
  if (length(fps) == 0) return(NULL)
  nrows <- nrow(read_tsv(fps[[1]], na = "-1", show_col_types = FALSE))
  n     <- min(LOCUS_N, nrows)
  set.seed(LOCUS_SEED)
  sort(sample.int(nrows, n))
})
names(locus_indices) <- LOCUS_ANNOTATIONS

read_locus <- function(fp) {
  fn  <- basename(fp)
  ann <- get_annotation(fn)
  idx <- locus_indices[[ann]]
  if (is.null(idx)) return(NULL)
  tmp <- read_tsv(fp, na = "-1", show_col_types = FALSE)
  tmp <- tmp[idx, , drop = FALSE]
  tibble(
    shannon    = tmp$shannon,
    avg_meth   = tmp$avg_meth,
    annotation = ann,
    sub_region = get_sub_region(fn),
    sub_type   = get_sub_type(fn)
  )
}

locus_data <- bplapply(det_paths_locus, read_locus, BPPARAM = param) |> bind_rows()
locus_data <- harmonize_levels(locus_data)
head(locus_data)
```

## Violin by sub-type

```{r locus_violin, fig.width = 12, fig.height = 8}
ggplot(locus_data, aes(x = sub_type, y = shannon, fill = sub_type)) +
  geom_violin(alpha = 0.5, draw_quantiles = 0.5) +
  geom_jitter(width = 0.2, size = 0.2, alpha = 0.05) +
  facet_wrap(~annotation, scales = "free_y") +
  labs(y = "per-locus Shannon entropy (H)") +
  theme_ng() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        legend.position = "none")
```

## Ridgelines by sub-type

```{r locus_ridges, fig.width = 12, fig.height = 8}
ggplot(locus_data, aes(x = shannon, y = sub_type, fill = sub_type)) +
  geom_density_ridges(alpha = 0.6, scale = 0.85) +
  facet_wrap(~annotation, scales = "free_x") +
  labs(x = "per-locus Shannon entropy (H)") +
  theme_ng() +
  theme(legend.position = "none")
```

## Methylation vs H

```{r locus_meth_vs_H, fig.width = 12, fig.height = 8}
locus_sub <- locus_data %>%
  slice_sample(n = min(2000, nrow(locus_data)), replace = FALSE)

ggplot(locus_sub, aes(x = avg_meth, y = shannon, color = sub_type)) +
  geom_point(size = 0.4, alpha = 0.3) +
  facet_wrap(~annotation) +
  labs(x = "avg. methylation", y = "per-locus Shannon entropy (H)") +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme_ng()
```

# UMAP: cell-type separability {.tabset .tabset-pills}

```{r umap_helpers}
library(uwot)

run_umap_wide <- function(wide_df, meta_cols,
                           n_hvf       = 1000L,
                           n_pcs       = 50L,
                           n_neighbors = 15L,
                           seed        = 42L) {
  if (is.null(wide_df) || nrow(wide_df) < 4L) return(NULL)
  meta <- wide_df %>% select(all_of(meta_cols))
  mat  <- wide_df %>% select(-all_of(meta_cols)) %>% as.matrix()

  mat[is.na(mat)] <- 0

  feat_var <- apply(mat, 2, var)
  n_keep   <- min(as.integer(n_hvf), sum(feat_var > 0))
  if (n_keep < 2L) {
    message("Too few variable features (", n_keep, "); skipping")
    return(NULL)
  }
  mat <- mat[, order(feat_var, decreasing = TRUE)[seq_len(n_keep)], drop = FALSE]
  message("HVF: ", n_keep, " of ", length(feat_var), " features selected")

  scaled <- scale(mat)
  bad    <- apply(scaled, 2, function(x) all(is.na(x) | is.nan(x)))
  scaled <- scaled[, !bad, drop = FALSE]
  if (ncol(scaled) == 0L) return(NULL)

  n_pcs_use <- min(as.integer(n_pcs), nrow(scaled) - 1L, ncol(scaled))
  if (n_pcs_use < 2L) {
    message("Too few PCs available (", n_pcs_use, "); skipping")
    return(NULL)
  }
  pca <- prcomp(scaled, rank. = n_pcs_use, scale. = FALSE)
  message("PCA: using ", n_pcs_use, " PCs")

  nn <- min(as.integer(n_neighbors), nrow(pca$x) - 1L)
  set.seed(seed)
  coords <- uwot::umap(pca$x, n_neighbors = nn, min_dist = 0.3,
                        n_epochs = 500, verbose = FALSE)
  bind_cols(meta, setNames(as.data.frame(coords), c("UMAP1", "UMAP2")))
}

## Standard plot: points colored by cell_class, labeled by sub_type
plot_umap_class <- function(df, title, subtitle = NULL) {
  if (is.null(df)) { message("No UMAP result for: ", title); return(invisible(NULL)) }
  ggplot(df, aes(x = UMAP1, y = UMAP2, color = cell_class)) +
    geom_point(size = 2.5, alpha = 0.85) +
    geom_text_repel(aes(label = sub_type), size = 2, max.overlaps = 25,
                    show.legend = FALSE) +
    scale_color_viridis_d(option = "turbo") +
    labs(title = title, subtitle = subtitle, color = "cell class") +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    theme_ng()
}
```

## Cell-level: adjS across annotations

Each point is one cell. Features are per-cell adjS for each annotation (one value
per annotation per cell). Within each (sub_region, sub_type) group the same cells
appear in the same row order across annotation files, so rows can be matched by
index. Color = CellClass from metadata.

```{r cell_umap_load}
short_paths_umap <- list.files(snakemake@params[["output_path"]],
                                pattern = "\\.out\\.gz$", full.names = TRUE)
short_paths_umap <- short_paths_umap[!grepl("\\.det\\.out\\.gz$",  short_paths_umap)]
short_paths_umap <- short_paths_umap[!grepl("\\.meth\\.out\\.gz$", short_paths_umap)]

## Read one column from all .out.gz files; return long table with cell_idx
read_cells_long <- function(fps, col) {
  bind_rows(bplapply(fps, function(fp) {
    fn   <- basename(fp)
    vals <- read_tsv(fp, na = "-1", show_col_types = FALSE)[[col]]
    data.frame(
      sub_region = get_sub_region(fn),
      sub_type   = get_sub_type(fn),
      annotation = get_annotation(fn),
      cell_idx   = seq_along(vals),
      value      = vals,
      stringsAsFactors = FALSE
    )
  }, BPPARAM = param))
}

cells_adjS_long <- read_cells_long(short_paths_umap, "sampen_norm")
cells_meth_long <- read_cells_long(short_paths_umap, "avg_meth")

## Pivot wide: rows = cells, columns = annotations
cells_adjS_wide <- pivot_wider(cells_adjS_long,
                                id_cols     = c(sub_region, sub_type, cell_idx),
                                names_from  = annotation,
                                values_from = value) %>%
  left_join(meta_grp %>% select(sub_region, sub_type, cell_class, major_type),
            by = c("sub_region", "sub_type"))

cells_meth_wide <- pivot_wider(cells_meth_long,
                                id_cols     = c(sub_region, sub_type, cell_idx),
                                names_from  = annotation,
                                values_from = value) %>%
  left_join(meta_grp %>% select(sub_region, sub_type, cell_class, major_type),
            by = c("sub_region", "sub_type"))

saveRDS(list(cells_adjS_wide = cells_adjS_wide,
             cells_meth_wide = cells_meth_wide),
        "ecker_cell_matrices.rds")

cat("Cells (adjS):", nrow(cells_adjS_wide),
    "| Annotations:", ncol(cells_adjS_wide) - 5L, "\n")
cat("Cell class breakdown:\n")
print(table(cells_adjS_wide$cell_class, useNA = "always"))
```

```{r cell_umap_adjS, fig.width = 8, fig.height = 7}
meta_cols_cell <- c("sub_region", "sub_type", "cell_idx", "cell_class", "major_type")

umap_cell_adjS <- run_umap_wide(cells_adjS_wide, meta_cols_cell, n_neighbors = 15L)
saveRDS(umap_cell_adjS, "ecker_umap_cell_adjS.rds")

plot_umap_class(umap_cell_adjS,
                title    = "Cell-level UMAP - adjS features",
                subtitle = "per-cell adjS per annotation as feature vector")
```

## Cell-level: avg_meth across annotations

Same as above but using average methylation as features, for direct comparison.

```{r cell_umap_meth, fig.width = 8, fig.height = 7}
umap_cell_meth <- run_umap_wide(cells_meth_wide, meta_cols_cell, n_neighbors = 15L)
saveRDS(umap_cell_meth, "ecker_umap_cell_meth.rds")

plot_umap_class(umap_cell_meth,
                title    = "Cell-level UMAP - avg. methylation features",
                subtitle = "per-cell avg. methylation per annotation as feature vector")
```

## Group-level: adjH across annotations

Each point is one (sub_region x sub_type) group. Features are median adjH per
annotation. Fewer points than the cell-level view but each point represents the
consensus entropy profile of a cell type.

```{r group_umap_adjH, fig.width = 8, fig.height = 7}
adjH_wide_grp <- all_grp_meta %>%
  select(sub_region, sub_type, cell_class, major_type, annotation, median_adjH) %>%
  pivot_wider(names_from = annotation, values_from = median_adjH)

umap_grp_adjH <- run_umap_wide(adjH_wide_grp,
                                meta_cols = c("sub_region", "sub_type",
                                              "cell_class", "major_type"),
                                n_neighbors = 5L)
saveRDS(umap_grp_adjH, "ecker_umap_grp_adjH.rds")

plot_umap_class(umap_grp_adjH,
                title    = "Group-level UMAP - adjH features",
                subtitle = "median adjH per annotation as feature vector")
```

## Group-level: avg_meth across annotations

```{r group_umap_meth, fig.width = 8, fig.height = 7}
meth_wide_grp <- all_grp_meta %>%
  select(sub_region, sub_type, cell_class, major_type, annotation, median_meth) %>%
  pivot_wider(names_from = annotation, values_from = median_meth)

umap_grp_meth <- run_umap_wide(meth_wide_grp,
                                meta_cols = c("sub_region", "sub_type",
                                              "cell_class", "major_type"),
                                n_neighbors = 5L)
plot_umap_class(umap_grp_meth,
                title    = "Group-level UMAP - avg. methylation features",
                subtitle = "median avg. methylation per annotation as feature vector")
```

## Group-level: gene loci, adjH

Each group embedded in the space of individual gene loci (per-locus adjH from
.norm.det.out.gz). Up to 2000 genes sampled. More features than the
annotation-summary view, at the cost of heavier imputation.

```{r gene_locus_umap_setup}
LOCUS_N_UMAP <- 2000L

norm_det_paths_genes <- list.files(snakemake@params[["output_path"]],
                                    pattern = "norm\\.det\\.out\\.gz$",
                                    full.names = TRUE)
det_paths_genes <- list.files(snakemake@params[["output_path"]],
                               pattern = "det\\.out\\.gz$", full.names = TRUE)
det_paths_genes <- det_paths_genes[!grepl("norm", det_paths_genes)]

build_gene_mat <- function(col, fps, n_loci = LOCUS_N_UMAP) {
  fps_g <- fps[get_annotation(basename(fps)) == "genes"]
  if (length(fps_g) < 4L) return(NULL)
  ref <- read_tsv(fps_g[[1]], na = "-1", show_col_types = FALSE)
  n   <- min(n_loci, nrow(ref))
  set.seed(LOCUS_SEED)
  idx <- sort(sample.int(nrow(ref), n))
  rows <- bplapply(fps_g, function(fp)
    read_tsv(fp, na = "-1", show_col_types = FALSE)[[col]][idx],
    BPPARAM = param)
  mat <- do.call(rbind, rows)
  colnames(mat) <- paste0("gene_", idx)
  df <- data.frame(
    sub_region = get_sub_region(basename(fps_g)),
    sub_type   = get_sub_type(basename(fps_g)),
    as.data.frame(mat),
    check.names = FALSE
  )
  left_join(df,
            meta_grp %>% select(sub_region, sub_type, cell_class, major_type),
            by = c("sub_region", "sub_type"))
}

gene_mat_adjH <- build_gene_mat("shannon_norm", fps = norm_det_paths_genes)
gene_mat_meth <- build_gene_mat("avg_meth",     fps = det_paths_genes)

saveRDS(list(gene_mat_adjH = gene_mat_adjH, gene_mat_meth = gene_mat_meth),
        "ecker_gene_matrices.rds")
```

```{r gene_locus_umap_adjH, fig.width = 8, fig.height = 7}
meta_cols_grp <- c("sub_region", "sub_type", "cell_class", "major_type")

umap_gene_adjH <- run_umap_wide(gene_mat_adjH, meta_cols_grp, n_neighbors = 5L)
plot_umap_class(umap_gene_adjH,
                title    = "Gene-locus UMAP - adjH features",
                subtitle = paste0(LOCUS_N_UMAP, " genes sampled"))
```

## Group-level: gene loci, avg_meth

```{r gene_locus_umap_meth, fig.width = 8, fig.height = 7}
umap_gene_meth <- run_umap_wide(gene_mat_meth, meta_cols_grp, n_neighbors = 5L)
plot_umap_class(umap_gene_meth,
                title    = "Gene-locus UMAP - avg. methylation features",
                subtitle = paste0(LOCUS_N_UMAP, " genes sampled"))
```

# Paper UMAP + our entropy overlay {.tabset .tabset-pills}

All cells from the original Ecker et al. metadata placed at their published UMAP
coordinates (L1UMAP), colored by the group-level entropy or methylation value
computed here. One panel per genomic annotation, so you can assess whether
entropy differences between cell types are annotation-specific.

```{r paper_umap_data}
## Join per-cell paper UMAP coords with group-level entropy for each annotation
cells_overlay <- meta_cells %>%
  left_join(
    all_grp_meta %>%
      select(sub_region, sub_type, annotation, median_adjH, median_meth,
             median_adjS, cell_class, major_type),
    by      = c("sub_region", "sub_type"),
    relationship = "many-to-many"
  ) %>%
  filter(!is.na(umap1), !is.na(annotation))

saveRDS(cells_overlay, "ecker_paper_umap_overlay.rds")
cat("Rows in overlay table:", nrow(cells_overlay), "\n")
```

## adjH per annotation

```{r paper_umap_adjH, fig.width = 14, fig.height = 10}
df_adjH <- cells_overlay %>% filter(!is.na(median_adjH)) %>% droplevels()
if (nrow(df_adjH) > 0) {
  ggplot(df_adjH, aes(x = umap1, y = umap2, color = median_adjH)) +
    geom_point(size = 0.4, alpha = 0.6) +
    scale_color_viridis_c(option = "plasma", na.value = "grey80") +
    facet_wrap(~annotation) +
    labs(title = "Paper UMAP coordinates - color = group adjH",
         x = "UMAP1 (Ecker et al.)", y = "UMAP2 (Ecker et al.)",
         color = "median adjH") +
    theme_ng() +
    theme(axis.text = element_blank(), axis.ticks = element_blank())
} else {
  message("No adjH overlay data available - check ecker_paper_umap_overlay.rds")
}
```

## avg_meth per annotation

```{r paper_umap_meth, fig.width = 14, fig.height = 10}
df_meth <- cells_overlay %>% filter(!is.na(median_meth)) %>% droplevels()
if (nrow(df_meth) > 0) {
  ggplot(df_meth, aes(x = umap1, y = umap2, color = median_meth)) +
    geom_point(size = 0.4, alpha = 0.6) +
    scale_color_viridis_c(option = "viridis", na.value = "grey80") +
    facet_wrap(~annotation) +
    labs(title = "Paper UMAP coordinates - color = group avg. methylation",
         x = "UMAP1 (Ecker et al.)", y = "UMAP2 (Ecker et al.)",
         color = "median meth") +
    theme_ng() +
    theme(axis.text = element_blank(), axis.ticks = element_blank())
} else {
  message("No methylation overlay data available - check ecker_paper_umap_overlay.rds")
}
```

## Cell class reference

```{r paper_umap_cellclass, fig.width = 8, fig.height = 7}
meta_cells %>%
  filter(!is.na(umap1)) %>%
  distinct(basename, .keep_all = TRUE) %>%
  ggplot(aes(x = umap1, y = umap2, color = cell_class)) +
  geom_point(size = 0.5, alpha = 0.6) +
  scale_color_viridis_d(option = "turbo") +
  labs(title = "Paper UMAP coordinates - cell class reference",
       x = "UMAP1 (Ecker et al.)", y = "UMAP2 (Ecker et al.)",
       color = "cell class") +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  theme_ng() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

# Session

```{r session, cache = FALSE}
sessionInfo()
```
