---
title: "Argelaguet figure"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    code_folding: hide
    number_sections: no
---

```{r setup}
suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
  library(dplyr)
  library(tidyr)
  library(ComplexHeatmap)
  library(circlize)
  library(ggrastr)
  library(ggrepel)
  library(viridisLite)
  library(grid)
})

source("rules/src/ggtheme.R")
source("rules/src/palettes.R")
source("rules/src/driver_utils.R")

knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)

log_con <- file(snakemake@log[[1]], open = "wt")
sink(log_con)
sink(log_con, type = "message")

out_dir <- snakemake@params$out_dir
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
```

```{r load}
cell_matrices  <- readRDS("argelaguet_cell_matrices.rds")
ent            <- readRDS("argelaguet_entropy.rds")
grp_meta       <- readRDS("argelaguet_groups_meta.rds")
cell_umap_adjS <- readRDS("argelaguet_umap_cell_adjS.rds")

cell_df <- if (file.exists("argelaguet_per_cell_summary.csv"))
  read.csv("argelaguet_per_cell_summary.csv", stringsAsFactors = FALSE) else NULL
ve_long <- if (file.exists("argelaguet_win_varexp.csv"))
  read.csv("argelaguet_win_varexp.csv", stringsAsFactors = FALSE) else NULL
```

# Compact figure (primary, A-M)

A) sampling and feature schema;
B) cell-level UMAP coloured by stage;
C) cell-level UMAP coloured by lineage;
D) per-cell mean methylation vs mean S, coloured by stage;
E) per-cell mean methylation vs mean S, coloured by lineage;
F) per-cell adjS at enhancers vs promoters, coloured by stage;
G) median variance explained per assay by stage and lineage;
H) time-course median adjS per annotation across stages;
I) time-course median adjH per annotation across stages;
J) heatmap of median adjS per lineage and annotation;
K) IQR adjS per lineage;
L) driver categorization across lineage classes;
M) driver categorization across stages.

```{r fa-panel-a-v2}
schema_classes <- c("Epiblast", "Mesoderm", "Endoderm", "ExE_ectoderm")
schema_layout <- list(
  Epiblast    = c("Epiblast", "Caudal_epiblast"),
  Mesoderm    = c("Nascent_mesoderm", "Mature_mesoderm"),
  Endoderm    = c("Embryonic_endoderm"),
  ExE_ectoderm = c("ExE_ectoderm")
)
schema_classes <- intersect(schema_classes, names(argelaguet_lineage_class_pal))
schema_layout  <- schema_layout[schema_classes]
schema_cells <- do.call(rbind, lapply(seq_along(schema_layout), function(i) {
  cls <- schema_classes[i]
  lns <- intersect(schema_layout[[i]], names(argelaguet_lineage_pal))
  if (length(lns) == 0) lns <- schema_layout[[i]][1]
  data.frame(lineage_class = cls, lineage = lns, stringsAsFactors = FALSE)
}))
schema_cells$col <- seq_len(nrow(schema_cells))
schema_cells$cls_color <- argelaguet_lineage_class_pal[schema_cells$lineage_class]
schema_cells$lin_color <- argelaguet_lineage_pal[schema_cells$lineage]

schema_regions <- c("windows", "Enh", "Prom", "marks")
n_schema_regions <- length(schema_regions)
n_schema_cols <- nrow(schema_cells)

set.seed(2)
schema_grid <- expand.grid(col = seq_len(n_schema_cols),
                           row = seq_len(n_schema_regions))
schema_grid$val <- runif(nrow(schema_grid), 0.1, 0.95)

bar_lin <- n_schema_regions + 0.9
bar_cls <- n_schema_regions + 1.55
ellipsis_y <- 0
cls_centers <- tapply(schema_cells$col, schema_cells$lineage_class, mean)
cls_short <- c(Epiblast = "Epiblast", Mesoderm = "Mesoderm",
               Endoderm = "Endoderm", ExE_ectoderm = "ExE-ect")
cls_labels <- cls_short[names(cls_centers)]

pA_v2 <- ggplot() +
  geom_tile(data = schema_grid,
            aes(x = col, y = row, fill = val),
            colour = "grey70", linewidth = 0.15) +
  scale_fill_gradient(low = "#FFFFFF", high = "#444444",
                      guide = "none", limits = c(0, 1)) +
  geom_tile(data = schema_cells,
            aes(x = col, y = bar_cls),
            fill = schema_cells$cls_color,
            height = 0.45, width = 0.95, inherit.aes = FALSE) +
  geom_tile(data = schema_cells,
            aes(x = col, y = bar_lin),
            fill = schema_cells$lin_color,
            height = 0.45, width = 0.95, inherit.aes = FALSE) +
  annotate("text",
           x = cls_centers, y = bar_cls + 0.5,
           label = cls_labels,
           size = 1.4) +
  annotate("text", x = -0.15, y = bar_cls,
           label = "lineage class", hjust = 1, size = 1.7) +
  annotate("text", x = -0.15, y = bar_lin,
           label = "lineage", hjust = 1, size = 1.7) +
  annotate("text", x = -0.15, y = seq_len(n_schema_regions),
           label = rev(schema_regions),
           hjust = 1, size = 1.7) +
  annotate("text", x = 1, y = ellipsis_y,
           label = "(...)", hjust = 0, size = 2.0) +
  annotate("text", x = n_schema_cols + 0.6, y = bar_cls,
           label = "(...)", hjust = 0, size = 2.0) +
  annotate("text", x = (n_schema_cols + 1) / 2, y = bar_cls + 1.1,
           label = "cells", size = 1.9) +
  annotate("text", x = -3.0, y = (n_schema_regions + 1) / 2,
           label = "regions", angle = 90, size = 1.9) +
  annotate("segment",
           x = n_schema_cols + 0.7, xend = n_schema_cols + 1.5,
           y = (n_schema_regions + 1) / 2,
           yend = (n_schema_regions + 1) / 2,
           arrow = arrow(length = unit(1.2, "mm"), type = "closed"),
           linewidth = 0.3) +
  annotate("text", x = n_schema_cols + 1.7,
           y = (n_schema_regions + 1) / 2,
           label = "adjS", hjust = 0, size = 2.0) +
  annotate("segment",
           x = (n_schema_cols + 1) / 2,
           xend = (n_schema_cols + 1) / 2,
           y = ellipsis_y - 0.4, yend = ellipsis_y - 1.05,
           arrow = arrow(length = unit(1.2, "mm"), type = "closed"),
           linewidth = 0.3) +
  annotate("text", x = (n_schema_cols + 1) / 2, y = ellipsis_y - 1.35,
           label = "adjH", size = 2.0) +
  coord_fixed(xlim = c(-3.4, n_schema_cols + 3.0),
              ylim = c(ellipsis_y - 1.7, bar_cls + 1.4),
              clip = "off") +
  theme_void(base_family = "Helvetica") +
  theme(plot.margin = margin(1, 1, 1, 1, "mm"))
```

```{r fa-panel-b-c-v2}
mk_cell_umap <- function(color_col, pal, lab) {
  ggplot(cell_umap_adjS,
         aes(x = UMAP1, y = UMAP2, color = .data[[color_col]])) +
    geom_point_rast(size = 0.2, alpha = 0.6, raster.dpi = 300) +
    scale_color_manual(values = pal, na.value = "grey80") +
    guides(color = guide_legend(override.aes = list(size = 1.0, alpha = 1),
                                ncol = 2)) +
    labs(x = "UMAP 1", y = "UMAP 2", color = lab) +
    theme_ng(base_size = 7) +
    theme(legend.position    = "bottom",
          legend.key.size    = unit(1.0, "mm"),
          legend.text        = element_text(size = 5),
          legend.title       = element_text(size = 6),
          legend.margin      = margin(0, 0, 0, 0),
          legend.box.margin  = margin(-6, 0, -2, 0),
          legend.box.spacing = unit(0.2, "mm"),
          plot.margin        = margin(0, 1, 0, 1, "mm"))
}

pB_v2 <- mk_cell_umap("stage",   argelaguet_stage_pal,           "stage")
pC_v2 <- mk_cell_umap("lineage", argelaguet_lineage_pal,         "lineage")
```

```{r fa-panel-d-e-v2}
mk_meth_s_scatter <- function(color_col, pal, lab) {
  if (is.null(cell_df)) {
    return(patchwork::plot_spacer() +
             labs(title = "argelaguet_per_cell_summary.csv missing"))
  }
  df <- cell_df
  df[[color_col]] <- factor(df[[color_col]])
  ggplot(df,
         aes(x = mean_meth, y = mean_sampen,
             color = .data[[color_col]])) +
    geom_point(size = 0.35, alpha = 0.6) +
    scale_color_manual(values = pal, na.value = "grey80") +
    guides(color = guide_legend(override.aes = list(size = 1.2, alpha = 1),
                                ncol = 2)) +
    labs(x = "mean methylation", y = "mean S", color = lab) +
    theme_ng(base_size = 7) +
    theme(legend.position    = "bottom",
          legend.key.size    = unit(1.0, "mm"),
          legend.text        = element_text(size = 5),
          legend.title       = element_text(size = 6),
          legend.margin      = margin(0, 0, 0, 0),
          legend.box.margin  = margin(-6, 0, -2, 0),
          legend.box.spacing = unit(0.2, "mm"),
          plot.margin        = margin(0, 1, 0, 1, "mm"))
}

pD_v2 <- mk_meth_s_scatter("stage",   argelaguet_stage_pal,   "stage")
pE_v2 <- mk_meth_s_scatter("lineage", argelaguet_lineage_pal, "lineage")
```

```{r fa-panel-f-v2}
enh_prom <- cell_matrices$cells_adjS_wide %>%
  select(stage, enh = `enh-E75-union`, prom = promoters) %>%
  filter(!is.na(enh), !is.na(prom))

pF_v2 <- ggplot(enh_prom, aes(x = prom, y = enh, color = stage)) +
  geom_point_rast(size = 0.35, alpha = 0.5, raster.dpi = 300) +
  scale_color_manual(values = argelaguet_stage_pal) +
  guides(color = guide_legend(override.aes = list(size = 1.2, alpha = 1),
                              nrow = 1)) +
  labs(x = "adjS promoters", y = "adjS enhancers", color = "stage") +
  theme_ng(base_size = 7) +
  theme(legend.position    = "bottom",
        legend.key.size    = unit(1.0, "mm"),
        legend.text        = element_text(size = 5),
        legend.title       = element_text(size = 6),
        legend.margin      = margin(0, 0, 0, 0),
        legend.box.margin  = margin(-6, 0, -2, 0),
        legend.box.spacing = unit(0.2, "mm"),
        plot.margin        = margin(0, 1, 0, 1, "mm"))
```

```{r fa-panel-g-v2}
if (!is.null(ve_long)) {
  ve_long$assay <- factor(ve_long$assay, levels = unique(ve_long$assay))
  pG_v2 <- ggplot(ve_long,
                  aes(x = assay, y = median_ve, fill = variable)) +
    geom_col(position = position_dodge(width = 0.8)) +
    scale_fill_manual(values = c("stage" = "#0072B2",
                                 "lineage" = "#E69F00")) +
    labs(x = "assay", y = "median % var. explained", fill = NULL) +
    theme_ng(base_size = 7) +
    theme(legend.key.size    = unit(1.5, "mm"),
          legend.position    = "top",
          legend.text        = element_text(size = 6),
          legend.margin      = margin(0, 0, 0, 0),
          legend.box.margin  = margin(0, 0, -4, 0),
          legend.box.spacing = unit(0.2, "mm"),
          plot.margin        = margin(0, 0, 0, 0, "mm"))
} else {
  pG_v2 <- patchwork::plot_spacer() +
    labs(title = "argelaguet_win_varexp.csv missing")
}
```

```{r fa-panel-h-v2}
tc_adjS <- ent$adjsampens %>%
  group_by(stage, annotation) %>%
  summarise(
    median_adjS = median(sampen, na.rm = TRUE),
    lo = quantile(sampen, 0.25, na.rm = TRUE),
    hi = quantile(sampen, 0.75, na.rm = TRUE),
    .groups = "drop"
  )
write.csv(tc_adjS, "argelaguet_fig_adjS_by_stage_annotation.csv", row.names = FALSE)

pH_v2 <- ggplot(tc_adjS,
                aes(x = stage, y = median_adjS,
                    color = annotation, group = annotation)) +
  geom_line(linewidth = 0.4) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = annotation),
              alpha = 0.1, color = NA) +
  scale_color_manual(values = argelaguet_annotation_pal) +
  scale_fill_manual(values = argelaguet_annotation_pal) +
  guides(color = guide_legend(ncol = 2, override.aes = list(linewidth = 1)),
         fill  = "none") +
  labs(x = "stage", y = "median adjS", color = "annotation") +
  theme_ng(base_size = 7) +
  theme(axis.text.x        = element_text(angle = 45, hjust = 1, size = 5.5),
        legend.position    = "right",
        legend.key.size    = unit(1.5, "mm"),
        legend.text        = element_text(size = 5),
        legend.title       = element_text(size = 6),
        legend.margin      = margin(0, 0, 0, 0),
        legend.box.margin  = margin(0, 0, 0, -3),
        legend.box.spacing = unit(0.2, "mm"),
        plot.margin        = margin(0, 0, 0, 0, "mm"))
```

```{r fa-panel-i-v2}
tc_adjH <- ent$adjshannons %>%
  group_by(stage, annotation) %>%
  summarise(
    median_adjH = median(median_shannon, na.rm = TRUE),
    lo = quantile(median_shannon, 0.25, na.rm = TRUE),
    hi = quantile(median_shannon, 0.75, na.rm = TRUE),
    .groups = "drop"
  )
write.csv(tc_adjH, "argelaguet_fig_adjH_by_stage_annotation.csv", row.names = FALSE)

pI_v2 <- ggplot(tc_adjH,
                aes(x = stage, y = median_adjH,
                    color = annotation, group = annotation)) +
  geom_line(linewidth = 0.4) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = annotation),
              alpha = 0.1, color = NA) +
  scale_color_manual(values = argelaguet_annotation_pal) +
  scale_fill_manual(values = argelaguet_annotation_pal) +
  guides(color = "none", fill = "none") +
  labs(x = "stage", y = "median adjH") +
  theme_ng(base_size = 7) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 5.5),
        plot.margin = margin(0, 0, 0, 0, "mm"))
```

```{r fa-panel-j-v2-data}
heat_df <- grp_meta$all_grp_meta %>%
  group_by(lineage, annotation, lineage_class) %>%
  summarise(median_adjS = median(median_adjS, na.rm = TRUE),
            .groups = "drop") %>%
  filter(!is.na(lineage_class))

wide <- heat_df %>%
  pivot_wider(names_from = annotation, values_from = median_adjS)

mat <- wide %>% select(-lineage, -lineage_class) %>% as.matrix()
rownames(mat) <- wide$lineage

brk <- c(min(mat, na.rm = TRUE),
         mean(range(mat, na.rm = TRUE)),
         max(mat, na.rm = TRUE))
col_fun_j <- colorRamp2(brk, c("navy", "white", "firebrick"))
lc_pal <- argelaguet_lineage_class_pal[sort(unique(wide$lineage_class))]
```

```{r fa-panel-j-v2}
ht_compact <- Heatmap(
  mat,
  name              = "median\nadjS",
  col               = col_fun_j,
  row_split         = wide$lineage_class,
  cluster_rows      = TRUE,
  cluster_row_slices = FALSE,
  cluster_columns   = TRUE,
  show_row_names    = TRUE,
  show_column_names = TRUE,
  row_names_gp      = gpar(fontsize = 4.5),
  column_names_gp   = gpar(fontsize = 5),
  row_title_gp      = gpar(fontsize = 6),
  row_title_rot     = 0,
  row_gap           = unit(0.6, "mm"),
  right_annotation  = rowAnnotation(
    lineage_class = wide$lineage_class,
    col = list(lineage_class = lc_pal),
    simple_anno_size = unit(1.8, "mm"),
    annotation_name_gp = gpar(fontsize = 5),
    show_legend = FALSE
  ),
  heatmap_legend_param = list(
    title_gp    = gpar(fontsize = 6),
    labels_gp   = gpar(fontsize = 5),
    grid_height = unit(2, "mm"),
    grid_width  = unit(2, "mm")
  )
)

pJ_v2 <- wrap_elements(full = grid::grid.grabExpr(draw(ht_compact)))
```

```{r fa-panel-k-v2}
var_lin <- ent$adjsampens %>%
  group_by(lineage) %>%
  summarise(iqr = IQR(sampen, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(iqr)) %>%
  mutate(lineage = factor(lineage, levels = lineage))
write.csv(var_lin, "argelaguet_fig_adjS_iqr_by_lineage.csv", row.names = FALSE)

pK_v2 <- ggplot(var_lin, aes(x = lineage, y = iqr)) +
  geom_col(fill = "#35b779") +
  coord_flip() +
  labs(x = NULL, y = "IQR adjS") +
  theme_ng(base_size = 7) +
  theme(axis.text.y = element_text(size = 5),
        plot.margin = margin(0, 0, 0, 0, "mm"))
```

```{r fa-panel-l-m-v2}
driver_df_lin <- categorize_drivers(grp_meta$all_grp_meta, "lineage_class")
driver_df_st  <- categorize_drivers(grp_meta$all_grp_meta, "stage")
write.csv(driver_df_lin, "argelaguet_driver_sd_range_lineage.csv", row.names = FALSE)
write.csv(driver_df_st,  "argelaguet_driver_sd_range_stage.csv",   row.names = FALSE)

mk_driver_panel <- function(df, x_lab, y_lab) {
  p <- plot_driver_scatter(df, x_label = x_lab, y_label = y_lab) +
    guides(color = guide_legend(nrow = 1, title.position = "left"),
           shape = guide_legend(nrow = 1, title.position = "left")) +
    theme_ng(base_size = 7) +
    theme(legend.position    = "bottom",
          legend.key.size    = unit(1.0, "mm"),
          legend.text        = element_text(size = 5),
          legend.title       = element_text(size = 6),
          legend.margin      = margin(0, 0, 0, 0),
          legend.box.margin  = margin(-6, 0, -2, 0),
          legend.box.spacing = unit(0.2, "mm"),
          plot.margin        = margin(0, 0, 0, 0, "mm"))
  repel_idx <- which(sapply(p$layers, function(l)
    inherits(l$geom, "GeomTextRepel")))
  if (length(repel_idx) > 0) p$layers[[repel_idx]] <- NULL
  p +
    ggrepel::geom_text_repel(size = 1.5, max.overlaps = Inf,
                             min.segment.length = 0,
                             segment.size = 0.2, box.padding = 0.25,
                             force = 3)
}

pL_v2 <- mk_driver_panel(driver_df_lin,
                         "SD adjH (lineage classes)",
                         "SD adjS (lineage classes)")
pM_v2 <- mk_driver_panel(driver_df_st,
                         "SD adjH (stages)",
                         "SD adjS (stages)")
```

```{r fa-main-v2, fig.width = 8.27, fig.height = 8.0}
# Main Fig 7 (manuscript, fig:argelaguet): panels A--I in three rows.
# J--M (heatmap, IQR, drivers) moved to single-panel supplementary figures
# per author preference (see fa-supp-* chunks below).
row1_v2 <- pA_v2 + pB_v2 + pC_v2 + plot_layout(widths = c(1.4, 2.0, 2.0))
row2_v2 <- pD_v2 + pE_v2 + pF_v2 + plot_layout(widths = c(1, 1, 1))
row3_v2 <- pG_v2 + pH_v2 + pI_v2 + plot_layout(widths = c(1, 1.5, 1.5))

main_fig <- row1_v2 / row2_v2 / row3_v2 +
  plot_layout(heights = c(1.4, 1.0, 1.0)) +
  plot_annotation(tag_levels = "A",
                  theme = theme(
                    plot.tag          = element_text(size = 9, face = "bold"),
                    plot.tag.position = c(0.01, 0.99),
                    plot.margin       = margin(1, 1, 1, 1, "mm")
                  ))

print(main_fig)

ggsave(
  file.path(out_dir, "fig_argelaguet_main.pdf"),
  plot   = main_fig,
  width  = 8.27,
  height = 8.0,
  units  = "in",
  device = cairo_pdf
)

# Legacy A--M compact figure kept for reference; can be dropped later.
row4_v2 <- pJ_v2 + pK_v2 + pL_v2 + pM_v2 +
  plot_layout(widths = c(1.5, 0.8, 1.0, 1.0))
big_fig <- row1_v2 / row2_v2 / row3_v2 / row4_v2 +
  plot_layout(heights = c(1.4, 1.0, 1.0, 1.4)) +
  plot_annotation(tag_levels = "A",
                  theme = theme(
                    plot.tag          = element_text(size = 9, face = "bold"),
                    plot.tag.position = c(0.01, 0.99),
                    plot.margin       = margin(1, 1, 1, 1, "mm")
                  ))
ggsave(
  file.path(out_dir, "fig_argelaguet_compact.pdf"),
  plot   = big_fig,
  width  = 8.27,
  height = 11.0,
  units  = "in",
  device = cairo_pdf
)
```

```{r fa-supp-iqr-lineage, fig.width = ng_fig_size(2, 1.2)$w, fig.height = ng_fig_size(2, 1.2)$h}
# Supp single-panel: IQR adjS per lineage. Manuscript ref sfig:argelaguet_iqr_lineage.
ggsave(
  file.path(out_dir, "fig_argelaguet_iqr_lineage.pdf"),
  plot   = pK_v2,
  width  = ng_fig_size(2, 1.2)$w,
  height = ng_fig_size(2, 1.2)$h,
  units  = "in",
  device = cairo_pdf
)
```

```{r fa-supp-drivers-lineage, fig.width = ng_fig_size(2, 1.5)$w, fig.height = ng_fig_size(2, 1.5)$h}
# Supp single-panel: driver categorization across lineage classes.
# Manuscript ref sfig:argelaguet_drivers_lineage.
ggsave(
  file.path(out_dir, "fig_argelaguet_drivers_lineage.pdf"),
  plot   = pL_v2,
  width  = ng_fig_size(2, 1.5)$w,
  height = ng_fig_size(2, 1.5)$h,
  units  = "in",
  device = cairo_pdf
)
```

```{r fa-supp-drivers-stage, fig.width = ng_fig_size(2, 1.5)$w, fig.height = ng_fig_size(2, 1.5)$h}
# Supp single-panel: driver categorization across stages.
# Manuscript ref sfig:argelaguet_drivers_stage.
ggsave(
  file.path(out_dir, "fig_argelaguet_drivers_stage.pdf"),
  plot   = pM_v2,
  width  = ng_fig_size(2, 1.5)$w,
  height = ng_fig_size(2, 1.5)$h,
  units  = "in",
  device = cairo_pdf
)
```

# Standalone heatmap (full size, supplementary)

```{r fa-heatmap-standalone, fig.width = 14, fig.height = 10}
ht_full <- Heatmap(
  mat,
  name              = "median adjS",
  col               = col_fun_j,
  row_split         = wide$lineage_class,
  cluster_rows      = TRUE,
  cluster_row_slices = FALSE,
  cluster_columns   = FALSE,
  show_row_names    = TRUE,
  show_column_names = TRUE,
  row_names_gp      = gpar(fontsize = 5),
  column_names_gp   = gpar(fontsize = 7),
  row_names_max_width = unit(4, "cm"),
  right_annotation  = rowAnnotation(
    lineage_class = wide$lineage_class,
    col = list(lineage_class = lc_pal),
    annotation_legend_param = list(lineage_class = list(title = "lineage class"))
  ),
  column_title = "median adjS per lineage and annotation"
)

draw(ht_full)
pdf(file.path(out_dir, "fig_argelaguet_heatmap.pdf"), width = 14, height = 10)
draw(ht_full)
invisible(dev.off())
```

# Legacy compact panels (kept for reference)

```{r fa-legacy-panels, fig.width = ng_fig_size(4, 2, panel_mm = 50)$w, fig.height = ng_fig_size(4, 2, panel_mm = 50)$h}
pa <- ggplot(cell_umap_adjS, aes(x = UMAP1, y = UMAP2, color = stage)) +
  geom_point_rast(size = 0.8, alpha = 0.7, raster.dpi = 300) +
  scale_color_manual(values = argelaguet_stage_pal) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(x = "UMAP 1", y = "UMAP 2", color = "stage") +
  theme_ng(base_size = 8)

pb <- ggplot(tc_adjS,
             aes(x = stage, y = median_adjS, color = annotation, group = annotation)) +
  geom_line(linewidth = 0.5) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = annotation),
              alpha = 0.1, color = NA) +
  scale_color_manual(values = argelaguet_annotation_pal) +
  scale_fill_manual(values = argelaguet_annotation_pal) +
  labs(x = "stage", y = "median adjS",
       color = "annotation", fill = "annotation") +
  theme_ng(base_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.key.size = unit(0.3, "cm"))

pc <- ggplot(tc_adjH,
             aes(x = stage, y = median_adjH, color = annotation, group = annotation)) +
  geom_line(linewidth = 0.5) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = annotation),
              alpha = 0.1, color = NA) +
  scale_color_manual(values = argelaguet_annotation_pal) +
  scale_fill_manual(values = argelaguet_annotation_pal) +
  labs(x = "stage", y = "median adjH",
       color = "annotation", fill = "annotation") +
  theme_ng(base_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.key.size = unit(0.3, "cm"))

pd <- ggplot(enh_prom, aes(x = prom, y = enh, color = stage)) +
  geom_point_rast(size = 0.5, alpha = 0.5, raster.dpi = 300) +
  scale_color_manual(values = argelaguet_stage_pal) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(x = "adjS promoters", y = "adjS enhancers", color = "stage") +
  theme_ng(base_size = 8)

pe <- ggplot(var_lin, aes(x = lineage, y = iqr)) +
  geom_col(fill = "#35b779") +
  coord_flip() +
  labs(x = NULL, y = "IQR adjS") +
  theme_ng(base_size = 8)

pf <- plot_driver_scatter(driver_df_lin,
                          x_label = "SD of adjH across lineage classes",
                          y_label = "SD of adjS across lineage classes") +
  theme_ng(base_size = 8) +
  theme(legend.key.size = unit(2, "mm"),
        legend.text     = element_text(size = 6),
        legend.title    = element_text(size = 7))

legend_theme <- theme(
  legend.key.size   = unit(2, "mm"),
  legend.text       = element_text(size = 5),
  legend.title      = element_text(size = 6),
  legend.position   = "right"
)
pb_legacy <- pb +
  guides(color = guide_legend(ncol = 1, override.aes = list(linewidth = 1)),
         fill  = guide_legend(ncol = 1)) + legend_theme
pc_legacy <- pc +
  guides(color = guide_legend(ncol = 1, override.aes = list(linewidth = 1)),
         fill  = guide_legend(ncol = 1)) + legend_theme
pe_legacy <- pe + theme(axis.text.y = element_text(size = 6))

gg_fig <- (pa | pb_legacy | pc_legacy) / (pd | pe_legacy | pf) +
  plot_layout(heights = c(1, 0.9)) +
  plot_annotation(tag_levels = "A",
                  theme = theme(plot.tag = element_text(size = 8, face = "bold")))

print(gg_fig)

ggsave(
  file.path(out_dir, "fig_argelaguet_panels.pdf"),
  plot   = gg_fig,
  width  = ng_fig_size(4, 2, panel_mm = 50)$w,
  height = ng_fig_size(4, 2, panel_mm = 50)$h,
  units  = "in",
  device = cairo_pdf
)
```
