---
title: "CRC entropy figure"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    code_folding: hide
    number_sections: no
---

```{r setup}
suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
  library(dplyr)
  library(tidyr)
  library(ComplexHeatmap)
  library(circlize)
  library(ggrastr)
  library(viridisLite)
  library(grid)
})

source("rules/src/ggtheme.R")
source("rules/src/palettes.R")

knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)

log_con <- file(snakemake@log[[1]], open = "wt")
sink(log_con)
sink(log_con, type = "message")

out_dir <- snakemake@params$out_dir
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

```

```{r load}
ent    <- readRDS("crc_entropy_summaries.rds")
embeds <- readRDS("crc_embeddings_debug.rds")
```

# Single-page CRC figure

Panels A-H assembled on one page:
A) schematic of feature spaces and biological groupings;
B) per-window UMAPs on S and adjS, coloured by location and patient;
C) per-variable median variance explained;
D) median adjS per biopsy location and annotation (heatmap);
E) per-cell entropy summaries by biopsy location;
F) differential entropy window counts per contrast vs NC;
G) genomic-annotation overlap in differential-entropy windows (dotplot);
H) annotation driver categorization (adjH vs adjS SD).

```{r single-page-setup}
suppressPackageStartupMessages({
  library(ggrepel)
})
source("rules/src/driver_utils.R")
```

```{r single-page-load}
de          <- readRDS(snakemake@input$de)
driver_df   <- readRDS("crc_driver_sd_range.rds")
ve_win_long <- read.csv("crc_win_varexp.csv", stringsAsFactors = FALSE)
cell_df_all <- read.csv("crc_per_cell_summary.csv", stringsAsFactors = FALSE)

cell_df_all$location <- factor(cell_df_all$location,
                               levels = c("NC", "PT", "LN", "ML", "MP", "MO"))
```

```{r fc-panel-a}
patients_A <- names(crc_patient_pal)
biopsies_A <- c("NC", "PT", "LN", "ML", "MP")

pat_dots <- data.frame(
  x = 2.6 + (seq_along(patients_A) - 1) * 1.2,
  y = 8.4,
  label = patients_A,
  col = crc_patient_pal[patients_A]
)
bio_dots <- data.frame(
  x = 2.6 + (seq_along(biopsies_A) - 1) * 1.5,
  y = 6.6,
  label = biopsies_A,
  col = crc_location_pal[biopsies_A]
)

win_tiles <- data.frame(
  xmin = 0.2 + (0:11) * 0.85,
  xmax = 0.2 + (0:11) * 0.85 + 0.72,
  ymin = 1.9,
  ymax = 2.3
)
feat_blocks <- data.frame(
  xmin = c(0.2, 1.8, 3.9, 5.5, 7.4, 9.0),
  xmax = c(1.3, 3.2, 4.9, 6.8, 8.4, 10.0),
  ymin = 0.3,
  ymax = 0.7
)

pA <- ggplot() +
  annotate("text", x = 0, y = 10.2, label = "Analysis design",
           hjust = 0, size = 3.2, fontface = "bold") +
  annotate("text", x = 0, y = 8.4, label = "patients", hjust = 0, size = 2.4) +
  geom_point(data = pat_dots, aes(x = x, y = y),
             color = pat_dots$col, size = 2.6) +
  geom_text(data = pat_dots, aes(x = x, y = y - 0.8, label = label),
            size = 1.6) +
  annotate("text", x = 0, y = 6.6, label = "biopsies", hjust = 0, size = 2.4) +
  geom_point(data = bio_dots, aes(x = x, y = y),
             color = bio_dots$col, size = 2.6) +
  geom_text(data = bio_dots, aes(x = x, y = y - 0.8, label = label),
            size = 1.6) +
  annotate("text", x = 0, y = 4.8,
           label = "unit = cells from one (patient, biopsy)",
           hjust = 0, size = 2.4) +
  annotate("text", x = 0.3, y = 3.9,
           label = "-> 1 adjS per region, 1 adjH per unit",
           hjust = 0, size = 2.4) +
  annotate("text", x = 0, y = 2.9,
           label = sprintf("windows (%s genomic tiles)",
                           format(nrow(embeds$row_data), big.mark = ",")),
           hjust = 0, size = 2.4) +
  geom_rect(data = win_tiles,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "grey40", color = NA) +
  annotate("text", x = 0, y = 1.3,
           label = sprintf("features (%d annotations)",
                           length(unique(ent$adjsampens$annotation))),
           hjust = 0, size = 2.4) +
  geom_rect(data = feat_blocks,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "grey40", color = NA) +
  coord_cartesian(xlim = c(0, 10.5), ylim = c(0.1, 11), clip = "off") +
  theme_void() +
  theme(plot.margin = margin(2, 2, 2, 2, "mm"))
```

```{r fc-panel-b}
mk_umap_df <- function(em, metric) {
  cd <- embeds$col_data[em$kept_cols, ]
  data.frame(
    UMAP1    = em$umap[, 1],
    UMAP2    = em$umap[, 2],
    location = cd$location,
    patient  = cd$patient,
    metric   = metric,
    stringsAsFactors = FALSE
  )
}

umap_df_B <- bind_rows(
  mk_umap_df(embeds$win_embeds$S,    "S"),
  mk_umap_df(embeds$win_embeds$adjS, "adjS")
)
umap_df_B$metric <- factor(umap_df_B$metric, levels = c("S", "adjS"))

p_loc <- ggplot(umap_df_B, aes(UMAP1, UMAP2, color = location)) +
  geom_point_rast(size = 0.2, alpha = 0.5, raster.dpi = 300) +
  scale_color_manual(values = crc_location_pal) +
  facet_wrap(~ metric, nrow = 1) +
  guides(color = guide_legend(override.aes = list(size = 1.3, alpha = 1),
                              nrow = 1)) +
  labs(x = "UMAP 1", y = "UMAP 2", color = "location") +
  theme_ng(base_size = 10) +
  theme(legend.position   = "bottom",
        legend.key.size   = unit(1.8, "mm"),
        legend.text       = element_text(size = 7),
        legend.title      = element_text(size = 8),
        legend.margin     = margin(0, 0, 0, 0),
        legend.box.margin = margin(-4, 0, 0, 0),
        plot.margin       = margin(2, 2, 0, 2, "mm"))

p_pat <- ggplot(umap_df_B, aes(UMAP1, UMAP2, color = patient)) +
  geom_point_rast(size = 0.2, alpha = 0.5, raster.dpi = 300) +
  scale_color_manual(values = crc_patient_pal) +
  facet_wrap(~ metric, nrow = 1) +
  guides(color = guide_legend(override.aes = list(size = 1.3, alpha = 1),
                              nrow = 1)) +
  labs(x = "UMAP 1", y = "UMAP 2", color = "patient") +
  theme_ng(base_size = 10) +
  theme(legend.position   = "bottom",
        legend.key.size   = unit(1.8, "mm"),
        legend.text       = element_text(size = 7),
        legend.title      = element_text(size = 8),
        legend.margin     = margin(0, 0, 0, 0),
        legend.box.margin = margin(-4, 0, 0, 0),
        plot.margin       = margin(0, 2, 2, 2, "mm"))

pB <- wrap_elements(p_loc / p_pat)
```

```{r fc-panel-c}
ve_win_long$assay <- factor(ve_win_long$assay,
                            levels = c("S", "adjS", "S_cellmeth", "methylation"))

pC <- ggplot(ve_win_long, aes(x = assay, y = median_ve, fill = variable)) +
  geom_col(position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = c("location" = "#0072B2", "patient" = "#E69F00")) +
  labs(x = "assay", y = "median % variance explained", fill = NULL) +
  theme_ng(base_size = 10) +
  theme(legend.key.size = unit(2, "mm"),
        legend.position = "top")
```

```{r fc-panel-d}
sample_df <- ent$adjsampens %>%
  filter(!is.na(patient), !is.na(location), !is.na(annotation)) %>%
  mutate(sample = paste0(patient, "_", location)) %>%
  group_by(annotation, sample, patient, location) %>%
  summarise(mean_adjS = mean(sampen, na.rm = TRUE), .groups = "drop")

write.csv(sample_df, "crc_fig_mean_adjS_by_sample.csv", row.names = FALSE)

mat_sample <- sample_df %>%
  select(annotation, sample, mean_adjS) %>%
  pivot_wider(names_from = sample, values_from = mean_adjS) %>%
  as.data.frame()
rownames(mat_sample) <- mat_sample$annotation
mat_sample$annotation <- NULL
mat_sample <- as.matrix(mat_sample)

col_meta_d <- sample_df %>%
  distinct(sample, patient, location)
col_meta_d <- col_meta_d[match(colnames(mat_sample), col_meta_d$sample), ]

col_fun_d <- colorRamp2(
  quantile(mat_sample, c(0.05, 0.5, 0.95), na.rm = TRUE),
  c("navy", "white", "firebrick")
)

ha_top <- HeatmapAnnotation(
  patient  = col_meta_d$patient,
  location = col_meta_d$location,
  col = list(
    patient  = crc_patient_pal[names(crc_patient_pal) %in% col_meta_d$patient],
    location = crc_location_pal[names(crc_location_pal) %in% col_meta_d$location]
  ),
  annotation_name_gp = gpar(fontsize = 7),
  annotation_legend_param = list(
    patient  = list(title_gp = gpar(fontsize = 8),
                    labels_gp = gpar(fontsize = 7)),
    location = list(title_gp = gpar(fontsize = 8),
                    labels_gp = gpar(fontsize = 7))
  )
)

annotation_category <- function(ann) {
  dplyr::case_when(
    grepl("^[0-9]+ ", ann)                         ~ "Segmentation",
    ann %in% c("H3K4me3", "H3K27me3", "H3K9me3")   ~ "Marks",
    ann %in% c("PMDs", "HMDs", "Lamin B1")         ~ "Domain",
    ann %in% c("Genes", "CpGi", "LINEs", "SINEs")  ~ "Features",
    grepl("^SCNA", ann)                             ~ "SCNA",
    TRUE                                            ~ NA_character_
  )
}

row_cat <- factor(annotation_category(rownames(mat_sample)),
                  levels = c("Segmentation", "Marks", "Domain",
                             "Features", "SCNA"))

ht_sample <- Heatmap(
  mat_sample,
  name            = "mean adjS",
  col             = col_fun_d,
  top_annotation  = ha_top,
  row_split       = row_cat,
  cluster_rows    = TRUE,
  cluster_row_slices = FALSE,
  cluster_columns = TRUE,
  show_row_names    = TRUE,
  show_column_names = TRUE,
  row_names_gp    = gpar(fontsize = 7),
  column_names_gp = gpar(fontsize = 6),
  row_title_gp    = gpar(fontsize = 8, fontface = "bold"),
  row_title_rot   = 0,
  heatmap_legend_param = list(
    title_gp  = gpar(fontsize = 8),
    labels_gp = gpar(fontsize = 7)
  )
)

pD <- wrap_elements(full = grid::grid.grabExpr(draw(ht_sample,
                                                    merge_legend = TRUE)))
```

```{r fc-panel-e}
pE <- ggplot(cell_df_all, aes(x = mean_meth, y = mean_adjS, color = location)) +
  geom_point(size = 0.5, alpha = 0.6) +
  geom_smooth(aes(group = location), method = "lm", se = FALSE,
              linewidth = 0.5) +
  scale_color_manual(values = crc_location_pal) +
  facet_wrap(~ patient, nrow = 1) +
  labs(x = "cell mean methylation", y = "cell mean adjS") +
  guides(x = guide_x_nolap(),
         color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  theme_ng(base_size = 10) +
  theme(legend.key.size = unit(2, "mm"),
        legend.text     = element_text(size = 8),
        legend.title    = element_text(size = 9))
```

```{r fc-panel-f}
de_nc <- de[c("pt_vs_nc", "ln_vs_nc", "ml_vs_nc", "mp_vs_nc")]

summarize_contrast <- function(df, alpha = 0.05) {
  sig <- df[df$adj_p < alpha, ]
  data.frame(
    upregulated   = sum(sig$moderated_t >= 0, na.rm = TRUE),
    downregulated = sum(sig$moderated_t < 0,  na.rm = TRUE)
  )
}

summary_table <- bind_rows(lapply(names(de_nc), function(nm) {
  cbind(contrast = nm, summarize_contrast(de_nc[[nm]]$coefs_df))
}))
write.csv(summary_table, "crc_de_counts.csv", row.names = FALSE)

contrast_short <- function(x) sub("_vs_nc$", "", x)

df_long_de <- pivot_longer(summary_table,
                           c("upregulated", "downregulated"),
                           names_to = "direction", values_to = "count") %>%
  mutate(contrast = contrast_short(contrast))

pF <- ggplot(df_long_de, aes(x = contrast, y = count + 1, fill = direction)) +
  geom_col(position = position_dodge(width = 0.8)) +
  geom_text(aes(label = count, y = count + 1),
            position = position_dodge(width = 0.8),
            vjust = -0.3, size = 2.2) +
  scale_y_log10() +
  scale_fill_manual(values = c(upregulated   = "#1b9e77",
                               downregulated = "#d95f02")) +
  labs(x = "contrast (vs NC)", y = "windows (log scale)", fill = NULL) +
  theme_ng(base_size = 9) +
  theme(legend.position   = "top",
        legend.key.size   = unit(2.5, "mm"),
        legend.text       = element_text(size = 7),
        legend.margin     = margin(0, 0, 0, 0),
        legend.box.margin = margin(0, 0, -2, 0))
```

```{r fc-panel-g}
rd          <- embeds$row_data
all_regions <- rownames(rd)

ann_pretty <- c(
  "pmds_pmd"                  = "PMDs",
  "hmds_pmd"                  = "HMDs",
  "H3K27me3_chip"             = "H3K27me3",
  "H3K9me3_chip"              = "H3K9me3",
  "H3K4me3_chip"              = "H3K4me3",
  "laminb1_lad"               = "Lamin B1",
  "genes_genes"               = "Genes",
  "cpgIslandExt_cpgIslandExt" = "CpGi",
  "crc01_gain_scna_scna"      = "SCNA gain",
  "crc01_lost_scna_scna"      = "SCNA loss"
)
annot_features <- paste0(names(ann_pretty), "_bin")

prop_df <- bind_rows(lapply(names(de_nc), function(nm) {
  cf   <- de_nc[[nm]]$coefs_df
  up   <- cf$region[cf$moderated_t >= 0 & cf$adj_p < 0.05]
  down <- cf$region[cf$moderated_t < 0  & cf$adj_p < 0.05]
  is_up   <- all_regions %in% up
  is_down <- all_regions %in% down

  bind_rows(lapply(annot_features, function(ann) {
    ann_vec <- rd[[ann]]
    if (is.null(ann_vec)) return(NULL)
    data.frame(
      annotation = gsub("_bin$", "", ann),
      contrast   = nm,
      direction  = c("up", "down"),
      proportion = c(
        if (sum(is_up)   > 0) sum(ann_vec[is_up]   == 1, na.rm = TRUE) / sum(is_up)   else NA_real_,
        if (sum(is_down) > 0) sum(ann_vec[is_down] == 1, na.rm = TRUE) / sum(is_down) else NA_real_
      ),
      stringsAsFactors = FALSE
    )
  }))
})) %>%
  filter(!is.na(proportion)) %>%
  mutate(contrast_dir = paste0(contrast_short(contrast), " (", direction, ")"),
         annotation   = factor(annotation,
                               levels = names(ann_pretty),
                               labels = unname(ann_pretty)))

write.csv(prop_df, "crc_de_annotation_overlap.csv", row.names = FALSE)

pG <- ggplot(prop_df, aes(x = contrast_dir, y = annotation)) +
  geom_point(aes(size = proportion, color = proportion)) +
  scale_size(range = c(0.8, 5)) +
  scale_color_gradient(low = "grey80", high = "red") +
  guides(size  = guide_legend(direction = "horizontal", title.position = "top",
                              nrow = 1),
         color = guide_colorbar(direction = "horizontal", title.position = "top",
                                barheight = unit(2, "mm"),
                                barwidth  = unit(28, "mm"))) +
  labs(x = "contrast (vs NC)", y = "annotation",
       size = "proportion", color = "proportion") +
  theme_ng(base_size = 9) +
  theme(aspect.ratio      = NULL,
        legend.position   = "bottom",
        legend.box        = "horizontal",
        legend.key.size   = unit(2.8, "mm"),
        legend.text       = element_text(size = 8),
        legend.title      = element_text(size = 9),
        legend.spacing.x  = unit(2, "mm"),
        legend.margin     = margin(0, 0, 0, 0),
        legend.box.margin = margin(-2, 0, 0, 0))
```

```{r fc-panel-h}
write.csv(driver_df, "crc_driver_sd_range.csv", row.names = FALSE)

pH <- plot_driver_scatter(driver_df,
                          x_label = "SD of adjH across biopsy locations",
                          y_label = "SD of adjS across biopsy locations") +
  guides(color = guide_legend(nrow = 2, byrow = TRUE,
                              title.position = "top")) +
  theme_ng(base_size = 9) +
  theme(legend.position   = "bottom",
        legend.key.size   = unit(2.5, "mm"),
        legend.text       = element_text(size = 7),
        legend.title      = element_text(size = 8),
        legend.margin     = margin(0, 0, 0, 0),
        legend.box.margin = margin(-2, 0, 0, 0))
```

# Half-A4 compact figure (primary)

Panels keep the A-H tag order but row composition is tight: row 1 = A,B,C;
row 2 = D,E (E reshaped with facet_wrap nrow=2); row 3 = F,G,H (G at ~50%
width of F and H).

```{r fc-panel-a-v2}
pA_v2 <- ggplot() +
  annotate("text", x = 0, y = 9.3, label = "Colon\ncancer",
           hjust = 0, vjust = 1, size = 2.4, fontface = "bold",
           lineheight = 0.9) +
  annotate("text", x = 0, y = 7.2,
           label = sprintf("%d patients,\n%d biopsies",
                           length(patients_A), length(biopsies_A)),
           hjust = 0, vjust = 1, size = 2.0, lineheight = 0.9) +
  annotate("text", x = 0, y = 5.3,
           label = "unit = cells\nfrom one\n(patient, biopsy)",
           hjust = 0, vjust = 1, size = 2.0, lineheight = 0.9) +
  annotate("text", x = 0, y = 2.6,
           label = "1 adjS per\nregion,\n1 adjH per\nunit",
           hjust = 0, vjust = 1, size = 2.0, lineheight = 0.9) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 10), clip = "off") +
  theme_void() +
  theme(plot.margin = margin(2, 1, 2, 1, "mm"))
```

```{r fc-panel-b-v2}
mk_compact_umap <- function(df, color_col, pal, title_lab) {
  ggplot(df, aes(UMAP1, UMAP2, color = .data[[color_col]])) +
    geom_point_rast(size = 0.2, alpha = 0.5, raster.dpi = 300) +
    scale_color_manual(values = pal) +
    facet_wrap(~ metric, nrow = 1) +
    guides(color = guide_legend(override.aes = list(size = 1.2, alpha = 1),
                                nrow = 1)) +
    labs(x = "UMAP 1", y = "UMAP 2", color = title_lab) +
    theme_ng(base_size = 7) +
    theme(legend.position     = "bottom",
          legend.key.size     = unit(1.2, "mm"),
          legend.text         = element_text(size = 5.5),
          legend.title        = element_text(size = 6.5),
          legend.margin       = margin(0, 0, 0, 0),
          legend.box.margin   = margin(-6, 0, -2, 0),
          legend.box.spacing  = unit(0.5, "mm"),
          legend.spacing.y    = unit(0, "mm"),
          strip.text          = element_text(size = 6.5,
                                             margin = margin(1, 0, 1, 0)),
          panel.spacing       = unit(0.8, "mm"),
          plot.margin         = margin(0, 0, 0, 0, "mm"))
}

p_loc_v2 <- mk_compact_umap(umap_df_B, "location",
                            crc_location_pal, "location")
p_pat_v2 <- mk_compact_umap(umap_df_B, "patient",
                            crc_patient_pal,  "patient")

pB_v2 <- wrap_elements(full = (p_loc_v2 / p_pat_v2) +
                         plot_layout(heights = c(1, 1)))
```

```{r fc-panel-c-v2}
pC_v2 <- ggplot(ve_win_long, aes(x = assay, y = median_ve, fill = variable)) +
  geom_col(position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = c("location" = "#0072B2", "patient" = "#E69F00")) +
  labs(x = "assay", y = "median % var. explained", fill = NULL) +
  theme_ng(base_size = 7) +
  theme(legend.key.size    = unit(1.5, "mm"),
        legend.position    = "top",
        legend.text        = element_text(size = 6),
        legend.margin      = margin(0, 0, 0, 0),
        legend.box.margin  = margin(0, 0, -4, 0),
        legend.box.spacing = unit(0.5, "mm"),
        plot.margin        = margin(0, 0, 0, 0, "mm"))
```

```{r fc-panel-d-v2}
ht_sample_v2 <- Heatmap(
  mat_sample,
  name            = "mean adjS",
  col             = col_fun_d,
  top_annotation  = HeatmapAnnotation(
    patient  = col_meta_d$patient,
    location = col_meta_d$location,
    col = list(
      patient  = crc_patient_pal[names(crc_patient_pal) %in% col_meta_d$patient],
      location = crc_location_pal[names(crc_location_pal) %in% col_meta_d$location]
    ),
    simple_anno_size = unit(2, "mm"),
    annotation_name_gp = gpar(fontsize = 5),
    show_legend = FALSE
  ),
  row_split       = row_cat,
  cluster_rows    = TRUE,
  cluster_row_slices = FALSE,
  cluster_columns = TRUE,
  show_row_names    = TRUE,
  show_column_names = TRUE,
  row_names_gp    = gpar(fontsize = 4.5),
  column_names_gp = gpar(fontsize = 4.5),
  row_title_gp    = gpar(fontsize = 6, fontface = "bold"),
  row_title_rot   = 0,
  row_gap         = unit(0.8, "mm"),
  heatmap_legend_param = list(
    title_gp    = gpar(fontsize = 6),
    labels_gp   = gpar(fontsize = 5),
    grid_height = unit(2, "mm"),
    grid_width  = unit(2, "mm")
  )
)

pD_v2 <- wrap_elements(full = grid::grid.grabExpr(draw(ht_sample_v2,
                                                       merge_legend = TRUE)))
```

```{r fc-panel-e-v2}
pE_v2 <- ggplot(cell_df_all, aes(x = mean_meth, y = mean_adjS, color = location)) +
  geom_point(size = 0.25, alpha = 0.6) +
  geom_smooth(aes(group = location), method = "lm", se = FALSE,
              linewidth = 0.35) +
  scale_color_manual(values = crc_location_pal) +
  facet_wrap(~ patient, nrow = 2) +
  labs(x = "cell mean methylation", y = "cell mean adjS") +
  guides(x = guide_x_nolap(),
         color = guide_legend(override.aes = list(size = 1.5, alpha = 1),
                              nrow = 1)) +
  theme_ng(base_size = 7) +
  theme(legend.position    = "bottom",
        legend.key.size    = unit(1.2, "mm"),
        legend.text        = element_text(size = 6),
        legend.title       = element_text(size = 7),
        legend.margin      = margin(0, 0, 0, 0),
        legend.box.margin  = margin(-6, 0, -2, 0),
        legend.box.spacing = unit(0.5, "mm"),
        strip.text         = element_text(size = 6,
                                          margin = margin(1, 0, 1, 0)),
        panel.spacing      = unit(0.8, "mm"),
        plot.margin        = margin(0, 0, 0, 0, "mm"))
```

```{r fc-panel-f-v2}
pF_v2 <- ggplot(df_long_de, aes(x = contrast, y = count + 1, fill = direction)) +
  geom_col(position = position_dodge(width = 0.8)) +
  geom_text(aes(label = count, y = count + 1),
            position = position_dodge(width = 0.8),
            vjust = -0.3, size = 1.8) +
  scale_y_log10() +
  scale_fill_manual(values = c(upregulated   = "#1b9e77",
                               downregulated = "#d95f02")) +
  labs(x = "contrast (vs NC)", y = "windows (log)", fill = NULL) +
  theme_ng(base_size = 7) +
  theme(legend.position    = "top",
        legend.key.size    = unit(1.5, "mm"),
        legend.text        = element_text(size = 5.5),
        legend.margin      = margin(0, 0, 0, 0),
        legend.box.margin  = margin(0, 0, -4, 0),
        legend.box.spacing = unit(0.5, "mm"),
        plot.margin        = margin(0, 0, 0, 0, "mm"))
```

```{r fc-panel-g-v2}
prop_df_v2 <- prop_df %>%
  mutate(contrast_dir = paste0(contrast_short(as.character(contrast)),
                               ifelse(direction == "up", "+", "-")))

pG_v2 <- ggplot(prop_df_v2, aes(x = contrast_dir, y = annotation)) +
  geom_point(aes(size = proportion, color = proportion)) +
  scale_size(range = c(0.4, 2.5)) +
  scale_color_gradient(low = "grey80", high = "red") +
  guides(size  = "none",
         color = guide_colorbar(direction = "vertical", title.position = "top",
                                barheight = unit(12, "mm"),
                                barwidth  = unit(1.8, "mm"))) +
  labs(x = "contrast (vs NC)", y = NULL,
       color = "prop.") +
  theme_ng(base_size = 6.5) +
  theme(aspect.ratio       = NULL,
        legend.position    = "right",
        legend.key.size    = unit(1.2, "mm"),
        legend.text        = element_text(size = 5),
        legend.title       = element_text(size = 6),
        legend.margin      = margin(0, 0, 0, 0),
        legend.box.margin  = margin(0, 0, 0, -4),
        legend.box.spacing = unit(0.5, "mm"),
        axis.text.y        = element_text(size = 5.5),
        axis.text.x        = element_text(size = 5.5, angle = 45,
                                          hjust = 1, vjust = 1),
        plot.margin        = margin(0, 0, 0, 0, "mm"))
```

```{r fc-panel-h-v2}
pH_v2 <- plot_driver_scatter(driver_df,
                             x_label = "SD adjH (locations)",
                             y_label = "SD adjS (locations)") +
  guides(color = guide_legend(nrow = 1, title.position = "left"),
         shape = guide_legend(nrow = 1, title.position = "left")) +
  theme_ng(base_size = 7) +
  theme(legend.position    = "bottom",
        legend.key.size    = unit(1.2, "mm"),
        legend.text        = element_text(size = 5.5),
        legend.title       = element_text(size = 6.5),
        legend.margin      = margin(0, 0, 0, 0),
        legend.box.margin  = margin(-6, 0, -2, 0),
        legend.box.spacing = unit(0.5, "mm"),
        plot.margin        = margin(0, 0, 0, 0, "mm"))

pH_v2$layers[[which(sapply(pH_v2$layers, function(l)
  inherits(l$geom, "GeomTextRepel")))]] <- NULL
pH_v2 <- pH_v2 +
  ggrepel::geom_text_repel(size = 1.6, max.overlaps = Inf,
                           min.segment.length = 0,
                           segment.size = 0.2, box.padding = 0.25,
                           force = 3)
```

```{r single-page-combined-v2, fig.width = 8.27, fig.height = 8.5}
row1_v2 <- pA_v2 + pB_v2 + pC_v2 + plot_layout(widths = c(0.5, 5.0, 1.1))
row2_v2 <- pD_v2 + pE_v2 + plot_layout(widths = c(1, 1))
row3_v2 <- pF_v2 + pG_v2 + pH_v2 + plot_layout(widths = c(2, 1, 2))

compact_fig <- row1_v2 / row2_v2 / row3_v2 +
  plot_layout(heights = c(2.3, 1.4, 1)) +
  plot_annotation(tag_levels = "A",
                  theme = theme(
                    plot.tag          = element_text(size = 9, face = "bold"),
                    plot.tag.position = c(0.01, 0.99),
                    plot.margin       = margin(1, 1, 1, 1, "mm")
                  ))

print(compact_fig)

ggsave(
  file.path(out_dir, "fig_crc_half_a4.pdf"),
  plot   = compact_fig,
  width  = 8.27,
  height = 8.5,
  units  = "in",
  device = cairo_pdf
)
```

# Legacy tall single-page figure

```{r single-page-combined, fig.width = 8.27, fig.height = 18.5}
design <- "
AABBBB
AABBBB
CCDDDD
EEEEEE
FFFHHH
#GGGG#
"

single_page_fig <- pA + pB + pC + pD + pE + pF + pG + pH +
  plot_layout(design  = design,
              heights = c(3.5, 3.5, 3.0, 2.0, 3.5, 3.0)) +
  plot_annotation(tag_levels = "A",
                  theme = theme(plot.tag = element_text(size = 10,
                                                         face = "bold")))

print(single_page_fig)

ggsave(
  file.path(out_dir, "fig_crc_single_page.pdf"),
  plot   = single_page_fig,
  width  = 8.27,
  height = 18.5,
  units  = "in",
  device = cairo_pdf
)
```
