---
title: "Argelaguet figure"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    code_folding: hide
    number_sections: no
---

```{r setup}
suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
  library(dplyr)
  library(tidyr)
  library(ComplexHeatmap)
  library(circlize)
  library(ggrastr)
  library(ggrepel)
  library(viridisLite)
  library(grid)
})

source("rules/src/ggtheme.R")
source("rules/src/palettes.R")
source("rules/src/driver_utils.R")

knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)

log_con <- file(snakemake@log[[1]], open = "wt")
sink(log_con)
sink(log_con, type = "message")

out_dir <- snakemake@params$out_dir
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

```

```{r load}
cell_matrices  <- readRDS("argelaguet_cell_matrices.rds")
ent            <- readRDS("argelaguet_entropy.rds")
grp_meta       <- readRDS("argelaguet_groups_meta.rds")
cell_umap_adjS <- readRDS("argelaguet_umap_cell_adjS.rds")
```

## Panel A: Cell-level UMAP colored by developmental stage

```{r panel-a}
pa <- ggplot(cell_umap_adjS, aes(x = UMAP1, y = UMAP2, color = stage)) +
  geom_point_rast(size = 0.8, alpha = 0.7, raster.dpi = 300) +
  scale_color_manual(values = argelaguet_stage_pal) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(x = "UMAP 1", y = "UMAP 2", color = "stage") +
  theme_ng(base_size = 8)
```

## Panel B: Median adjS per annotation across developmental stages

```{r panel-b}
tc_adjS <- ent$adjsampens %>%
  group_by(stage, annotation) %>%
  summarise(
    median_adjS = median(sampen, na.rm = TRUE),
    lo  = quantile(sampen, 0.25, na.rm = TRUE),
    hi  = quantile(sampen, 0.75, na.rm = TRUE),
    .groups = "drop"
  )
write.csv(tc_adjS, "argelaguet_fig_adjS_by_stage_annotation.csv", row.names = FALSE)

pb <- ggplot(tc_adjS, aes(x = stage, y = median_adjS, color = annotation, group = annotation)) +
  geom_line(linewidth = 0.5) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = annotation), alpha = 0.1, color = NA) +
  scale_color_manual(values = argelaguet_annotation_pal) +
  scale_fill_manual(values = argelaguet_annotation_pal) +
  labs(x = "stage", y = "median adjS", color = "annotation", fill = "annotation") +
  theme_ng(base_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.key.size = unit(0.3, "cm"))
```

## Panel C: Median adjH per annotation across developmental stages

```{r panel-c}
tc_adjH <- ent$adjshannons %>%
  group_by(stage, annotation) %>%
  summarise(
    median_adjH = median(median_shannon, na.rm = TRUE),
    lo  = quantile(median_shannon, 0.25, na.rm = TRUE),
    hi  = quantile(median_shannon, 0.75, na.rm = TRUE),
    .groups = "drop"
  )
write.csv(tc_adjH, "argelaguet_fig_adjH_by_stage_annotation.csv", row.names = FALSE)

pc <- ggplot(tc_adjH, aes(x = stage, y = median_adjH, color = annotation, group = annotation)) +
  geom_line(linewidth = 0.5) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = annotation), alpha = 0.1, color = NA) +
  scale_color_manual(values = argelaguet_annotation_pal) +
  scale_fill_manual(values = argelaguet_annotation_pal) +
  labs(x = "stage", y = "median adjH", color = "annotation", fill = "annotation") +
  theme_ng(base_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.key.size = unit(0.3, "cm"))
```

## Panel D: Per-cell adjS at enhancers vs promoters

```{r panel-d}
enh_prom <- cell_matrices$cells_adjS_wide %>%
  select(stage, enh = `enh-E75-union`, prom = promoters) %>%
  filter(!is.na(enh), !is.na(prom))

pd <- ggplot(enh_prom, aes(x = prom, y = enh, color = stage)) +
  geom_point_rast(size = 0.5, alpha = 0.5, raster.dpi = 300) +
  scale_color_manual(values = argelaguet_stage_pal) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(x = "adjS promoters", y = "adjS enhancers", color = "stage") +
  theme_ng(base_size = 8)
```

## Panel E: adjS variability (IQR) per lineage

```{r panel-e}
var_lin <- ent$adjsampens %>%
  group_by(lineage) %>%
  summarise(iqr = IQR(sampen, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(iqr)) %>%
  mutate(lineage = factor(lineage, levels = lineage))
write.csv(var_lin, "argelaguet_fig_adjS_iqr_by_lineage.csv", row.names = FALSE)

pe <- ggplot(var_lin, aes(x = lineage, y = iqr)) +
  geom_col(fill = "#35b779") +
  coord_flip() +
  labs(x = NULL, y = "IQR adjS") +
  theme_ng(base_size = 8)
```

## Panel F: Driver categorization (adjH vs adjS across lineage class)

For each annotation, SDs of group-level median adjH and adjS are computed
across lineage classes and categorized (see driver_utils.R).

```{r panel-f-driver}
driver_df_arg <- categorize_drivers(grp_meta$all_grp_meta, "lineage_class")
write.csv(driver_df_arg, "argelaguet_driver_sd_range.csv", row.names = FALSE)

pf <- plot_driver_scatter(driver_df_arg,
                          x_label = "SD of adjH across lineage classes",
                          y_label = "SD of adjS across lineage classes") +
  theme_ng(base_size = 8) +
  theme(legend.key.size = unit(2, "mm"),
        legend.text     = element_text(size = 6),
        legend.title    = element_text(size = 7))
```

## Heatmap: Median adjS per lineage and annotation

```{r heatmap}
heat_df <- grp_meta$all_grp_meta %>%
  group_by(lineage, annotation, lineage_class) %>%
  summarise(median_adjS = median(median_adjS, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(lineage_class))

wide <- heat_df %>%
  pivot_wider(names_from = annotation, values_from = median_adjS)

mat <- wide %>% select(-lineage, -lineage_class) %>% as.matrix()
rownames(mat) <- wide$lineage

brk <- c(min(mat, na.rm = TRUE),
          mean(range(mat, na.rm = TRUE)),
          max(mat, na.rm = TRUE))

lc_pal <- argelaguet_lineage_class_pal[sort(unique(wide$lineage_class))]

ht <- Heatmap(
  mat,
  name              = "median adjS",
  col               = colorRamp2(brk, c("navy", "white", "firebrick")),
  row_split         = wide$lineage_class,
  cluster_rows      = TRUE,
  cluster_columns   = FALSE,
  show_row_names    = TRUE,
  show_column_names = TRUE,
  row_names_gp      = gpar(fontsize = 5),
  column_names_gp   = gpar(fontsize = 7),
  row_names_max_width = unit(4, "cm"),
  right_annotation  = rowAnnotation(
    lineage_class = wide$lineage_class,
    col = list(lineage_class = lc_pal),
    annotation_legend_param = list(lineage_class = list(title = "lineage class"))
  ),
  column_title = "median adjS per lineage and annotation"
)
```

## Combined figure

- A: Cell-level UMAP (adjS across annotations) colored by developmental stage.
- B: Median adjS per annotation across stages (line plot with IQR ribbon).
- C: Median adjH per annotation across stages (line plot with IQR ribbon).
- D: Per-cell adjS at enhancers (E7.5 union) vs promoters.
- E: adjS variability (IQR) per lineage.
- F: Driver categorization (SD adjH vs SD adjS across lineage classes).
- Heatmap (separate): median adjS per lineage and annotation, split by lineage class.

```{r combined, fig.width = ng_fig_size(4, 2, panel_mm = 50)$w, fig.height = ng_fig_size(4, 2, panel_mm = 50)$h}
legend_theme <- theme(
  legend.key.size   = unit(2, "mm"),
  legend.text       = element_text(size = 5),
  legend.title      = element_text(size = 6),
  legend.position   = "right"
)
pb <- pb + guides(color = guide_legend(ncol = 1, override.aes = list(linewidth = 1)),
                  fill  = guide_legend(ncol = 1)) + legend_theme
pc <- pc + guides(color = guide_legend(ncol = 1, override.aes = list(linewidth = 1)),
                  fill  = guide_legend(ncol = 1)) + legend_theme
pe <- pe + theme(axis.text.y = element_text(size = 6))

fig_w <- ng_fig_size(4, 2, panel_mm = 50)$w
fig_h <- ng_fig_size(4, 2, panel_mm = 50)$h

gg_fig <- (pa | pb | pc) / (pd | pe | pf) +
  plot_layout(heights = c(1, 0.9)) +
  plot_annotation(tag_levels = "A",
                  theme = theme(plot.tag = element_text(size = 8, face = "bold")))

print(gg_fig)

ggsave(
  file.path(out_dir, "fig_argelaguet_panels.pdf"),
  plot   = gg_fig,
  width  = fig_w,
  height = fig_h,
  units  = "in",
  device = cairo_pdf
)
```

```{r heatmap-save, fig.width = 14, fig.height = 10}
draw(ht)
pdf(file.path(out_dir, "fig_argelaguet_heatmap.pdf"), width = 14, height = 10)
draw(ht)
invisible(dev.off())
```
