---
title: "Ecker entropy figure"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    code_folding: hide
    number_sections: no
---

```{r setup}
suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
  library(dplyr)
  library(tidyr)
  library(ComplexHeatmap)
  library(circlize)
  library(ggrastr)
  library(ggrepel)
  library(viridisLite)
  library(grid)
})

source("rules/src/ggtheme.R")
source("rules/src/palettes.R")
source("rules/src/driver_utils.R")

knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)

log_con <- file(snakemake@log[[1]], open = "wt")
sink(log_con)
sink(log_con, type = "message")

out_dir <- snakemake@params$out_dir
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
```

```{r load}
meta         <- readRDS("ecker_groups_meta.rds")
ent          <- readRDS("ecker_entropy.rds")
umap_overlay <- readRDS("ecker_paper_umap_overlay.rds")
win_umap     <- readRDS("ecker_umap_windows_adjS.rds")

ve_long <- if (file.exists("ecker_win_varexp.csv"))
             read.csv("ecker_win_varexp.csv", stringsAsFactors = FALSE) else NULL
cell_df <- if (file.exists("ecker_per_cell_summary.csv"))
             read.csv("ecker_per_cell_summary.csv", stringsAsFactors = FALSE) else NULL
```

```{r embedding_diagnostics_report}
diag_path <- "ecker_embedding_diagnostics.csv"
if (file.exists(diag_path)) {
  cat("Ecker embedding diagnostics (from ecker_embeddings.Rmd):\n")
  print(read.csv(diag_path, stringsAsFactors = FALSE))
}
if (!is.null(ve_long)) {
  cat("\nPer-window variance explained by cell_class and sub_type:\n")
  print(ve_long)
}
```

```{r paper-umap-data}
umap_cells <- umap_overlay %>%
  filter(annotation == "Genes") %>%
  select(sub_type,
         cell_class = cell_class.x,
         major_type = major_type.x,
         umap1, umap2,
         median_adjS, median_adjH, median_meth) %>%
  distinct(sub_type, umap1, umap2, .keep_all = TRUE)
```

# Half-A4 compact figure (primary)

A) sampling and feature schema;
B) paper UMAP coloured by cell class, adjS, adjH, methylation;
C) our windows-based UMAP coloured by cell class for adjS and methylation;
D) median adjS heatmap per cell type and annotation;
E) per-cell mean methylation vs mean adjS;
F) median variance explained per assay by cell_class and sub_type;
G) R-squared of adjS by cell_class per annotation;
H) driver categorization (SD adjH vs SD adjS across cell classes).

```{r fe-panel-a-v2}
schema_classes <- c("Exc", "Inh", "NonN")
schema_layout <- list(
  Exc  = c("CT-L6", "IT-L23", "PT-L5"),
  Inh  = c("MGE-Sst", "CGE-Vip"),
  NonN = c("ASC", "OPC")
)
schema_classes <- intersect(schema_classes, names(ecker_cell_class_pal))
schema_layout  <- schema_layout[schema_classes]
schema_layout  <- lapply(seq_along(schema_layout), function(i) {
  cls <- schema_classes[i]
  mts <- intersect(schema_layout[[i]], names(ecker_major_type_pal))
  if (length(mts) == 0) {
    keep <- meta$all_grp_meta %>%
      filter(cell_class == cls) %>%
      pull(major_type) %>% unique()
    mts <- intersect(keep, names(ecker_major_type_pal))[1:2]
  }
  data.frame(cell_class = cls, major_type = mts, stringsAsFactors = FALSE)
})
schema_cells <- do.call(rbind, schema_layout)
schema_cells$col       <- seq_len(nrow(schema_cells))
schema_cells$cls_color <- ecker_cell_class_pal[schema_cells$cell_class]
schema_cells$mt_color  <- ecker_major_type_pal[schema_cells$major_type]

schema_regions <- c("windows", "CpG isl.", "HMM", "marks")
n_schema_regions <- length(schema_regions)
n_schema_cols <- nrow(schema_cells)

set.seed(2)
schema_grid <- expand.grid(col = seq_len(n_schema_cols),
                           row = seq_len(n_schema_regions))
schema_grid$val <- runif(nrow(schema_grid), 0.1, 0.95)

bar_mt  <- n_schema_regions + 0.9
bar_cls <- n_schema_regions + 1.55
ellipsis_y <- 0
cls_centers <- tapply(schema_cells$col, schema_cells$cell_class, mean)

pA_v2 <- ggplot() +
  geom_tile(data = schema_grid,
            aes(x = col, y = row, fill = val),
            colour = "grey70", linewidth = 0.15) +
  scale_fill_gradient(low = "#FFFFFF", high = "#444444",
                      guide = "none", limits = c(0, 1)) +
  geom_tile(data = schema_cells,
            aes(x = col, y = bar_cls),
            fill = schema_cells$cls_color,
            height = 0.45, width = 0.95, inherit.aes = FALSE) +
  geom_tile(data = schema_cells,
            aes(x = col, y = bar_mt),
            fill = schema_cells$mt_color,
            height = 0.45, width = 0.95, inherit.aes = FALSE) +
  annotate("text",
           x = cls_centers, y = bar_cls + 0.5,
           label = names(cls_centers),
           size = 1.7) +
  annotate("text", x = -0.15, y = bar_cls,
           label = "cell class", hjust = 1, size = 1.7) +
  annotate("text", x = -0.15, y = bar_mt,
           label = "major type", hjust = 1, size = 1.7) +
  annotate("text", x = -0.15, y = seq_len(n_schema_regions),
           label = rev(schema_regions),
           hjust = 1, size = 1.7) +
  annotate("text", x = 1, y = ellipsis_y,
           label = "(...)", hjust = 0, size = 2.0) +
  annotate("text", x = n_schema_cols + 0.6, y = bar_cls,
           label = "(...)", hjust = 0, size = 2.0) +
  annotate("text", x = (n_schema_cols + 1) / 2, y = bar_cls + 1.1,
           label = "cells", size = 1.9) +
  annotate("text", x = -3.0, y = (n_schema_regions + 1) / 2,
           label = "regions", angle = 90, size = 1.9) +
  annotate("segment",
           x = n_schema_cols + 0.7, xend = n_schema_cols + 1.5,
           y = (n_schema_regions + 1) / 2,
           yend = (n_schema_regions + 1) / 2,
           arrow = arrow(length = unit(1.2, "mm"), type = "closed"),
           linewidth = 0.3) +
  annotate("text", x = n_schema_cols + 1.7,
           y = (n_schema_regions + 1) / 2,
           label = "adjS", hjust = 0, size = 2.0) +
  annotate("segment",
           x = (n_schema_cols + 1) / 2,
           xend = (n_schema_cols + 1) / 2,
           y = ellipsis_y - 0.4, yend = ellipsis_y - 1.05,
           arrow = arrow(length = unit(1.2, "mm"), type = "closed"),
           linewidth = 0.3) +
  annotate("text", x = (n_schema_cols + 1) / 2, y = ellipsis_y - 1.35,
           label = "adjH", size = 2.0) +
  coord_fixed(xlim = c(-3.4, n_schema_cols + 3.0),
              ylim = c(ellipsis_y - 1.7, bar_cls + 1.4),
              clip = "off") +
  theme_void(base_family = "Helvetica") +
  theme(plot.margin = margin(1, 1, 1, 1, "mm"))
```

```{r fe-panel-b-v2}
mk_paper_umap <- function(color_col, pal, lab, viridis_opt = NULL) {
  p <- ggplot(umap_cells, aes(x = umap1, y = umap2)) +
    geom_point_rast(aes(color = .data[[color_col]]),
                    size = 0.18, alpha = 0.6, raster.dpi = 300) +
    labs(x = "UMAP 1", y = "UMAP 2", color = lab) +
    theme_ng(base_size = 7) +
    theme(legend.key.size    = unit(1.2, "mm"),
          legend.text        = element_text(size = 5.5),
          legend.title       = element_text(size = 6.5),
          legend.margin      = margin(0, 0, 0, 0),
          legend.box.margin  = margin(-3, 0, -2, 0),
          legend.box.spacing = unit(0.2, "mm"),
          plot.margin        = margin(0, 1, 0, 1, "mm"))
  if (!is.null(pal)) {
    p <- p + scale_color_manual(values = pal) +
      guides(color = guide_legend(override.aes = list(size = 1.0, alpha = 1),
                                  ncol = 2))
  } else {
    p <- p + scale_color_viridis_c(option = viridis_opt %||% "viridis",
                                   n.breaks = 3) +
      guides(color = guide_colorbar(barwidth = unit(12, "mm"),
                                    barheight = unit(1.2, "mm")))
  }
  p
}

`%||%` <- function(a, b) if (is.null(a)) b else a

p_b_class <- mk_paper_umap("cell_class", ecker_cell_class_pal, "cell class")
p_b_adjS  <- mk_paper_umap("median_adjS", NULL, "adjS",  viridis_opt = "viridis")
p_b_adjH  <- mk_paper_umap("median_adjH", NULL, "adjH",  viridis_opt = "plasma")
p_b_meth  <- mk_paper_umap("median_meth", NULL, "meth.", viridis_opt = "mako")

pB_v2 <- wrap_elements(full = (p_b_class + p_b_adjS) /
                                (p_b_adjH + p_b_meth) +
                                plot_layout(heights = c(1, 1)))
```

```{r fe-panel-c-v2}
mk_our_umap <- function(em, label) {
  if (is.null(em)) return(NULL)
  df <- data.frame(
    UMAP1      = em$umap[, 1],
    UMAP2      = em$umap[, 2],
    cell_class = win_umap$col_data$cell_class[em$kept_cols],
    assay      = label,
    stringsAsFactors = FALSE
  )
  ggplot(df, aes(UMAP1, UMAP2, color = cell_class)) +
    geom_point_rast(size = 0.18, alpha = 0.6, raster.dpi = 300) +
    scale_color_manual(values = ecker_cell_class_pal) +
    guides(color = guide_legend(override.aes = list(size = 1.0, alpha = 1),
                                ncol = 2)) +
    labs(x = "UMAP 1", y = "UMAP 2", color = "cell class",
         subtitle = paste0("ours: ", label)) +
    theme_ng(base_size = 7) +
    theme(plot.subtitle      = element_text(size = 6.5, hjust = 0.5,
                                             margin = margin(0, 0, 1, 0)),
          legend.position    = "bottom",
          legend.key.size    = unit(1.2, "mm"),
          legend.text        = element_text(size = 5.5),
          legend.title       = element_text(size = 6.5),
          legend.margin      = margin(0, 0, 0, 0),
          legend.box.margin  = margin(-6, 0, -2, 0),
          legend.box.spacing = unit(0.2, "mm"),
          plot.margin        = margin(0, 1, 0, 1, "mm"))
}

p_c_adjS <- mk_our_umap(win_umap$win_embeds$adjS,        "adjS")
p_c_meth <- mk_our_umap(win_umap$win_embeds$methylation, "meth.")
pC_v2 <- wrap_elements(full = (p_c_adjS + p_c_meth) +
                                plot_layout(guides = "collect") &
                              theme(legend.position = "bottom"))
```

```{r fe-panel-d-v2-data}
grp <- meta$all_grp_meta

ht_wide <- grp %>%
  select(sub_type, annotation, cell_class, major_type, median_adjS) %>%
  group_by(sub_type, annotation, cell_class, major_type) %>%
  summarise(median_adjS = median(median_adjS, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = annotation, values_from = median_adjS)

mat <- ht_wide %>% select(-sub_type, -cell_class, -major_type) %>% as.matrix()
rownames(mat) <- ht_wide$sub_type

cc_ht_pal <- ecker_cell_class_pal[sort(unique(ht_wide$cell_class))]
mt_pal    <- ecker_major_type_pal[sort(unique(ht_wide$major_type))]

brk <- c(min(mat, na.rm = TRUE),
         mean(range(mat, na.rm = TRUE)),
         max(mat, na.rm = TRUE))
col_fun_d <- colorRamp2(brk, c("navy", "white", "firebrick"))
```

```{r fe-panel-d-v2}
ht_compact <- Heatmap(
  mat,
  name              = "median\nadjS",
  col               = col_fun_d,
  row_split         = ht_wide$cell_class,
  cluster_rows      = TRUE,
  cluster_row_slices = FALSE,
  cluster_columns   = TRUE,
  show_row_names    = FALSE,
  show_column_names = TRUE,
  row_names_gp      = gpar(fontsize = 4),
  column_names_gp   = gpar(fontsize = 5),
  row_title_gp      = gpar(fontsize = 6),
  row_title_rot     = 0,
  row_gap           = unit(0.6, "mm"),
  right_annotation  = rowAnnotation(
    cell_class = ht_wide$cell_class,
    major_type = ht_wide$major_type,
    col        = list(cell_class = cc_ht_pal, major_type = mt_pal),
    simple_anno_size = unit(1.8, "mm"),
    annotation_name_gp = gpar(fontsize = 5),
    show_legend = FALSE
  ),
  heatmap_legend_param = list(
    title_gp    = gpar(fontsize = 6),
    labels_gp   = gpar(fontsize = 5),
    grid_height = unit(2, "mm"),
    grid_width  = unit(2, "mm")
  )
)

pD_v2 <- wrap_elements(full = grid::grid.grabExpr(draw(ht_compact)))
```

```{r fe-panel-e-v2}
if (!is.null(cell_df)) {
  cell_df$cell_class <- factor(cell_df$cell_class,
                               levels = names(ecker_cell_class_pal))
  pE_v2 <- ggplot(cell_df,
                  aes(x = mean_meth, y = mean_adjS, color = cell_class)) +
    geom_point(size = 0.25, alpha = 0.6) +
    geom_smooth(aes(group = cell_class), method = "lm", se = FALSE,
                linewidth = 0.35) +
    scale_color_manual(values = ecker_cell_class_pal) +
    facet_wrap(~ major_type, nrow = 3) +
    labs(x = "cell mean methylation", y = "cell mean adjS") +
    guides(x = guide_x_nolap(),
           color = guide_legend(override.aes = list(size = 1.5, alpha = 1),
                                nrow = 1)) +
    theme_ng(base_size = 7) +
    theme(legend.position    = "bottom",
          legend.key.size    = unit(1.2, "mm"),
          legend.text        = element_text(size = 6),
          legend.title       = element_text(size = 7),
          legend.margin      = margin(0, 0, 0, 0),
          legend.box.margin  = margin(-6, 0, -2, 0),
          legend.box.spacing = unit(0.2, "mm"),
          strip.text         = element_text(size = 5.5,
                                            margin = margin(1, 0, 1, 0)),
          panel.spacing.x    = unit(1.2, "mm"),
          panel.spacing.y    = unit(0.6, "mm"),
          plot.margin        = margin(0, 0, 0, 0, "mm"))
} else {
  pE_v2 <- patchwork::plot_spacer() +
    labs(title = "ecker_per_cell_summary.csv missing")
}
```

```{r fe-panel-f-v2}
if (!is.null(ve_long)) {
  ve_filt <- ve_long[ve_long$assay != "adjS_yamet", , drop = FALSE]
  ve_filt$assay <- factor(ve_filt$assay,
                          levels = c("S", "adjS", "methylation"))
  pF_v2 <- ggplot(ve_filt,
                  aes(x = assay, y = median_ve, fill = variable)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.7) +
    scale_fill_manual(values = c("cell_class" = "#0072B2",
                                 "sub_type"   = "#E69F00")) +
    labs(x = "assay", y = "median % var. explained", fill = NULL) +
    theme_ng(base_size = 7) +
    theme(legend.position      = "top",
          legend.justification = "left",
          legend.background    = element_rect(fill = "white", colour = NA),
          legend.key.size      = unit(1.8, "mm"),
          legend.text          = element_text(size = 6),
          legend.margin        = margin(0, 1, 0, 1),
          legend.box.margin    = margin(0, 0, -2, 0),
          plot.margin          = margin(1, 0, 0, 0, "mm"))
} else {
  pF_v2 <- patchwork::plot_spacer() +
    labs(title = "ecker_win_varexp.csv missing")
}
```

```{r fe-panel-g-v2}
r2_by_ann <- grp %>%
  group_by(annotation) %>%
  summarise(
    r2 = summary(lm(median_adjS ~ cell_class, data = cur_data()))$r.squared,
    .groups = "drop"
  )
write.csv(r2_by_ann, "ecker_fig_adjS_r2_by_cell_class.csv", row.names = FALSE)

pG_v2 <- ggplot(r2_by_ann, aes(x = reorder(annotation, r2), y = r2)) +
  geom_col(fill = "#0072B2") +
  coord_flip() +
  labs(x = NULL, y = "R^2 (adjS by cell class)") +
  theme_ng(base_size = 7) +
  theme(axis.text.y = element_text(size = 5.5),
        plot.margin = margin(0, 0, 0, 0, "mm"))
```

```{r fe-panel-h-v2}
driver_df_ecker <- categorize_drivers(grp, "cell_class")
write.csv(driver_df_ecker, "ecker_driver_sd_range.csv", row.names = FALSE)

pH_v2 <- plot_driver_scatter(driver_df_ecker,
                             x_label = "SD adjH (cell classes)",
                             y_label = "SD adjS (cell classes)") +
  guides(color = guide_legend(nrow = 1, title.position = "left"),
         shape = guide_legend(nrow = 1, title.position = "left")) +
  theme_ng(base_size = 7) +
  theme(legend.position    = "bottom",
        legend.key.size    = unit(1.2, "mm"),
        legend.text        = element_text(size = 5.5),
        legend.title       = element_text(size = 6.5),
        legend.margin      = margin(0, 0, 0, 0),
        legend.box.margin  = margin(-6, 0, -2, 0),
        legend.box.spacing = unit(0.2, "mm"),
        plot.margin        = margin(0, 0, 0, 0, "mm"))

pH_v2$layers[[which(sapply(pH_v2$layers, function(l)
  inherits(l$geom, "GeomTextRepel")))]] <- NULL
pH_v2 <- pH_v2 +
  ggrepel::geom_text_repel(size = 1.6, max.overlaps = Inf,
                           min.segment.length = 0,
                           segment.size = 0.2, box.padding = 0.25,
                           force = 3)
```

```{r fe-combined-v2, fig.width = 8.27, fig.height = 8.5}
row1_v2 <- pA_v2 + pB_v2 + pC_v2 +
  plot_layout(widths = c(1.4, 3.6, 2.4))
row2_v2 <- pD_v2 + pE_v2 + plot_layout(widths = c(1, 1))
row3_v2 <- pF_v2 + pG_v2 + pH_v2 + plot_layout(widths = c(1, 1, 1))

compact_fig <- row1_v2 / row2_v2 / row3_v2 +
  plot_layout(heights = c(2.0, 1.82, 1)) +
  plot_annotation(tag_levels = "A",
                  theme = theme(
                    plot.tag          = element_text(size = 9, face = "bold"),
                    plot.tag.position = c(0.01, 0.99),
                    plot.margin       = margin(1, 1, 1, 1, "mm")
                  ))

print(compact_fig)

ggsave(
  file.path(out_dir, "fig_ecker_half_a4.pdf"),
  plot   = compact_fig,
  width  = 8.27,
  height = 9.5,
  units  = "in",
  device = cairo_pdf
)
```

# Standalone heatmap (full size)

The compact heatmap in panel D fits roughly half a page. The standalone
file below renders the same matrix at a larger size for supplementary use.

```{r fe-heatmap-standalone, fig.width = 10, fig.height = 10}
ht_full <- Heatmap(
  mat,
  name              = "median adjS",
  col               = col_fun_d,
  row_split         = ht_wide$cell_class,
  cluster_rows      = TRUE,
  cluster_row_slices = FALSE,
  cluster_columns   = FALSE,
  show_row_names    = TRUE,
  show_column_names = TRUE,
  row_names_gp      = gpar(fontsize = 6),
  column_names_gp   = gpar(fontsize = 7),
  right_annotation  = rowAnnotation(
    cell_class = ht_wide$cell_class,
    major_type = ht_wide$major_type,
    col        = list(cell_class = cc_ht_pal, major_type = mt_pal),
    annotation_legend_param = list(
      cell_class = list(title = "cell class"),
      major_type = list(title = "major type", ncol = 2)
    )
  ),
  column_title = "median adjS per cell type and annotation"
)

draw(ht_full)
pdf(file.path(out_dir, "fig_ecker_heatmap.pdf"), width = 10, height = 10)
draw(ht_full)
invisible(dev.off())
```

# Legacy panels (kept for reference)

```{r fe-legacy-panels, fig.width = ng_fig_size(4, 3, panel_mm = 45)$w, fig.height = ng_fig_size(4, 3, panel_mm = 45)$h}
pa <- ggplot(umap_cells, aes(x = umap1, y = umap2, color = cell_class)) +
  geom_point_rast(size = 0.15, alpha = 0.5, raster.dpi = 300) +
  scale_color_manual(values = ecker_cell_class_pal) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(x = "UMAP 1", y = "UMAP 2", color = "cell class") +
  theme_ng(base_size = 8)

pb <- ggplot(umap_cells, aes(x = umap1, y = umap2, color = median_adjS)) +
  geom_point_rast(size = 0.15, alpha = 0.5, raster.dpi = 300) +
  scale_color_viridis_c(option = "viridis") +
  labs(x = "UMAP 1", y = "UMAP 2", color = "adjS") +
  theme_ng(base_size = 8)

pc <- ggplot(umap_cells, aes(x = umap1, y = umap2, color = median_adjH)) +
  geom_point_rast(size = 0.15, alpha = 0.5, raster.dpi = 300) +
  scale_color_viridis_c(option = "plasma") +
  labs(x = "UMAP 1", y = "UMAP 2", color = "adjH") +
  theme_ng(base_size = 8)

pd <- ggplot(umap_cells, aes(x = umap1, y = umap2, color = median_meth)) +
  geom_point_rast(size = 0.15, alpha = 0.5, raster.dpi = 300) +
  scale_color_distiller(palette = "RdBu", direction = -1) +
  labs(x = "UMAP 1", y = "UMAP 2", color = "avg. meth.") +
  theme_ng(base_size = 8)

var_class <- ent$adjsampens %>%
  group_by(cell_class) %>%
  summarise(iqr = IQR(sampen, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(iqr)) %>%
  mutate(cell_class = factor(cell_class, levels = cell_class))
write.csv(var_class, "ecker_fig_adjS_iqr_by_cell_class.csv", row.names = FALSE)

pe <- ggplot(var_class, aes(x = cell_class, y = iqr, fill = cell_class)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = ecker_cell_class_pal) +
  labs(x = "cell class", y = "IQR adjS") +
  theme_ng(base_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

pf <- ggplot(r2_by_ann, aes(x = reorder(annotation, r2), y = r2)) +
  geom_col(fill = "#0072B2") +
  coord_flip() +
  labs(x = NULL, y = "R^2 (adjS by cell class)") +
  theme_ng(base_size = 8)

ph <- plot_driver_scatter(driver_df_ecker,
                          x_label = "SD adjH across cell classes",
                          y_label = "SD adjS across cell classes") +
  theme_ng(base_size = 8) +
  theme(legend.key.size = unit(2, "mm"),
        legend.text     = element_text(size = 6),
        legend.title    = element_text(size = 7))

em_adjS_legacy <- win_umap$win_embeds[["adjS"]]
if (!is.null(em_adjS_legacy)) {
  wins_df <- data.frame(
    umap1      = em_adjS_legacy$umap[, 1],
    umap2      = em_adjS_legacy$umap[, 2],
    cell_class = win_umap$col_data$cell_class[em_adjS_legacy$kept_cols]
  )
  pg <- ggplot(wins_df, aes(x = umap1, y = umap2, color = cell_class)) +
    geom_point_rast(size = 0.15, alpha = 0.5, raster.dpi = 300) +
    scale_color_manual(values = ecker_cell_class_pal) +
    guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
    labs(x = "UMAP 1", y = "UMAP 2", color = "cell class",
         title = "windows-based adjS UMAP (cells)") +
    theme_ng(base_size = 8)
} else {
  pg <- NULL
}

row2_legacy <- if (!is.null(pg)) (pe | pf | pg | ph) else (pe | pf | ph)
gg_fig <- (pa | pb | pc | pd) / row2_legacy +
  plot_layout(heights = c(1, 0.9)) +
  plot_annotation(tag_levels = "A",
                  theme = theme(plot.tag = element_text(size = 8, face = "bold")))

print(gg_fig)

ggsave(
  file.path(out_dir, "fig_ecker_panels.pdf"),
  plot   = gg_fig,
  width  = ng_fig_size(4, 3, panel_mm = 45)$w,
  height = ng_fig_size(4, 3, panel_mm = 45)$h,
  units  = "in",
  device = cairo_pdf
)
```
