---
title: "Combined within and across cell variability – adjusted entropies (adjS / adjH)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    code_folding: hide
    number_sections: no
    fig_crop: yes
---

```{r setup}
suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
  library(svglite)
})

source("rules/src/ggtheme.R")

knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.path = "simulations/output/08_plots/",
  dev      = c("png", "svg"),
  dpi      = 300
)

log_con <- file(snakemake@log[[1]], open = "wt")
sink(log_con)
sink(log_con, type = "message")

## ── Colour scales ─────────────────────────────────────────────────────────────
wc_scale <- scale_color_gradient(low = "grey90", high = "#000080")   # within-cell
ac_scale <- scale_color_gradient(low = "grey90", high = "#800080")   # across-cell

## ── Theme for this figure (labels ~10–11 pt for A4 output) ───────────────────
fig_base <- 14   # base font size

## Square aspect ratio (panels A–D)
fig_theme_sq <- theme(
  axis.title      = element_text(size = fig_base + 1, face = "bold", color = "black"),
  axis.text       = element_text(size = fig_base,       color = "black"),
  legend.text     = element_text(size = fig_base - 1),
  legend.title    = element_text(size = fig_base,       face = "bold"),
  legend.key.size = unit(0.35, "cm")
)

## Free aspect ratio (panels E–H subplots)
fig_theme_free <- fig_theme_sq + theme(aspect.ratio = 1)

## Smaller theme for performance sub-panels (G/H)
perf_theme <- theme(
  aspect.ratio    = NULL,
  axis.title      = element_text(size = fig_base,     face = "bold", color = "black"),
  axis.text       = element_text(size = fig_base - 1, color = "black"),
  plot.title      = element_text(size = fig_base,     face = "bold", hjust = 0.5),
  legend.text     = element_text(size = fig_base - 1),
  legend.title    = element_text(size = fig_base - 1, face = "bold"),
  legend.key.size = unit(0.3, "cm")
)

## patchwork annotation theme
pa_theme <- theme(
  plot.title = element_text(size = fig_base + 1, face = "bold", hjust = 0.5)
)

## Helper: add "simul. X" label inside the upper-right corner of a plot
add_simul_label <- function(p, label) {
  p + annotate(
    "text",
    x = Inf, y = Inf,
    label    = label,
    hjust    = 1.08, vjust = 1.6,
    size     = 3.2,
    fontface = "italic",
    color    = "grey35"
  )
}
```

```{r load-data}
## ── Unadjusted (S / H) ────────────────────────────────────────────────────────
all_long_wc  <- readRDS("simulations/rds/all.long_within.rds")
jnt_wc       <- readRDS("simulations/rds/jnt_within.rds")
s_plots_wc   <- readRDS("simulations/rds/s_plots_within.rds")
n_plots_wc   <- readRDS("simulations/rds/n_plots_within.rds")

all_long_ac  <- readRDS("simulations/rds/all.long_across.rds")
jnt_ac       <- readRDS("simulations/rds/jnt_across.rds")
s_plots_ac   <- readRDS("simulations/rds/s_plots_across.rds")
n_plots_ac   <- readRDS("simulations/rds/n_plots_across.rds")

## ── Adjusted (adjS / adjH) ────────────────────────────────────────────────────
all_long_wc_adj <- readRDS("simulations/rds/all.long_within_adj.rds")
jnt_wc_adj      <- readRDS("simulations/rds/jnt_within_adj.rds")
s_plots_wc_adj  <- readRDS("simulations/rds/s_plots_within_adj.rds")
n_plots_wc_adj  <- readRDS("simulations/rds/n_plots_within_adj.rds")

all_long_ac_adj <- readRDS("simulations/rds/all.long_across_adj.rds")
jnt_ac_adj      <- readRDS("simulations/rds/jnt_across_adj.rds")
s_plots_ac_adj  <- readRDS("simulations/rds/s_plots_across_adj.rds")
n_plots_ac_adj  <- readRDS("simulations/rds/n_plots_across_adj.rds")
```

# Figure 1 – Adjusted entropies (adjS / adjH)

Same layout as the main combined figure but using adjusted sample entropy (adjS)
and adjusted Shannon entropy (adjH) in place of their unadjusted counterparts.

```{r adj-panels-ABCD}
## ── Row 1: scatter plots ──────────────────────────────────────────────────────

A <- add_simul_label(
  ggplot(all_long_wc_adj, aes(x = meth, y = adj_sampen, color = wc_vi)) +
    geom_point(alpha = 0.3, size = 0.5) +
    theme_ng(base_size = fig_base) + wc_scale + fig_theme_sq +
    labs(x = "avg. methylation", y = "adj. sample entropy (adjS)", color = "wcVI"),
  "simul. II"
)

B <- add_simul_label(
  ggplot(all_long_wc_adj, aes(x = norm_shannon, y = adj_sampen, color = wc_vi)) +
    geom_point(alpha = 0.7, size = 0.5) +
    theme_ng(base_size = fig_base) + wc_scale + fig_theme_sq +
    labs(x = "adj. Shannon entropy (adjH)", y = "adj. sample entropy (adjS)", color = "wcVI"),
  "simul. II"
)

C <- add_simul_label(
  ggplot(all_long_ac_adj, aes(x = meth, y = adj_sampen, color = bc_vi)) +
    geom_point(alpha = 0.3, size = 0.5) +
    theme_ng(base_size = fig_base) + ac_scale + fig_theme_sq +
    labs(x = "avg. methylation", y = "adj. sample entropy (adjS)", color = "acVI"),
  "simul. III"
)

D <- add_simul_label(
  ggplot(all_long_ac_adj, aes(x = norm_shannon, y = adj_sampen, color = bc_vi)) +
    geom_point(alpha = 0.7, size = 0.5) +
    theme_ng(base_size = fig_base) + ac_scale + fig_theme_sq +
    labs(x = "adj. Shannon entropy (adjH)", y = "adj. sample entropy (adjS)", color = "acVI"),
  "simul. III"
)
```

```{r adj-panels-EF}
## ── Row 2: scMET panels (same metric – scMET output is independent of adjS/adjH)

## E: within-cell (wcVI)
e_mu <- ggplot(jnt_wc_adj, aes(x = avg_meth, y = scmet_mu, color = wc_vi)) +
  geom_point(alpha = 0.7, size = 0.5) +
  theme_ng(base_size = fig_base) + wc_scale + fig_theme_free +
  labs(x = "avg. methylation", y = expression(scMET ~ mu), color = "wcVI")

e_gamma <- ggplot(jnt_wc_adj, aes(x = avg_meth, y = scmet_gamma, color = wc_vi)) +
  geom_point(alpha = 0.7, size = 0.5) +
  theme_ng(base_size = fig_base) + wc_scale + fig_theme_free +
  labs(x = "avg. methylation", y = expression(scMET ~ gamma), color = "wcVI")

e_epsilon <- ggplot(jnt_wc_adj, aes(x = avg_meth, y = scmet_epsilon, color = wc_vi)) +
  geom_point(alpha = 0.7, size = 0.5) +
  theme_ng(base_size = fig_base) + wc_scale + fig_theme_free +
  labs(x = "avg. methylation", y = expression(scMET ~ epsilon), color = "wcVI")

E_group <- wrap_plots(e_mu, e_gamma, e_epsilon, ncol = 3, guides = "collect") +
  plot_annotation(title = "simul. II", theme = pa_theme) &
  theme(legend.position = "right")

## F: across-cell (acVI)
f_mu <- ggplot(jnt_ac_adj, aes(x = avg_meth, y = scmet_mu, color = bc_vi)) +
  geom_point(alpha = 0.7, size = 0.5) +
  theme_ng(base_size = fig_base) + ac_scale + fig_theme_free +
  labs(x = "avg. methylation", y = expression(scMET ~ mu), color = "acVI")

f_gamma <- ggplot(jnt_ac_adj, aes(x = avg_meth, y = scmet_gamma, color = bc_vi)) +
  geom_point(alpha = 0.7, size = 0.5) +
  theme_ng(base_size = fig_base) + ac_scale + fig_theme_free +
  labs(x = "avg. methylation", y = expression(scMET ~ gamma), color = "acVI")

f_epsilon <- ggplot(jnt_ac_adj, aes(x = avg_meth, y = scmet_epsilon, color = bc_vi)) +
  geom_point(alpha = 0.7, size = 0.5) +
  theme_ng(base_size = fig_base) + ac_scale + fig_theme_free +
  labs(x = "avg. methylation", y = expression(scMET ~ epsilon), color = "acVI")

F_group <- wrap_plots(f_mu, f_gamma, f_epsilon, ncol = 3, guides = "collect") +
  plot_annotation(title = "simul. III", theme = pa_theme) &
  theme(legend.position = "right")
```

```{r adj-panels-GH}
## ── Row 3: performance grids (adjS / adjH NMI) ────────────────────────────────

make_perf_grid <- function(s_plots, n_plots) {
  g_s_nmi <- s_plots$nmi     + perf_theme +
    labs(title = "NMI",           x = "",        y = "performance")
  g_s_cpu <- s_plots$seconds + perf_theme +
    labs(title = "CPU time (s)",  x = "# cells", y = "")
  g_s_mem <- s_plots$mem     + perf_theme +
    labs(title = "memory (MB)",   x = "",        y = "")

  g_n_nmi <- n_plots$nmi     + perf_theme +
    labs(title = "NMI",           x = "",           y = "performance")
  g_n_cpu <- n_plots$seconds + perf_theme +
    labs(title = "CPU time (s)",  x = "# features", y = "")
  g_n_mem <- n_plots$mem     + perf_theme +
    labs(title = "memory (MB)",   x = "",           y = "")

  wrap_plots(
    g_s_nmi, g_s_cpu, g_s_mem,
    g_n_nmi, g_n_cpu, g_n_mem,
    ncol    = 3,
    guides  = "collect"
  ) & theme(legend.position = "bottom")
}

G_inner <- make_perf_grid(s_plots_wc_adj, n_plots_wc_adj)
H_inner <- make_perf_grid(s_plots_ac_adj, n_plots_ac_adj)

G_group <- G_inner + plot_annotation(title = "simul. II",  theme = pa_theme)
H_group <- H_inner + plot_annotation(title = "simul. III", theme = pa_theme)
```

```{r adj-combined, fig.width=15, fig.height=15}
## ── Assemble adjS/adjH figure ─────────────────────────────────────────────────
row1 <- (A | B | C | D)
row2 <- wrap_plots(E_group, F_group, ncol = 2)
row3 <- wrap_plots(G_group, H_group, ncol = 2)

fig_adj <- row1 / row2 / row3 +
  plot_layout(heights = c(1.8, 0.7, 2)) +
  plot_annotation(
    tag_levels = list(c("A", "B", "C", "D", "E", "", "", "F", "", "", "G", "H")),
    theme = theme(plot.tag = element_text(size = 14, face = "bold"))
  )

print(fig_adj)
```

```{r save-adj-figure}
ggsave(
  filename = "simulations/output/figure_simulations_combined_adj.svg",
  plot     = fig_adj,
  width    = 8.27,
  height   = 11.69,
  units    = "in"
)

ggsave(
  filename = "simulations/output/figure_simulations_combined_adj.pdf",
  plot     = fig_adj,
  width    = 8.27,
  height   = 11.69,
  units    = "in"
)
```

# Figure 2 – Pairwise comparison: adjusted vs unadjusted entropies

These panels directly compare S vs adjS and H vs adjH, revealing how the
coverage-normalisation shifts entropy values relative to within/across-cell
variability indices.

```{r pairwise-data}
## Merge unadjusted and adjusted long-format data for pairwise plots.
## Both data frames have the same row structure (same features × samples),
## so we can cbind the adj columns directly.
wc_cmp <- data.frame(
  sampen       = all_long_wc$sampen,
  adj_sampen   = all_long_wc_adj$adj_sampen,
  shannon      = all_long_wc$shannon,
  norm_shannon = all_long_wc_adj$norm_shannon,
  meth         = all_long_wc$meth,
  avg_meth     = all_long_wc$avg_meth,
  wc_vi        = all_long_wc$wc_vi
)

ac_cmp <- data.frame(
  sampen       = all_long_ac$sampen,
  adj_sampen   = all_long_ac_adj$adj_sampen,
  shannon      = all_long_ac$shannon,
  norm_shannon = all_long_ac_adj$norm_shannon,
  meth         = all_long_ac$meth,
  avg_meth     = all_long_ac$avg_meth,
  bc_vi        = all_long_ac$bc_vi
)

## Feature-level (jnt) comparison for H vs adjH
wc_jnt_cmp <- data.frame(
  sampen_avg   = jnt_wc$sampen_avg,
  adj_sampen_avg = jnt_wc_adj$adj_sampen_avg,
  shannon      = jnt_wc$shannon,
  norm_shannon = jnt_wc_adj$shannon_norm,
  avg_meth     = jnt_wc$avg_meth,
  wc_vi        = jnt_wc$wc_vi
)

ac_jnt_cmp <- data.frame(
  sampen_avg   = jnt_ac$sampen_avg,
  adj_sampen_avg = jnt_ac_adj$adj_sampen_avg,
  shannon      = jnt_ac$shannon,
  norm_shannon = jnt_ac_adj$shannon_norm,
  avg_meth     = jnt_ac$avg_meth,
  bc_vi        = jnt_ac$bc_vi
)
```

```{r pairwise-panels-wc}
## ── Within-cell (simul. II) ───────────────────────────────────────────────────

P1 <- add_simul_label(
  ggplot(wc_cmp, aes(x = sampen, y = adj_sampen, color = wc_vi)) +
    geom_point(alpha = 0.3, size = 0.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
    theme_ng(base_size = fig_base) + wc_scale + fig_theme_sq +
    labs(x = "sample entropy (S)", y = "adj. sample entropy (adjS)", color = "wcVI"),
  "simul. II"
)

P2 <- add_simul_label(
  ggplot(wc_jnt_cmp, aes(x = shannon, y = norm_shannon, color = wc_vi)) +
    geom_point(alpha = 0.7, size = 0.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
    theme_ng(base_size = fig_base) + wc_scale + fig_theme_sq +
    labs(x = "Shannon entropy (H)", y = "adj. Shannon entropy (adjH)", color = "wcVI"),
  "simul. II"
)

P3 <- add_simul_label(
  ggplot(wc_cmp, aes(x = meth, y = sampen - adj_sampen, color = wc_vi)) +
    geom_point(alpha = 0.3, size = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    theme_ng(base_size = fig_base) + wc_scale + fig_theme_sq +
    labs(x = "avg. methylation", y = "S \u2212 adjS", color = "wcVI"),
  "simul. II"
)

P4 <- add_simul_label(
  ggplot(wc_jnt_cmp, aes(x = avg_meth, y = shannon - norm_shannon, color = wc_vi)) +
    geom_point(alpha = 0.7, size = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    theme_ng(base_size = fig_base) + wc_scale + fig_theme_sq +
    labs(x = "avg. methylation", y = "H \u2212 adjH", color = "wcVI"),
  "simul. II"
)
```

```{r pairwise-panels-ac}
## ── Across-cell (simul. III) ──────────────────────────────────────────────────

P5 <- add_simul_label(
  ggplot(ac_cmp, aes(x = sampen, y = adj_sampen, color = bc_vi)) +
    geom_point(alpha = 0.3, size = 0.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
    theme_ng(base_size = fig_base) + ac_scale + fig_theme_sq +
    labs(x = "sample entropy (S)", y = "adj. sample entropy (adjS)", color = "acVI"),
  "simul. III"
)

P6 <- add_simul_label(
  ggplot(ac_jnt_cmp, aes(x = shannon, y = norm_shannon, color = bc_vi)) +
    geom_point(alpha = 0.7, size = 0.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
    theme_ng(base_size = fig_base) + ac_scale + fig_theme_sq +
    labs(x = "Shannon entropy (H)", y = "adj. Shannon entropy (adjH)", color = "acVI"),
  "simul. III"
)

P7 <- add_simul_label(
  ggplot(ac_cmp, aes(x = meth, y = sampen - adj_sampen, color = bc_vi)) +
    geom_point(alpha = 0.3, size = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    theme_ng(base_size = fig_base) + ac_scale + fig_theme_sq +
    labs(x = "avg. methylation", y = "S \u2212 adjS", color = "acVI"),
  "simul. III"
)

P8 <- add_simul_label(
  ggplot(ac_jnt_cmp, aes(x = avg_meth, y = shannon - norm_shannon, color = bc_vi)) +
    geom_point(alpha = 0.7, size = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    theme_ng(base_size = fig_base) + ac_scale + fig_theme_sq +
    labs(x = "avg. methylation", y = "H \u2212 adjH", color = "acVI"),
  "simul. III"
)
```

```{r pairwise-combined, fig.width=15, fig.height=8}
## ── Assemble pairwise comparison figure ───────────────────────────────────────
## Row 1: within-cell (P1-P4), Row 2: across-cell (P5-P8)
fig_cmp <- (P1 | P2 | P3 | P4) / (P5 | P6 | P7 | P8) +
  plot_annotation(
    tag_levels = "A",
    theme = theme(plot.tag = element_text(size = 14, face = "bold"))
  )

print(fig_cmp)
```

```{r save-pairwise-figure}
ggsave(
  filename = "simulations/output/figure_simulations_comparison_adj_unadj.svg",
  plot     = fig_cmp,
  width    = 16,
  height   = 8,
  units    = "in"
)

ggsave(
  filename = "simulations/output/figure_simulations_comparison_adj_unadj.pdf",
  plot     = fig_cmp,
  width    = 16,
  height   = 8,
  units    = "in"
)
```

# Figure 3: adjS within methylation bins

Binning average methylation into 0.1-wide intervals and showing adjS per bin
confirms that the residual association between adjS and variability indices is
not an artifact of methylation level.

```{r bins-wc, fig.width = ng_fig_size(5, 2)$w, fig.height = ng_fig_size(5, 2)$h}
all_long_wc_adj %>%
  mutate(meth_bin = cut(meth, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>%
  filter(!is.na(meth_bin), !is.na(adj_sampen)) %>%
  ggplot(aes(x = meth_bin, y = adj_sampen, fill = cut(wc_vi, 5))) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  scale_fill_viridis_d(option = "viridis", name = "wcVI quintile") +
  labs(
    x = "avg. methylation bin",
    y = "adj. sample entropy (adjS)",
    title = "simul. II within-cell variability"
  ) +
  theme_ng(base_size = fig_base) + fig_theme_sq +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

```{r bins-ac, fig.width = ng_fig_size(5, 2)$w, fig.height = ng_fig_size(5, 2)$h}
all_long_ac_adj %>%
  mutate(meth_bin = cut(meth, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>%
  filter(!is.na(meth_bin), !is.na(adj_sampen)) %>%
  ggplot(aes(x = meth_bin, y = adj_sampen, fill = cut(bc_vi, 5))) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  scale_fill_viridis_d(option = "plasma", name = "acVI quintile") +
  labs(
    x = "avg. methylation bin",
    y = "adj. sample entropy (adjS)",
    title = "simul. III across-cell variability"
  ) +
  theme_ng(base_size = fig_base) + fig_theme_sq +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```
