---
title: "Within Cell Variability"
author: "Atreya Choudhury"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    code_folding: hide
    number_sections: no
    fig_crop: yes
---

## Strategy

The goal is to generate features with various heterogeneity levels for the purpose of within cell differentiation.
That means features should be differentiable, not by the heterogeneity between different samples, but by heterogeneity between individual lines.
This means we want features to be differentiable in terms of their sample entropy but not their 2-mer shannon entropy.

Consider the sequence $0011\; 0$.
Notice that every 2-mer is seen exactly once.
This can be generalised to longer sequences, say $0011\; 0011\; 0011\; 0$.
Here, all the 2-mers have the same frequency.
Similary, the sequence $0110\; 0$ has the same property.
We will refer to these short sequences as snips.
Putting these two together, observe the two sequences below.

$$0011\; 0011\; 0110\; 0110\; 0$$
$$0011\; 0110\; 0011\; 0110\; 0$$

They both have the same distribution of 2-mers but differ in their sample entropy (the second one having higher sample entropy).
This provides the basis of our simulation framework.
For each feature, we start with a sequence where the first half is made up of $0011$s and the second is made up of $0110$s.
For each feature, we pick a certain fraction of snips (equally from the two halves) in proportion to a true within cell variability index (wcVI).
These snip positions differ between different features (even ones with the same wcVI) but are the same for different samples within the same feature.
The snips at the selected positions are then shuffled in the different samples.
The idea is that features where a higher fraction of snips were shuffled exhibit higher sample entropy compared to those where a lower fraction of snips were shuffled.
We also add some stochastic methylation sampling from a beta-binomial distribution to control the level to which a feature in a particular sample is over/under-methylated.

We start by creating a template for a cell.
We have a fixed number of features, the length of each feature is taken to be the same and the number of simulated CpG sites is the same as well.
Every feature is given a **within cell Variability Index (wcVI)** between 1 and 10 and stored in the `wc_vi` column.
As described above we pick a fraction of snips in proportion to the feature's `wc_vi`.
This is stored in the `snip_pos` column.
For every feature, we sample from a bernoulli distribution indicating whether the feature will be over/under-methylated compared to the baseline.
This is stored in the `higher` column.
The extent to which each sample in each feature is over/under-methylated is sampled from a beta binomial model which is biased towards lower deviations.
This difference is stored in the `delta` column.
This template file can now be used for generating new samples and also for running `yamet` on the simulated data.

With the template, we now generate samples as described above.
For samples within the same feature, we shuffle the snips at the positions described in `snip_pos`.
At this stage, observe that the all the features in all the samples are approximately $50\%$ methylated and that this value is the same for all the samples.
We would like to introduce some randomness to the average methylation level which is done using the data generated in the `higher` and `delta` column.
For every sample in every feature, we sample $35\%$ of the snips.
The value in the `higher` column for that feature tells us whether this sample should be over-methylated or under-methylated.
From this sampled set of snips, we further subsample for two sets of snips such that the difference in their size is in proportion to the value in the `delta` column at that sample-feature intersection.
Then, one set of these snips is changed to $1110$ while the other is changed to $0001$.
This not only introduces the difference in methylation but also adds more regularity at higher/lower methylation, simulating the quadratic relationship between sample entropy and average methylation.


```{r, setup}
suppressPackageStartupMessages({
  library(ggplot2)
  library(ggpubr)
  library(svglite)
})

source("rules/src/mi_table.R")
source("rules/src/make_grid_plots.R")
source("rules/src/ggtheme.R")

knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.path = "simulations/output/05_plots/",
    dev = c("png", "svg"),
    dpi=500
)

data_dir <- dirname(snakemake@input[["disp_scmet"]])
S_GRID <- snakemake@params[["S_GRID"]]
N_GRID <- snakemake@params[["N_GRID"]]
DISP <- snakemake@params[["DISP"]]

## Large axis labels applied to every plot
big_labels <- theme(
  axis.title = element_text(size = 22, face = "bold", color = "black"),
  axis.text  = element_text(size = 18, color = "black")
)

## Within-cell colour gradient (navy)
wc_scale <- scale_color_gradient(low = "grey90", high = "#000080")
```


```{r}
dir.create('simulations/rds', showWarnings = FALSE)
```

## Variation
For benchmarking against the number of samples, we used a grid with `r length(S_GRID$S)` sample sizes (S = `r paste(S_GRID$S, collapse = ", ")`), `r S_GRID$N` features where each feature has `r S_GRID$f` CpGs (all covered).
```{r}
s_plots <- make_grid_plots(S_GRID, "S", "wc_vi", data_dir)
```

For benchmarking against the number of features, we used a grid with `r N_GRID$S` samples, `r length(N_GRID$N)` feature sizes (N = `r paste(N_GRID$N, collapse = ", ")`) where each feature has `r N_GRID$f` CpGs (all covered).
```{r}
n_plots <- make_grid_plots(N_GRID, "N", "wc_vi", data_dir)
```

## One Example
We show one example with `r DISP$S` samples, `r DISP$N` features where each feature has `r DISP$f` CpGs (all covered).
```{r}
S <- DISP$S
N <- DISP$N
f <- DISP$f

intervals <- read.table(
  paste0(data_dir, paste("/intervals", S, N, f, "tsv", sep = ".")),
  header = T
)
intervals <- intervals["wc_vi"]
sampens <- read.table(
  paste0(data_dir, paste("/yamet", S, N, f, "det.out", sep = ".")),
  header = T
)
jnt <- cbind(sampens, intervals)
rownames(jnt) <- paste(jnt$start, jnt$end, sep = ":")
output_cols <- grep("^output", colnames(jnt), value = TRUE)
jnt$sampen_avg <- rowMeans(jnt[, output_cols])
jnt.long <- do.call(rbind, lapply(output_cols, function(col) {
  data.frame(
    start = jnt$start,
    end = jnt$end,
    sample = col,
    sampen = jnt[[col]],
    shannon = jnt$shannon,
    avg_meth = jnt$avg_meth,
    wc_vi = jnt$wc_vi
  )
}))

meth <- read.table(
  paste0(data_dir, paste("/yamet", S, N, f, "meth.out", sep = ".")),
  header = TRUE
)
meth.long <- do.call(rbind, lapply(output_cols, function(col) {
  data.frame(
    meth = meth[[col]]
  )
}))

all.long <- cbind(jnt.long, meth.long)

samp_meth <- ggplot(all.long, aes(x = meth, y = sampen, color = wc_vi)) +
  geom_point(alpha = 0.3) +
  theme_ng() +
  wc_scale +
  big_labels +
  labs(
    x = "avg. methylation", y = "sample entropy", color = "wcVI"
  )

samp_vi <- ggplot(all.long, aes(x = as.factor(wc_vi), y = sampen)) +
  geom_point(alpha = 0.7, position = position_jitter(width = 0.2)) +
  geom_violin(alpha = 0.3, position = position_dodge(0.8)) +
  theme_ng() +
  big_labels +
  labs(
    x = "wcVI", y = "sample entropy"
  )

shannon_meth <- ggplot(jnt, aes(x = avg_meth, y = shannon, color = wc_vi)) +
  geom_point(alpha = 0.7) +
  theme_ng() +
  wc_scale +
  big_labels +
  labs(
    x = "avg. methylation", y = "Shannon entropy", color = "wcVI"
  )

shannon_vi <- ggplot(jnt, aes(x = as.factor(wc_vi), y = shannon, color = avg_meth)) +
  geom_point(alpha = 0.7) +
  theme_ng() +
  big_labels +
  labs(
    x = "wcVI", y = "Shannon entropy", color = "meth"
  )

samp_shan <- ggplot(all.long, aes(x = shannon, y = sampen, color = wc_vi)) +
  geom_point(alpha = 0.7) +
  theme_ng() +
  wc_scale +
  big_labels +
  labs(
    x = "Shannon entropy", y = "sample entropy", color = "wcVI"
  )
```

```{r}
scmet_fit <- readRDS(paste0(data_dir, paste("/scmet", S, N, f, "rds", sep = ".")))
tmp <- colMeans(scmet_fit$posterior$mu)
jnt$scmet_mu <- tmp[match(rownames(jnt), names(tmp))]
tmp <- colMeans(scmet_fit$posterior$gamma)
jnt$scmet_gamma <- tmp[match(rownames(jnt), names(tmp))]
tmp <- colMeans(scmet_fit$posterior$epsilon)
jnt$scmet_epsilon <- tmp[match(rownames(jnt), names(tmp))]
```

```{r}
scmet_mu_vi <- ggplot(jnt, aes(x = as.factor(wc_vi), y = scmet_mu, color = avg_meth)) +
  geom_point(alpha = 0.7) +
  theme_ng() +
  big_labels +
  labs(
    x = "wcVI", y = expression("scMET " ~ mu), color = "meth"
  )

scmet_mu_meth <- ggplot(jnt, aes(x = avg_meth, y = scmet_mu, color = wc_vi)) +
  geom_point(alpha = 0.7) +
  theme_ng() +
  wc_scale +
  big_labels +
  labs(
    x = "avg. methylation", y = expression("scMET " ~mu), color = "wcVI"
  )

scmet_gamma_meth <- ggplot(jnt, aes(x = avg_meth, y = scmet_gamma, color = wc_vi)) +
  geom_point(alpha = 0.7) +
  theme_ng() +
  wc_scale +
  big_labels +
  labs(
    x = "avg. methylation", y = expression("scMET " ~gamma), color = "wcVI"
  )

scmet_epsilon_meth <- ggplot(jnt, aes(x = avg_meth, y = scmet_epsilon, color = wc_vi)) +
  geom_point(alpha = 0.7) +
  theme_ng() +
  wc_scale +
  big_labels +
  labs(
    x = "avg. methylation", y = expression("scMET " ~epsilon), color = "wcVI"
  )

sampen_avg_vi <- ggplot(jnt, aes(x = as.factor(wc_vi), y = sampen_avg, color = avg_meth)) +
  geom_point(alpha = 0.7) +
  theme_ng() +
  big_labels +
  labs(
    x = "wcVI", y = "sampen (average)", color = "meth"
  )

sampen_avg_meth <- ggplot(jnt, aes(x = avg_meth, y = sampen_avg, color = wc_vi)) +
  geom_point(alpha = 0.7) +
  theme_ng() +
  wc_scale +
  big_labels +
  labs(
    x = "avg. methylation", y = "sampen (average)", color = "wcVI"
  )
```

```{r}
mi_table <- mi_table_gen(jnt, "wc_vi", table = T)
mi_fd <- mi_table_gen(jnt, "wc_vi", table = F)
saveRDS(mi_fd,     'simulations/rds/perf_within.rds')
saveRDS(all.long,  'simulations/rds/all.long_within.rds')
saveRDS(jnt,       'simulations/rds/jnt_within.rds')
saveRDS(s_plots,   'simulations/rds/s_plots_within.rds')
saveRDS(n_plots,   'simulations/rds/n_plots_within.rds')
```

```{r adj-within}
## ── Adjusted entropy (adjS = norm. sample entropy, adjH = norm. Shannon) ──────

## Load norm.det.out for the display example
norm_det <- read.table(
  paste0(data_dir, paste("/yamet", S, N, f, "norm.det.out", sep = ".")),
  header = T
)
rownames(norm_det) <- paste(norm_det$start, norm_det$end, sep = ":")
norm_output_cols <- grep("^output", colnames(norm_det), value = TRUE)

## Per-cell per-feature long format with adj columns
jnt.adj.long <- do.call(rbind, lapply(output_cols, function(col) {
  data.frame(
    start        = jnt$start,
    end          = jnt$end,
    sample       = col,
    adj_sampen   = norm_det[[col]],
    norm_shannon = norm_det$shannon_norm,
    avg_meth     = norm_det$avg_meth,
    wc_vi        = jnt$wc_vi
  )
}))
## Attach per-cell methylation (same as unadjusted all.long)
all.long_adj <- cbind(jnt.adj.long, meth.long)

## Add adj summary columns to jnt, then save jnt_within_adj.rds
jnt$adj_sampen_avg <- rowMeans(norm_det[rownames(jnt), norm_output_cols])
jnt$shannon_norm   <- norm_det[rownames(jnt), "shannon_norm"]

## Adj grid performance plots
s_plots_adj <- make_grid_plots(S_GRID, "S", "wc_vi", data_dir, adj = TRUE)
n_plots_adj <- make_grid_plots(N_GRID, "N", "wc_vi", data_dir, adj = TRUE)
mi_fd_adj   <- mi_table_gen(jnt, "wc_vi", table = FALSE, adj = TRUE)

saveRDS(mi_fd_adj,    'simulations/rds/perf_within_adj.rds')
saveRDS(all.long_adj, 'simulations/rds/all.long_within_adj.rds')
saveRDS(jnt,          'simulations/rds/jnt_within_adj.rds')
saveRDS(s_plots_adj,  'simulations/rds/s_plots_within_adj.rds')
saveRDS(n_plots_adj,  'simulations/rds/n_plots_within_adj.rds')
```

## Plots

```{r samp_meth}
print(samp_meth)
```

```{r samp_vi}
print(samp_vi)
```

```{r samp_shan}
print(samp_shan)
```

```{r shannon_meth}
print(shannon_meth)
```

```{r shannon_vi}
print(shannon_vi)
```

```{r scmet_mu_vi}
print(scmet_mu_vi)
```

```{r scmet_mu_meth}
print(scmet_mu_meth)
```

```{r scmet_gamma_meth}
print(scmet_gamma_meth)
```

```{r scmet_epsilon_meth}
print(scmet_epsilon_meth)
```

```{r sampen_avg_vi}
print(sampen_avg_vi)
```

```{r sampen_avg_meth}
print(sampen_avg_meth)
```

## Performance

```{r nmi}
metric_labels <- c(
  "Sample Entropy" = "sample entropy",
  "Shannon Entropy" = "Shannon entropy",
  "Average Methylation" = "avg. methylation",
  "scMET (mu)" = expression("scMET" ~ mu),
  "scMET (gamma)" = expression("scMET" ~ gamma),
  "scMET (epsilon)" = expression("scMET" ~ epsilon)
)

nmi <- ggplot(mi_fd, aes(x = reorder(Metric, -NMI), y = NMI)) +
    geom_point(size = 3) +
    geom_text(aes(label = round(NMI, 3)), vjust = -0.5, size = 5) +
    theme_ng() +
    big_labels +
    scale_x_discrete(labels = metric_labels) +
    labs(
        x = NULL,
        y = "Normalized Mutual Information (to wcVI)"
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 16))

print(nmi)
```

```{r s_nmi}
print(s_plots$nmi + big_labels + xlab("# cells") + ylab("performance (metric NMI)"))
```

```{r s_seconds}
print(s_plots$seconds + big_labels + xlab("# cells") + ylab("CPU time (s)"))
```

```{r s_mem}
print(s_plots$mem + big_labels + xlab("# cells") + ylab("memory usage (MB)"))
```

```{r n_nmi}
print(n_plots$nmi + big_labels + xlab("# features") + ylab("performance (metric NMI)"))
```

```{r n_seconds}
print(n_plots$seconds + big_labels + xlab("# features") + ylab("CPU time (s)"))
```

```{r n_mem}
print(n_plots$mem + big_labels + xlab("# features") + ylab("memory usage (MB)"))
```
