---
title: "Combined within and across cell variability – adjusted entropies (adjS / adjH)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    code_folding: hide
    number_sections: no
    fig_crop: yes
---

```{r setup}
suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
  library(svglite)
  library(dplyr)
})

source("rules/src/ggtheme.R")

knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.path = "simulations/output/08_plots/",
  dev      = c("png", "svg"),
  dpi      = 300
)

log_con <- file(snakemake@log[[1]], open = "wt")
sink(log_con)
sink(log_con, type = "message")

## ── Colour scales ─────────────────────────────────────────────────────────────
wc_scale <- scale_color_gradient(low = "grey90", high = "#000080")   # within-cell
ac_scale <- scale_color_gradient(low = "grey90", high = "#800080")   # across-cell

## ── Theme for this figure ────────────────────────────────────────────────────
fig_base <- 9   # base font size; theme_ng handles all sizing

panel_theme <- theme_ng(base_size = fig_base) +
  theme(plot.title       = element_text(size = fig_base, face = "italic",
                                         colour = "grey35", hjust = 0.98,
                                         margin = margin(b = -8)),
        legend.key.size  = unit(3, "mm"),
        plot.margin      = margin(2, 2, 2, 2))

## scMET sub-panels: numeric μ/γ/ε titles, no per-panel y-label
scmet_panel_theme <- theme_ng(base_size = fig_base) +
  theme(plot.title       = element_text(hjust = 0.5, face = "plain",
                                         size = fig_base),
        legend.key.size  = unit(3, "mm"),
        plot.margin      = margin(2, 2, 2, 2))

## Performance sub-panels (G/H): square aspect inherited from theme_ng,
## centred title.
perf_theme <- theme_ng(base_size = fig_base) +
  theme(plot.title       = element_text(size = fig_base, hjust = 0.5,
                                         face = "plain"),
        legend.key.size  = unit(3, "mm"),
        plot.margin      = margin(2, 2, 2, 2))

## Legacy helpers retained for the Figure 2 (pairwise comparison) chunks below.
fig_theme_sq <- theme_ng(base_size = fig_base) +
  theme(legend.key.size = unit(3, "mm"))
add_simul_label <- function(p, label) {
  p + annotate("text", x = Inf, y = Inf, label = label,
               hjust = 1.1, vjust = 1.6, size = 2.8,
               fontface = "italic", colour = "grey35")
}
```

```{r load-data}
## ── Unadjusted (S / H) ────────────────────────────────────────────────────────
all_long_wc  <- readRDS("simulations/rds/all.long_within.rds")
jnt_wc       <- readRDS("simulations/rds/jnt_within.rds")
s_plots_wc   <- readRDS("simulations/rds/s_plots_within.rds")
n_plots_wc   <- readRDS("simulations/rds/n_plots_within.rds")

all_long_ac  <- readRDS("simulations/rds/all.long_across.rds")
jnt_ac       <- readRDS("simulations/rds/jnt_across.rds")
s_plots_ac   <- readRDS("simulations/rds/s_plots_across.rds")
n_plots_ac   <- readRDS("simulations/rds/n_plots_across.rds")

## ── Adjusted (adjS / adjH) ────────────────────────────────────────────────────
all_long_wc_adj <- readRDS("simulations/rds/all.long_within_adj.rds")
jnt_wc_adj      <- readRDS("simulations/rds/jnt_within_adj.rds")
s_plots_wc_adj  <- readRDS("simulations/rds/s_plots_within_adj.rds")
n_plots_wc_adj  <- readRDS("simulations/rds/n_plots_within_adj.rds")

all_long_ac_adj <- readRDS("simulations/rds/all.long_across_adj.rds")
jnt_ac_adj      <- readRDS("simulations/rds/jnt_across_adj.rds")
s_plots_ac_adj  <- readRDS("simulations/rds/s_plots_across_adj.rds")
n_plots_ac_adj  <- readRDS("simulations/rds/n_plots_across_adj.rds")
```

# Figure 1 – Adjusted entropies (adjS / adjH)

Same layout as the main combined figure but using adjusted sample entropy (adjS)
and adjusted Shannon entropy (adjH) in place of their unadjusted counterparts.

```{r adj-panels-ABCD}
## ── Row 1: scatter plots ──────────────────────────────────────────────────────
A <- ggplot(all_long_wc_adj, aes(meth, adj_sampen, colour = wc_vi)) +
  geom_point(alpha = 0.3, size = 0.4) + wc_scale + panel_theme +
  labs(x = "avg. methylation", y = "adj. sample entropy (adjS)",
       colour = "wcVI", title = "simul. II")
B <- ggplot(all_long_wc_adj, aes(norm_shannon, adj_sampen, colour = wc_vi)) +
  geom_point(alpha = 0.7, size = 0.4) + wc_scale + panel_theme +
  labs(x = "adj. Shannon entropy (adjH)", y = "adj. sample entropy (adjS)",
       colour = "wcVI", title = "simul. II")
C <- ggplot(all_long_ac_adj, aes(meth, adj_sampen, colour = bc_vi)) +
  geom_point(alpha = 0.3, size = 0.4) + ac_scale + panel_theme +
  labs(x = "avg. methylation", y = "adj. sample entropy (adjS)",
       colour = "acVI", title = "simul. III")
D <- ggplot(all_long_ac_adj, aes(norm_shannon, adj_sampen, colour = bc_vi)) +
  geom_point(alpha = 0.7, size = 0.4) + ac_scale + panel_theme +
  labs(x = "adj. Shannon entropy (adjH)", y = "adj. sample entropy (adjS)",
       colour = "acVI", title = "simul. III")
```

```{r adj-panels-EF}
## ── Rows 2 and 3: scMET panels (one full-width row per simulation)
## scMET μ/γ/ε as titles; "scMET score" only on the leftmost sub-panel y-axis;
## legend at bottom so sub-panels are full-width within the row.
mk_scmet <- function(df, vi_col, scale, vi_lab) {
  mk <- function(yvar, sub) {
    ggplot(df, aes(avg_meth, .data[[yvar]], colour = .data[[vi_col]])) +
      geom_point(alpha = 0.7, size = 0.4) + scale + scmet_panel_theme +
      labs(x = "avg. methylation", y = NULL, title = sub, colour = vi_lab)
  }
  m <- mk("scmet_mu",      expression(scMET ~ mu)) +
    labs(y = "scMET score")
  g <- mk("scmet_gamma",   expression(scMET ~ gamma))
  e <- mk("scmet_epsilon", expression(scMET ~ epsilon))
  wrap_plots(m, g, e, ncol = 3, guides = "collect") &
    theme(legend.position = "bottom")
}

E_row <- mk_scmet(jnt_wc_adj, "wc_vi", wc_scale, "wcVI")
F_row <- mk_scmet(jnt_ac_adj, "bc_vi", ac_scale, "acVI")
```

```{r adj-panels-GH}
## ── Row 4: performance grids (G simul. II, H simul. III)
## Each is a 3×2 grid of square panels (NMI/CPU/memory × cells/features).
mk_perf <- function(s_plots, n_plots) {
  panels <- list(
    s_plots$nmi     + perf_theme + labs(title = "NMI",          x = NULL,         y = "performance"),
    s_plots$seconds + perf_theme + labs(title = "CPU time (s)", x = "# cells",    y = NULL),
    s_plots$mem     + perf_theme + labs(title = "memory (MB)",  x = NULL,         y = NULL),
    n_plots$nmi     + perf_theme + labs(title = NULL,           x = NULL,         y = "performance"),
    n_plots$seconds + perf_theme + labs(title = NULL,           x = "# features", y = NULL),
    n_plots$mem     + perf_theme + labs(title = NULL,           x = NULL,         y = NULL)
  )
  wrap_plots(panels, ncol = 3, guides = "collect") &
    theme(legend.position = "bottom")
}

G_inner <- mk_perf(s_plots_wc_adj, n_plots_wc_adj)
H_inner <- mk_perf(s_plots_ac_adj, n_plots_ac_adj)
```

```{r adj-combined, fig.width=11.69, fig.height=14}
## ── Assemble adjS/adjH figure ─────────────────────────────────────────────────
## Equal-area sub-panels in rows 2-4 (E sub-panels = F sub-panels = G/H sub-panels).
## Layout: A B C D / E (50% width centred) / F (50% width centred) / G | H.
A <- A + labs(tag = "A")
B <- B + labs(tag = "B")
C <- C + labs(tag = "C")
D <- D + labs(tag = "D")
E_row[[1]] <- E_row[[1]] + labs(tag = "E")
F_row[[1]] <- F_row[[1]] + labs(tag = "F")
G_inner[[1]] <- G_inner[[1]] + labs(tag = "G")
H_inner[[1]] <- H_inner[[1]] + labs(tag = "H")

strip_legends <- function(p) p & theme(legend.position = "none")
GH_row <- wrap_plots(G_inner, strip_legends(H_inner), ncol = 2)

E_padded <- wrap_plots(plot_spacer(), E_row, plot_spacer(),
                       ncol = 3, widths = c(0.25, 0.5, 0.25))
F_padded <- wrap_plots(plot_spacer(), F_row, plot_spacer(),
                       ncol = 3, widths = c(0.25, 0.5, 0.25))

fig_adj <- (A | B | C | D) / E_padded / F_padded / GH_row +
  plot_layout(heights = c(1.5, 1, 1, 2)) &
  theme(plot.tag = element_text(size = fig_base + 3, face = "bold",
                                hjust = 0, vjust = 1))

print(fig_adj)
```

```{r save-adj-figure}
ggsave(
  filename = "simulations/output/figure_simulations_combined_adj.svg",
  plot     = fig_adj,
  width    = 11.69,
  height   = 14,
  units    = "in"
)

ggsave(
  filename = "simulations/output/figure_simulations_combined_adj.pdf",
  plot     = fig_adj,
  width    = 11.69,
  height   = 14,
  units    = "in"
)
```

# Figure 2 – Pairwise comparison: adjusted vs unadjusted entropies

These panels directly compare S vs adjS and H vs adjH, revealing how the
coverage-normalisation shifts entropy values relative to within/across-cell
variability indices.

```{r pairwise-data}
## Merge unadjusted and adjusted long-format data for pairwise plots.
## Both data frames have the same row structure (same features × samples),
## so we can cbind the adj columns directly.
wc_cmp <- data.frame(
  sampen       = all_long_wc$sampen,
  adj_sampen   = all_long_wc_adj$adj_sampen,
  shannon      = all_long_wc$shannon,
  norm_shannon = all_long_wc_adj$norm_shannon,
  meth         = all_long_wc$meth,
  avg_meth     = all_long_wc$avg_meth,
  wc_vi        = all_long_wc$wc_vi
)

ac_cmp <- data.frame(
  sampen       = all_long_ac$sampen,
  adj_sampen   = all_long_ac_adj$adj_sampen,
  shannon      = all_long_ac$shannon,
  norm_shannon = all_long_ac_adj$norm_shannon,
  meth         = all_long_ac$meth,
  avg_meth     = all_long_ac$avg_meth,
  bc_vi        = all_long_ac$bc_vi
)

## Feature-level (jnt) comparison for H vs adjH
wc_jnt_cmp <- data.frame(
  sampen_avg   = jnt_wc$sampen_avg,
  adj_sampen_avg = jnt_wc_adj$adj_sampen_avg,
  shannon      = jnt_wc$shannon,
  norm_shannon = jnt_wc_adj$shannon_norm,
  avg_meth     = jnt_wc$avg_meth,
  wc_vi        = jnt_wc$wc_vi
)

ac_jnt_cmp <- data.frame(
  sampen_avg   = jnt_ac$sampen_avg,
  adj_sampen_avg = jnt_ac_adj$adj_sampen_avg,
  shannon      = jnt_ac$shannon,
  norm_shannon = jnt_ac_adj$shannon_norm,
  avg_meth     = jnt_ac$avg_meth,
  bc_vi        = jnt_ac$bc_vi
)
```

```{r pairwise-panels-wc}
## ── Within-cell (simul. II) ───────────────────────────────────────────────────

P1 <- add_simul_label(
  ggplot(wc_cmp, aes(x = sampen, y = adj_sampen, color = wc_vi)) +
    geom_point(alpha = 0.3, size = 0.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
    theme_ng(base_size = fig_base) + wc_scale + fig_theme_sq +
    labs(x = "sample entropy (S)", y = "adj. sample entropy (adjS)", color = "wcVI"),
  "simul. II"
)

P2 <- add_simul_label(
  ggplot(wc_jnt_cmp, aes(x = shannon, y = norm_shannon, color = wc_vi)) +
    geom_point(alpha = 0.7, size = 0.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
    theme_ng(base_size = fig_base) + wc_scale + fig_theme_sq +
    labs(x = "Shannon entropy (H)", y = "adj. Shannon entropy (adjH)", color = "wcVI"),
  "simul. II"
)

P3 <- add_simul_label(
  ggplot(wc_cmp, aes(x = meth, y = sampen - adj_sampen, color = wc_vi)) +
    geom_point(alpha = 0.3, size = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    theme_ng(base_size = fig_base) + wc_scale + fig_theme_sq +
    labs(x = "avg. methylation", y = "S - adjS", color = "wcVI"),
  "simul. II"
)

P4 <- add_simul_label(
  ggplot(wc_jnt_cmp, aes(x = avg_meth, y = shannon - norm_shannon, color = wc_vi)) +
    geom_point(alpha = 0.7, size = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    theme_ng(base_size = fig_base) + wc_scale + fig_theme_sq +
    labs(x = "avg. methylation", y = "H - adjH", color = "wcVI"),
  "simul. II"
)
```

```{r pairwise-panels-ac}
## ── Across-cell (simul. III) ──────────────────────────────────────────────────

P5 <- add_simul_label(
  ggplot(ac_cmp, aes(x = sampen, y = adj_sampen, color = bc_vi)) +
    geom_point(alpha = 0.3, size = 0.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
    theme_ng(base_size = fig_base) + ac_scale + fig_theme_sq +
    labs(x = "sample entropy (S)", y = "adj. sample entropy (adjS)", color = "acVI"),
  "simul. III"
)

P6 <- add_simul_label(
  ggplot(ac_jnt_cmp, aes(x = shannon, y = norm_shannon, color = bc_vi)) +
    geom_point(alpha = 0.7, size = 0.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
    theme_ng(base_size = fig_base) + ac_scale + fig_theme_sq +
    labs(x = "Shannon entropy (H)", y = "adj. Shannon entropy (adjH)", color = "acVI"),
  "simul. III"
)

P7 <- add_simul_label(
  ggplot(ac_cmp, aes(x = meth, y = sampen - adj_sampen, color = bc_vi)) +
    geom_point(alpha = 0.3, size = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    theme_ng(base_size = fig_base) + ac_scale + fig_theme_sq +
    labs(x = "avg. methylation", y = "S - adjS", color = "acVI"),
  "simul. III"
)

P8 <- add_simul_label(
  ggplot(ac_jnt_cmp, aes(x = avg_meth, y = shannon - norm_shannon, color = bc_vi)) +
    geom_point(alpha = 0.7, size = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    theme_ng(base_size = fig_base) + ac_scale + fig_theme_sq +
    labs(x = "avg. methylation", y = "H - adjH", color = "acVI"),
  "simul. III"
)
```

```{r pairwise-combined, fig.width=15, fig.height=8}
## ── Assemble pairwise comparison figure ───────────────────────────────────────
## Row 1: within-cell (P1-P4), Row 2: across-cell (P5-P8)
fig_cmp <- (P1 | P2 | P3 | P4) / (P5 | P6 | P7 | P8) +
  plot_annotation(
    tag_levels = "A",
    theme = theme(plot.tag = element_text(size = 14, face = "bold"))
  )

print(fig_cmp)
```

```{r save-pairwise-figure}
ggsave(
  filename = "simulations/output/figure_simulations_comparison_adj_unadj.svg",
  plot     = fig_cmp,
  width    = 12.8,
  height   = 6.4,
  units    = "in"
)

ggsave(
  filename = "simulations/output/figure_simulations_comparison_adj_unadj.pdf",
  plot     = fig_cmp,
  width    = 12.8,
  height   = 6.4,
  units    = "in"
)
```

# Figure 3: adjS within methylation bins

Binning average methylation into 0.1-wide intervals and showing adjS per bin
confirms that the residual association between adjS and variability indices is
not an artifact of methylation level.

```{r bins-wc, fig.width = ng_fig_size(5, 2)$w, fig.height = ng_fig_size(5, 2)$h}
all_long_wc_adj %>%
  mutate(meth_bin = cut(meth, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>%
  filter(!is.na(meth_bin), !is.na(adj_sampen)) %>%
  ggplot(aes(x = meth_bin, y = adj_sampen, fill = cut(wc_vi, 5))) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  scale_fill_viridis_d(option = "viridis", name = "wcVI quintile") +
  labs(
    x = "avg. methylation bin",
    y = "adj. sample entropy (adjS)",
    title = "simul. II within-cell variability"
  ) +
  theme_ng(base_size = fig_base) + fig_theme_sq +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

```{r bins-ac, fig.width = ng_fig_size(5, 2)$w, fig.height = ng_fig_size(5, 2)$h}
all_long_ac_adj %>%
  mutate(meth_bin = cut(meth, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>%
  filter(!is.na(meth_bin), !is.na(adj_sampen)) %>%
  ggplot(aes(x = meth_bin, y = adj_sampen, fill = cut(bc_vi, 5))) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  scale_fill_viridis_d(option = "plasma", name = "acVI quintile") +
  labs(
    x = "avg. methylation bin",
    y = "adj. sample entropy (adjS)",
    title = "simul. III across-cell variability"
  ) +
  theme_ng(base_size = fig_base) + fig_theme_sq +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```
